### 2017-02-01 output/summary.Rnw
Initial commit

\documentclass{article}
\usepackage[margins=1in]{geometry}
\usepackage{Sweave}

\title{Summary of CA Rice Breeding}
\author{Matthew Espe}

\begin{document}

\maketitle

<<echo = FALSE>>=
library(rstan)
library(ggplot2)
library(lme4)
library(rstanarm)

load("../data/all_vt_weather.Rda")
##fit <- read_stan_csv(list.files("~/Dropbox/vt_results", "^samples[0-9].csv",
##                                full.names = TRUE))

## Get the number of years in the trial
n_years <- aggregate(year ~ id, data = all_vt_result,
          function(x) {
              length(unique(x))})

first_yr <- aggregate(year ~ id, data = all_vt_result,
          min)

i <- match(all_vt_result$id, first_yr$id)
all_vt_result$first_yr <- first_yr$year[i]
all_vt_result$yrs_in_trial <- all_vt_result$year - first_yr$year[i]

@

\section{Background}

In general, rice yields in CA have been flat over the last 20+ years,
compared to the preceeding period. Yet, annual reports from the CA
Cooperative Rice Research Foundation show that each released variety
has a yield advantage over the previous varieties, typically on the
order of 4-6\%. This raises the question of how to reconcile these
facts. Namely, there are two possibilities:

\begin{enumerate}
\item{There is a mistake in the annual reports, and there is little or no
    yield advantage of newer varieties over old.}
\item{The comparison to old varieties is flawed, allowing an
    appearant yield increase to manifest without an accompanied
    increase in statewide yields.}
\end{enumerate}

\subsection{Easy Facts}

\begin{enumerate}
\item At the year of release (1984), M-202 averaged 9,600 lbs/ac.
\item In 2015, the yield of M-209 was between 9,000-10,000 lbs/ac.
\item Therefore, there does not seem to have been an appreciatable
  increase in yield over this period.
\end{enumerate}

\begin{figure}

<<label=plot1, fig=TRUE, echo=FALSE>>=
boxplot(yield_lb ~ year, data = all_vt_result,
        xlab = "Year", ylab = expression("Yield ( lb ac"^-1~")"))
@

  \caption{Distribution of yield over the period 1995 -- 2015.} \label{fig:plot}
\end{figure}

\section{Do newer cultivars have a yield advantage?}

Looking at the numbers, it seems that each new medium-grain cultivar
has an advantage over the previous ones, but that the advantage is not
huge. Typical improvements of 5\% are reported by the breeders at the
RES, and the data show perhaps a bit less than this. There does not
appear to be a yield advantage of M-206 of M-205. It is worth
noting that there are fewer years of data for M-209, and those years
were typically higher yielding, which might lead ot a false optimism
regarding M-209.

\begin{figure}

<<label=plot2, fig=TRUE, echo=FALSE>>=
vars <- c("M206","M202","M204","M205", "M209")
i <- all_vt_result$id %in% vars

boxplot(yield_lb ~ id, data = all_vt_result[i,],
        xlab = "Cultivar",
        ylab = "Yield (lbs/ac)")
@
\caption{Boxplot of yields for four common medium grain cultivars for
  the period 1995--2015.}
\end{figure}

However, the question remains how CA yield have not been improving
more after 20 years of breeding effort.

\section{Evidence of yield decreases over time}

There is some evidence of a yield decrease over time, though this
turns out to be a tricky thing to estimate. This is because there are
several plausible and related mechanisms at work that are potentially
confounding eachother.

\begin{figure}

  <<fig = TRUE, echo=FALSE>>=
  plot(yield_lb ~ year, data = all_vt_result, type = "n")
  mods <- lmList(yield_lb ~ (1|year) | id, data = all_vt_result[i,])
  pred <- lapply(mods, function(x){
      data.frame(pred = unique(predict(x)),
                 year = unique(x$model$year))
  })
  sapply(seq_along(pred), function(i) points(x = pred[[i]]$year, y = pred[[i]]$pred,
                                             pch = i))
  legend("bottomright", legend = vars, pch = 1:5)

@

<<echo=FALSE>>=
meds_split <- split(all_vt_result[i,], paste(all_vt_result[i, "year"], all_vt_result[i,"site"]))
meds_scaled <- lapply(meds_split, function(x){
    m202 <- mean(x$yield_lb[x$id == "M202"])
    x$yield_lb <- x$yield_lb - m202
    return(x)
    })
meds_scaled <- do.call(rbind, meds_scaled)
ggplot(meds_scaled, aes(y = yield_lb, x = year, group = id)) + geom_smooth(aes(color = id), method = "lm") + geom_point()

mod <- stan_glm(yield_lb ~ factor(year) + site + factor(year):site +
                    yrs_in_trial + id + id:yrs_in_trial,
                data = all_vt_result[i,], prior = normal(), cores = 2L, iter = 1000)

@ 
\begin{enumerate}
  \item Technology improvements over time, including advances in
    nutrient, pest, and water management
  \item Variability over time 
  \item Variability between sites
\end{enumerate}

These factors make it difficult to estimate any one factor separate
from the others. Looking at the raw data, there appears to be a slight
increase in yield over time. From the above, we would expect this due
to technology improvements.

\begin{figure}
<<fig=TRUE, echo=FALSE>>=

by_year <- aggregate(year ~ id, data = all_vt_result, function(x) length(unique(x)))
morethan3 <- by_year$year > 3

i <- all_vt_result$id %in% by_year$id[morethan3]

plot(yield_lb ~ year, all_vt_result[i,],
     xlab = "Year", ylab = "Yield (lbs/ac)")

abline(lm(yield_lb ~ year, all_vt_result[i,]))

@
\caption{Yield over time for all cultivars during the period 1995 --
  2015. Only named (released) entries are included.}
\end{figure}

\begin{figure}
<<fig=TRUE, echo=FALSE>>=

named <- grepl("^[A-Z]", all_vt_result$id)

named_split <- split(all_vt_result[named,], all_vt_result$id[named])
ans <- lapply(named_split, function(x){
    minYr <- min(x$year, na.rm = TRUE)
    data.frame(yld = x$yield_lb,
               yr = minYr)
})
ans <- do.call(rbind, ans)

plot(yld ~ yr, data = ans,
     xlab = "Year of introduction",
     ylab = "Yield (lbs/ac)")
m1 <- lm(yld ~ yr, data = ans)
abline(m1)
@ 
\caption{The yield of named (released varieties), grouped by year of
  introduction. There is a significant increase over time, but the
  estimate is small (\Sexpr{round(coef(m1)[2], 1)} lbs per year increase).}
\end{figure}

There is evidence of increased yield potential over time, but it is
small, much smaller than the estimated 5\% increase per
release. According to this estimate, there was only a
\Sexpr{round(coef(m1)[2] * 21, 1)} lb increase over
the 21 year period.

\begin{figure}
<<fig=TRUE, echo=FALSE>>=
mod <- lm(yield_lb ~ year, data = all_vt_result)
resids <- residuals(mod)

m202 <- all_vt_result$id == "M202"
par(mfrow = c(1,2))
plot(all_vt_result$yield_lb[m202] ~ all_vt_result$year[m202],
     xlab = "Year",
     ylab = "Yield (raw)",
     main = "M-202") 
abline(lm(all_vt_result$yield_lb[m202] ~ all_vt_result$year[m202]))

plot(resids[m202] ~ all_vt_result$year[m202],
      xlab = "Year",
     ylab = "Yield (de-trended)",
     main = "M-202") 
abline(lm(resids[m202] ~ all_vt_result$year[m202]))
@
\caption{Comparison between the raw yield of cultivar M-202, and the
  yield normalized to remove the annual trends. Amounts less than 0
  are the amount lost compared to using a more modern variety.}
\end{figure}

\end{document}

### 2017-02-01 src/EDA.R

@@ -1,0 +1,59 @@
+## EDA on VT data
+## M. Espe
+## Jan 2017
+
+## Cleaned data from the other project
+load("../data/all_vt_weather.Rda")
+
+## Look at varieties over time
+n_years <- aggregate(year ~ id, data = all_vt_result,
+          function(x) {
+              length(unique(x))})
+
+first_yr <- aggregate(year ~ id, data = all_vt_result,
+          min)
+
+i <- match(all_vt_result$id, first_yr$id)
+all_vt_result$first_yr <- first_yr$year[i]
+all_vt_result$yrs_in_trial <- all_vt_result$year - first_yr$year[i]
+
+j <- n_years$year >= 7
+
+long_ids <- all_vt_result$id %in% n_years$id[j]
+
+## Look by varieties
+vt_split <- split(all_vt_result[long_ids,], all_vt_result$id[long_ids])
+
+par(mfrow = c(5, 5))
+sapply(vt_split, function(x){
+    plot(x$yield_lb ~ x$year,
+         xlim = c(1995, 2015),
+         main = unique(x$id))
+    abline(lm(x$yield_lb ~ x$year))
+})
+
+library(ggplot2)
+ggplot(data = all_vt_result[long_ids,],
+       aes(x = yrs_in_trial, y = yield_lb, group = id)) +
+    geom_point(aes(color = id)) +
+    geom_smooth(method = lm, aes(color = id)) +
+    facet_wrap(~id) +
+    theme_bw()
+
+ggplot(data = all_vt_result[long_ids,],
+       aes(x = yrs_in_trial, y = yield_lb, group = id)) +
+    geom_point(aes(color = id)) +
+    geom_smooth(method = lm, aes(color = id)) +
+    facet_wrap(~id) +
+    theme_bw()
+
+library(rstanarm)
+options(mc.cores = 2L)
+fit <- stan_glmer(scale(yield_lb) ~ yrs_in_trial + (1 + yrs_in_trial|id),
+                  data = all_vt_result[long_ids,],
+                  prior = normal(), iter = 500)
+
+## Look at side by side differences
+
+fit <- stan_glmer(scale(yield_lb) ~ 
+    

### 2017-02-01 output/summary.Rnw
Commit before start re-arranging from EDA to more directed

@@
 \begin{figure}
+<<label=plot2, fig=TRUE, echo=FALSE>>=
+  vars <- grepl("^M[0-9]{3}$", all_vt_result$id)
+  mod_data <- all_vt_result[vars,]
 
-<<label=plot2, fig=TRUE, echo=FALSE>>=
-vars <- c("M206","M202","M204","M205", "M209")
-i <- all_vt_result$id %in% vars
-
-boxplot(yield_lb ~ id, data = all_vt_result[i,],
+  boxplot(yield_lb ~ id, data = mod_data,
           xlab = "Cultivar",
@@
 
-\section{Evidence of yield decreases over time}
+\subsection{Hypotheses}
+
+\begin{enumerate}
+  \item After a cultivar has been released, the yield decreases over
+    time.
+  \item Each cultivar release is recovering to the old yield, not
+    increasing yield potential.
+\end{enumerate}
+
+\section{H1: Do yields decreases over time once a cultivar is released?}
+
+\subsection{Methods}
+
+Linear model of yield of a cultivar as a response to the number of years since it
+was entered into the trials. Corrections for year, site, year:site,
+and cultivar differences are made according to:
+
+$ yield = \alpha + \beta_{years in trial j}*x_{years in trials j} +
+\beta_{site} + \beta_{year} + \beta_{site:year} + \beta_{cultivar j}$
+
+Only medium-grained, released varieties were included in the analysis,
+reducing the data from \Sexpr{nrow(all_vt_result)} to
+\Sexpr{nrow(mod_data)} observations. Included varieties are:
+\Sexpr{paste(gsub('^M', 'M-', unique(mod_data$id)), collapse = ', ')}.
 
@@
 
-\begin{figure}
+% \begin{figure}
 
-  <<fig = TRUE, echo=FALSE>>=
-  plot(yield_lb ~ year, data = all_vt_result, type = "n")
-  mods <- lmList(yield_lb ~ (1|year) | id, data = all_vt_result[i,])
-  pred <- lapply(mods, function(x){
-      data.frame(pred = unique(predict(x)),
-                 year = unique(x$model$year))
-  })
-  sapply(seq_along(pred), function(i) points(x = pred[[i]]$year, y = pred[[i]]$pred,
-                                             pch = i))
-  legend("bottomright", legend = vars, pch = 1:5)
+%   <<fig = TRUE, echo=FALSE>>=
+%   ## plot(yield_lb ~ year, data = all_vt_result, type = "n")
+%   ## mods <- lmList(yield_lb ~ (1|year) | id, data = mod_data)
+%   ## pred <- lapply(mods, function(x){
+%   ##     data.frame(pred = unique(predict(x)),
+%   ##                year = unique(x$model$year))
+%   ## })
+%   ## sapply(seq_along(pred), function(i) points(x = pred[[i]]$year, y = pred[[i]]$pred,
+%   ##                                            pch = i))
+%   ## legend("bottomright", legend = vars, pch = 1:5)
+
+% @
+
+% <<echo=FALSE>>=
+% ## meds_split <- split(mod_data, paste(all_vt_result[i, "year"], all_vt_result[i,"site"]))
+% ## meds_scaled <- lapply(meds_split, function(x){
+% ##     m202 <- mean(x$yield_lb[x$id == "M202"])
+% ##     x$yield_lb <- x$yield_lb - m202
+% ##     return(x)
+% ##     })
+% ## meds_scaled <- do.call(rbind, meds_scaled)
+% ## ggplot(meds_scaled, aes(y = yield_lb, x = year, group = id)) + geom_smooth(aes(color = id), method = "lm") + geom_point()
+
+% ## mod <- stan_glm(yield_lb ~ factor(year) + site + factor(year):site +
+% ##                     yrs_in_trial + id + id:yrs_in_trial,
+% ##                 data = mod_data, prior = normal(), cores = 2L, iter = 1000)
 
@@
 
-<<echo=FALSE>>=
-meds_split <- split(all_vt_result[i,], paste(all_vt_result[i, "year"], all_vt_result[i,"site"]))
-meds_scaled <- lapply(meds_split, function(x){
-    m202 <- mean(x$yield_lb[x$id == "M202"])
-    x$yield_lb <- x$yield_lb - m202
-    return(x)
-    })
-meds_scaled <- do.call(rbind, meds_scaled)
-ggplot(meds_scaled, aes(y = yield_lb, x = year, group = id)) + geom_smooth(aes(color = id), method = "lm") + geom_point()
-
-mod <- stan_glm(yield_lb ~ factor(year) + site + factor(year):site +
-                    yrs_in_trial + id + id:yrs_in_trial,
-                data = all_vt_result[i,], prior = normal(), cores = 2L, iter = 1000)
-
-@ 
 \begin{enumerate}
@@
 
-plot(yield_lb ~ year, all_vt_result[i,],
+plot(yield_lb ~ year, mod_data,
      xlab = "Year", ylab = "Yield (lbs/ac)")
 
-abline(lm(yield_lb ~ year, all_vt_result[i,]))
+abline(lm(yield_lb ~ year, mod_data))
 

### 2017-02-06 src/h1_model.R
Forgot to commit before changing to custom Stan model

@@ -1,0 +1,12 @@
+## Run model and save output as text files
+##.args = list(commandArgs(trailingOnly=TRUE))
+
+library(rstanarm)
+source("prep_model_data.R")
+
+mod <- stan_glmer(yield_cs ~ (1|yr_fact) + (1|site) + (1|yr_site_fact) +
+                      yrs_in_trial + yrs_in_trial:id,
+                  data = mod_data, prior = normal(), cores = 2L, iter = 2000,## ... = .args
+                  )
+
+save(mod, file = "../output/h1_fit.Rda")

### 2017-02-06 src/prep_model_data.R

@@ -1,0 +1,28 @@
+## Prepares data for write-up
+
+load("../data/all_vt_weather.Rda")
+##fit <- read_stan_csv(list.files("~/Dropbox/vt_results", "^samples[0-9].csv",
+##                                full.names = TRUE))
+
+## Get the number of years in the trial
+n_years <- aggregate(year ~ id, data = all_vt_result,
+          function(x) {
+              length(unique(x))})
+
+first_yr <- aggregate(year ~ id, data = all_vt_result,
+          min)
+
+i <- match(all_vt_result$id, first_yr$id)
+all_vt_result$first_yr <- first_yr$year[i]
+all_vt_result$yrs_in_trial <- all_vt_result$year - first_yr$year[i]
+
+
+## Get some centered and scaled vars for model
+all_vt_result$yield_cs <- (all_vt_result$yield - 9000) / 1000
+all_vt_result$yr_fact <- factor(all_vt_result$year)
+all_vt_result$yr_site_fact <- factor(paste(all_vt_result$year, all_vt_result$site))
+
+## Subset to medium grains
+vars <- grepl("^M[0-9]{3}$", all_vt_result$id)
+mod_data <- all_vt_result[vars,]
+

### 2017-02-06 src/h1_model.R
Includes site:yr, but not cult:dur

@@ -1,12 +1,30 @@
 ## Run model and save output as text files
 ##.args = list(commandArgs(trailingOnly=TRUE))
 
-library(rstanarm)
+library(rstan)
+##library(rstanarm)
 source("prep_model_data.R")
 
-mod <- stan_glmer(yield_cs ~ (1|yr_fact) + (1|site) + (1|yr_site_fact) +
-                      yrs_in_trial + yrs_in_trial:id,
-                  data = mod_data, prior = normal(), cores = 2L, iter = 2000,## ... = .args
-                  )
+mm <- stan_model(file = "yield_model.stan")
 
-save(mod, file = "../output/h1_fit.Rda")
+n.level <- function(x){
+    length(unique(x))
+    }
+to.idx <- function(x){
+    as.integer(factor(x))
+    }
+
+dd <- list(N = nrow(mod_data),
+           n_cult = n.level(mod_data$id),
+           n_site = n.level(mod_data$site),
+           n_year = n.level(mod_data$year),
+           n_var = 3L,
+           cult = to.idx(mod_data$id),
+           site = to.idx(mod_data$site),
+           year = to.idx(mod_data$year),
+           duration = mod_data$yrs_in_trial,
+           yield = mod_data$yield_cs)
+
+fit <- sampling(mm, data = dd, iter = 1000, cores = 2L, control = list(adapt_delta = 0.9))
+
+save(fit, file = "../output/h1_fit.Rda")

### 2017-02-07 output/summary.Rnw
Should work

@@
 \documentclass{article}
-\usepackage[margins=1in]{geometry}
+\usepackage[margin=1in]{geometry}
 \usepackage{Sweave}
@@
 <<echo = FALSE>>=
+
 library(rstan)
@@
 
-load("../data/all_vt_weather.Rda")
-##fit <- read_stan_csv(list.files("~/Dropbox/vt_results", "^samples[0-9].csv",
-##                                full.names = TRUE))
-
-## Get the number of years in the trial
-n_years <- aggregate(year ~ id, data = all_vt_result,
-          function(x) {
-              length(unique(x))})
-
-first_yr <- aggregate(year ~ id, data = all_vt_result,
-          min)
-
-i <- match(all_vt_result$id, first_yr$id)
-all_vt_result$first_yr <- first_yr$year[i]
-all_vt_result$yrs_in_trial <- all_vt_result$year - first_yr$year[i]
-
+source("../src/prep_model_data.R")
+load("h1_fit.Rda")
 @
@@
 has a yield advantage over the previous varieties, typically on the
-order of 4-6\%. This raises the question of how to reconcile these
+order of 4--6\%. This raises the question of how to reconcile these
 facts. Namely, there are two possibilities:
@@
 
-\begin{figure}
-
-<<label=plot1, fig=TRUE, echo=FALSE>>=
-boxplot(yield_lb ~ year, data = all_vt_result,
-        xlab = "Year", ylab = expression("Yield ( lb ac"^-1~")"))
-@
-
-  \caption{Distribution of yield over the period 1995 -- 2015.} \label{fig:plot}
-\end{figure}
-
-\section{Do newer cultivars have a yield advantage?}
-
-Looking at the numbers, it seems that each new medium-grain cultivar
-has an advantage over the previous ones, but that the advantage is not
-huge. Typical improvements of 5\% are reported by the breeders at the
-RES, and the data show perhaps a bit less than this. There does not
-appear to be a yield advantage of M-206 of M-205. It is worth
-noting that there are fewer years of data for M-209, and those years
-were typically higher yielding, which might lead ot a false optimism
-regarding M-209.
-
-\begin{figure}
-<<label=plot2, fig=TRUE, echo=FALSE>>=
-  vars <- grepl("^M[0-9]{3}$", all_vt_result$id)
-  mod_data <- all_vt_result[vars,]
-
-  boxplot(yield_lb ~ id, data = mod_data,
-          xlab = "Cultivar",
-          ylab = "Yield (lbs/ac)")
-@
-\caption{Boxplot of yields for four common medium grain cultivars for
-  the period 1995--2015.}
-\end{figure}
-
-However, the question remains how CA yield have not been improving
-more after 20 years of breeding effort.
-
 \subsection{Hypotheses}
@@
 \begin{enumerate}
-  \item After a cultivar has been released, the yield decreases over
+  \item H1: After a cultivar has been released, its yield decreases over
     time.
-  \item Each cultivar release is recovering to the old yield, not
+  \item H2: Each cultivar release is recovering to the old yield benchmark, not
     increasing yield potential.
@@
 \Sexpr{nrow(mod_data)} observations. Included varieties are:
-\Sexpr{paste(gsub('^M', 'M-', unique(mod_data$id)), collapse = ', ')}.
+\Sexpr{paste(gsub('^M', 'M-', unique(mod_data[,'id'])), collapse = ', ')}.
 
-There is some evidence of a yield decrease over time, though this
-turns out to be a tricky thing to estimate. This is because there are
-several plausible and related mechanisms at work that are potentially
-confounding eachother.
+\subsection{Results}
+
+<<fig=TRUE, echo=FALSE>>=
+plot(mod, regex_pars = "yrs_in_trial")
+@ 
+\section{H2: Do newer cultivars have a yield advantage?}
+
+\subsection{Methods}
+
+In the above model, the estimate for $\beta_{cultivar}$ is
+the estimated advantage for cultivar X at year 0 (first year in the
+trial) averaged over variability attributable to differences in years,
+sites, and the year:site interaction.
+
+% Notes: the expected result here is that there is no difference
+% between varieties in year 0. This might be complicated by the fact
+% that we have M-202 15 years into its release. We might need to get a
+% better way to cross reference these to the true start year.
+
+\subsection{Results}
+
+The raw numbers generally support the conclusion that each new
+medium-grain cultivar has an advantage over the previous ones, but
+that advantage is not large (Fig. \ref{fig:raw_yield}). Typical
+improvements of 5\% are reported by the breeders at the RES, and the
+data do not support gains this large.
+
+\begin{figure}
+
+<<label=plot1, fig=TRUE, echo=FALSE>>=
+boxplot(yield_lb ~ year, data = mod_data,
+        xlab = "Year", ylab = expression("Yield ( lb ac"^-1~")"))
+@
+  \caption{Distribution of yield over the period 1995 -- 2015.} \label{fig:plot}
+\end{figure}
+
+\begin{figure}
+<<label=plot2, fig=TRUE, echo=FALSE>>=
+
+  boxplot(yield_lb ~ id, data = mod_data,
+          xlab = "Cultivar",
+          ylab = "Yield (lbs/ac)")
+@
+\caption{Boxplot of yields for released medium grain cultivars for
+  the period 1995--2015.}
+\label{fig:raw_yield}
+\end{figure}
+
+However, the question remains how CA yield have not been improving
+more after 20 years of breeding effort.
 
@@
 % ## ggplot(meds_scaled, aes(y = yield_lb, x = year, group = id)) + geom_smooth(aes(color = id), method = "lm") + geom_point()
-
-% ## mod <- stan_glm(yield_lb ~ factor(year) + site + factor(year):site +
-% ##                     yrs_in_trial + id + id:yrs_in_trial,
-% ##                 data = mod_data, prior = normal(), cores = 2L, iter = 1000)
 

### 2017-02-07 src/getYears.R

@@ -1,0 +1,24 @@
+
+doc <- readLines("../data/var_list.txt")
+first <- grep("^Cultivar", doc)
+last <- grep("^$", doc)
+last <- last[last > first][1]
+
+txt <- strsplit(doc[first:(last-1)], " {2,}")
+txt <- do.call(rbind, txt)
+
+yrTbl <- as.data.frame(txt[-1,], stringsAsFactors = FALSE)
+colnames(yrTbl) <- txt[1,]
+
+yrTbl$Year <- as.numeric(yrTbl$Year)
+yrTbl$Cultivar <- gsub("M-", "M", yrTbl$Cultivar)
+
+## Add M105, M207, and M209 by hand
+
+yrTbl <- rbind(yrTbl, data.frame(Cultivar = c("M105","M207","M209"),
+                                 Grain = c("M","M","M"),
+                                 Year = c(2012, 2005, 2015),
+                                 Maturity = c("E","E","E"),
+                                 Parents = ""))
+
+save(yrTbl, file = "../data/yrTbl.Rda")

### 2017-02-07 src/h1_model.R

@@ -5,7 +5,7 @@ library(rstan)
 ##library(rstanarm)
 source("prep_model_data.R")
 
-mm <- stan_model(file = "yield_model.stan")
+mm <- stan_model(file = "yield_model.stan", verbose = TRUE)
 
 n.level <- function(x){
     length(unique(x))
@@ -18,11 +18,14 @@ dd <- list(N = nrow(mod_data),
            n_cult = n.level(mod_data$id),
            n_site = n.level(mod_data$site),
            n_year = n.level(mod_data$year),
-           n_var = 3L,
+           n_yr_site = n.level(mod_data$yr_site_fact),
+           n_var = 5L,
            cult = to.idx(mod_data$id),
            site = to.idx(mod_data$site),
            year = to.idx(mod_data$year),
+           yr_site = to.idx(mod_data$yr_site_fact),
            duration = mod_data$yrs_in_trial,
+           release_yr = mod_data$release_yr - 2000,
            yield = mod_data$yield_cs)
 
 fit <- sampling(mm, data = dd, iter = 1000, cores = 2L, control = list(adapt_delta = 0.9))

### 2017-02-07 src/prep_model_data.R

@@ -1,6 +1,10 @@
 ## Prepares data for write-up
 
 load("../data/all_vt_weather.Rda")
+
+## Replace 08Y3269 with M209
+all_vt_result$id[all_vt_result$id == "08Y3269"] <- "M209"
+
 ##fit <- read_stan_csv(list.files("~/Dropbox/vt_results", "^samples[0-9].csv",
 ##                                full.names = TRUE))
 
@@ -9,20 +13,24 @@ n_years <- aggregate(year ~ id, data = all_vt_result,
           function(x) {
               length(unique(x))})
 
-first_yr <- aggregate(year ~ id, data = all_vt_result,
-          min)
-
-i <- match(all_vt_result$id, first_yr$id)
-all_vt_result$first_yr <- first_yr$year[i]
-all_vt_result$yrs_in_trial <- all_vt_result$year - first_yr$year[i]
-
-
-## Get some centered and scaled vars for model
-all_vt_result$yield_cs <- (all_vt_result$yield - 9000) / 1000
-all_vt_result$yr_fact <- factor(all_vt_result$year)
-all_vt_result$yr_site_fact <- factor(paste(all_vt_result$year, all_vt_result$site))
-
 ## Subset to medium grains
 vars <- grepl("^M[0-9]{3}$", all_vt_result$id)
-mod_data <- all_vt_result[vars,]
+
+## Do not include San Joaquin sites - known to hammer with cold and
+## Not managed similar to most of CA rice production
+no_sj <- all_vt_result$county != "SAN JOAQUIN"
+mod_data <- all_vt_result[(vars & no_sj),]
+
+## Use officially published years instead of years in trial
+load("../data/yrTbl.Rda")
+i <- match(mod_data$id, yrTbl$Cultivar)
+
+mod_data$release_yr <- yrTbl$Year[i]
+mod_data$yrs_in_trial <- mod_data$year - mod_data$release_yr
+
+## Get some centered and scaled vars for model
+mod_data$yield_cs <- (mod_data$yield - 9000) / 1000
+mod_data$yr_fact <- factor(mod_data$year)
+mod_data$yr_site_fact <- factor(paste(mod_data$year, mod_data$site))
+
 

### 2017-02-08 output/summary.Rnw
Updated summary with new models and revised figures

@@
 \usepackage{Sweave}
+\linespread{1.3}
 
@@
 library(rstan)
-library(ggplot2)
-library(lme4)
 library(rstanarm)
+source("../src/prep_model_data.R")
+load("../output/h1_fit.Rda")
 
-source("../src/prep_model_data.R")
-load("h1_fit.Rda")
 @
@@
 In general, rice yields in CA have been flat over the last 20+ years,
-compared to the preceeding period. Yet, annual reports from the CA
-Cooperative Rice Research Foundation show that each released variety
-has a yield advantage over the previous varieties, typically on the
+compared to the preceding period. Yet, annual reports from the CA
+Cooperative Rice Research Foundation show that each released cultivar
+has a yield advantage over the previous cultivars, typically on the
 order of 4--6\%. This raises the question of how to reconcile these
@@
 \item{There is a mistake in the annual reports, and there is little or no
-    yield advantage of newer varieties over old.}
-\item{The comparison to old varieties is flawed, allowing an
-    appearant yield increase to manifest without an accompanied
+    yield advantage of newer cultivars over old.}
+\item{The comparison to old cultivars is flawed, allowing an
+    apparent yield increase to manifest without an accompanied
     increase in statewide yields.}
@@
 
-\subsection{Easy Facts}
+\subsection{Motivating Evidence}
 
-\begin{enumerate}
-\item At the year of release (1984), M-202 averaged 9,600 lbs/ac.
-\item In 2015, the yield of M-209 was between 9,000-10,000 lbs/ac.
-\item Therefore, there does not seem to have been an appreciatable
+\begin{itemize}
+\item At the year of release (1984), M-202 averaged 9,600 lbs acre\textsuperscript{-1}.
+\item In 2015, the yield of M-209 was between 9,000-10,000 lbs acre\textsuperscript{-1}.
+\item Therefore, there does not seem to have been an appreciable
   increase in yield over this period.
-\end{enumerate}
+\end{itemize}
 
@@
 
-$ yield = \alpha + \beta_{years in trial j}*x_{years in trials j} +
-\beta_{site} + \beta_{year} + \beta_{site:year} + \beta_{cultivar j}$
+$$ yield = \alpha + \beta_{years\ in\ trial}* x_{years\ in\ trials} + \beta_{site} + \beta_{year} + \beta_{site:year} + \beta_{cultivar}$$
 
-Only medium-grained, released varieties were included in the analysis,
-reducing the data from \Sexpr{nrow(all_vt_result)} to
-\Sexpr{nrow(mod_data)} observations. Included varieties are:
-\Sexpr{paste(gsub('^M', 'M-', unique(mod_data[,'id'])), collapse = ', ')}.
+Only medium-grained, released cultivars were included in the analysis
+and data from the San Joaquin trials, which are not representative of
+typical CA rice production (frequent exposure to cold, drill-seeded,
+etc.) were excluded. This reduced the data from \Sexpr{nrow(all_vt_result)} to
+\Sexpr{nrow(mod_data)} observations. Included cultivars are:
+\Sexpr{paste(gsub('^M', 'M-', unique(mod_data[,'id'])), collapse = ', ')}. The
+official release years were used for each cultivar for the year of
+release (i.e., not the year seed was first commercially available).
 
@@
 
+\begin{figure}[h]
+
 <<fig=TRUE, echo=FALSE>>=
-plot(mod, regex_pars = "yrs_in_trial")
+yrs_in <- -6:34
+pp <- posterior_predict(yrs_fit, re.form = NA, newdata = data.frame(yrs_in_trial = yrs_in))
+pp <-  pp * 1000 + 9000
+
+ints <- apply(pp, 2, quantile, p = c(0.05, 0.5, 0.95))
+
+plot(yield_lb ~ yrs_in_trial, data = mod_data, pch = 16, col = rgb(0,0,0,0.25),
+     xlab = "Year since entry",
+     ylab = expression("Yield ( lbs acre"^-1~")"))
+lines(x = yrs_in, y = ints[2,], lty = 2) 
+polygon(x = c(yrs_in, rev(yrs_in)), y = c(ints[1,],rev(ints[3,])), col = rgb(0,0,0,0.2), border = NA)
+text(20, 3500, bquote(.(round(fixef(yrs_fit)[2] * 1000, 1))~lbs~year^-1))
+
 @ 
+\caption{There is a small decrease in yield over the time since a
+  cultivar was entered into the trial.}
+\label{fig:yrs}
+\end{figure}
+
+As seen in Fig. \ref{fig:yrs}, there is a small decrease in yield
+over time once a cultivar is released. This decrease is estimated to be
+\Sexpr{round(fixef(yrs_fit)[2] * 1000, 1)} lbs
+year\textsuperscript{-1}. The oldest cultivar in the trials has lost
+an estimated \Sexpr{round(fixef(yrs_fit)[2] * 1000 * 36, 1)} lbs
+acre\textsuperscript{-1} since release.
+
 \section{H2: Do newer cultivars have a yield advantage?}
@@
 
-In the above model, the estimate for $\beta_{cultivar}$ is
-the estimated advantage for cultivar X at year 0 (first year in the
-trial) averaged over variability attributable to differences in years,
-sites, and the year:site interaction.
+A similar but separate model was used to estimate whether newer
+releases had a yield advantage over older ones:
 
-% Notes: the expected result here is that there is no difference
-% between varieties in year 0. This might be complicated by the fact
-% that we have M-202 15 years into its release. We might need to get a
-% better way to cross reference these to the true start year.
+$$ yield = \alpha + \beta_{release\ year}*x_{release\ year} +
+\beta_{site} + \beta_{year} + \beta_{site:year} + \beta_{cultivar}$$
+
+Again, the estimate was conditioned on differences between cultivars,
+sites, years, and site:years.
 
@@
 
-The raw numbers generally support the conclusion that each new
-medium-grain cultivar has an advantage over the previous ones, but
-that advantage is not large (Fig. \ref{fig:raw_yield}). Typical
-improvements of 5\% are reported by the breeders at the RES, and the
-data do not support gains this large.
+\begin{figure}[h]
 
-\begin{figure}
+<<fig=TRUE, echo=FALSE>>=
+yrs <- 1980:2015
+pp <- posterior_predict(release_fit, re.form = NA, newdata = data.frame(release_yr = yrs))
+pp <-  pp * 1000 + 9000
+
+ints <- apply(pp, 2, quantile, p = c(0.05, 0.5, 0.95))
+
+plot(yield_lb ~ release_yr, data = mod_data, pch = 16, col = rgb(0,0,0,0.25),
+     xlab = "Year of Release",
+     ylab =  expression("Yield ( lbs acre"^-1~")"))
+lines(x = yrs, y = ints[2,], lty = 2) 
+polygon(x = c(yrs, rev(yrs)), y = c(ints[1,],rev(ints[3,])), col = rgb(0,0,0,0.2), border = NA)
+text(1990, 3500, bquote(.(round(fixef(release_fit)[2] * 1000, 1))~lbs~year^-1))
+@
+
+\caption{There is an estimated increase in yield over time, but it is
+  smaller than the previously estimated 54 lbs
+  year\textsuperscript{-1}}
+\label{fig:release}
+\end{figure}
+
+Contrary to the estimates for decreases once a cultivar is released, there was a
+small increase over time (Fig. \ref{fig:release}), though it was small
+(\Sexpr{round(fixef(release_fit)[2] * 1000, 1)} lbs
+year\textsuperscript{-1}). This suggests that the overall progress on
+yield over the period from 1980 -- 2016 is approximately
+\Sexpr{round(fixef(release_fit)[2] * 1000 * 36, 1)} lbs acre\textsuperscript{-1}.
+
+\section{Conclusions}
+
+\begin{itemize}
+  \item H1: There is evidence of a small decrease in yield once a cultivar is
+    released. This complicates comparisons between new entries and
+    checks, since the check will decrease over time. This decrease
+    will manifest as an apparent larger yield advantage of the new
+    entry over the check.
+  \item H2: There is evidence of a small increase in yield with each
+    new released cultivar. Therefore, it appears that successive
+    cultivar releases are not merely recovering to the old benchmark
+    but are surpassing it. However, progress is slower than previously
+    estimated (\Sexpr{round(fixef(release_fit)[2] * 1000, 1)}
+    lbs acre\textsuperscript{-1} year\textsuperscript{-1} increase
+    verses a previous estimate of 54 lbs acre\textsuperscript{-1}
+    year\textsuperscript{-1}).
+\end{itemize}
+
+The practical impact of these results is:
+
+\begin{enumerate}
+\item Breeders need to be aware of the potential for the yield of
+  checks to decrease over time. Due to this, comparisons to older checks
+  might give a false sense of improvement. In order to have greater
+  improvements in yield, the target yield improvement for a new cultivar
+  needs to be higher to offset the impact of yield decrease once a
+  cultivar is released.
+\item Breeding is essential to maintaining yield in rice. Yield is not
+  stable and without continuous effort, yield potential will decrease
+  over time.
+
+\end{enumerate}
+
+\section{Complications}
+
+There is strong collinearity present between the year of release and
+the number of years in the trials which is complicating the
+simultaneous estimation of these effects. With the current size of the
+data set (N = \Sexpr{nrow(mod_data)}), it is impossible to estimate
+the decrease over time since release and the improvement in yield over
+time with each new release in the same model. This is why each effect was
+estimated using a separate regression.
+
+Increasing the size of the data to the full data set (N =
+\Sexpr{nrow(all_vt_result)}), allows the simultaneous estimation
+and results in similar estimated effect sizes for each of these
+mechanisms. However, this necessitates the inclusion of many entries
+which have characteristics not similar to conventional cultivars.
+
+\newpage
+
+\section{Supplemental figures}
+
+\begin{figure}[h]
 
@@
 
-\begin{figure}
+\begin{figure}[h]
 <<label=plot2, fig=TRUE, echo=FALSE>>=
@@
 
-However, the question remains how CA yield have not been improving
-more after 20 years of breeding effort.
-
-% \begin{figure}
-
-%   <<fig = TRUE, echo=FALSE>>=
-%   ## plot(yield_lb ~ year, data = all_vt_result, type = "n")
-%   ## mods <- lmList(yield_lb ~ (1|year) | id, data = mod_data)
-%   ## pred <- lapply(mods, function(x){
-%   ##     data.frame(pred = unique(predict(x)),
-%   ##                year = unique(x$model$year))
-%   ## })
-%   ## sapply(seq_along(pred), function(i) points(x = pred[[i]]$year, y = pred[[i]]$pred,
-%   ##                                            pch = i))
-%   ## legend("bottomright", legend = vars, pch = 1:5)
-
-% @
-
-% <<echo=FALSE>>=
-% ## meds_split <- split(mod_data, paste(all_vt_result[i, "year"], all_vt_result[i,"site"]))
-% ## meds_scaled <- lapply(meds_split, function(x){
-% ##     m202 <- mean(x$yield_lb[x$id == "M202"])
-% ##     x$yield_lb <- x$yield_lb - m202
-% ##     return(x)
-% ##     })
-% ## meds_scaled <- do.call(rbind, meds_scaled)
-% ## ggplot(meds_scaled, aes(y = yield_lb, x = year, group = id)) + geom_smooth(aes(color = id), method = "lm") + geom_point()
-
-@
-
-\begin{enumerate}
-  \item Technology improvements over time, including advances in
-    nutrient, pest, and water management
-  \item Variability over time 
-  \item Variability between sites
-\end{enumerate}
-
-These factors make it difficult to estimate any one factor separate
-from the others. Looking at the raw data, there appears to be a slight
-increase in yield over time. From the above, we would expect this due
-to technology improvements.
-
-\begin{figure}
-<<fig=TRUE, echo=FALSE>>=
-
-by_year <- aggregate(year ~ id, data = all_vt_result, function(x) length(unique(x)))
-morethan3 <- by_year$year > 3
-
-i <- all_vt_result$id %in% by_year$id[morethan3]
-
-plot(yield_lb ~ year, mod_data,
-     xlab = "Year", ylab = "Yield (lbs/ac)")
-
-abline(lm(yield_lb ~ year, mod_data))
-
-@
-\caption{Yield over time for all cultivars during the period 1995 --
-  2015. Only named (released) entries are included.}
-\end{figure}
-
-\begin{figure}
-<<fig=TRUE, echo=FALSE>>=
-
-named <- grepl("^[A-Z]", all_vt_result$id)
-
-named_split <- split(all_vt_result[named,], all_vt_result$id[named])
-ans <- lapply(named_split, function(x){
-    minYr <- min(x$year, na.rm = TRUE)
-    data.frame(yld = x$yield_lb,
-               yr = minYr)
-})
-ans <- do.call(rbind, ans)
-
-plot(yld ~ yr, data = ans,
-     xlab = "Year of introduction",
-     ylab = "Yield (lbs/ac)")
-m1 <- lm(yld ~ yr, data = ans)
-abline(m1)
-@ 
-\caption{The yield of named (released varieties), grouped by year of
-  introduction. There is a significant increase over time, but the
-  estimate is small (\Sexpr{round(coef(m1)[2], 1)} lbs per year increase).}
-\end{figure}
-
-There is evidence of increased yield potential over time, but it is
-small, much smaller than the estimated 5\% increase per
-release. According to this estimate, there was only a
-\Sexpr{round(coef(m1)[2] * 21, 1)} lb increase over
-the 21 year period.
-
-\begin{figure}
-<<fig=TRUE, echo=FALSE>>=
-mod <- lm(yield_lb ~ year, data = all_vt_result)
-resids <- residuals(mod)
-
-m202 <- all_vt_result$id == "M202"
-par(mfrow = c(1,2))
-plot(all_vt_result$yield_lb[m202] ~ all_vt_result$year[m202],
-     xlab = "Year",
-     ylab = "Yield (raw)",
-     main = "M-202") 
-abline(lm(all_vt_result$yield_lb[m202] ~ all_vt_result$year[m202]))
-
-plot(resids[m202] ~ all_vt_result$year[m202],
-      xlab = "Year",
-     ylab = "Yield (de-trended)",
-     main = "M-202") 
-abline(lm(resids[m202] ~ all_vt_result$year[m202]))
-@
-\caption{Comparison between the raw yield of cultivar M-202, and the
-  yield normalized to remove the annual trends. Amounts less than 0
-  are the amount lost compared to using a more modern variety.}
-\end{figure}
-
 \end{document}

### 2017-02-08 src/h1_model.R

@@ -2,7 +2,7 @@
 ##.args = list(commandArgs(trailingOnly=TRUE))
 
 library(rstan)
-##library(rstanarm)
+library(rstanarm)
 source("prep_model_data.R")
 
 mm <- stan_model(file = "yield_model.stan", verbose = TRUE)
@@ -28,6 +28,15 @@ dd <- list(N = nrow(mod_data),
            release_yr = mod_data$release_yr - 2000,
            yield = mod_data$yield_cs)
 
-fit <- sampling(mm, data = dd, iter = 1000, cores = 2L, control = list(adapt_delta = 0.9))
+## Rprof("mod.Rprof")
+fit <- sampling(mm, data = dd, iter = 100, cores = 1L, control = list(adapt_delta = 0.9))
+## Rprof(NULL)
+
+fit <- stan_glmer(yield_cs ~ (1|year) + (1|site) + (1|id) + (1|yr_site_fact) + yrs_in_trial,
+                  data = mod_data, iter = 1000, cores = 2L)
+
+fit2 <- stan_glmer(yield_cs ~ (1|year) + (1|site) + (1|id) + (1|yr_site_fact) + release_yr,
+                  data = mod_data, iter = 1000, cores = 2L)
+
 
 save(fit, file = "../output/h1_fit.Rda")

### 2017-02-15 output/summary.Rnw
Updated to have estimates from just RES

@@
 library(rstanarm)
+library(lme4)
+
 source("../src/prep_model_data.R")
-load("../output/h1_fit.Rda")
+load("h1_fit.Rda")
 
@@
 yrs <- 1980:2015
-pp <- posterior_predict(release_fit, re.form = NA, newdata = data.frame(release_yr = yrs))
+pp <- posterior_predict(release_fit, re.form = NA, newdata = data.frame(release_yr_c = yrs - 2000))
 pp <-  pp * 1000 + 9000
@@
 
+\begin{figure}
+<<fig=TRUE, fig.width = 8, fig.height = 5, out.width = '80%', echo=FALSE>>=
+tars <- c("M202","M206","M209")
+par(mfrow = c(1,3),
+    mar = c(4,0,2,0),
+    oma = c(4,4,0.5,0.5))
+
+invisible(sapply(tars, function(x){
+    plot(yield_lb ~ year, data = mod_data[(mod_data$id == x & mod_data$site == "RES"),],
+         xlim = c(1995, 2016), ylim = c(6000,12000),
+         pch = 16, main = x,
+         ylab = "",
+         yaxt = "n",
+         xlab = "Year")
+    if(x == "M202"){
+        axis(2)
+        mtext(expression("Yield ( lbs acre"^-1~")"), 2, line = 2.25, cex.lab = 0.6)
+    }
+    abline(lm(yield_lb ~ year, data = mod_data[(mod_data$id == x & mod_data$site == "RES"),]), lty = 2)
+
+}))
+
+@
+\caption{The performance of three major cultivars at the RES site over
+time.}
+\label{fig:res_yrs}
+\end{figure}
+
+\begin{figure}
+
+<<fig=TRUE, echo=FALSE>>=
+par(mfrow = c(1,3),
+    mar = c(4,0,2,0),
+    oma = c(4,4,0.5,0.5))
+
+invisible(sapply(tars, function(x){
+    plot(yield_lb ~ yrs_in_trial,
+         data = mod_data[(mod_data$id == x & mod_data$site == "RES"),],
+         xlim = c(-6, 36), ylim = c(6000,12000),
+         pch = 16, main = x,
+         ylab = "",
+         yaxt = "n",
+         xlab = "Years after release")
+    if(x == "M202"){
+        axis(2)
+        mtext(expression("Yield ( lbs acre"^-1~")"), 2, line = 2.25, cex.lab = 0.6)
+    }
+    abline(lm(yield_lb ~ yrs_in_trial,
+              data = mod_data[(mod_data$id == x & mod_data$site == "RES"),]), lty = 2)
+}))
+
+@
+\caption{The performance of three major cultivars at the RES site
+  since the year each cultivar was released.}
+\label{fig:res_since_release}
+\end{figure}
+
 \end{document}

### 2017-02-15 src/getYears.R

@@ -1,3 +1,5 @@
+## Get published years of release for varieties
+## to see if there is an improvement over time
 
 doc <- readLines("../data/var_list.txt")
 first <- grep("^Cultivar", doc)
@@ -20,5 +22,8 @@ yrTbl <- rbind(yrTbl, data.frame(Cultivar = c("M105","M207","M209"),
                                  Year = c(2012, 2005, 2015),
                                  Maturity = c("E","E","E"),
                                  Parents = ""))
+
+## Date for M-206 is wrong - should be 2004
+yrTbl$Year[yrTbl$Cultivar == "M206"] <- 2004
 
 save(yrTbl, file = "../data/yrTbl.Rda")

### 2017-02-15 src/h1_model.R

@@ -5,38 +5,52 @@ library(rstan)
 library(rstanarm)
 source("prep_model_data.R")
 
-mm <- stan_model(file = "yield_model.stan", verbose = TRUE)
+## mm <- stan_model(file = "yield_model.stan", verbose = TRUE)
 
-n.level <- function(x){
-    length(unique(x))
-    }
-to.idx <- function(x){
-    as.integer(factor(x))
-    }
+## n.level <- function(x){
+##     length(unique(x))
+##     }
+## to.idx <- function(x){
+##     as.integer(factor(x))
+##     }
 
-dd <- list(N = nrow(mod_data),
-           n_cult = n.level(mod_data$id),
-           n_site = n.level(mod_data$site),
-           n_year = n.level(mod_data$year),
-           n_yr_site = n.level(mod_data$yr_site_fact),
-           n_var = 5L,
-           cult = to.idx(mod_data$id),
-           site = to.idx(mod_data$site),
-           year = to.idx(mod_data$year),
-           yr_site = to.idx(mod_data$yr_site_fact),
-           duration = mod_data$yrs_in_trial,
-           release_yr = mod_data$release_yr - 2000,
-           yield = mod_data$yield_cs)
+## Pull M401, M402 - remove Tucker 2015
+## look at only RES vs.  
+## M202 vs M206 vs. M209
+
+## dd <- list(N = nrow(mod_data),
+##            n_cult = n.level(mod_data$id),
+##            n_site = n.level(mod_data$site),
+##            n_year = n.level(mod_data$year),
+##            n_yr_site = n.level(mod_data$yr_site_fact),
+##            n_var = 4L,
+##            cult = to.idx(mod_data$id),
+##            site = to.idx(mod_data$site),
+##            year = to.idx(mod_data$year),
+##            yr_site = to.idx(mod_data$yr_site_fact),
+##            yield = mod_data$yield_cs)
 
 ## Rprof("mod.Rprof")
-fit <- sampling(mm, data = dd, iter = 100, cores = 1L, control = list(adapt_delta = 0.9))
-## Rprof(NULL)
+## yrs_fit <- sampling(mm, data = c(dd, list(x = mod_data$yrs_in_trial)),
+##                     iter = 500, cores = 2L, control = list(adapt_delta = 0.9))
[…]

### 2017-02-15 src/prep_model_data.R

@@ -1,17 +1,8 @@
 ## Prepares data for write-up
-
 load("../data/all_vt_weather.Rda")
 
 ## Replace 08Y3269 with M209
 all_vt_result$id[all_vt_result$id == "08Y3269"] <- "M209"
-
-##fit <- read_stan_csv(list.files("~/Dropbox/vt_results", "^samples[0-9].csv",
-##                                full.names = TRUE))
-
-## Get the number of years in the trial
-n_years <- aggregate(year ~ id, data = all_vt_result,
-          function(x) {
-              length(unique(x))})
 
 ## Subset to medium grains
 vars <- grepl("^M[0-9]{3}$", all_vt_result$id)
@@ -21,11 +12,22 @@ vars <- grepl("^M[0-9]{3}$", all_vt_result$id)
 no_sj <- all_vt_result$county != "SAN JOAQUIN"
 mod_data <- all_vt_result[(vars & no_sj),]
 
+## Exclude M-401/402 and Tucker site, per conversation with Bruce
+## 10 Feb 2017:
+## M-401 and M-402 are often mis-managed in trials
+## Tucker was a problematic site with high variability and low yields
+not4s <- !mod_data$id %in% c("M401","M402")
+notTuck <- mod_data$site != "TUCKER"
+
+mod_data <- mod_data[(notTuck & not4s),]
+
 ## Use officially published years instead of years in trial
 load("../data/yrTbl.Rda")
 i <- match(mod_data$id, yrTbl$Cultivar)
 
 mod_data$release_yr <- yrTbl$Year[i]
+mod_data$release_yr_c <- mod_data$release_yr - 2000
+
 mod_data$yrs_in_trial <- mod_data$year - mod_data$release_yr
 
 ## Get some centered and scaled vars for model

### 2017-04-25 output/breed_paper.Rnw
Draft of paper before removing report details

--- 2017-02-08 output/summary.Rnw
@@
 \usepackage[margin=1in]{geometry}
+\usepackage[font=it]{caption}
 \usepackage{Sweave}
@@
 
-\section{Background}
+\section{Abstract}
 
-In general, rice yields in CA have been flat over the last 20+ years,
-compared to the preceding period. Yet, annual reports from the CA
+Using the variety trial data, we find consistent estimates of the
+reduction of yield once a variety is released, estimated to be between
+-12 to -22 kg per year. This effect persists using all data, only
+named/released cultivars, or only released cultivars only at the Rice
+experiment station. This effect is offset by yield improvements over
+the years, estimated to be small but positive effect of
+
+\section{Introduction}
+
+Constraints on ariable land are increasing simultaneous with the need to
+increase total production to meet growing demand, which has necessitated
+harvesting more grain per unit land area. Historically,
+agricultural research has been successful in staving off the Mathesian
+disaster through continued yield improvement of staple crops, however more and more
+production systems are experiencing
+plateaus in grain yield \citep{cassman}. If continued improvements per
+unit land area are no longer possible, increasing the area under
+cultivation will be needed to meet demand, which carries with it
+undesirable ecological implications. Therefore, it is of critical
+importance to better understand why grain yields have not continued to
+increase.
+
+CA rice production represents one such rice production
+system. Rice yields in CA during the period from 1920 - 1990 increased rapidly,
+but since 1990 have increased only slowly if at all (Fig. 1
+\ref{fig:avg_yields}). However, continued improvement of grain yield as the
+result of plant breeding has been demonstrated since 1990
+\citep{rrb}, with each new cultivar havings a yield advantage over the
+previous cultivars. We sought to better understand how grain yields
+have increased in the breeders' fields without a corresponding
+increase in yield in the farmers' fields. Two competing hypotheses
+were tested: (1)
+In the course of
+investigating the drivers of yield in this system, we were compelled
+to explore how yields could increase with each release of improved
+cultivars, yet average yield for the region remain flat. Specifically,
+we offered two competing hypotheses to explain this phenomenon: (1)
+the yield advantage offered by newly developed cultivars is not
+greater than previous cultivars, or (2) yield increases gained through
+plant breeding are being offset by decreases in yield of released
+cultivars.
+
+Yet, annual reports from the CA
 Cooperative Rice Research Foundation show that each released cultivar
@@
 polygon(x = c(yrs_in, rev(yrs_in)), y = c(ints[1,],rev(ints[3,])), col = rgb(0,0,0,0.2), border = NA)
-text(20, 3500, bquote(.(round(fixef(yrs_fit)[2] * 1000, 1))~lbs~year^-1))
+text(30, 4500, bquote(.(round(fixef(yrs_fit)[2] * 1000, 1))~lbs~year^-1))
 
@@
 \caption{There is a small decrease in yield over the time since a
-  cultivar was entered into the trial.}
+  cultivar was entered into the trial using the full data
+  set. However, the effect is small (\Sexpr{round(fixef(yrs_fit)[2] * 1000, 1)}
+  lbs year\textsuperscript{-1}) in relation to the variability
+  between sites and years.}
 \label{fig:yrs}
@@
 
+<<echo=FALSE>>=
+all_CI <- round(posterior_interval(yrs_fit, pars = "yrs_in_trial") * 1000, 1)
+RES_CI <- round(posterior_interval(RES_yrs_fit, pars = "yrs_in_trial") * 1000, 1)
+@
+
 As seen in Fig. \ref{fig:yrs}, there is a small decrease in yield
-over time once a cultivar is released. This decrease is estimated to be
+over time once a cultivar is released, though the estimated effect is
+small compared to the variability in the data set. This decrease is estimated to be
 \Sexpr{round(fixef(yrs_fit)[2] * 1000, 1)} lbs
@@
 an estimated \Sexpr{round(fixef(yrs_fit)[2] * 1000 * 36, 1)} lbs
-acre\textsuperscript{-1} since release.
+acre\textsuperscript{-1} since release. Using just data from the Rice
+Experiment Station (RES), the estimated effect is more drastic
+(an estimated \Sexpr{round(fixef(RES_yrs_fit)[2] * 1000, 1)} lbs
+year\textsuperscript{-1}), but with a wider credible interval
+(\Sexpr{RES_CI[1]} -- \Sexpr{RES_CI[2]} vs \Sexpr{all_CI[1]} --
+\Sexpr{all_CI[2]}).
+
+\begin{figure}[h]
+
+<<fig=TRUE, echo=FALSE>>=
+yrs_in <- -6:34
+pp <- posterior_predict(RES_yrs_fit, re.form = NA, newdata = data.frame(yrs_in_trial = yrs_in))
+pp <-  pp * 1000 + 9000
+
+ints <- apply(pp, 2, quantile, p = c(0.05, 0.5, 0.95))
+RES <- mod_data[,'site'] == "RES"
+
+plot(yield_lb ~ yrs_in_trial, data = mod_data[RES,], pch = 16, col = rgb(0,0,0,0.25),
+     xlab = "Year since entry",
+     ylab = expression("Yield ( lbs acre"^-1~")"))
+lines(x = yrs_in, y = ints[2,], lty = 2) 
+polygon(x = c(yrs_in, rev(yrs_in)), y = c(ints[1,],rev(ints[3,])), col = rgb(0,0,0,0.2), border = NA)
+text(30, 5000, bquote(.(round(fixef(RES_yrs_fit)[2] * 1000, 1))~lbs~year^-1))
+
+@
+\caption{The relationship between yield and years since entry for only
+a single site, the Rice Experiment Station (RES). RES is considered
+the best managed site, and has the most plots per year of any of the
+sites. The estimated decrease over time is estimated to be stronger
+using only these data (\Sexpr{round(fixef(RES_yrs_fit)[2] * 1000, 1)}
+lbs year\textsuperscript{-1}), primarly due to lower variability.}
+\label{fig:RES_yrs}
+\end{figure}
+
+\begin{figure}
+<<fig=TRUE, fig.width = 8, fig.height = 5, out.width = '80%', echo=FALSE>>=
+tars <- c("M202","M206")
+par(mfrow = c(1, length(tars)),
+    mar = c(4,0,2,0),
+    oma = c(4,4,0.5,0.5))
+
+invisible(sapply(tars, function(x){
+    plot(yield_lb ~ year, data = mod_data[(mod_data$id == x & mod_data$site == "RES"),],
+         xlim = c(1995, 2016), ylim = c(5000,14000),
+         pch = 16, main = x,
+         ylab = "",
+         yaxt = "n",
+         xlab = "Year")
+    if(x == "M202"){
+        axis(2)
+        mtext(expression("Yield ( lbs acre"^-1~")"), 2, line = 2.25, cex.lab = 0.6)
+    }
+    abline(lm(yield_lb ~ year, data = mod_data[(mod_data$id == x & mod_data$site == "RES"),]), lty = 2)
+
+}))
+
+@
+\caption{The performance of two major cultivars at the RES site over
+time. There is evidence that a cultivar's performance decreases over
+time.}
+\label{fig:res_yrs}
+\end{figure}
+
+\begin{figure}
+
+<<fig=TRUE, echo=FALSE>>=
+par(mfrow = c(1, length(tars)),
+    mar = c(4,0,2,0),
+    oma = c(4,4,0.5,0.5))
+
+invisible(sapply(tars, function(x){
+    plot(yield_lb ~ yrs_in_trial,
+         data = mod_data[(mod_data$id == x & mod_data$site == "RES"),],
+         xlim = c(-6, 36), ylim = c(5000,14000),
+         pch = 16, main = x,
+         ylab = "",
+         yaxt = "n",
+         xlab = "Years after release")
+    if(x == "M202"){
+        axis(2)
+        mtext(expression("Yield ( lbs acre"^-1~")"), 2, line = 2.25, cex.lab = 0.6)
+    }
+    abline(lm(yield_lb ~ yrs_in_trial,
+              data = mod_data[(mod_data$id == x & mod_data$site == "RES"),]), lty = 2)
+}))
+
+@
+\caption{The performance of two major cultivars at the RES site
+  since the year each cultivar was released. Both M-202 and M-206 have
+  similar intercepts, suggesting similar yield at the time of release.}
+\label{fig:res_since_release}
+\end{figure}
 
@@
 polygon(x = c(yrs, rev(yrs)), y = c(ints[1,],rev(ints[3,])), col = rgb(0,0,0,0.2), border = NA)
-text(1990, 3500, bquote(.(round(fixef(release_fit)[2] * 1000, 1))~lbs~year^-1))
+text(2010, 4500, bquote(.(round(fixef(release_fit)[2] * 1000, 1))~lbs~year^-1))
 @
 
-\caption{There is an estimated increase in yield over time, but it is
-  smaller than the previously estimated 54 lbs
+\caption{The effect of year of release on yield. There is an estimated
+  increase in yield over time, but it is smaller than the previously estimated 54 lbs
   year\textsuperscript{-1}}
@@
 \Sexpr{round(fixef(release_fit)[2] * 1000 * 36, 1)} lbs acre\textsuperscript{-1}.
+
+\begin{figure}[h]
+
+<<fig=TRUE, echo=FALSE>>=
+yrs <- 1980:2015
+pp <- posterior_predict(RES_rel_fit, re.form = NA,
+                        newdata = data.frame(release_yr_c = yrs - 2000))
+pp <-  pp * 1000 + 9000
+
+ints <- apply(pp, 2, quantile, p = c(0.05, 0.5, 0.95))
+
+plot(yield_lb ~ release_yr, data = mod_data[RES,], pch = 16, col = rgb(0,0,0,0.25),
+     xlab = "Year of Release",
+     ylab =  expression("Yield ( lbs acre"^-1~")"))
+lines(x = yrs, y = ints[2,], lty = 2) 
+polygon(x = c(yrs, rev(yrs)), y = c(ints[1,],rev(ints[3,])), col = rgb(0,0,0,0.2), border = NA)
+text(2010, 4500, bquote(.(round(fixef(RES_rel_fit)[2] * 1000, 1))~lbs~year^-1))
+@
+
+\caption{The effect of year of release on yield for only data from the
+  Rice Experiment Station (RES). There is an estimated
+  increase in yield over time, but it is smaller than the previously estimated 54 lbs
+  year\textsuperscript{-1}}
+\label{fig:RES_release}
+\end{figure}
 

### 2017-04-28 output/breed_paper.Rnw
Updates figures

--- 2017-02-08 output/summary.Rnw
@@
-\documentclass{article}
+\documentclass[12pt,]{elsarticle}
 \usepackage[margin=1in]{geometry}
+\usepackage{hyperref}
+\usepackage{amsmath}
+\usepackage{longtable,booktabs}
+\usepackage{graphicx,grffile, textcomp}
+\usepackage{setspace}
+\usepackage{etoolbox}
+\doublespacing
+\usepackage{lineno}
+\usepackage{lscape} %% Harvard
+\bibliographystyle{model2-names}\biboptions{authoryear}
+\pagenumbering{gobble}
+%% For figure numbers in supplementary info
+%% from http://bytesizebio.net/2013/03/11/adding-supplementary-tables-and-figures-in-latex/
+\newcommand{\beginsupplement}{%
+  \setcounter{table}{0}
+  \renewcommand{\thetable}{S\arabic{table}}%
+  \setcounter{figure}{0}
+  \renewcommand{\thefigure}{S\arabic{figure}}%
+}
 \usepackage{Sweave}
-\linespread{1.3}
-
-\title{Summary of CA Rice Breeding}
-\author{Matthew Espe}
+\SweaveOpts{prefix.string=graphics/report}
 
@@
 
-\maketitle
+\begin{frontmatter}
+
+\title{Rice yield improvement by plant breeding are offset by
+  decreases over time}
+\author[ucd]{Matthew B. Espe\corref{corresauthor}}
+\cortext[corresauthor]{Corresponding author} \ead{mespe@ucdavis.edu}
+
+\address[ucd]{Dept. of Plant Sciences, University of California --
+Davis, Davis CA 95616}
+
+\begin{abstract}
+
+  Using the variety trial data, we find consistent estimates of the
+  reduction of yield once a variety is released, estimated to be between
+  -12 to -22 kg per year. This effect persists using all data, only
+  named/released cultivars, or only released cultivars only at the Rice
+  experiment station. This effect is offset by yield improvements over
+  the years, estimated to be small but positive effect of
+
+\end{abstract}
+
+\begin{keyword} rice, yield potential, temperature
+\end{keyword}
+
+\end{frontmatter}
+
+\newpage
+
+\textbf{Abbreviations}
+
+CA: California
+
+CI: Credible Interval
+
+\newpage
+
+%\linenumbers
 
@@
 
-\section{Background}
+\section{Introduction}\label{intro}
 
-In general, rice yields in CA have been flat over the last 20+ years,
-compared to the preceding period. Yet, annual reports from the CA
-Cooperative Rice Research Foundation show that each released cultivar
-has a yield advantage over the previous cultivars, typically on the
-order of 4--6\%. This raises the question of how to reconcile these
-facts. Namely, there are two possibilities:
+\setlength{\parindent}{1cm}
 
-\begin{enumerate}
-\item{There is a mistake in the annual reports, and there is little or no
-    yield advantage of newer cultivars over old.}
-\item{The comparison to old cultivars is flawed, allowing an
-    apparent yield increase to manifest without an accompanied
-    increase in statewide yields.}
-\end{enumerate}
+Constraints on ariable land are increasing simultaneous with the need
+to increase total production to meet growing demand, which has
+necessitated harvesting more grain per unit land area. Historically,
+agricultural research has been successful in staving off the Mathesian
+disaster through continued yield improvement of staple crops, however
+more and more production systems are experiencing plateaus in grain
+yield \citep{Peng2010}. If continued improvements per unit land area
+are no longer possible, increasing the area under cultivation will be
+needed to meet demand, which carries with it undesirable ecological
+implications. Therefore, it is of critical importance to better
+understand why grain yields have not continued to increase.
 
-\subsection{Motivating Evidence}
+CA rice production represents one such rice production system. Rice
+yields in CA during the period from 1920 -- 1990 increased rapidly, but
+since 1990 have increased only slowly if at all (Fig. 1
+\ref{fig:avg_yields}). However, continued improvement of grain yield
+as the result of plant breeding has been demonstrated since 1990
+\citep{rrb}, with each new cultivar having a yield advantage over the
+previous cultivars. Yet, these same reports show stagnant yields; for
+example, at the year of release (1984), cultivar 'M-202' averaged
+9,600 lbs acre\textsuperscript{-1}. In 2015, the yield of 'M-209' was
+between 9,000 -- 10,000 lbs acre\textsuperscript{-1}. We sought to better
+understand how grain yields have increased in the breeders' fields
+without a corresponding increase in yield in the farmers' fields. Two
+competing hypotheses were tested: (1) Grain yields have not improved
+since 1990, or (2) grain yield improvements have been offset by an
+associated decrease in yield once a cultivar was selected.
 
-\begin{itemize}
-\item At the year of release (1984), M-202 averaged 9,600 lbs acre\textsuperscript{-1}.
-\item In 2015, the yield of M-209 was between 9,000-10,000 lbs acre\textsuperscript{-1}.
-\item Therefore, there does not seem to have been an appreciable
-  increase in yield over this period.
-\end{itemize}
+\begin{figure}[h]
+<<fig=TRUE, echo=FALSE>>=
+CA_data = read.csv("../data/CA_rice_yield.csv", stringsAsFactors = FALSE)
+CA_data$Value = as.numeric(gsub(",", "", CA_data$Value))
+plot(Value * 1.12 ~ Year, CA_data,
+     type = "l",
+     xlab = "Year",
+     ylab = expression("Average grain yield (kg ha"^-1~")"))
 
-\subsection{Hypotheses}
+@
+\caption{Average rice yields over the period \Sexpr{min(CA_data$Year)}
+  -- \Sexpr{max(CA_data$Year)} for California, USA. Rapid gains in
+    grain yield were experienced from approximately 1950 -- 1990, but
+    annual gains since 1990 have been slower.}
+\label{fig:state_yield}
+\end{figure}
 
-\begin{enumerate}
-  \item H1: After a cultivar has been released, its yield decreases over
-    time.
-  \item H2: Each cultivar release is recovering to the old yield benchmark, not
-    increasing yield potential.
-\end{enumerate}
+In the course of investigating the drivers of yield in this system, we
+were compelled to explore how yields could increase with each release
+of improved cultivars, yet average yield for the region remain
+flat. Specifically, we offered two competing hypotheses to explain
+this phenomenon: (1) the yield advantage offered by newly developed
+cultivars is not greater than previous cultivars, or (2) yield
+increases gained through plant breeding are being offset by decreases
+in yield of released cultivars.
 
-\section{H1: Do yields decreases over time once a cultivar is released?}
+Yet, annual reports from the CA Cooperative Rice Research Foundation
+show that each released cultivar has a yield advantage over the
+previous cultivars, typically on the order of 4 -- 6\%. This raises the
+question of how to reconcile these facts. Namely, there are two
+possibilities:
 
-\subsection{Methods}
+\section{Methods}\label{methods}
+
+\subsection{Site description}
+
+\subsection{Statistical analysis}
 
@@
 
-$$ yield = \alpha + \beta_{years\ in\ trial}* x_{years\ in\ trials} + \beta_{site} + \beta_{year} + \beta_{site:year} + \beta_{cultivar}$$
+\begin{equation}
+ yield = \alpha + \beta_{years\ in\ trial}* x_{years\ in\ trials} + \beta_{site} + \beta_{year} + \beta_{site:year} + \beta_{cultivar}
+\end{equation}
 
@@
 
-\subsection{Results}
-
-\begin{figure}[h]
-
-<<fig=TRUE, echo=FALSE>>=
-yrs_in <- -6:34
-pp <- posterior_predict(yrs_fit, re.form = NA, newdata = data.frame(yrs_in_trial = yrs_in))
-pp <-  pp * 1000 + 9000
-
-ints <- apply(pp, 2, quantile, p = c(0.05, 0.5, 0.95))
-
-plot(yield_lb ~ yrs_in_trial, data = mod_data, pch = 16, col = rgb(0,0,0,0.25),
-     xlab = "Year since entry",
-     ylab = expression("Yield ( lbs acre"^-1~")"))
-lines(x = yrs_in, y = ints[2,], lty = 2) 
-polygon(x = c(yrs_in, rev(yrs_in)), y = c(ints[1,],rev(ints[3,])), col = rgb(0,0,0,0.2), border = NA)
-text(20, 3500, bquote(.(round(fixef(yrs_fit)[2] * 1000, 1))~lbs~year^-1))
-
-@ 
-\caption{There is a small decrease in yield over the time since a
-  cultivar was entered into the trial.}
-\label{fig:yrs}
-\end{figure}
-
-As seen in Fig. \ref{fig:yrs}, there is a small decrease in yield
-over time once a cultivar is released. This decrease is estimated to be
-\Sexpr{round(fixef(yrs_fit)[2] * 1000, 1)} lbs
-year\textsuperscript{-1}. The oldest cultivar in the trials has lost
-an estimated \Sexpr{round(fixef(yrs_fit)[2] * 1000 * 36, 1)} lbs
-acre\textsuperscript{-1} since release.
-
-\section{H2: Do newer cultivars have a yield advantage?}
-
-\subsection{Methods}
-
 A similar but separate model was used to estimate whether newer
@@
 
-$$ yield = \alpha + \beta_{release\ year}*x_{release\ year} +
-\beta_{site} + \beta_{year} + \beta_{site:year} + \beta_{cultivar}$$
+\begin{equation}
+yield = \alpha + \beta_{release\ year}*x_{release\ year} +
+\beta_{site} + \beta_{year} + \beta_{site:year} + \beta_{cultivar}
+\end{equation}
 
@@
 
-\subsection{Results}
+\section{Results}\label{results}
+
+\subsection{Yield improvements over time}
 
@@
 yrs <- 1980:2015
-pp <- posterior_predict(release_fit, re.form = NA, newdata = data.frame(release_yr = yrs))
-pp <-  pp * 1000 + 9000
+pp <- posterior_predict(release_fit, re.form = NA, newdata = data.frame(release_yr_c = yrs - 2000))
+pp <-  (pp * 1000 + 9000) * 1.12
 
@@
 
-plot(yield_lb ~ release_yr, data = mod_data, pch = 16, col = rgb(0,0,0,0.25),
-     xlab = "Year of Release",
-     ylab =  expression("Yield ( lbs acre"^-1~")"))
-lines(x = yrs, y = ints[2,], lty = 2) 
-polygon(x = c(yrs, rev(yrs)), y = c(ints[1,],rev(ints[3,])), col = rgb(0,0,0,0.2), border = NA)
-text(1990, 3500, bquote(.(round(fixef(release_fit)[2] * 1000, 1))~lbs~year^-1))
+res = mod_data$site == "RES"
+plot_data = as.data.frame(tapply(mod_data$yield_lb[res] * 1.12,
+                                 mod_data$id[res], mean))
+plot_data$se = tapply(mod_data$yield_lb[res] * 1.12,
+                      mod_data$id[res],
+                      function(x)
+                          sd(x)/sqrt(length(x)))
+
+plot_data$tmp = tapply(mod_data$release_yr[res],
+                       mod_data$id[res], unique)
+colnames(plot_data) = c("yield_kg", "yield_se", "release_yr")
+
+plot(yield_kg ~ release_yr, data = plot_data, pch = 16, #col = rgb(0,0,0,0.25),
+     xlab = "Release year",
+     ylab = expression("Grain yield ( kg ha"^-1~")"),
+     ylim = c(8000, 13000),
+     xlim = c(1980,2019))
+abline(a = (((fixef(RES_rel_fit)[1] * 1000) + 9000) * 1.12) -
+           (2000 * fixef(RES_rel_fit)[2] * 1000),
+       b = fixef(RES_rel_fit)[2] * 1000, col = "grey")
+with(plot_data, segments(x0 = release_yr, y0=yield_kg + yield_se,
+                         y1 = yield_kg - yield_se))
+text(x = plot_data$release_yr, y = plot_data$yield_kg,
+     labels = gsub("M", "M-", row.names(plot_data)),
+     pos = 4, cex = 0.8)
+
+# ints <- apply(pp, 2, quantile, p = c(0.05, 0.5, 0.95))
+# polygon(x = c(yrs_in, rev(yrs_in)), y = c(ints[1,],rev(ints[3,])), col = rgb(0,0,0,0.2), border = NA)
+# text(0, 7000, bquote(.(round(fixef(RES_yrs_fit)[2] * 1000, 1))~lbs~year^-1))
+
+# plot(yield_lb ~ release_yr, data = mod_data, pch = 16, col = rgb(0,0,0,0.25),
+#      xlab = "Year of Release",
+#      ylab =  expression("Yield ( lbs acre"^-1~")"))
+# lines(x = yrs, y = ints[2,], lty = 2) 
+# polygon(x = c(yrs, rev(yrs)), y = c(ints[1,],rev(ints[3,])), col = rgb(0,0,0,0.2), border = NA)
+text(1990, 8000, bquote(.(round(fixef(release_fit)[2] * 1000 * 1.12, 1))~kg~ha^-1~year^-1))
 @
 
-\caption{There is an estimated increase in yield over time, but it is
-  smaller than the previously estimated 54 lbs
-  year\textsuperscript{-1}}
+\caption{The yield increase over time of medium grain rice varieties
+  in California, USA. There is an estimated
+  increase in yield over time, but it is smaller than a previously
+  estimate of 54 lbs year\textsuperscript{-1}. This estimate is
+  potentially misleading, because it only
+  considers the average yield of each variety during the period 1995
+  -- 2015, and the older varieties have experiencing greater losses
+  since being released.}
 \label{fig:release}
+\end{figure}
+
+\subsection{Yield decreases over time}
+
+% \begin{figure}[h]
+
+% <<fig=TRUE, echo=FALSE>>=
+% yrs_in <- -6:34
+% pp <- posterior_predict(yrs_fit, re.form = NA, newdata = data.frame(yrs_in_trial = yrs_in))
+% pp <-  (pp * 1000 + 9000) * 1.12
+
+% ints <- apply(pp, 2, quantile, p = c(0.05, 0.5, 0.95))
+
+% plot_data = as.data.frame(tapply(mod_data$yield_lb * 1.12, mod_data$yrs_in_trial, mean))
+% plot_data$se = tapply(mod_data$yield_lb * 1.12, mod_data$yrs_in_trial, function(x)
+%     sd(x)/sqrt(length(x)))
+% plot_data$tmp = as.numeric(row.names(plot_data))
+% colnames(plot_data) = c("yield_kg", "yield_se", "yrs_in_trial")
+
+% plot(yield_kg ~ yrs_in_trial, data = plot_data, pch = 16, #col = rgb(0,0,0,0.25),
+%      xlab = "Year since entry",
+%      ylab = expression("Yield ( kg ha"^-1~")"),
+%      ylim = c(8700, 11200))
+% with(plot_data, segments(x0 = yrs_in_trial, y0=yield_kg + yield_se,
+%                          y1 = yield_kg - yield_se))
+% abline(a = ((fixef(yrs_fit)[1] * 1000) + 9000) * 1.12, b = fixef(yrs_fit)[2] * 1000)
+% # lines(x = yrs_in, y = ints[2,], lty = 2) 
+% # polygon(x = c(yrs_in, rev(yrs_in)), y = c(ints[1,],rev(ints[3,])), col = rgb(0,0,0,0.2), border = NA)
+% text(30, 4500, bquote(.(round(fixef(yrs_fit)[2] * 1000, 1))~lbs~year^-1))
+
+% @ 
+% \caption{There is a small decrease in yield over the time since a
+  % cultivar was entered into the trial using the full data
+  % set. However, the effect is small (\Sexpr{round(fixef(yrs_fit)[2] * 1000, 1)}
+  % lbs year\textsuperscript{-1}) in relation to the variability
+  % between sites and years.}
+% \label{fig:yrs}
+% \end{figure}
+
+<<echo=FALSE>>=
+all_CI <- round(posterior_interval(yrs_fit, pars = "yrs_in_trial") * 1000, 1)
+RES_CI <- round(posterior_interval(RES_yrs_fit, pars = "yrs_in_trial") * 1000, 1)
+@
+
+As seen in Fig. \ref{fig:yrs}, there is a small decrease in yield
+over time once a cultivar is released, though the estimated effect is
+small compared to the variability in the data set. This decrease is estimated to be
+\Sexpr{round(fixef(yrs_fit)[2] * 1000, 1)} lbs
+year\textsuperscript{-1}. The oldest cultivar in the trials has lost
+an estimated \Sexpr{round(fixef(yrs_fit)[2] * 1000 * 36 * 1.12, 1)} kg ha\textsuperscript{-1} since release. Using just data from the Rice
+Experiment Station (RES), the estimated effect is more drastic
+(an estimated \Sexpr{round(fixef(RES_yrs_fit)[2] * 1000 * 1.12, 1)} kg
+ha\textsuperscript{-1} year\textsuperscript{-1}), but with a wider credible interval
+(\Sexpr{RES_CI[1]} -- \Sexpr{RES_CI[2]} vs \Sexpr{all_CI[1]} --
+\Sexpr{all_CI[2]}).
+
+\begin{figure}[h]
+
+<<fig=TRUE, echo=FALSE>>=
+yrs_in <- -7:35
+pp <- posterior_predict(RES_yrs_fit, re.form = NA, newdata = data.frame(yrs_in_trial = yrs_in))
+pp <-  (pp * 1000 + 9000) * 1.12
+
+RES = res = mod_data$site == "RES"
+plot_data = as.data.frame(tapply(mod_data$yield_lb[res] * 1.12,
+                                 mod_data$yrs_in_trial[res], mean))
+plot_data$se = tapply(mod_data$yield_lb[res] * 1.12,
+                      mod_data$yrs_in_trial[res],
+                      function(x)
+                          sd(x)/sqrt(length(x)))
+
+plot_data$tmp = as.numeric(row.names(plot_data))
+colnames(plot_data) = c("yield_kg", "yield_se", "yrs_in_trial")
+
+plot(yield_kg ~ yrs_in_trial, data = plot_data, pch = 16, #col = rgb(0,0,0,0.25),
+     xlab = "Year since entry",
+     ylab = expression("Grain yield ( kg ha"^-1~")"),
+     ylim = c(7000, 13000),
+     xlim = c(-5,33))
+with(plot_data, segments(x0 = yrs_in_trial, y0=yield_kg + yield_se,
+                         y1 = yield_kg - yield_se))
+abline(a = ((fixef(RES_yrs_fit)[1] * 1000) + 9000) * 1.12,
+       b = fixef(RES_yrs_fit)[2] * 1000)
+ints <- apply(pp, 2, quantile, p = c(0.05, 0.5, 0.95))
+#polygon(x = c(yrs_in, rev(yrs_in)), y = c(ints[1,],rev(ints[3,])), col = rgb(0,0,0,0.2), border = NA)
+text(0, 7000, bquote(.(round(fixef(RES_yrs_fit)[2] * 1000 * 1.12, 1))~kg~ha^-1~year^-1))
+
+@
+\caption{The relationship between yield and the years since entry for
+  medium grain rice varieties planted at the Rice Experiment Station
+  (RES), California, USA. RES is considered
+  the best managed site, and has the most plots per year of any of the
+  sites. The estimated decrease over time is estimated to be stronger
+  using only these data (\Sexpr{round(fixef(RES_yrs_fit)[2] * 1000, 1)}
+  lbs year\textsuperscript{-1}), primarly due to lower variability.}
+\label{fig:RES_yrs}
+\end{figure}
+
+\begin{figure}
+<<fig=TRUE, fig.width = 10, fig.height = 5, out.width = '80%', echo=FALSE>>=
+tars <- c("M202","M206")
+par(mfrow = c(1, length(tars)),
+    mar = c(4,0,2,0),
+    oma = c(4,4,0.5,0.5))
+
+invisible(sapply(tars, function(x){
+    plot(yield_lb * 1.12 ~ year,
+         data = mod_data[(mod_data$id == x & mod_data$site == "RES"),],
+         xlim = c(1995, 2016), ylim = c(5000,14000),
+         pch = 16, main = gsub("M", "M-", x),
+         ylab = "",
+         yaxt = "n",
+         xlab = "Year")
+    if(x == "M202"){
+        axis(2)
+        mtext(expression("Yield ( kg ha"^-1~")"), 2, line = 2.25, cex.lab = 0.6)
+    }
+    # abline(lm(yield_lb ~ year, data = mod_data[(mod_data$id == x & mod_data$site == "RES"),]), lty = 2)
+    abline(h = mean(mod_data$yield_lb[(mod_data$id == x & mod_data$site == "RES")] * 1.12),
+           lty = 2)
+}))
+
+@
+\caption{The performance of two major cultivars at the CA Rice
+  Experiment Station (RES) over the period 1995 -- 2015. The dashed line is the average
+  yield for each cultivar over the period which it was planted at the
+  RES. The newer cultivar, M-209, appears to have a slight yield
+  advantage over the older cultivar, M-202.}
+\label{fig:res_yrs}
+\end{figure}
+
+\begin{figure}
+
+<<fig=TRUE, echo=FALSE>>=
+par(mfrow = c(1, length(tars)),
+    mar = c(4,0,2,0),
+    oma = c(4,4,0.5,0.5))
+
+invisible(sapply(tars, function(x){
+    plot(yield_lb*1.12 ~ yrs_in_trial,
+         data = mod_data[(mod_data$id == x & mod_data$site == "RES"),],
+         xlim = c(-6, 36), ylim = c(5000,14000),
+         pch = 16, main = gsub("M","M-", x),
+         ylab = "",
+         yaxt = "n",
+         xlab = "Years after release")
+    if(x == "M202"){
+        axis(2)
+        mtext(expression("Yield ( kg ha"^-1~")"), 2, line = 2.25, cex.lab = 0.6)
+    }
+    abline(lm(yield_lb*1.12 ~ yrs_in_trial,
+              data = mod_data[(mod_data$id == x & mod_data$site == "RES"),]), lty = 2)
+}))
+
+@
+\caption{The performance of two major cultivars at the CA Rice
+  Experiment Station (RES) since the year each cultivar was
+  released. Both M-202 and M-206 have similar intercepts, suggesting
+  similar yield at the time of release and an apparent yield advantage
+  of the newer cultivar, M-209, might be explained by a yield loss of
+  M-202 over time.}
+\label{fig:res_since_release}
 \end{figure}
@@
 \Sexpr{round(fixef(release_fit)[2] * 1000 * 36, 1)} lbs acre\textsuperscript{-1}.
+
+% \begin{figure}[h]
+
+% <<fig=TRUE, echo=FALSE>>=
+% yrs <- 1980:2015
+% pp <- posterior_predict(RES_rel_fit, re.form = NA,
+%                         newdata = data.frame(release_yr_c = yrs - 2000))
+% pp <-  pp * 1000 + 9000
+
+% ints <- apply(pp, 2, quantile, p = c(0.05, 0.5, 0.95))
+
+% plot(yield_lb ~ release_yr, data = mod_data[RES,], pch = 16, col = rgb(0,0,0,0.25),
+%      xlab = "Year of Release",
+%      ylab =  expression("Yield ( lbs acre"^-1~")"))
+% lines(x = yrs, y = ints[2,], lty = 2) 
+% polygon(x = c(yrs, rev(yrs)), y = c(ints[1,],rev(ints[3,])), col = rgb(0,0,0,0.2), border = NA)
+% text(2010, 4500, bquote(.(round(fixef(RES_rel_fit)[2] * 1000, 1))~lbs~year^-1))
+% @
+
+% \caption{The effect of year of release on yield for only data from the
+%   Rice Experiment Station (RES). There is an estimated
+%   increase in yield over time, but it is smaller than the previously estimated 54 lbs
+%   year\textsuperscript{-1}}
+% \label{fig:RES_release}
+% \end{figure}
+
+\section{Discussion}
+
+\subsection{Complications}
+
+There is strong collinearity present between the year of release and
+the number of years in the trials which is complicating the
+simultaneous estimation of these effects. With the current size of the
+data set (N = \Sexpr{nrow(mod_data)}), it is impossible to estimate
+the decrease over time since release and the improvement in yield over
+time with each new release in the same model. This is why each effect was
+estimated using a separate regression.
+
+Increasing the size of the data to the full data set (N =
+\Sexpr{nrow(all_vt_result)}), allows the simultaneous estimation
+and results in similar estimated effect sizes for each of these
+mechanisms. However, this necessitates the inclusion of many entries
+which have characteristics not similar to conventional cultivars.
 
@@
 
-\section{Complications}
-
-There is strong collinearity present between the year of release and
-the number of years in the trials which is complicating the
-simultaneous estimation of these effects. With the current size of the
-data set (N = \Sexpr{nrow(mod_data)}), it is impossible to estimate
-the decrease over time since release and the improvement in yield over
-time with each new release in the same model. This is why each effect was
-estimated using a separate regression.
-
-Increasing the size of the data to the full data set (N =
-\Sexpr{nrow(all_vt_result)}), allows the simultaneous estimation
-and results in similar estimated effect sizes for each of these
-mechanisms. However, this necessitates the inclusion of many entries
-which have characteristics not similar to conventional cultivars.
-
 \newpage
@@
 
+\section{References}\label{references}
+
+\setlength{\parindent}{-0.2in} \setlength{\leftskip}{0.2in}
+\setlength{\parskip}{8pt}
+
+\bibliography{breeding}
+
+\newpage
+
+\section{Supplemental tables and
+figures}\label{supplemental-tables-and-figures}
+
+\beginsupplement
+
 \end{document}

### 2017-04-28 output/summary.Rnw
Updates figures

--- 2017-02-08 output/summary.Rnw
@@
 
+\section{Summary}
+
+Using the variety trial data, we find consistent estimates of the
+reduction of yield once a variety is released, estimated to be between
+-12 to -22 kg per year. This effect persists using all data, only
+named/released cultivars, or only released cultivars only at the Rice
+experiment station. This effect is offset by yield improvements over
+the years, estimated to be small but positive effect of
+
 \section{Background}
 
+Increased constraints on ariable land simultaneous with the need to
+increase total production to meet growing demand have necessitated
+harvesting more grain per unit land area over time. Historically,
+agricultural research has been successful in staving off the Mathesian
+disaster through continued yield improvement, however more and more
+production systems are experiencing
+plateaus in grain yield \citep{cassman}. If continued improvements per
+unit land area are no longer possible, increasing the total land under
+cultivation will be needed to meet demand, which carries with it
+undesirable ecological implications. Therefore, it is of critical
+importance to better understand why grain yields have not continued to
+increase.
+
 In general, rice yields in CA have been flat over the last 20+ years,
-compared to the preceding period. Yet, annual reports from the CA
+compared to the preceding period, despite a demonstrated improvement
+in grain yield via plant breeding \citep{rrb}. In the course of
+investigating the drivers of yield in this system, we were compelled
+to explore how yields could increase with each release of improved
+cultivars, yet average yield for the region remain flat. Specifically,
+we offered two competing hypotheses to explain this phenomenon: (1)
+the yield advantage offered by newly developed cultivars is not
+greater than previous cultivars, or (2) yield increases gained through
+plant breeding are being offset by decreases in yield of released
+cultivars.
+
+Yet, annual reports from the CA
 Cooperative Rice Research Foundation show that each released cultivar

### 2017-04-28 src/prep_model_data.R

@@ -14,12 +14,14 @@ mod_data <- all_vt_result[(vars & no_sj),]
 
 ## Exclude M-401/402 and Tucker site, per conversation with Bruce
 ## 10 Feb 2017:
-## M-401 and M-402 are often mis-managed in trials
+## M-401 and M-402 are often mis-managed in trials, however they are also old varieties
+## Question if excluding them is not creating more issues
+## not4s <- !mod_data$id %in% c("M401","M402")
+
 ## Tucker was a problematic site with high variability and low yields
-not4s <- !mod_data$id %in% c("M401","M402")
 notTuck <- mod_data$site != "TUCKER"
 
-mod_data <- mod_data[(notTuck & not4s),]
+mod_data <- mod_data[(notTuck),]
 
 ## Use officially published years instead of years in trial
 load("../data/yrTbl.Rda")

### 2017-04-30 output/breed_paper.Rnw
Adjusts figure 1

@@
 
-\title{Rice yield improvement by plant breeding are offset by
+\title{Rice yield improvements by plant breeding are offset by
   decreases over time}
@@
 
-%\linenumbers
+\linenumbers
 
@@
 
+prep_plot_data = function(x, idx)
+{
+    as.data.frame(do.call(rbind, tapply(x, idx, function(x){
+        c(mean = mean(x), se = (sd(x)/sqrt(length(x))))
+    })))
+}
+
 @
@@
 Constraints on ariable land are increasing simultaneous with the need
-to increase total production to meet growing demand, which has
-necessitated harvesting more grain per unit land area. Historically,
+to increase total production to meet growing demand \citep{Foley2011a,
+  Godfray2010, Mueller2012}, which has
+necessitated harvesting more grain per unit land area
+\citep{Lobell2009, Tittonell2014}. Historically,
 agricultural research has been successful in staving off the Mathesian
@@
 more and more production systems are experiencing plateaus in grain
-yield \citep{Peng2010}. If continued improvements per unit land area
+yield \citep{Grassini2013, Peng2010}. If continued improvements per unit land area
 are no longer possible, increasing the area under cultivation will be
@@
 implications. Therefore, it is of critical importance to better
-understand why grain yields have not continued to increase.
+understand why grain yields have not continued to increase in
+instensified production systems.
 
-CA rice production represents one such rice production system. Rice
-yields in CA during the period from 1920 -- 1990 increased rapidly, but
-since 1990 have increased only slowly if at all (Fig. 1
-\ref{fig:avg_yields}). However, continued improvement of grain yield
+California (CA) rice production represents one such rice production
+system. Rice yields in CA during the period from 1920 -- 1990
+increased rapidly, but since 1990 have increased more slowly (Fig. 1
+\ref{fig:state_yield}). However, continued improvement of grain yield
 as the result of plant breeding has been demonstrated since 1990
@@
 previous cultivars. Yet, these same reports show stagnant yields; for
-example, at the year of release (1984), cultivar 'M-202' averaged
-9,600 lbs acre\textsuperscript{-1}. In 2015, the yield of 'M-209' was
-between 9,000 -- 10,000 lbs acre\textsuperscript{-1}. We sought to better
-understand how grain yields have increased in the breeders' fields
-without a corresponding increase in yield in the farmers' fields. Two
-competing hypotheses were tested: (1) Grain yields have not improved
-since 1990, or (2) grain yield improvements have been offset by an
-associated decrease in yield once a cultivar was selected.
+example, at the year of release (1985), cultivar 'M-202' averaged
+11,060 kg ha\textsuperscript{-1} \citep[14\% MC;][]{m202}. In 2015,
+the average yield of the newest released cultivar 'M-209' was 10,841
+kg ha\textsuperscript{-1} \citep{m209}. This suggests little to no
+progress via breeding, however 'M-209' demonstrated an overall 10\%
+yield advantage over 'M-202' in side-by-side comparisons \citep{m209}. We sought to
+better understand how grain yields have increased in the breeders'
+fields without a corresponding increase in yield in the farmers'
+fields. Two competing hypotheses were tested: (1) Grain yields have
+not improved since 1990, or (2) grain yield improvements have been
+offset by an associated decrease in yield once a cultivar was
+selected.
 
-\begin{figure}[h]
+\begin{figure}[ht]
 <<fig=TRUE, echo=FALSE>>=
@@
 CA_data$Value = as.numeric(gsub(",", "", CA_data$Value))
+CA_data = CA_data[order(CA_data$Year),]
 plot(Value * 1.12 ~ Year, CA_data,
-     type = "l",
+     # type = "l",
+     pch = 16,
      xlab = "Year",
-     ylab = expression("Average grain yield (kg ha"^-1~")"))
+     ylab = expression("Average grain yield ( kg ha"^-1~")"))
+mv_avg = filter(c(rep(mean(CA_data$Value[1:10] * 1.12), 5),
+                  CA_data$Value * 1.12,
+                  rep(mean(CA_data$Value[(nrow(CA_data) - 10):nrow(CA_data)] * 1.12), 5)),
+                rep(1/11,11))
+lines(x = c(1907:1911, CA_data$Year, 2017:2021), y = mv_avg)
 
@@
     grain yield were experienced from approximately 1950 -- 1990, but
-    annual gains since 1990 have been slower.}
+    annual gains since 1990 have been slower. The line represents the
+    10 year moving average.}
 \label{fig:state_yield}
@@
 \subsection{Site description}
+
+The Rice Experiment Station (RES) of the California Cooperative Rice
+Research Foundation
 
@@
 res = mod_data$site == "RES"
-plot_data = as.data.frame(tapply(mod_data$yield_lb[res] * 1.12,
-                                 mod_data$id[res], mean))
-plot_data$se = tapply(mod_data$yield_lb[res] * 1.12,
-                      mod_data$id[res],
-                      function(x)
-                          sd(x)/sqrt(length(x)))
+plot_data = prep_plot_data(mod_data$yield_lb[res] * 1.12,
+                           mod_data$id[res])
 
@@
                        mod_data$id[res], unique)
+
 colnames(plot_data) = c("yield_kg", "yield_se", "release_yr")
@@
            (2000 * fixef(RES_rel_fit)[2] * 1000),
-       b = fixef(RES_rel_fit)[2] * 1000, col = "grey")
+       b = fixef(RES_rel_fit)[2] * 1000, col = "grey50")
 with(plot_data, segments(x0 = release_yr, y0=yield_kg + yield_se,
@@
 
-# ints <- apply(pp, 2, quantile, p = c(0.05, 0.5, 0.95))
-# polygon(x = c(yrs_in, rev(yrs_in)), y = c(ints[1,],rev(ints[3,])), col = rgb(0,0,0,0.2), border = NA)
-# text(0, 7000, bquote(.(round(fixef(RES_yrs_fit)[2] * 1000, 1))~lbs~year^-1))
+text(1990, 8000, bquote(.(round(fixef(release_fit)[2] * 1000 * 1.12, 1))~kg~ha^-1~year^-1))
 
-# plot(yield_lb ~ release_yr, data = mod_data, pch = 16, col = rgb(0,0,0,0.25),
-#      xlab = "Year of Release",
-#      ylab =  expression("Yield ( lbs acre"^-1~")"))
-# lines(x = yrs, y = ints[2,], lty = 2) 
-# polygon(x = c(yrs, rev(yrs)), y = c(ints[1,],rev(ints[3,])), col = rgb(0,0,0,0.2), border = NA)
-text(1990, 8000, bquote(.(round(fixef(release_fit)[2] * 1000 * 1.12, 1))~kg~ha^-1~year^-1))
 @
@@
 RES = res = mod_data$site == "RES"
-plot_data = as.data.frame(tapply(mod_data$yield_lb[res] * 1.12,
-                                 mod_data$yrs_in_trial[res], mean))
-plot_data$se = tapply(mod_data$yield_lb[res] * 1.12,
-                      mod_data$yrs_in_trial[res],
-                      function(x)
-                          sd(x)/sqrt(length(x)))
+plot_data = prep_plot_data(mod_data$yield_lb[res] * 1.12,
+                           mod_data$yrs_in_trial[res])
 
@@
      ylab = expression("Grain yield ( kg ha"^-1~")"),
-     ylim = c(7000, 13000),
-     xlim = c(-5,33))
+     ylim = c(7000, 13000))
 with(plot_data, segments(x0 = yrs_in_trial, y0=yield_kg + yield_se,
@@
 abline(a = ((fixef(RES_yrs_fit)[1] * 1000) + 9000) * 1.12,
-       b = fixef(RES_yrs_fit)[2] * 1000)
+       b = fixef(RES_yrs_fit)[2] * 1000, col = "grey50")
 ints <- apply(pp, 2, quantile, p = c(0.05, 0.5, 0.95))
-#polygon(x = c(yrs_in, rev(yrs_in)), y = c(ints[1,],rev(ints[3,])), col = rgb(0,0,0,0.2), border = NA)
-text(0, 7000, bquote(.(round(fixef(RES_yrs_fit)[2] * 1000 * 1.12, 1))~kg~ha^-1~year^-1))
+text(5, 7000, bquote(.(round(fixef(RES_yrs_fit)[2] * 1000 * 1.12, 1))~kg~ha^-1~year^-1))
 
@@
   lbs year\textsuperscript{-1}), primarly due to lower variability.}
-\label{fig:RES_yrs}
+\label{fig:RES_year}
 \end{figure}
 
-\begin{figure}
-<<fig=TRUE, fig.width = 10, fig.height = 5, out.width = '80%', echo=FALSE>>=
+\begin{figure}[ht]
+<<fig=TRUE, width = 8, height = 5, out.width = '85%', echo=FALSE>>=
 tars <- c("M202","M206")
@@
 
-invisible(sapply(tars, function(x){
-    plot(yield_lb * 1.12 ~ year,
-         data = mod_data[(mod_data$id == x & mod_data$site == "RES"),],
-         xlim = c(1995, 2016), ylim = c(5000,14000),
-         pch = 16, main = gsub("M", "M-", x),
+plot_data = lapply(tars, function(x) {
+    dd = mod_data[mod_data$id == x & mod_data$site == "RES",]
+    tmp = prep_plot_data(dd$yield_lb * 1.12, dd$year)
+    tmp$year = as.numeric(row.names(tmp))
+    tmp
+    })
+
+invisible(sapply(1:2, function(i){
+    plot(mean ~ year,
+         data = plot_data[[i]],
+         xlim = c(1995, 2016), ylim = c(6500,14500),
+         pch = 16, main = gsub("M","M-", tars[i]),
          ylab = "",
          yaxt = "n",
-         xlab = "Year")
-    if(x == "M202"){
+         xlab = "Years after release")
+    segments(x0 = plot_data[[i]]$year,
+             y0 = plot_data[[i]]$mean + plot_data[[i]]$se,
+             y1 = plot_data[[i]]$mean - plot_data[[i]]$se)
+    if(i == 1){
         axis(2)
@@
     }
-    # abline(lm(yield_lb ~ year, data = mod_data[(mod_data$id == x & mod_data$site == "RES"),]), lty = 2)
-    abline(h = mean(mod_data$yield_lb[(mod_data$id == x & mod_data$site == "RES")] * 1.12),
-           lty = 2)
+#     abline(lm(yield_lb*1.12 ~ year,
+    #               data = mod_data[(mod_data$id == tars[i] & mod_data$site == "RES"),]),
+    # lty = 2)
+
+    abline(h = mean(mod_data$yield_lb[(mod_data$id == tars[i] &
+                                       mod_data$site == "RES")] * 1.12),
+           lty = 1, col = "grey50")
+
 }))
+
+# invisible(sapply(tars, function(x){
+#     plot(yield_lb * 1.12 ~ year,
+#          data = mod_data[(mod_data$id == x & mod_data$site == "RES"),],
+#          xlim = c(1995, 2016), ylim = c(6500,14500),
+#          pch = 16,
+#          col = rgb(0,0,0,0.5),
+#          main = gsub("M", "M-", x),
+#          ylab = "",
+#          yaxt = "n",
+#          xlab = "Year")
+#     if(x == "M202"){
+#         axis(2)
+#         mtext(expression("Yield ( kg ha"^-1~")"), 2, line = 2.25, cex.lab = 0.6)
+#     }
+#     # abline(lm(yield_lb ~ year, data = mod_data[(mod_data$id == x & mod_data$site == "RES"),]), lty = 2)
+#     abline(h = mean(mod_data$yield_lb[(mod_data$id == x & mod_data$site == "RES")] * 1.12),
+#            lty = 2)
+# }))
 
@@
 
-<<fig=TRUE, echo=FALSE>>=
+<<fig=TRUE, echo=FALSE, width = 8, height = 5>>=
 par(mfrow = c(1, length(tars)),
@@
 
-invisible(sapply(tars, function(x){
-    plot(yield_lb*1.12 ~ yrs_in_trial,
-         data = mod_data[(mod_data$id == x & mod_data$site == "RES"),],
-         xlim = c(-6, 36), ylim = c(5000,14000),
-         pch = 16, main = gsub("M","M-", x),
+plot_data = lapply(tars, function(x) {
+    dd = mod_data[mod_data$id == x & mod_data$site == "RES",]
+    tmp = prep_plot_data(dd$yield_lb * 1.12, dd$yrs_in_trial)
+    tmp$yrs_in_trial = as.numeric(row.names(tmp))
+    tmp
+    })
+
+invisible(sapply(1:2, function(i){
+    plot(mean ~ yrs_in_trial,
+         data = plot_data[[i]],
+         xlim = c(-6, 36), ylim = c(6500,14500),
+         pch = 16, main = gsub("M","M-", tars[i]),
          ylab = "",
@@
          xlab = "Years after release")
-    if(x == "M202"){
+    segments(x0 = plot_data[[i]]$yrs_in_trial,
+             y0 = plot_data[[i]]$mean + plot_data[[i]]$se,
+             y1 = plot_data[[i]]$mean - plot_data[[i]]$se)
+    if(i == 1){
         axis(2)
@@
     abline(lm(yield_lb*1.12 ~ yrs_in_trial,
-              data = mod_data[(mod_data$id == x & mod_data$site == "RES"),]), lty = 2)
+              data = mod_data[(mod_data$id == tars[i] & mod_data$site == "RES"),]),
+           lty = 1, col = "grey50")
 }))
@@
 
-\begin{itemize}
-  \item H1: There is evidence of a small decrease in yield once a cultivar is
-    released. This complicates comparisons between new entries and
-    checks, since the check will decrease over time. This decrease
-    will manifest as an apparent larger yield advantage of the new
-    entry over the check.
-  \item H2: There is evidence of a small increase in yield with each
-    new released cultivar. Therefore, it appears that successive
-    cultivar releases are not merely recovering to the old benchmark
-    but are surpassing it. However, progress is slower than previously
-    estimated (\Sexpr{round(fixef(release_fit)[2] * 1000, 1)}
-    lbs acre\textsuperscript{-1} year\textsuperscript{-1} increase
-    verses a previous estimate of 54 lbs acre\textsuperscript{-1}
+There is evidence of a small decrease in yield over time once a cultivar is
+released. This complicates comparisons between new entries and checks,
+since the check will decrease over time. This decrease will manifest
+as an apparent larger yield advantage of the new entry over the check.
+There is evidence of a small increase in yield with each new released
+cultivar. Therefore, it appears that successive cultivar releases are
+not merely recovering to the old benchmark but are surpassing
+it. However, progress is slower than previously estimated
+(\Sexpr{round(fixef(release_fit)[2] * 1000, 1)} lbs
+acre\textsuperscript{-1} year\textsuperscript{-1} increase verses a
+previous estimate of 54 lbs acre\textsuperscript{-1}
 year\textsuperscript{-1}).
-\end{itemize}
 
-The practical impact of these results is:
+The practical impact of these results is: Breeders need to be aware of
+the potential for the yield of checks to decrease over time. Due to
+this, comparisons to older checks might give a false sense of
+improvement. In order to have greater improvements in yield, the
+target yield improvement for a new cultivar needs to be higher to
+offset the impact of yield decrease once a cultivar is
+released. Breeding is essential to maintaining yield in rice. Yield is
+not stable and without continuous effort, yield potential will
+decrease over time.
 
-\begin{enumerate}
-\item Breeders need to be aware of the potential for the yield of
-  checks to decrease over time. Due to this, comparisons to older checks
-  might give a false sense of improvement. In order to have greater
-  improvements in yield, the target yield improvement for a new cultivar
-  needs to be higher to offset the impact of yield decrease once a
-  cultivar is released.
-\item Breeding is essential to maintaining yield in rice. Yield is not
-  stable and without continuous effort, yield potential will decrease
-  over time.
+\newpage
+\section{References}\label{references}
 
-\end{enumerate}
+\setlength{\parindent}{-0.2in} \setlength{\leftskip}{0.2in}
+\setlength{\parskip}{8pt}
+
+\bibliography{breeding}
 
@@
 
-\section{Supplemental figures}
+\section{Supplemental tables and
+figures}\label{supplemental-tables-and-figures}
+
+\beginsupplement
 
@@
 
-\section{References}\label{references}
-
-\setlength{\parindent}{-0.2in} \setlength{\leftskip}{0.2in}
-\setlength{\parskip}{8pt}
-
-\bibliography{breeding}
-
-\newpage
-
-\section{Supplemental tables and
-figures}\label{supplemental-tables-and-figures}
-
-\beginsupplement
-
 \end{document}

### 2017-05-04 output/breed_paper.Rnw
Adjust to only show RES

@@
 \usepackage{hyperref}
+\usepackage{amssymb}
 \usepackage{amsmath}
@@
 \author[ucd]{Matthew B. Espe\corref{corresauthor}}
+\author[ucd]{David MacKill}
+\author[ucd]{Jim E. Hill}
+\author[res]{Kent McKenzie}
+\author[ucd]{Bruce A. Linquist}
+
 \cortext[corresauthor]{Corresponding author} \ead{mespe@ucdavis.edu}
@@
 Davis, Davis CA 95616}
+
+\address[res]{Rice Experiment Station, California Cooperative Rice
+Research Foundation, Biggs, CA 95917}
 
@@
 
-\subsection{Site description}
+\subsection{Site and data description}
 
 The Rice Experiment Station (RES) of the California Cooperative Rice
-Research Foundation
+Research Foundation (CCRRF) is located northwest of Biggs, CA (39.4648,
+-121.7342), and has been the central location for the development of
+improved rice varieties suited for CA since 1969. The climate at the
+RES is Mediterranian, characterized by mild winters during which most
+of the annual precipitation occurs (\Sexpr{18*24.5} -- \Sexpr{25*24.5}
+mm annually) and warm summers largely free of precipitation
+events. Soils at the RES are classified as Esquon-Neerdobe clays with
+roughly 2\% SOM \citep{ssurgo} in the top 15cm. As the central
+location for the breeding program, the RES is researcher-managed (as
+opposed to farmer-managed trials) and conducts the bulk of the trials
+for the Statewide Variety Trials (approximately 30\% of the total
+plots are planted at RES). For these reasons, we focused our
+investigation on the RES to avoid complications due to site, climate, and
+management differences at the other 6 -- 8 farmer-managed trials.
+
+As part of the annual Statewide Variety Trials, newly developed
+entries are tested against named and released cultivars (checks) at
+the RES in multiple trials spanning multiple planting dates per
+year. For the purposes of this study, we concentrated on released
+medium grain cultivars, as medium grain cultivars constitute roughly
+90\% of the planted area in CA. This lead to a final data set of
+\Sexpr{format(nrow(mod_data), big.mark = ',')} observations with
+\Sexpr{length(unique(mod_data[,'id']))} cultivars represented,
+constituting every publicly released medium grain variety developed by
+the CCRRF from \Sexpr{min(mod_data[,'release_yr'])} --
+\Sexpr{max(mod_data[,'release_yr'])}. For each cultivar, the official
+year of release (typically 3 -- 5 years after introduction to the
+trials but prior to widespread commercial seed availability) and the
+time elapse (in years) since release was calculated.
 
@@
 
-Linear model of yield of a cultivar as a response to the number of years since it
-was entered into the trials. Corrections for year, site, year:site,
-and cultivar differences are made according to:
+Yield was modeled as a response to the number of years since the
+variety was officially released using a mixed-effects linear
+model. Differences between years and cultivars were included in the
+model as random effects according to:
 
 \begin{equation}
- yield = \alpha + \beta_{years\ in\ trial}* x_{years\ in\ trials} + \beta_{site} + \beta_{year} + \beta_{site:year} + \beta_{cultivar}
+ yield_i = \alpha + \beta_{years\ since\ release\ i}* x_{years\ since\
+   release\ i} + \beta_{year\ i} + \beta_{cultivar\ i}
 \end{equation}
 
-Only medium-grained, released cultivars were included in the analysis
-and data from the San Joaquin trials, which are not representative of
-typical CA rice production (frequent exposure to cold, drill-seeded,
-etc.) were excluded. This reduced the data from \Sexpr{nrow(all_vt_result)} to
-\Sexpr{nrow(mod_data)} observations. Included cultivars are:
-\Sexpr{paste(gsub('^M', 'M-', unique(mod_data[,'id'])), collapse = ', ')}. The
-official release years were used for each cultivar for the year of
-release (i.e., not the year seed was first commercially available).
-
-A similar but separate model was used to estimate whether newer
-releases had a yield advantage over older ones:
+Where $yield_i$ is the grain yield for plot $i$, $\alpha$ is the
+average yield for the RES across varieties and years (intercept),
+$\beta_{years\ since\ release\ i}$ is effect of $x_{years\ since\
+release\ i}$, the number of years $i$'s cultivar has been in the
+trials (negative for years prior to release, zero at the year of
+release), $\beta_{year}$ is the random effect of year, and
+$\beta_{cultivar}$ is the random effect of cultivar. A similar but
+separate model was used to estimate the yield advantage of newer
+releases over older ones:
 
 \begin{equation}
-yield = \alpha + \beta_{release\ year}*x_{release\ year} +
-\beta_{site} + \beta_{year} + \beta_{site:year} + \beta_{cultivar}
+yield_i = \alpha + \beta_{release\ year\ i}*x_{release\ year\ i} +
+\beta_{year\ i} + \beta_{cultivar\ i}
 \end{equation}
 
-Again, the estimate was conditioned on differences between cultivars,
-sites, years, and site:years.
+Where $\beta_{release\ year\ i}$ is effect of $x_{release\ year\ i}$,
+the year of release for $i$'s cultivar. All other parameters are
+similar to Eq. 1.
+
+Data was processed and models were fit in R, an environment for
+statistical computing \citep{R}. Models were fit using the 'rstanarm'
+package \citep{rstanarm}, an interface to Stan, a language for
+probabilistic programming \citep{stan}. All model diagnostics,
+including $\hat{R}$, effective size, and posterior predictive checks
+were examined before reporting results.
 
@@
 
-\begin{figure}[h]
+<<echo=FALSE>>=
+rel_CI = round(posterior_interval(RES_rel_fit, pars = "release_yr_c") * scl, 1)
+
+posterior_interval(RES_rel_fit, pars = "release_yr_c") * scl
+m202_v_206 = ((posterior_predict(RES_rel_fit, newdata = data.frame(release_yr_c = c(-15, 4)),
+                                 re.form = NA) * scl) + center)
+m202_v_206 = m202_v_206[,2] - m202_v_206[,1]
+m202_v_206 = round(quantile(m202_v_206, p = c(0.05, 0.5, 0.95)), 1)
+@
+
+There was a positive relationship between the year of release and
+yield, with an estimated \Sexpr{round(fixef(RES_rel_fit)[2] * scl, 1)}
+kg ha\textsuperscript{-1} year\textsuperscript{-1} increase (90\% CI:
+\Sexpr{rel_CI[1]} -- \Sexpr{rel_CI[2]})
+(Fig. \ref{fig:release}). Comparing the two dominate cultivars M-202,
+which was widely planted from the late 1980s until the early 2000s,
+and M-206, currently the most widely planted CRRFS cultivar
+(approximately 50\% of planted area), there is an estimated
+\Sexpr{m202_v_206[2]} kg ha\textsuperscript{-1} yield advantage of M-206
+over M-202 (Figs \ref{fig:release} \& \ref{fig:compare_yrs}).
+
+\begin{figure}[ht]
 
@@
 yrs <- 1980:2015
-pp <- posterior_predict(release_fit, re.form = NA, newdata = data.frame(release_yr_c = yrs - 2000))
-pp <-  (pp * 1000 + 9000) * 1.12
+pp <- posterior_predict(RES_rel_fit, re.form = NA, newdata = data.frame(release_yr_c = yrs - 2000))
+pp <-  (pp * scl + center)
 
@@
 res = mod_data$site == "RES"
-plot_data = prep_plot_data(mod_data$yield_lb[res] * 1.12,
+plot_data = prep_plot_data(mod_data$yield_kg[res] ,
                            mod_data$id[res])
@@
      ylab = expression("Grain yield ( kg ha"^-1~")"),
-     ylim = c(8000, 13000),
+     ylim = c(8000, 12500),
      xlim = c(1980,2019))
-abline(a = (((fixef(RES_rel_fit)[1] * 1000) + 9000) * 1.12) -
-           (2000 * fixef(RES_rel_fit)[2] * 1000),
-       b = fixef(RES_rel_fit)[2] * 1000, col = "grey50")
+abline(a = (((fixef(RES_rel_fit)[1] * scl) + center) ) -
+           (2000 * fixef(RES_rel_fit)[2] * scl),
+       b = fixef(RES_rel_fit)[2] * scl, col = "grey50",
+       lty = 2)
 with(plot_data, segments(x0 = release_yr, y0=yield_kg + yield_se,
@@
 
-text(1990, 8000, bquote(.(round(fixef(release_fit)[2] * 1000 * 1.12, 1))~kg~ha^-1~year^-1))
+text(1990, 8000, bquote(.(round(fixef(RES_rel_fit)[2] * scl , 1))~kg~ha^-1~year^-1))
 
@@
 
-\subsection{Yield decreases over time}
-
-% \begin{figure}[h]
-
-% <<fig=TRUE, echo=FALSE>>=
-% yrs_in <- -6:34
-% pp <- posterior_predict(yrs_fit, re.form = NA, newdata = data.frame(yrs_in_trial = yrs_in))
-% pp <-  (pp * 1000 + 9000) * 1.12
-
-% ints <- apply(pp, 2, quantile, p = c(0.05, 0.5, 0.95))
-
-% plot_data = as.data.frame(tapply(mod_data$yield_lb * 1.12, mod_data$yrs_in_trial, mean))
-% plot_data$se = tapply(mod_data$yield_lb * 1.12, mod_data$yrs_in_trial, function(x)
-%     sd(x)/sqrt(length(x)))
-% plot_data$tmp = as.numeric(row.names(plot_data))
-% colnames(plot_data) = c("yield_kg", "yield_se", "yrs_in_trial")
-
-% plot(yield_kg ~ yrs_in_trial, data = plot_data, pch = 16, #col = rgb(0,0,0,0.25),
-%      xlab = "Year since entry",
-%      ylab = expression("Yield ( kg ha"^-1~")"),
-%      ylim = c(8700, 11200))
-% with(plot_data, segments(x0 = yrs_in_trial, y0=yield_kg + yield_se,
-%                          y1 = yield_kg - yield_se))
-% abline(a = ((fixef(yrs_fit)[1] * 1000) + 9000) * 1.12, b = fixef(yrs_fit)[2] * 1000)
-% # lines(x = yrs_in, y = ints[2,], lty = 2) 
-% # polygon(x = c(yrs_in, rev(yrs_in)), y = c(ints[1,],rev(ints[3,])), col = rgb(0,0,0,0.2), border = NA)
-% text(30, 4500, bquote(.(round(fixef(yrs_fit)[2] * 1000, 1))~lbs~year^-1))
-
-% @ 
-% \caption{There is a small decrease in yield over the time since a
-  % cultivar was entered into the trial using the full data
-  % set. However, the effect is small (\Sexpr{round(fixef(yrs_fit)[2] * 1000, 1)}
-  % lbs year\textsuperscript{-1}) in relation to the variability
-  % between sites and years.}
-% \label{fig:yrs}
-% \end{figure}
-
-<<echo=FALSE>>=
-all_CI <- round(posterior_interval(yrs_fit, pars = "yrs_in_trial") * 1000, 1)
-RES_CI <- round(posterior_interval(RES_yrs_fit, pars = "yrs_in_trial") * 1000, 1)
-@
-
-As seen in Fig. \ref{fig:yrs}, there is a small decrease in yield
-over time once a cultivar is released, though the estimated effect is
-small compared to the variability in the data set. This decrease is estimated to be
-\Sexpr{round(fixef(yrs_fit)[2] * 1000, 1)} lbs
-year\textsuperscript{-1}. The oldest cultivar in the trials has lost
-an estimated \Sexpr{round(fixef(yrs_fit)[2] * 1000 * 36 * 1.12, 1)} kg ha\textsuperscript{-1} since release. Using just data from the Rice
-Experiment Station (RES), the estimated effect is more drastic
-(an estimated \Sexpr{round(fixef(RES_yrs_fit)[2] * 1000 * 1.12, 1)} kg
-ha\textsuperscript{-1} year\textsuperscript{-1}), but with a wider credible interval
-(\Sexpr{RES_CI[1]} -- \Sexpr{RES_CI[2]} vs \Sexpr{all_CI[1]} --
-\Sexpr{all_CI[2]}).
-
-\begin{figure}[h]
-
-<<fig=TRUE, echo=FALSE>>=
-yrs_in <- -7:35
-pp <- posterior_predict(RES_yrs_fit, re.form = NA, newdata = data.frame(yrs_in_trial = yrs_in))
-pp <-  (pp * 1000 + 9000) * 1.12
-
-RES = res = mod_data$site == "RES"
-plot_data = prep_plot_data(mod_data$yield_lb[res] * 1.12,
-                           mod_data$yrs_in_trial[res])
-
-plot_data$tmp = as.numeric(row.names(plot_data))
-colnames(plot_data) = c("yield_kg", "yield_se", "yrs_in_trial")
-
-plot(yield_kg ~ yrs_in_trial, data = plot_data, pch = 16, #col = rgb(0,0,0,0.25),
-     xlab = "Year since entry",
-     ylab = expression("Grain yield ( kg ha"^-1~")"),
-     ylim = c(7000, 13000))
-with(plot_data, segments(x0 = yrs_in_trial, y0=yield_kg + yield_se,
-                         y1 = yield_kg - yield_se))
-abline(a = ((fixef(RES_yrs_fit)[1] * 1000) + 9000) * 1.12,
-       b = fixef(RES_yrs_fit)[2] * 1000, col = "grey50")
-ints <- apply(pp, 2, quantile, p = c(0.05, 0.5, 0.95))
-text(5, 7000, bquote(.(round(fixef(RES_yrs_fit)[2] * 1000 * 1.12, 1))~kg~ha^-1~year^-1))
-
-@
-\caption{The relationship between yield and the years since entry for
-  medium grain rice varieties planted at the Rice Experiment Station
-  (RES), California, USA. RES is considered
-  the best managed site, and has the most plots per year of any of the
-  sites. The estimated decrease over time is estimated to be stronger
-  using only these data (\Sexpr{round(fixef(RES_yrs_fit)[2] * 1000, 1)}
-  lbs year\textsuperscript{-1}), primarly due to lower variability.}
-\label{fig:RES_year}
-\end{figure}
-
 \begin{figure}[ht]
@@
     dd = mod_data[mod_data$id == x & mod_data$site == "RES",]
-    tmp = prep_plot_data(dd$yield_lb * 1.12, dd$year)
+    tmp = prep_plot_data(dd$yield_kg , dd$year)
     tmp$year = as.numeric(row.names(tmp))
@@
          data = plot_data[[i]],
-         xlim = c(1995, 2016), ylim = c(6500,14500),
+         xlim = c(1995, 2016), ylim = c(7500,13500),
          pch = 16, main = gsub("M","M-", tars[i]),
@@
     }
-#     abline(lm(yield_lb*1.12 ~ year,
+#     abline(lm(yield_kg*1.12 ~ year,
     #               data = mod_data[(mod_data$id == tars[i] & mod_data$site == "RES"),]),
@@
 
-    abline(h = mean(mod_data$yield_lb[(mod_data$id == tars[i] &
-                                       mod_data$site == "RES")] * 1.12),
+    abline(h = mean(mod_data$yield_kg[(mod_data$id == tars[i] &
+                                       mod_data$site == "RES")] ),
            lty = 1, col = "grey50")
@@
 # invisible(sapply(tars, function(x){
-#     plot(yield_lb * 1.12 ~ year,
+#     plot(yield_kg  ~ year,
 #          data = mod_data[(mod_data$id == x & mod_data$site == "RES"),],
@@
 #     }
-#     # abline(lm(yield_lb ~ year, data = mod_data[(mod_data$id == x & mod_data$site == "RES"),]), lty = 2)
-#     abline(h = mean(mod_data$yield_lb[(mod_data$id == x & mod_data$site == "RES")] * 1.12),
+#     # abline(lm(yield_kg ~ year, data = mod_data[(mod_data$id == x & mod_data$site == "RES"),]), lty = 2)
+#     abline(h = mean(mod_data$yield_kg[(mod_data$id == x & mod_data$site == "RES")] ),
 #            lty = 2)
@@
 \caption{The performance of two major cultivars at the CA Rice
-  Experiment Station (RES) over the period 1995 -- 2015. The dashed line is the average
+  Experiment Station (RES) over the period 1995 -- 2015. The line is the average
   yield for each cultivar over the period which it was planted at the
-  RES. The newer cultivar, M-209, appears to have a slight yield
+  RES. The newer cultivar, M-206, appears to have a slight yield
   advantage over the older cultivar, M-202.}
-\label{fig:res_yrs}
+\label{fig:compare_yrs}
+\end{figure}
+
+\subsection{Yield decreases over time}
+
+% \begin{figure}[h]
+
+% <<fig=TRUE, echo=FALSE>>=
+% yrs_in <- -6:34
+% pp <- posterior_predict(RES_yrs_fit, re.form = NA, newdata = data.frame(yrs_in_trial = yrs_in))
+% pp <-  (pp * scl + center)
+
+% ints <- apply(pp, 2, quantile, p = c(0.05, 0.5, 0.95))
+
+% plot_data = as.data.frame(tapply(mod_data$yield_kg , mod_data$yrs_in_trial, mean))
+% plot_data$se = tapply(mod_data$yield_kg , mod_data$yrs_in_trial, function(x)
+%     sd(x)/sqrt(length(x)))
+% plot_data$tmp = as.numeric(row.names(plot_data))
+% colnames(plot_data) = c("yield_kg", "yield_se", "yrs_in_trial")
+
+% plot(yield_kg ~ yrs_in_trial, data = plot_data, pch = 16, #col = rgb(0,0,0,0.25),
+%      xlab = "Year since entry",
+%      ylab = expression("Yield ( kg ha"^-1~")"),
+%      ylim = c(8700, 11200))
+% with(plot_data, segments(x0 = yrs_in_trial, y0=yield_kg + yield_se,
+%                          y1 = yield_kg - yield_se))
+% abline(a = ((fixef(RES_yrs_fit)[1] * scl) + center) , b = fixef(RES_yrs_fit)[2] * scl)
+% # lines(x = yrs_in, y = ints[2,], lty = 2) 
+% # polygon(x = c(yrs_in, rev(yrs_in)), y = c(ints[1,],rev(ints[3,])), col = rgb(0,0,0,0.2), border = NA)
+% text(30, 4500, bquote(.(round(fixef(RES_yrs_fit)[2] * scl, 1))~lbs~year^-1))
+
+% @ 
+% \caption{There is a small decrease in yield over the time since a
+  % cultivar was entered into the trial using the full data
+  % set. However, the effect is small (\Sexpr{round(fixef(RES_yrs_fit)[2] * scl, 1)}
+  % lbs year\textsuperscript{-1}) in relation to the variability
+  % between sites and years.}
+% \label{fig:yrs}
+% \end{figure}
+
+<<echo=FALSE>>=
+all_CI <- round(posterior_interval(RES_yrs_fit, pars = "yrs_in_trial") * scl, 1)
+RES_CI <- round(posterior_interval(RES_yrs_fit, pars = "yrs_in_trial") * scl, 1)
+@
+
+As seen in Fig. \ref{fig:yrs}, there is a small decrease in yield
+over time once a cultivar is released, though the estimated effect is
+small compared to the variability in the data set. This decrease is estimated to be
+\Sexpr{round(fixef(RES_yrs_fit)[2] * scl, 1)} lbs
+year\textsuperscript{-1}. The oldest cultivar in the trials has lost
+an estimated \Sexpr{round(fixef(RES_yrs_fit)[2] * scl * 36 , 1)} kg ha\textsuperscript{-1} since release. Using just data from the Rice
+Experiment Station (RES), the estimated effect is more drastic
+(an estimated \Sexpr{round(fixef(RES_yrs_fit)[2] * scl , 1)} kg
+ha\textsuperscript{-1} year\textsuperscript{-1}), but with a wider credible interval
+(\Sexpr{RES_CI[1]} -- \Sexpr{RES_CI[2]} vs \Sexpr{all_CI[1]} --
+\Sexpr{all_CI[2]}).
+
+\begin{figure}[ht]
+
+<<fig=TRUE, echo=FALSE>>=
+yrs_in <- -7:35
+pp <- posterior_predict(RES_yrs_fit, re.form = NA, newdata = data.frame(yrs_in_trial = yrs_in))
+pp <-  (pp * scl + center)
+
+RES = res = mod_data$site == "RES"
+plot_data = prep_plot_data(mod_data$yield_kg[res] ,
+                           mod_data$yrs_in_trial[res])
+
+plot_data$tmp = as.numeric(row.names(plot_data))
+colnames(plot_data) = c("yield_kg", "yield_se", "yrs_in_trial")
+
+plot(yield_kg ~ yrs_in_trial, data = plot_data, pch = 16, #col = rgb(0,0,0,0.25),
+     xlab = "Year since release",
+     ylab = expression("Grain yield ( kg ha"^-1~")"),
+     ylim = c(8000, 12500))
+with(plot_data, segments(x0 = yrs_in_trial, y0=yield_kg + yield_se,
+                         y1 = yield_kg - yield_se))
+abline(a = ((fixef(RES_yrs_fit)[1] * scl) + center) ,
+       b = fixef(RES_yrs_fit)[2] * scl, col = "grey50",
+       lty = 2)
+ints <- apply(pp, 2, quantile, p = c(0.05, 0.5, 0.95))
+text(5, 8000, bquote(.(round(fixef(RES_yrs_fit)[2] * scl , 1))~kg~ha^-1~year^-1))
+
+@
+\caption{The relationship between yield and the years since release for
+  medium grain rice varieties planted at the Rice Experiment Station
+  (RES), California, USA. RES is considered
+  the best managed site, and has the most plots per year of any of the
+  sites. The estimated decrease over time is estimated to be stronger
+  using only these data (\Sexpr{round(fixef(RES_yrs_fit)[2] * scl, 1)}
+  lbs year\textsuperscript{-1}), primarly due to lower variability.}
+\label{fig:RES_year}
 \end{figure}
@@
 
-<<fig=TRUE, echo=FALSE, width = 8, height = 5>>=
-par(mfrow = c(1, length(tars)),
-    mar = c(4,0,2,0),
-    oma = c(4,4,0.5,0.5))
+<<fig=TRUE, echo=FALSE, width = 6, height = 5>>=
+# par(mfrow = c(1, length(tars)),
+#     mar = c(4,0,2,0),
+#     oma = c(4,4,0.5,0.5))
 
@@
     dd = mod_data[mod_data$id == x & mod_data$site == "RES",]
-    tmp = prep_plot_data(dd$yield_lb * 1.12, dd$yrs_in_trial)
+    tmp = prep_plot_data(dd$yield_kg , dd$yrs_in_trial)
     tmp$yrs_in_trial = as.numeric(row.names(tmp))
@@
 
-invisible(sapply(1:2, function(i){
+# invisible(sapply(1:2, function(i){
+#     plot(mean ~ yrs_in_trial,
+#          data = plot_data[[i]],
+#          xlim = c(-6, 36), ylim = c(6500,14500),
+#          pch = 16, main = gsub("M","M-", tars[i]),
+#          ylab = "",
+#          yaxt = "n",
+#          xlab = "Years after release")
+#     segments(x0 = plot_data[[i]]$yrs_in_trial,
+#              y0 = plot_data[[i]]$mean + plot_data[[i]]$se,
+#              y1 = plot_data[[i]]$mean - plot_data[[i]]$se)
+#     if(i == 1){
+#         axis(2)
+#         mtext(expression("Yield ( kg ha"^-1~")"), 2, line = 2.25, cex.lab = 0.6)
+#     }
+#     abline(lm(yield_kg*1.12 ~ yrs_in_trial,
+#               data = mod_data[(mod_data$id == tars[i] & mod_data$site == "RES"),]),
+#            lty = 1, col = "grey50")
+# }))
+
 plot(mean ~ yrs_in_trial,
-         data = plot_data[[i]],
-         xlim = c(-6, 36), ylim = c(6500,14500),
+     data = plot_data[[1]],
+     xlim = c(-6, 32), ylim = c(6500,14500),
      pch = 16, main = gsub("M","M-", tars[i]),
-         ylab = "",
-         yaxt = "n",
+     ylab = expression("Yield ( kg ha"^-1~")"),
      xlab = "Years after release")
-    segments(x0 = plot_data[[i]]$yrs_in_trial,
-             y0 = plot_data[[i]]$mean + plot_data[[i]]$se,
-             y1 = plot_data[[i]]$mean - plot_data[[i]]$se)
-    if(i == 1){
-        axis(2)
-        mtext(expression("Yield ( kg ha"^-1~")"), 2, line = 2.25, cex.lab = 0.6)
-    }
-    abline(lm(yield_lb*1.12 ~ yrs_in_trial,
-              data = mod_data[(mod_data$id == tars[i] & mod_data$site == "RES"),]),
+segments(x0 = plot_data[[1]]$yrs_in_trial,
+         y0 = plot_data[[1]]$mean + plot_data[[1]]$se,
+         y1 = plot_data[[1]]$mean - plot_data[[1]]$se)
+points(mean ~ yrs_in_trial,
+       data = plot_data[[2]],
+       pch = 17, col = "grey40")
+segments(x0 = plot_data[[2]]$yrs_in_trial,
+         y0 = plot_data[[2]]$mean + plot_data[[2]]$se,
+         y1 = plot_data[[2]]$mean - plot_data[[2]]$se,
+         col = "grey40")
+abline(lm(yield_kg ~ yrs_in_trial,
+          data = mod_data[(mod_data$id %in% tars & mod_data$site == "RES"),]),
        lty = 1, col = "grey50")
-}))
+legend("topright", legend = gsub("M","M-", tars),
+       pch = c(16, 17), col = c("black","grey40"))
 
@@
   Experiment Station (RES) since the year each cultivar was
-  released. Both M-202 and M-206 have similar intercepts, suggesting
-  similar yield at the time of release and an apparent yield advantage
-  of the newer cultivar, M-209, might be explained by a yield loss of
-  M-202 over time.}
+  released. Both M-202 ($\bullet$) and M-206 ($\blacktriangle$) have similar
+  intercepts, suggesting similar yield at the time of release and an
+  apparent yield advantage of the newer cultivar, M-206, might be
+  explained by a yield loss of M-202 over time. Points represent the
+  mean, and bars (|) the standard error.}
 \label{fig:res_since_release}
@@
 small increase over time (Fig. \ref{fig:release}), though it was small
-(\Sexpr{round(fixef(release_fit)[2] * 1000, 1)} lbs
+(\Sexpr{round(fixef(RES_rel_fit)[2] * scl, 1)} lbs
 year\textsuperscript{-1}). This suggests that the overall progress on
 yield over the period from 1980 -- 2016 is approximately
-\Sexpr{round(fixef(release_fit)[2] * 1000 * 36, 1)} lbs acre\textsuperscript{-1}.
+\Sexpr{round(fixef(RES_rel_fit)[2] * scl * 36, 1)} lbs acre\textsuperscript{-1}.
 
@@
 %                         newdata = data.frame(release_yr_c = yrs - 2000))
-% pp <-  pp * 1000 + 9000
+% pp <-  pp * scl + center
 
@@
 
-% plot(yield_lb ~ release_yr, data = mod_data[RES,], pch = 16, col = rgb(0,0,0,0.25),
+% plot(yield_kg ~ release_yr, data = mod_data[RES,], pch = 16, col = rgb(0,0,0,0.25),
 %      xlab = "Year of Release",
@@
 % polygon(x = c(yrs, rev(yrs)), y = c(ints[1,],rev(ints[3,])), col = rgb(0,0,0,0.2), border = NA)
-% text(2010, 4500, bquote(.(round(fixef(RES_rel_fit)[2] * 1000, 1))~lbs~year^-1))
+% text(2010, 4500, bquote(.(round(fixef(RES_rel_fit)[2] * scl, 1))~lbs~year^-1))
 % @
@@
 it. However, progress is slower than previously estimated
-(\Sexpr{round(fixef(release_fit)[2] * 1000, 1)} lbs
+(\Sexpr{round(fixef(RES_rel_fit)[2] * scl, 1)} lbs
 acre\textsuperscript{-1} year\textsuperscript{-1} increase verses a
@@
 <<label=plot1, fig=TRUE, echo=FALSE>>=
-boxplot(yield_lb ~ year, data = mod_data,
+boxplot(yield_kg ~ year, data = mod_data,
         xlab = "Year", ylab = expression("Yield ( lb ac"^-1~")"))
@@
 
-  boxplot(yield_lb ~ id, data = mod_data,
+  boxplot(yield_kg ~ id, data = mod_data,
           xlab = "Cultivar",

### 2017-05-04 src/h1_model.R

@@ -36,13 +36,13 @@ source("prep_model_data.R")
 ## release_fit <- sampling(mm, data = c(dd, list(x = mod_data$release_yr - 2000)),
 ##                     iter = 500, cores = 2L, control = list(adapt_delta = 0.9))
 
-yrs_fit <- stan_glmer(yield_cs ~ (1|year) + (1|site) + (1|id) +
-                          (1|yr_site_fact) + yrs_in_trial,
-                      data = mod_data, iter = 1000, cores = 2L)
+# yrs_fit <- stan_glmer(yield_cs ~ (1|year) + (1|site) + (1|id) +
+#                           (1|yr_site_fact) + yrs_in_trial,
+#                       data = mod_data, iter = 1000, cores = 2L)
 
-release_fit <- stan_glmer(yield_cs ~ (1|year) + (1|site) +
-                              (1|id) + (1|yr_site_fact) + release_yr_c,
-                          data = mod_data, iter = 1000, cores = 2L)
+# release_fit <- stan_glmer(yield_cs ~ (1|year) + (1|site) +
+#                               (1|id) + (1|yr_site_fact) + release_yr_c,
+#                           data = mod_data, iter = 1000, cores = 2L)
 
 i <- mod_data$site == "RES"
 
@@ -53,4 +53,5 @@ RES_rel_fit <- stan_glmer(yield_cs ~ (1|year) + (1|id) + release_yr_c,
                           data = mod_data[i,], iter = 1000, cores = 2L)
 
 
-save(yrs_fit, release_fit, RES_yrs_fit, RES_rel_fit, file = "../output/h1_fit.Rda")
+save( #yrs_fit, release_fit,
+     RES_yrs_fit, RES_rel_fit, file = "../output/h1_fit.Rda")

### 2017-05-04 src/prep_model_data.R

@@ -32,8 +32,14 @@ mod_data$release_yr_c <- mod_data$release_yr - 2000
 
 mod_data$yrs_in_trial <- mod_data$year - mod_data$release_yr
 
+# convert yield to kg/ha here once, rather than repeatedly in paper
+mod_data$yield_kg = mod_data$yield_lb * 1.12
+
 ## Get some centered and scaled vars for model
-mod_data$yield_cs <- (mod_data$yield - 9000) / 1000
+center = 10000
+scl = 1000
+
+mod_data$yield_cs <- (mod_data$yield_kg - center) / scl
 mod_data$yr_fact <- factor(mod_data$year)
 mod_data$yr_site_fact <- factor(paste(mod_data$year, mod_data$site))
 

### 2017-05-04 output/breed_paper.Rnw
Cleaned up unused code

@@
     }
-#     abline(lm(yield_kg*1.12 ~ year,
-    #               data = mod_data[(mod_data$id == tars[i] & mod_data$site == "RES"),]),
-    # lty = 2)
 
@@
 }))
-
-# invisible(sapply(tars, function(x){
-#     plot(yield_kg  ~ year,
-#          data = mod_data[(mod_data$id == x & mod_data$site == "RES"),],
-#          xlim = c(1995, 2016), ylim = c(6500,14500),
-#          pch = 16,
-#          col = rgb(0,0,0,0.5),
-#          main = gsub("M", "M-", x),
-#          ylab = "",
-#          yaxt = "n",
-#          xlab = "Year")
-#     if(x == "M202"){
-#         axis(2)
-#         mtext(expression("Yield ( kg ha"^-1~")"), 2, line = 2.25, cex.lab = 0.6)
-#     }
-#     # abline(lm(yield_kg ~ year, data = mod_data[(mod_data$id == x & mod_data$site == "RES"),]), lty = 2)
-#     abline(h = mean(mod_data$yield_kg[(mod_data$id == x & mod_data$site == "RES")] ),
-#            lty = 2)
-# }))
 
@@
 \subsection{Yield decreases over time}
-
-% \begin{figure}[h]
-
-% <<fig=TRUE, echo=FALSE>>=
-% yrs_in <- -6:34
-% pp <- posterior_predict(RES_yrs_fit, re.form = NA, newdata = data.frame(yrs_in_trial = yrs_in))
-% pp <-  (pp * scl + center)
-
-% ints <- apply(pp, 2, quantile, p = c(0.05, 0.5, 0.95))
-
-% plot_data = as.data.frame(tapply(mod_data$yield_kg , mod_data$yrs_in_trial, mean))
-% plot_data$se = tapply(mod_data$yield_kg , mod_data$yrs_in_trial, function(x)
-%     sd(x)/sqrt(length(x)))
-% plot_data$tmp = as.numeric(row.names(plot_data))
-% colnames(plot_data) = c("yield_kg", "yield_se", "yrs_in_trial")
-
-% plot(yield_kg ~ yrs_in_trial, data = plot_data, pch = 16, #col = rgb(0,0,0,0.25),
-%      xlab = "Year since entry",
-%      ylab = expression("Yield ( kg ha"^-1~")"),
-%      ylim = c(8700, 11200))
-% with(plot_data, segments(x0 = yrs_in_trial, y0=yield_kg + yield_se,
-%                          y1 = yield_kg - yield_se))
-% abline(a = ((fixef(RES_yrs_fit)[1] * scl) + center) , b = fixef(RES_yrs_fit)[2] * scl)
-% # lines(x = yrs_in, y = ints[2,], lty = 2) 
-% # polygon(x = c(yrs_in, rev(yrs_in)), y = c(ints[1,],rev(ints[3,])), col = rgb(0,0,0,0.2), border = NA)
-% text(30, 4500, bquote(.(round(fixef(RES_yrs_fit)[2] * scl, 1))~lbs~year^-1))
-
-% @ 
-% \caption{There is a small decrease in yield over the time since a
-  % cultivar was entered into the trial using the full data
-  % set. However, the effect is small (\Sexpr{round(fixef(RES_yrs_fit)[2] * scl, 1)}
-  % lbs year\textsuperscript{-1}) in relation to the variability
-  % between sites and years.}
-% \label{fig:yrs}
-% \end{figure}
 
@@
 <<fig=TRUE, echo=FALSE, width = 6, height = 5>>=
-# par(mfrow = c(1, length(tars)),
-#     mar = c(4,0,2,0),
-#     oma = c(4,4,0.5,0.5))
 
@@
     })
-
-# invisible(sapply(1:2, function(i){
-#     plot(mean ~ yrs_in_trial,
-#          data = plot_data[[i]],
-#          xlim = c(-6, 36), ylim = c(6500,14500),
-#          pch = 16, main = gsub("M","M-", tars[i]),
-#          ylab = "",
-#          yaxt = "n",
-#          xlab = "Years after release")
-#     segments(x0 = plot_data[[i]]$yrs_in_trial,
-#              y0 = plot_data[[i]]$mean + plot_data[[i]]$se,
-#              y1 = plot_data[[i]]$mean - plot_data[[i]]$se)
-#     if(i == 1){
-#         axis(2)
-#         mtext(expression("Yield ( kg ha"^-1~")"), 2, line = 2.25, cex.lab = 0.6)
-#     }
-#     abline(lm(yield_kg*1.12 ~ yrs_in_trial,
-#               data = mod_data[(mod_data$id == tars[i] & mod_data$site == "RES"),]),
-#            lty = 1, col = "grey50")
-# }))
 
@@
 \Sexpr{round(fixef(RES_rel_fit)[2] * scl * 36, 1)} lbs acre\textsuperscript{-1}.
-
-% \begin{figure}[h]
-
-% <<fig=TRUE, echo=FALSE>>=
-% yrs <- 1980:2015
-% pp <- posterior_predict(RES_rel_fit, re.form = NA,
-%                         newdata = data.frame(release_yr_c = yrs - 2000))
-% pp <-  pp * scl + center
-
-% ints <- apply(pp, 2, quantile, p = c(0.05, 0.5, 0.95))
-
-% plot(yield_kg ~ release_yr, data = mod_data[RES,], pch = 16, col = rgb(0,0,0,0.25),
-%      xlab = "Year of Release",
-%      ylab =  expression("Yield ( lbs acre"^-1~")"))
-% lines(x = yrs, y = ints[2,], lty = 2) 
-% polygon(x = c(yrs, rev(yrs)), y = c(ints[1,],rev(ints[3,])), col = rgb(0,0,0,0.2), border = NA)
-% text(2010, 4500, bquote(.(round(fixef(RES_rel_fit)[2] * scl, 1))~lbs~year^-1))
-% @
-
-% \caption{The effect of year of release on yield for only data from the
-%   Rice Experiment Station (RES). There is an estimated
-%   increase in yield over time, but it is smaller than the previously estimated 54 lbs
-%   year\textsuperscript{-1}}
-% \label{fig:RES_release}
-% \end{figure}
 

### 2017-05-04 output/breed_paper.Rnw
Changes

@@
   RES. The newer cultivar, M-206, appears to have a slight yield
-  advantage over the older cultivar, M-202.}
+  advantage (an estimated \Sexpr{m202_v_206[2]} kg ha\textsuperscript{-1})over the older cultivar, M-202.}
 \label{fig:compare_yrs}
@@
 
-As seen in Fig. \ref{fig:yrs}, there is a small decrease in yield
-over time once a cultivar is released, though the estimated effect is
-small compared to the variability in the data set. This decrease is estimated to be
-\Sexpr{round(fixef(RES_yrs_fit)[2] * scl, 1)} lbs
-year\textsuperscript{-1}. The oldest cultivar in the trials has lost
-an estimated \Sexpr{round(fixef(RES_yrs_fit)[2] * scl * 36 , 1)} kg ha\textsuperscript{-1} since release. Using just data from the Rice
-Experiment Station (RES), the estimated effect is more drastic
-(an estimated \Sexpr{round(fixef(RES_yrs_fit)[2] * scl , 1)} kg
-ha\textsuperscript{-1} year\textsuperscript{-1}), but with a wider credible interval
-(\Sexpr{RES_CI[1]} -- \Sexpr{RES_CI[2]} vs \Sexpr{all_CI[1]} --
-\Sexpr{all_CI[2]}).
+When yields are arranged according to the years since release, there
+is a small decrease in yield over time once a cultivar is released,
+though the estimated effect is small compared to the variability in
+the data set. This decrease is estimated to be
+\Sexpr{round(fixef(RES_yrs_fit)[2] * scl, 1)} kg
+ha\textsuperscript{-1} year\textsuperscript{-1} (90\% CI: \Sexpr{RES_CI[2]} --
+\Sexpr{RES_CI[1]}). The oldest cultivar
+in the trials is estimated to have lost
+\Sexpr{round(fixef(RES_yrs_fit)[2] * scl * 36 , 1)} kg
+ha\textsuperscript{-1} since release.
+
+Again comparing the yield of two prominant cultivars, M-202 and M-206,
+arranged not by the year but rather the years since release, we find
+that the two cultivars are compariable and that the estimated yield
+advantage (\Sexpr{m202_v_206[2]} kg ha\textsuperscript{-1}) can be
+explained by the differences in the amount of time that has passed
+since each cultivar was released
+(Fig. \ref{fig:compare_since_release})
 
@@
   mean, and bars (|) the standard error.}
-\label{fig:res_since_release}
+\label{fig:compare_since_release}
 \end{figure}

### 2017-05-10 output/breed_paper.Rnw
Updates paper to use RES data back to 1984

@@
 
+RES: Rice Experiment Station
+
+CCRRF: California Cooperative Rice Research Foundation
+
 \newpage
@@
 
-<<echo = FALSE>>=
+<<echo=FALSE>>=
 
@@
 library(lme4)
+library(xtable)
 
-source("../src/prep_model_data.R")
+load("../data/model_data.rda")
 load("h1_fit.Rda")
+load("../data/all_vt_weather.Rda")
 
@@
 to increase total production to meet growing demand \citep{Foley2011a,
-  Godfray2010, Mueller2012}, which has
-necessitated harvesting more grain per unit land area
-\citep{Lobell2009, Tittonell2014}. Historically,
-agricultural research has been successful in staving off the Mathesian
-disaster through continued yield improvement of staple crops, however
-more and more production systems are experiencing plateaus in grain
-yield \citep{Grassini2013, Peng2010}. If continued improvements per unit land area
-are no longer possible, increasing the area under cultivation will be
-needed to meet demand, which carries with it undesirable ecological
+Godfray2010, Mueller2012}, which has necessitated harvesting more
+grain per unit land area \citep{Lobell2009,
+Tittonell2014}. Historically, agricultural research has been
+successful in staving off the Mathesian disaster through continued
+yield improvement of staple crops, however more and more production
+systems are experiencing plateaus in grain yield \citep{Grassini2013,
+Peng2010}. If continued improvements per unit land area are no longer
+possible, increasing the area under cultivation will be needed to meet
+demand, which carries with it undesirable ecological
 implications. Therefore, it is of critical importance to better
@@
 system. Rice yields in CA during the period from 1920 -- 1990
-increased rapidly, but since 1990 have increased more slowly (Fig. 1
-\ref{fig:state_yield}). However, continued improvement of grain yield
-as the result of plant breeding has been demonstrated since 1990
-\citep{rrb}, with each new cultivar having a yield advantage over the
-previous cultivars. Yet, these same reports show stagnant yields; for
-example, at the year of release (1985), cultivar 'M-202' averaged
-11,060 kg ha\textsuperscript{-1} \citep[14\% MC;][]{m202}. In 2015,
-the average yield of the newest released cultivar 'M-209' was 10,841
-kg ha\textsuperscript{-1} \citep{m209}. This suggests little to no
-progress via breeding, however 'M-209' demonstrated an overall 10\%
-yield advantage over 'M-202' in side-by-side comparisons \citep{m209}. We sought to
-better understand how grain yields have increased in the breeders'
-fields without a corresponding increase in yield in the farmers'
-fields. Two competing hypotheses were tested: (1) Grain yields have
-not improved since 1990, or (2) grain yield improvements have been
-offset by an associated decrease in yield once a cultivar was
-selected.
+increased rapidly, in part due to improvements in rice genetics (i.e.,
+the Green Revolution) through the efforts of the California
+Cooperative Rice Research Foundation (CCRRF; a collaboration between
+the University of California, the USDA-Agricultural Research Service,
+and farmer-funded research). However, since 1990 CA state average
+yields have increased more slowly (Fig. \ref{fig:state_yield}) despite
+demonstrated improvements of grain yield as the result of plant
+breeding \citep{ccrrf}, with each new cultivar having a yield
+advantage over the previous cultivars in side-by-side comparisons in
+the annual Statewide Variety Trials. Yet, taken across years, these
+same reports show stagnant yields; for example, at the year of release
+(1985), cultivar M-202 averaged 11 Mg ha\textsuperscript{-1}
+\citep[14\% MC;][]{m202}. In 2015, the average yield of the newest
+released cultivar M-209 was 10.8 Mg ha\textsuperscript{-1}
+\citep{m209}. In other systems, it has been shown that rice cultivars
+experience a lack of adaptation to changing biotic and abiotic
+conditions, resulting in a decrease of a cultivar's performance over
+time \citep{Peng2010, Peng2000, Datta1995}. We sought to explore this
+phenomonon and better understand how grain yields have increased in
+the breeders' fields without a corresponding increase in yield in the
+farmers' fields. Two competing hypotheses were tested: (1) the yield
+advantage offered by newly developed cultivars is not greater than
+previous cultivars, or (2) yield increases gained through plant
+breeding are being offset by decreases in the yield of released
+cultivars over time.
 
@@
 CA_data = read.csv("../data/CA_rice_yield.csv", stringsAsFactors = FALSE)
-CA_data$Value = as.numeric(gsub(",", "", CA_data$Value))
+CA_data$Value = as.numeric(gsub(",", "", CA_data$Value))/1000
 CA_data = CA_data[order(CA_data$Year),]
@@
      xlab = "Year",
-     ylab = expression("Average grain yield ( kg ha"^-1~")"))
+     ylab = expression("Average grain yield ( Mg ha"^-1~")"))
 mv_avg = filter(c(rep(mean(CA_data$Value[1:10] * 1.12), 5),
@@
 
-In the course of investigating the drivers of yield in this system, we
-were compelled to explore how yields could increase with each release
-of improved cultivars, yet average yield for the region remain
-flat. Specifically, we offered two competing hypotheses to explain
-this phenomenon: (1) the yield advantage offered by newly developed
-cultivars is not greater than previous cultivars, or (2) yield
-increases gained through plant breeding are being offset by decreases
-in yield of released cultivars.
-
-Yet, annual reports from the CA Cooperative Rice Research Foundation
-show that each released cultivar has a yield advantage over the
-previous cultivars, typically on the order of 4 -- 6\%. This raises the
-question of how to reconcile these facts. Namely, there are two
-possibilities:
-
 \section{Methods}\label{methods}
@@
 
-The Rice Experiment Station (RES) of the California Cooperative Rice
-Research Foundation (CCRRF) is located northwest of Biggs, CA (39.4648,
--121.7342), and has been the central location for the development of
-improved rice varieties suited for CA since 1969. The climate at the
-RES is Mediterranian, characterized by mild winters during which most
-of the annual precipitation occurs (\Sexpr{18*24.5} -- \Sexpr{25*24.5}
-mm annually) and warm summers largely free of precipitation
-events. Soils at the RES are classified as Esquon-Neerdobe clays with
-roughly 2\% SOM \citep{ssurgo} in the top 15cm. As the central
-location for the breeding program, the RES is researcher-managed (as
-opposed to farmer-managed trials) and conducts the bulk of the trials
-for the Statewide Variety Trials (approximately 30\% of the total
-plots are planted at RES). For these reasons, we focused our
-investigation on the RES to avoid complications due to site, climate, and
-management differences at the other 6 -- 8 farmer-managed trials.
+The Rice Experiment Station (RES) of the CCRRF is located northwest of
+Biggs, CA (39.4648, -121.7342), and has been the central location for
+the development of improved rice varieties adapted for CA since
+1969. The climate at the RES is Mediterranian, characterized by mild
+winters during which most of the annual precipitation occurs
+(\Sexpr{18*24.5} -- \Sexpr{25*24.5} mm annually) and warm summers
+largely free of precipitation events. Soils at the RES are classified
+as Esquon-Neerdobe clays with roughly 2\% SOM \citep{ssurgo} in the
+top 15cm. As the central location for the breeding program, the RES is
+researcher-managed (as opposed to farmer-managed trials) and conducts
+the bulk of the trials for the Statewide Variety Trials (approximately
+3 -- 4 times the number of plots compared to any other location). For
+these reasons, we focused our investigation on the RES to avoid
+potential complications due to site, climate, and management
+differences at the other 6 -- 8 farmer-managed trials per year.
 
@@
 medium grain cultivars, as medium grain cultivars constitute roughly
-90\% of the planted area in CA. This lead to a final data set of
+90\% of the planted area in CA. For these cultivars planted in
+breeding trials at the RES from \Sexpr{min(mod_data[,'year'])} --
+\Sexpr{max(mod_data[,'year'])} plot level observations were
+collected. This lead to a final data set of
 \Sexpr{format(nrow(mod_data), big.mark = ',')} observations with
 \Sexpr{length(unique(mod_data[,'id']))} cultivars represented,
-constituting every publicly released medium grain variety developed by
+including every publicly released medium grain variety developed by
 the CCRRF from \Sexpr{min(mod_data[,'release_yr'])} --
@@
 time elapse (in years) since release was calculated.
+
+Plots at the RES are managed similar to perscribed best management
+practice, including land preparation, fertility management, and pest
+and disease control. For planting, rice seed is pre-germinated prior
+to direct-seeding into pre-flooded fields. Plots were either 14.0 or
+18.5 m\textsuperscript{2}, and were harvested after physiological
+maturity using a small-plot combine. All yields were converted to 14\%
+moisture content prior to reporting.
 
@@
  yield_i = \alpha + \beta_{years\ since\ release\ i}* x_{years\ since\
-   release\ i} + \beta_{year\ i} + \beta_{cultivar\ i}
+   release\ i} + \beta_{release\ year\ i}*x_{release\ year\ i} + \beta_{year\ i} + \beta_{cultivar\ i}
 \end{equation}
@@
 trials (negative for years prior to release, zero at the year of
-release), $\beta_{year}$ is the random effect of year, and
-$\beta_{cultivar}$ is the random effect of cultivar. A similar but
-separate model was used to estimate the yield advantage of newer
-releases over older ones:
-
-\begin{equation}
-yield_i = \alpha + \beta_{release\ year\ i}*x_{release\ year\ i} +
-\beta_{year\ i} + \beta_{cultivar\ i}
-\end{equation}
-
-Where $\beta_{release\ year\ i}$ is effect of $x_{release\ year\ i}$,
-the year of release for $i$'s cultivar. All other parameters are
-similar to Eq. 1.
+release), $\beta_{release\ year\ i}$ is the effect of $x_{release\
+year\ i}$, the year of release for $i$'s cultivar, $\beta_{year}$ is
+the random effect of year, and $\beta_{cultivar}$ is the random effect
+of cultivar.
 
@@
 <<echo=FALSE>>=
-rel_CI = round(posterior_interval(RES_rel_fit, pars = "release_yr_c") * scl, 1)
+rel_CI = round(posterior_interval(RES_full, pars = "release_yr_c") * scl, 1)
 
-posterior_interval(RES_rel_fit, pars = "release_yr_c") * scl
-m202_v_206 = ((posterior_predict(RES_rel_fit, newdata = data.frame(release_yr_c = c(-15, 4)),
-                                 re.form = NA) * scl) + center)
-m202_v_206 = m202_v_206[,2] - m202_v_206[,1]
-m202_v_206 = round(quantile(m202_v_206, p = c(0.05, 0.5, 0.95)), 1)
+m202_v_206 = round(fixef(RES_full)[2] * scl * 19, 0)
+# m202_v_206 = round((posterior_interval(RES_full, pars = "release_yr_c") * scl) * 19, 0)
+
+# m202_v_206 = ((posterior_predict(RES_full, newdata = data.frame(release_yr_c = c(-15, 4)),
+#                                  re.form = NA) * scl) + center)
+# m202_v_206 = m202_v_206[,2] - m202_v_206[,1]
+# m202_v_206 = round(quantile(m202_v_206, p = c(0.05, 0.5, 0.95)), 1)
+
 @
@@
 There was a positive relationship between the year of release and
-yield, with an estimated \Sexpr{round(fixef(RES_rel_fit)[2] * scl, 1)}
+yield, with an estimated \Sexpr{round(fixef(RES_full)[2] * scl, 1)}
 kg ha\textsuperscript{-1} year\textsuperscript{-1} increase (90\% CI:
@@
 (approximately 50\% of planted area), there is an estimated
-\Sexpr{m202_v_206[2]} kg ha\textsuperscript{-1} yield advantage of M-206
-over M-202 (Figs \ref{fig:release} \& \ref{fig:compare_yrs}).
+\Sexpr{m202_v_206} kg ha\textsuperscript{-1} yield advantage of
+M-206 over M-202 (Figs \ref{fig:release} \& \ref{fig:compare_yrs}).
+
+\subsection{Yield decreases over time}
 
@@
 <<fig=TRUE, echo=FALSE>>=
-yrs <- 1980:2015
-pp <- posterior_predict(RES_rel_fit, re.form = NA, newdata = data.frame(release_yr_c = yrs - 2000))
-pp <-  (pp * scl + center)
+yrs_in <- -10:40
+pp <- posterior_linpred(RES_full, re.form = NA,
+                        newdata = data.frame(yrs_in_trial = yrs_in,
+                                             release_yr_c = 0))
+pp <-  (pp * scl + center) / 1000 
+ints = apply(pp, 2, quantile, p = c(0.05, 0.95))
 
-ints <- apply(pp, 2, quantile, p = c(0.05, 0.5, 0.95))
+RES = res = mod_data$site == "RES"
+plot_data = prep_plot_data(mod_data$yield_kg[res] / 1000,
+                           mod_data$yrs_in_trial[res])
+
+plot_data$tmp = as.numeric(row.names(plot_data))
+colnames(plot_data) = c("yield_kg", "yield_se", "yrs_in_trial")
+
+plot(yield_kg ~ yrs_in_trial, data = plot_data, pch = 16, #col = rgb(0,0,0,0.25),
+     xlab = "Year since release",
+     ylab = expression("Grain yield ( kg ha"^-1~")"),
+     ylim = c(8.0, 12.5))
+with(plot_data, segments(x0 = yrs_in_trial, y0=yield_kg + yield_se,
+                         y1 = yield_kg - yield_se))
+abline(a = ((fixef(RES_full)[1] * scl) + center) / 1000 ,
+       b = fixef(RES_full)[3] * scl/1000)
+polygon(x = c(yrs_in, rev(yrs_in)),
+        y = c(ints[1,],rev(ints[2,])),
+        border = NA,
+        col = rgb(0,0,0,0.1))
+text(5, 8.0, bquote(.(round(fixef(RES_full)[3] * scl , 1))~kg~ha^-1~year^-1))
+
+@
+\caption{The relationship between yield and the years since release
+  for medium grain rice varieties planted at the Rice Experiment Station
+  (RES), California, USA. The estimated decrease over time once a
+  cultivar is release is estimated to be
+  \Sexpr{round(fixef(RES_full)[3] * scl, 1)} kg ha\textsuperscript{-1}
+  year\textsuperscript{-1}. The shaded area is the 90\% credible
+  interval.}
+\label{fig:RES_year}
+\end{figure}
+
+\begin{figure}[ht]
+<<fig=TRUE, echo=FALSE>>=
+yrs <- 1975:2025
+pp <- posterior_linpred(RES_full, re.form = NA,
+                        newdata = data.frame(release_yr_c = yrs - 2000,
+                                             yrs_in_trial = 0),
+                        prob = 0.5)
+pp <-  (pp * scl + center) / 1000
+ints = apply(pp, 2, quantile, p = c(0.05, 0.95))
 
 res = mod_data$site == "RES"
-plot_data = prep_plot_data(mod_data$yield_kg[res] ,
+plot_data = prep_plot_data(mod_data$yield_kg[res]/1000 ,
                            mod_data$id[res])
@@
      xlab = "Release year",
-     ylab = expression("Grain yield ( kg ha"^-1~")"),
-     ylim = c(8000, 12500),
+     ylab = expression("Grain yield ( Mg ha"^-1~")"),
+     ylim = c(8, 12.5),
      xlim = c(1980,2019))
-abline(a = (((fixef(RES_rel_fit)[1] * scl) + center) ) -
-           (2000 * fixef(RES_rel_fit)[2] * scl),
-       b = fixef(RES_rel_fit)[2] * scl, col = "grey50",
-       lty = 2)
+abline(a = ((((fixef(RES_full)[1] * scl) + center) ) -
+           (2000 * fixef(RES_full)[2] * scl)) / 1000,
+       b = fixef(RES_full)[2] * scl/1000)
+abline(lm(yield_kg/1000 ~ release_yr, data = mod_data[res,]), lty = 2)
 with(plot_data, segments(x0 = release_yr, y0=yield_kg + yield_se,
                          y1 = yield_kg - yield_se))
-text(x = plot_data$release_yr, y = plot_data$yield_kg,
+text(x = plot_data$release_yr, y = plot_data$yield_kg + 0.15,
      labels = gsub("M", "M-", row.names(plot_data)),
@@
 
-text(1990, 8000, bquote(.(round(fixef(RES_rel_fit)[2] * scl , 1))~kg~ha^-1~year^-1))
+legend("bottomright", legend = c("w/ years since release",
+                                 "w/o years since release",
+                                 "90% Credible interval"),
+       lty = c(1,2, NA),
+       bty = "n",
+       fill = c(NA,NA, "grey"),
+       border = NA,
+       merge = TRUE)
+polygon(x = c(yrs, rev(yrs)),
+        y = c(ints[1,],rev(ints[2,])),
+        border = NA,
+        col = rgb(0,0,0,0.1))
+
+# text(1990, 8.0, bquote(.(round(fixef(RES_full)[2] * scl , 1))~kg~ha^-1~year^-1))
 
@@
 \caption{The yield increase over time of medium grain rice varieties
-  in California, USA. There is an estimated
-  increase in yield over time, but it is smaller than a previously
-  estimate of 54 lbs year\textsuperscript{-1}. This estimate is
-  potentially misleading, because it only
-  considers the average yield of each variety during the period 1995
-  -- 2015, and the older varieties have experiencing greater losses
-  since being released.}
+  in California, USA. Ignoring the years since release (dashed line)
+  suggests an increase in yield potential over time. However, correcting
+  the comparison to account for differences in the length of time since
+  each cultivar was released (solid line) shows no increase in yield over time.}
 \label{fig:release}
@@
     dd = mod_data[mod_data$id == x & mod_data$site == "RES",]
-    tmp = prep_plot_data(dd$yield_kg , dd$year)
+    tmp = prep_plot_data(dd$yield_kg/1000 , dd$year)
     tmp$year = as.numeric(row.names(tmp))
@@
          data = plot_data[[i]],
-         xlim = c(1995, 2016), ylim = c(7500,13500),
+         xlim = c(1985, 2016), ylim = c(7.5,13.5),
          pch = 16, main = gsub("M","M-", tars[i]),
@@
          yaxt = "n",
-         xlab = "Years after release")
+         xlab = "Year",
+         cex.axis = 0.8)
     segments(x0 = plot_data[[i]]$year,
@@
     if(i == 1){
-        axis(2)
-        mtext(expression("Yield ( kg ha"^-1~")"), 2, line = 2.25, cex.lab = 0.6)
+        axis(2, cex.axis = 0.8,)
+        mtext(expression("Yield ( Mg ha"^-1~")"), 2, line = 2.25, cex.lab = 0.6)
     }
 
-    abline(h = mean(mod_data$yield_kg[(mod_data$id == tars[i] &
-                                       mod_data$site == "RES")] ),
-           lty = 1, col = "grey50")
-
+    # abline(h = mean(mod_data$yield_kg[(mod_data$id == tars[i] &
+    #                                    mod_data$site == "RES")] ),
+    #        lty = 1, col = "grey50")
+    mod = lm(mean ~ year, data = plot_data[[i]])
+    tmp = predict(mod)
+    if(i == 1){
+        abline(mod)
+    }else{
+        lines(as.numeric(names(tmp)), tmp)
+    }
+    abline(h = 11.2, lty = 2)
+    text(2000, 7.5, bquote(.(round(coef(mod)[2] * 1000, 1))~kg~ha^-1~year^-1))
 }))
@@
 \caption{The performance of two major cultivars at the CA Rice
-  Experiment Station (RES) over the period 1995 -- 2015. The line is the average
-  yield for each cultivar over the period which it was planted at the
-  RES. The newer cultivar, M-206, appears to have a slight yield
-  advantage (an estimated \Sexpr{m202_v_206[2]} kg ha\textsuperscript{-1})over the older cultivar, M-202.}
+  Experiment Station (RES) over the period 1995 -- 2015. The solid line
+  represents a simple linear regression of yield for each cultivar over
+  the period which it was planted at the RES (including only that
+  cultivar's data), while the dashed line is the average yield of M-202
+  at its year of release (1985). Both cultivars have a similar yield at
+  their respective year of release, and similar decreases over
+  time. The slope for the individual regressions are comparible to the
+slope estimated from all medium grain cultivars
+(\Sexpr{round(fixef(RES_full)[3] * scl, 1)} kg ha\textsuperscript{-1} year\textsuperscript{-1}).}
 \label{fig:compare_yrs}
@@
 
-\subsection{Yield decreases over time}
-
 <<echo=FALSE>>=
-all_CI <- round(posterior_interval(RES_yrs_fit, pars = "yrs_in_trial") * scl, 1)
-RES_CI <- round(posterior_interval(RES_yrs_fit, pars = "yrs_in_trial") * scl, 1)
+all_CI <- round(posterior_interval(RES_full, pars = "yrs_in_trial") * scl, 1)
+RES_CI <- round(posterior_interval(RES_full, pars = "yrs_in_trial") * scl, 1)
 @
@@
 the data set. This decrease is estimated to be
-\Sexpr{round(fixef(RES_yrs_fit)[2] * scl, 1)} kg
+\Sexpr{round(fixef(RES_full)[3] * scl, 1)} kg
 ha\textsuperscript{-1} year\textsuperscript{-1} (90\% CI: \Sexpr{RES_CI[2]} --
@@
 in the trials is estimated to have lost
-\Sexpr{round(fixef(RES_yrs_fit)[2] * scl * 36 , 1)} kg
+\Sexpr{round(fixef(RES_full)[3] * scl * 36 , 1)} kg
 ha\textsuperscript{-1} since release.
@@
 that the two cultivars are compariable and that the estimated yield
-advantage (\Sexpr{m202_v_206[2]} kg ha\textsuperscript{-1}) can be
+advantage (\Sexpr{m202_v_206} kg ha\textsuperscript{-1}) can be
 explained by the differences in the amount of time that has passed
 since each cultivar was released
-(Fig. \ref{fig:compare_since_release})
+(Fig. \ref{fig:compare_yrs} \& \ref{fig:compare_since_release})
 
-\begin{figure}[ht]
-
-<<fig=TRUE, echo=FALSE>>=
-yrs_in <- -7:35
-pp <- posterior_predict(RES_yrs_fit, re.form = NA, newdata = data.frame(yrs_in_trial = yrs_in))
-pp <-  (pp * scl + center)
-
-RES = res = mod_data$site == "RES"
-plot_data = prep_plot_data(mod_data$yield_kg[res] ,
-                           mod_data$yrs_in_trial[res])
-
-plot_data$tmp = as.numeric(row.names(plot_data))
-colnames(plot_data) = c("yield_kg", "yield_se", "yrs_in_trial")
-
-plot(yield_kg ~ yrs_in_trial, data = plot_data, pch = 16, #col = rgb(0,0,0,0.25),
-     xlab = "Year since release",
-     ylab = expression("Grain yield ( kg ha"^-1~")"),
-     ylim = c(8000, 12500))
-with(plot_data, segments(x0 = yrs_in_trial, y0=yield_kg + yield_se,
-                         y1 = yield_kg - yield_se))
-abline(a = ((fixef(RES_yrs_fit)[1] * scl) + center) ,
-       b = fixef(RES_yrs_fit)[2] * scl, col = "grey50",
-       lty = 2)
-ints <- apply(pp, 2, quantile, p = c(0.05, 0.5, 0.95))
-text(5, 8000, bquote(.(round(fixef(RES_yrs_fit)[2] * scl , 1))~kg~ha^-1~year^-1))
-
-@
-\caption{The relationship between yield and the years since release for
-  medium grain rice varieties planted at the Rice Experiment Station
-  (RES), California, USA. RES is considered
-  the best managed site, and has the most plots per year of any of the
-  sites. The estimated decrease over time is estimated to be stronger
-  using only these data (\Sexpr{round(fixef(RES_yrs_fit)[2] * scl, 1)}
-  lbs year\textsuperscript{-1}), primarly due to lower variability.}
-\label{fig:RES_year}
-\end{figure}
-
-\begin{figure}
-
-<<fig=TRUE, echo=FALSE, width = 6, height = 5>>=
-
-plot_data = lapply(tars, function(x) {
-    dd = mod_data[mod_data$id == x & mod_data$site == "RES",]
-    tmp = prep_plot_data(dd$yield_kg , dd$yrs_in_trial)
-    tmp$yrs_in_trial = as.numeric(row.names(tmp))
-    tmp
-    })
-
-plot(mean ~ yrs_in_trial,
-     data = plot_data[[1]],
-     xlim = c(-6, 32), ylim = c(6500,14500),
-     pch = 16, main = gsub("M","M-", tars[i]),
-     ylab = expression("Yield ( kg ha"^-1~")"),
-     xlab = "Years after release")
-segments(x0 = plot_data[[1]]$yrs_in_trial,
-         y0 = plot_data[[1]]$mean + plot_data[[1]]$se,
-         y1 = plot_data[[1]]$mean - plot_data[[1]]$se)
-points(mean ~ yrs_in_trial,
-       data = plot_data[[2]],
-       pch = 17, col = "grey40")
-segments(x0 = plot_data[[2]]$yrs_in_trial,
-         y0 = plot_data[[2]]$mean + plot_data[[2]]$se,
-         y1 = plot_data[[2]]$mean - plot_data[[2]]$se,
-         col = "grey40")
-abline(lm(yield_kg ~ yrs_in_trial,
-          data = mod_data[(mod_data$id %in% tars & mod_data$site == "RES"),]),
-       lty = 1, col = "grey50")
-legend("topright", legend = gsub("M","M-", tars),
-       pch = c(16, 17), col = c("black","grey40"))
-
-@
-\caption{The performance of two major cultivars at the CA Rice
-  Experiment Station (RES) since the year each cultivar was
-  released. Both M-202 ($\bullet$) and M-206 ($\blacktriangle$) have similar
-  intercepts, suggesting similar yield at the time of release and an
-  apparent yield advantage of the newer cultivar, M-206, might be
-  explained by a yield loss of M-202 over time. Points represent the
-  mean, and bars (|) the standard error.}
-\label{fig:compare_since_release}
-\end{figure}
-
-Contrary to the estimates for decreases once a cultivar is released, there was a
-small increase over time (Fig. \ref{fig:release}), though it was small
-(\Sexpr{round(fixef(RES_rel_fit)[2] * scl, 1)} lbs
+Contrary to the estimates for decreases once a cultivar is released,
+there was no increase in yield potential over time once the
+differences between the years since release were accounted for (Fig. \ref{fig:release})
+(\Sexpr{round(fixef(RES_full)[2] * scl, 1)} kg ha\textsuperscript{-1}
 year\textsuperscript{-1}). This suggests that the overall progress on
 yield over the period from 1980 -- 2016 is approximately
-\Sexpr{round(fixef(RES_rel_fit)[2] * scl * 36, 1)} lbs acre\textsuperscript{-1}.
+\Sexpr{round(fixef(RES_full)[3] * scl * 36, 1)} lbs acre\textsuperscript{-1}.
 
@@
 
+ We postulate that true
+yield increases are only possible if the improvement in yield of each
+new cultivar exceeds former's benchmark yield at the time of release,
+and therefore side-by-side comparisons are potentially misleading.
+
 There is evidence of a small decrease in yield over time once a cultivar is
@@
 it. However, progress is slower than previously estimated
-(\Sexpr{round(fixef(RES_rel_fit)[2] * scl, 1)} lbs
+(\Sexpr{round(fixef(RES_full)[2] * scl, 1)} lbs
 acre\textsuperscript{-1} year\textsuperscript{-1} increase verses a
@@
 \newpage
+
+\section{Tables}
+
+<<echo=FALSE, results=tex>>=
+CIs = posterior_interval(RES_full)[1:3,] * scl
+CIs[1,] = round(CIs[1,] + center, -2)
+CIs = round(CIs, 1)
+med_eff = fixef(RES_full) * scl
+med_eff[1] = round(med_eff[1] + center, -2)
+med_eff = round(med_eff, 1)
+
+RES_coef = cbind(med_eff, paste(CIs[,1],CIs[,2], sep = " -- "))
+rownames(RES_coef) = c("Intercept (kg ha\\textsuperscript{-1})",
+                       "Release year (kg ha\\textsuperscript{-1} year\\textsuperscript{-1})",
+                       "Years since release (kg ha\\textsuperscript{-1} year\\textsuperscript{-1})")
+colnames(RES_coef) = c("Median", "90\\% Credible interval")
+
+print(xtable(RES_coef,
+             caption = "Estimated median effects and 90\\% credible intervals.",
+             label = "tbl:coef",
+             align = "lcc"),
+      caption.placement="top",
+      booktabs=TRUE,
+      sanitize.colnames.function = function(x){x},
+      sanitize.rownames.function = function(x){x})
+@
+\newpage
 \section{References}\label{references}
@@
 
-\begin{figure}[h]
+\begin{figure}[ht]
 
-<<label=plot1, fig=TRUE, echo=FALSE>>=
-boxplot(yield_kg ~ year, data = mod_data,
-        xlab = "Year", ylab = expression("Yield ( lb ac"^-1~")"))
+<<fig=TRUE, echo=FALSE, width = 6, height = 5>>=
+
+plot_data = lapply(tars, function(x) {
+    dd = mod_data[mod_data$id == x & mod_data$site == "RES",]
+    tmp = prep_plot_data(dd$yield_kg / 1000, dd$yrs_in_trial)
+    tmp$yrs_in_trial = as.numeric(row.names(tmp))
+    tmp
+    })
+
+plot(mean ~ yrs_in_trial,
+     data = plot_data[[1]],
+     xlim = c(-6, 32), ylim = c(6.5,14.5),
+     pch = 16, #main = gsub("M","M-", tars[i]),
+     ylab = expression("Yield ( kg ha"^-1~")"),
+     xlab = "Years after release")
+segments(x0 = plot_data[[1]]$yrs_in_trial,
+         y0 = plot_data[[1]]$mean + plot_data[[1]]$se,
+         y1 = plot_data[[1]]$mean - plot_data[[1]]$se)
+points(mean ~ yrs_in_trial,
+       data = plot_data[[2]],
+       pch = 17, col = "grey40")
+segments(x0 = plot_data[[2]]$yrs_in_trial,
+         y0 = plot_data[[2]]$mean + plot_data[[2]]$se,
+         y1 = plot_data[[2]]$mean - plot_data[[2]]$se,
+         col = "grey40")
+abline(lm(yield_kg/1000 ~ yrs_in_trial,
+          data = mod_data[(mod_data$id %in% tars & mod_data$site == "RES"),]),
+       lty = 1, col = "grey50")
+legend("topright", legend = gsub("M","M-", tars),
+       pch = c(16, 17), col = c("black","grey40"))
+
 @
-  \caption{Distribution of yield over the period 1995 -- 2015.} \label{fig:plot}
+\caption{The performance of two major cultivars at the CA Rice
+  Experiment Station (RES) since the year each cultivar was
+  released. Both M-202 ($\bullet$) and M-206 ($\blacktriangle$) have similar
+  intercepts, suggesting similar yield at the time of release and an
+  apparent yield advantage of the newer cultivar, M-206, might be
+  explained by a yield loss of M-202 over time. Points represent the
+  mean, and bars (|) the standard error.}
+\label{fig:compare_since_release}
 \end{figure}
 
-\begin{figure}[h]
-<<label=plot2, fig=TRUE, echo=FALSE>>=
+\newpage
 
-  boxplot(yield_kg ~ id, data = mod_data,
-          xlab = "Cultivar",
-          ylab = "Yield (lbs/ac)")
+\begin{figure}[ht]
+<<fig=TRUE, echo=FALSE>>=
+tars <- c("M202","M206")
+
+boxplot(yield_kg/1000 ~ id, data = mod_data[mod_data$id %in% tars,],
+        ylab = expression("Grain yield ( Mg ha"^-1~")"))
+
 @
-\caption{Boxplot of yields for released medium grain cultivars for
-  the period 1995--2015.}
-\label{fig:raw_yield}
+\caption{The performance of two major cultivars at the CA Rice
+Experiment Station (RES) over the period 1995 -- 2015. The newer
+cultivar, M-206, appears to have a slight yield advantage (an
+estimated \Sexpr{m202_v_206} kg ha\textsuperscript{-1}) over the older
+cultivar, M-202, though this comparison does not account for the 10
+year difference in years since release between the cultivars.}
+\label{fig:compare_vars}
 \end{figure}
 
+% \begin{figure}[h]
+
+% <<label=plot1, fig=TRUE, echo=FALSE>>=
+% boxplot(yield_kg ~ year, data = mod_data,
+%         xlab = "Year", ylab = expression("Yield ( kg ha"^-1~")"))
+% @
+%   \caption{Distribution of yield over the period 1995 -- 2015.} \label{fig:plot}
+% \end{figure}
+
+% \begin{figure}[h]
+% <<label=plot2, fig=TRUE, echo=FALSE>>=
+
+%   boxplot(yield_kg ~ id, data = mod_data,
+%           xlab = "Cultivar",
+%           ylab = "Yield ( kg/ha")
+% @
+% \caption{Boxplot of yields for released medium grain cultivars for
+%   the period 1995--2015.}
+% \label{fig:raw_yield}
+% \end{figure}
+
 \end{document}

### 2017-05-10 src/h1_model.R

@@ -3,8 +3,8 @@
 
 library(rstan)
 library(rstanarm)
-source("prep_model_data.R")
-
+# source("prep_model_data.R")
+load("../data/model_data.rda")
 ## mm <- stan_model(file = "yield_model.stan", verbose = TRUE)
 
 ## n.level <- function(x){
@@ -46,12 +46,14 @@ source("prep_model_data.R")
 
 i <- mod_data$site == "RES"
 
-RES_yrs_fit <- stan_glmer(yield_cs ~ (1|year) + (1|id) + yrs_in_trial,
+# RES_yrs_fit <- stan_glmer(yield_cs ~ (1|year) + (1|id) + yrs_in_trial,
+#                           data = mod_data[i,], iter = 1000, cores = 2L)
+
+# RES_rel_fit <- stan_glmer(yield_cs ~ (1|year) + (1|id) + release_yr_c,
+#                           data = mod_data[i,], iter = 1000, cores = 2L)
+
+RES_full = stan_glmer(yield_cs ~ (1|year) + (1|id) + release_yr_c + yrs_in_trial,
                           data = mod_data[i,], iter = 1000, cores = 2L)
 
-RES_rel_fit <- stan_glmer(yield_cs ~ (1|year) + (1|id) + release_yr_c,
-                          data = mod_data[i,], iter = 1000, cores = 2L)
 
-
-save( #yrs_fit, release_fit,
-     RES_yrs_fit, RES_rel_fit, file = "../output/h1_fit.Rda")
+save(RES_full, file = "../output/h1_fit.Rda")

### 2017-05-10 src/munge_pre95.R

@@ -1,0 +1,46 @@
+# Use data compiled pre-1995
+
+library(readxl)
+
+ff = "../data/RES Variety Trail 1984-1994.xls"
+n_sheets = length(excel_sheets("../data/RES Variety Trail 1984-1994.xls"))
+
+pre95_head = lapply(1:n_sheets, read_excel, path = ff, n_max = 5, col_types = "text", col_names = FALSE)
+
+lapply(pre95_head, function(x) grep(pattern = "Location", as.data.frame(x)[,1]))
+
+pre95 = lapply(1:n_sheets, read_excel, path = ff, skip = 3)
+
+pre95 = lapply(pre95, as.data.frame, stringsAsFactor = FALSE)
+
+pre95 = do.call(rbind, pre95)
+
+
+# fix colnames by hand
+cc = c(
+"site"       ,    "trial_type"     ,     "county"              ,
+ "year"           ,    "planted"  ,     "Advanced/Prelminary" ,
+ "#"              ,    "rep"           ,     "#__2"                ,
+ "Experimental ID",    "id"           ,     "#__3"                ,
+ "#__4"           ,    "Source"         ,     "grain_type"                ,
+ "seedling_vigor"          ,    "Date"           ,     "dth"                ,
+ "height"           ,    "lodging"         ,     "(sq ft)"             ,
+ "(lbs)"          ,    "moisture"              ,     "yield_lb"       ,
+ "Notes"          )   
+
+colnames(pre95) = cc
+
+tmp = lapply(seq_along(all_vt_result), function(i){
+    idx = match(colnames(all_vt_result)[i], cc)
+    # browser()
+    if(is.na(idx)){
+        c(all_vt_result[,i], rep(NA, nrow(pre95)))
+    } else {
+        c(all_vt_result[,i], pre95[,idx])
+    }
+    # print("hi")
+})
+
+all_data = as.data.frame(tmp, stringsAsFactors = FALSE)
+
+colnames(all_data) = colnames(all_vt_result)

### 2017-05-10 src/prep_model_data.R

@@ -4,13 +4,16 @@ load("../data/all_vt_weather.Rda")
 ## Replace 08Y3269 with M209
 all_vt_result$id[all_vt_result$id == "08Y3269"] <- "M209"
 
+## Merge with pre95 data
+source("../src/munge_pre95.R")
+
 ## Subset to medium grains
-vars <- grepl("^M[0-9]{3}$", all_vt_result$id)
+vars <- grepl("^M[0-9]{3}$", all_data$id)
 
 ## Do not include San Joaquin sites - known to hammer with cold and
 ## Not managed similar to most of CA rice production
-no_sj <- all_vt_result$county != "SAN JOAQUIN"
-mod_data <- all_vt_result[(vars & no_sj),]
+no_sj <- all_data$county != "SAN JOAQUIN"
+mod_data <- all_data[(vars & no_sj),]
 
 ## Exclude M-401/402 and Tucker site, per conversation with Bruce
 ## 10 Feb 2017:
@@ -43,4 +46,5 @@ mod_data$yield_cs <- (mod_data$yield_kg - center) / scl
 mod_data$yr_fact <- factor(mod_data$year)
 mod_data$yr_site_fact <- factor(paste(mod_data$year, mod_data$site))
 
-
+save(mod_data, scl, center, file = "../data/model_data.rda")
+################################################################################

### 2017-05-11 output/breed_paper.Rnw
Updates paper

@@
 model. Differences between years and cultivars were included in the
-model as random effects according to:
 
-\begin{equation}
- yield_i = \alpha + \beta_{years\ since\ release\ i}* x_{years\ since\
-   release\ i} + \beta_{release\ year\ i}*x_{release\ year\ i} + \beta_{year\ i} + \beta_{cultivar\ i}
-\end{equation}
+%
+% model as random effects according to:
 
-Where $yield_i$ is the grain yield for plot $i$, $\alpha$ is the
-average yield for the RES across varieties and years (intercept),
-$\beta_{years\ since\ release\ i}$ is effect of $x_{years\ since\
-release\ i}$, the number of years $i$'s cultivar has been in the
-trials (negative for years prior to release, zero at the year of
-release), $\beta_{release\ year\ i}$ is the effect of $x_{release\
-year\ i}$, the year of release for $i$'s cultivar, $\beta_{year}$ is
-the random effect of year, and $\beta_{cultivar}$ is the random effect
-of cultivar.
+% \begin{equation}
+%  yield_i = \alpha + \beta_{years\ since\ release\ i}* x_{years\ since\
+%    release\ i} + \beta_{release\ year\ i}*x_{release\ year\ i} + \beta_{year\ i} + \beta_{cultivar\ i}
+% \end{equation}
+
+% Where $yield_i$ is the grain yield for plot $i$, $\alpha$ is the
+% average yield for the RES across varieties and years (intercept),
+% $\beta_{years\ since\ release\ i}$ is effect of $x_{years\ since\
+% release\ i}$, the number of years $i$'s cultivar has been in the
+% trials (negative for years prior to release, zero at the year of
+% release), $\beta_{release\ year\ i}$ is the effect of $x_{release\
+% year\ i}$, the year of release for $i$'s cultivar, $\beta_{year}$ is
+% the random effect of year, and $\beta_{cultivar}$ is the random effect
+% of cultivar.
 
@@
 
-legend("bottomright", legend = c("w/ years since release",
-                                 "w/o years since release",
+legend("bottomright", legend = c("w/ years since release adj.",
+                                 "w/o years since release adj.",
                                  "90% Credible interval"),
@@
        border = NA,
-       merge = TRUE)
+       merge = TRUE,
+       cex = 0.8)
 polygon(x = c(yrs, rev(yrs)),
@@
 
+\newpage
+
+<<echo=FALSE, results=tex>>=
+ids = aggregate(release_yr ~ id, data = mod_data, unique)
+colnames(ids) = c("Cultivar", "Release Year")
+ids = ids[order(ids[,2]),]
+ids[,2] = as.integer(ids[,2])
+ids[,1] = gsub("M", "M-", ids[,1])
+
+print(xtable(ids,
+             caption = "All named medium grain rice cultivars and their associated release year planted at the Rice Experiment Station (RES), Biggs, California from 1984 -- 2015.",
+             label = "tbl:ids"),
+      include.rownames = FALSE,
+      booktabs = TRUE,
+      caption.placement = "top")
+@ 
 % \begin{figure}[h]

### 2017-05-15 output/breed_paper.Rnw
Adds sources, re-arranged again

@@
 
+RES = res = mod_data$site == "RES"
+
 prep_plot_data = function(x, idx)
@@
 }
+
+release_yr_coef = round(fixef(RES_full)[2] * scl, 1)
+yrs_coef = round(fixef(RES_full)[3] * scl, 1)
+
+rel_CI = round(posterior_interval(RES_full, pars = "release_yr_c") * scl, 1)
+m202_v_206 = round(release_yr_coef * 19, 0)
 
@@
 
-The Rice Experiment Station (RES) of the CCRRF is located northwest of
+The Rice Experiment Station (RES) of the CCRRF is located north-west of
 Biggs, CA (39.4648, -121.7342), and has been the central location for
-the development of improved rice varieties adapted for CA since
+CCRRF's efforts to develop improved rice varieties adapted for CA since
 1969. The climate at the RES is Mediterranian, characterized by mild
@@
 
-\subsection{Yield improvements over time}
-
-<<echo=FALSE>>=
-rel_CI = round(posterior_interval(RES_full, pars = "release_yr_c") * scl, 1)
-
-m202_v_206 = round(fixef(RES_full)[2] * scl * 19, 0)
-# m202_v_206 = round((posterior_interval(RES_full, pars = "release_yr_c") * scl) * 19, 0)
-
-# m202_v_206 = ((posterior_predict(RES_full, newdata = data.frame(release_yr_c = c(-15, 4)),
-#                                  re.form = NA) * scl) + center)
-# m202_v_206 = m202_v_206[,2] - m202_v_206[,1]
-# m202_v_206 = round(quantile(m202_v_206, p = c(0.05, 0.5, 0.95)), 1)
-
-@
-
-There was a positive relationship between the year of release and
-yield, with an estimated \Sexpr{round(fixef(RES_full)[2] * scl, 1)}
-kg ha\textsuperscript{-1} year\textsuperscript{-1} increase (90\% CI:
-\Sexpr{rel_CI[1]} -- \Sexpr{rel_CI[2]})
-(Fig. \ref{fig:release}). Comparing the two dominate cultivars M-202,
-which was widely planted from the late 1980s until the early 2000s,
-and M-206, currently the most widely planted CRRFS cultivar
-(approximately 50\% of planted area), there is an estimated
-\Sexpr{m202_v_206} kg ha\textsuperscript{-1} yield advantage of
-M-206 over M-202 (Figs \ref{fig:release} \& \ref{fig:compare_yrs}).
-
 \subsection{Yield decreases over time}
@@
 
-RES = res = mod_data$site == "RES"
 plot_data = prep_plot_data(mod_data$yield_kg[res] / 1000,
@@
 abline(a = ((fixef(RES_full)[1] * scl) + center) / 1000 ,
-       b = fixef(RES_full)[3] * scl/1000)
+       b = yrs_coef/1000)
 polygon(x = c(yrs_in, rev(yrs_in)),
@@
         col = rgb(0,0,0,0.1))
-text(5, 8.0, bquote(.(round(fixef(RES_full)[3] * scl , 1))~kg~ha^-1~year^-1))
+text(5, 8.0, bquote(.(yrs_coef)~kg~ha^-1~year^-1))
 
@@
   cultivar is release is estimated to be
-  \Sexpr{round(fixef(RES_full)[3] * scl, 1)} kg ha\textsuperscript{-1}
+  \Sexpr{yrs_coef} kg ha\textsuperscript{-1}
   year\textsuperscript{-1}. The shaded area is the 90\% credible
@@
 abline(a = ((((fixef(RES_full)[1] * scl) + center) ) -
-           (2000 * fixef(RES_full)[2] * scl)) / 1000,
-       b = fixef(RES_full)[2] * scl/1000)
+           (2000 * release_yr_coef)) / 1000,
+       b = release_yr_coef/1000)
 abline(lm(yield_kg/1000 ~ release_yr, data = mod_data[res,]), lty = 2)
@@
 
-# text(1990, 8.0, bquote(.(round(fixef(RES_full)[2] * scl , 1))~kg~ha^-1~year^-1))
+# text(1990, 8.0, bquote(.(round(release_yr_coef))~kg~ha^-1~year^-1))
 
@@
 slope estimated from all medium grain cultivars
-(\Sexpr{round(fixef(RES_full)[3] * scl, 1)} kg ha\textsuperscript{-1} year\textsuperscript{-1}).}
+(\Sexpr{yrs_coef} kg ha\textsuperscript{-1} year\textsuperscript{-1}).}
 \label{fig:compare_yrs}
@@
 <<echo=FALSE>>=
-all_CI <- round(posterior_interval(RES_full, pars = "yrs_in_trial") * scl, 1)
 RES_CI <- round(posterior_interval(RES_full, pars = "yrs_in_trial") * scl, 1)
@@
 
-When yields are arranged according to the years since release, there
-is a small decrease in yield over time once a cultivar is released,
-though the estimated effect is small compared to the variability in
-the data set. This decrease is estimated to be
-\Sexpr{round(fixef(RES_full)[3] * scl, 1)} kg
-ha\textsuperscript{-1} year\textsuperscript{-1} (90\% CI: \Sexpr{RES_CI[2]} --
-\Sexpr{RES_CI[1]}). The oldest cultivar
-in the trials is estimated to have lost
-\Sexpr{round(fixef(RES_full)[3] * scl * 36 , 1)} kg
-ha\textsuperscript{-1} since release.
+When yields are compared accounting for differences in the years since
+release at the time of comparision, there is a small decrease in yield
+over time once a cultivar is released
+(\Sexpr{yrs_coef}
+kg ha\textsuperscript{-1} year\textsuperscript{-1}, 90\%
+CI: \Sexpr{RES_CI[2]} -- \Sexpr{RES_CI[1]}). The oldest cultivar in
+the trials is estimated to have lost \Sexpr{round(yrs_coef*36,1)}
+kg ha\textsuperscript{-1} since release. Again
+comparing the yield of two prominant cultivars, M-202 and M-206, the
+two cultivars are experiencing a similar trend over time to eachother
+(Fig. \ref{fig:compare_yrs}) and the estimate from all released medium
+grains. The estimated yield advantage of the newer cultivar (M-206)
+(\Sexpr{m202_v_206} kg ha\textsuperscript{-1}) can be explained by the
+differences in the amount of time that has passed since each cultivar
+was released (Fig. \ref{fig:compare_yrs} \&
+\ref{fig:compare_since_release})
 
-Again comparing the yield of two prominant cultivars, M-202 and M-206,
-arranged not by the year but rather the years since release, we find
-that the two cultivars are compariable and that the estimated yield
-advantage (\Sexpr{m202_v_206} kg ha\textsuperscript{-1}) can be
-explained by the differences in the amount of time that has passed
-since each cultivar was released
-(Fig. \ref{fig:compare_yrs} \& \ref{fig:compare_since_release})
+\subsection{Yield improvements over time}
+
+<<echo=FALSE>>=
+# m202_v_206 = round((posterior_interval(RES_full, pars = "release_yr_c") * scl) * 19, 0)
+
+# m202_v_206 = ((posterior_predict(RES_full, newdata = data.frame(release_yr_c = c(-15, 4)),
+#                                  re.form = NA) * scl) + center)
+# m202_v_206 = m202_v_206[,2] - m202_v_206[,1]
+# m202_v_206 = round(quantile(m202_v_206, p = c(0.05, 0.5, 0.95)), 1)
+
+@
 
@@
 there was no increase in yield potential over time once the
-differences between the years since release were accounted for (Fig. \ref{fig:release})
-(\Sexpr{round(fixef(RES_full)[2] * scl, 1)} kg ha\textsuperscript{-1}
-year\textsuperscript{-1}). This suggests that the overall progress on
-yield over the period from 1980 -- 2016 is approximately
-\Sexpr{round(fixef(RES_full)[3] * scl * 36, 1)} lbs acre\textsuperscript{-1}.
+differences between the years since release were accounted for
+(Fig. \ref{fig:release}) (\Sexpr{release_yr_coef}
+kg ha\textsuperscript{-1} year\textsuperscript{-1}). This suggests
+that the overall progress on yield over the period from 1980 -- 2016
+is approximately \Sexpr{round(release_yr_coef * 36, 0)} kg
+ha\textsuperscript{-1}.
+
+When differences between the number of years since each cultivar's
+release are taken into account, the estimated increase in grain yield
+over time is estimated to be \Sexpr{release_yr_coef}
+kg ha\textsuperscript{-1} year\textsuperscript{-1} (90\% CI:
+\Sexpr{rel_CI[1]} -- \Sexpr{rel_CI[2]}). Therefore, the evidence is
+consistent with no increase in yield potential over the study
+period. If the differences in years since release of each cultivar is
+ignored, there appears to be a small increase over time (approximately
+\Sexpr{round(coef(lm(yield_kg ~ release_yr, mod_data[res,])), 1)})
+(Fig. \ref{fig:release}), much less than previous estimates of
+\Sexpr{54 * 1.12} kg ha\textsuperscript{-1} year\textsuperscript{-1}
+(CITE).  Comparing the two dominate cultivars M-202, which was widely
+planted from the late 1980s until the early 2000s, and M-206,
+currently the most widely planted CRRFS cultivar (approximately 50\%
+of planted area), there is an estimated \Sexpr{m202_v_206} kg
+ha\textsuperscript{-1} yield advantage of M-206 over M-202 (Figs
+\ref{fig:release} \& \ref{fig:compare_vars}), though both cultivars
+yielded similarly at their respective year of release
+(Fig. \ref{fig:compare_yrs}).
 
@@
 
-\subsection{Complications}
+\subsection{Improving yield potential over time}
 
-There is strong collinearity present between the year of release and
-the number of years in the trials which is complicating the
-simultaneous estimation of these effects. With the current size of the
-data set (N = \Sexpr{nrow(mod_data)}), it is impossible to estimate
-the decrease over time since release and the improvement in yield over
-time with each new release in the same model. This is why each effect was
-estimated using a separate regression.
+Overall, we found evidence consistent with a no improvements in yield
+potential in medium grain rice cultivars released by CCRRF's breeding
+improvement program over the period \Sexpr{min(mod_data[,'year'])} --
+\Sexpr{max(mod_data[,'year'])}. Rather, it seems possible that the
+breeding program has been maintaining yield potential in the face of
+changing biotic and abiotic pressures, which are expected to decrease
+yield over time \citep{Peng2010}. This highlights both the need
+for maintanence breeding and the difficulty of increasing genetic
+yield potential as yield approaches the physiological limit. Without
+maintanence breeding continuously selecting cultivars well adapted to
+current conditions, CA rice production systems are predicted to have
+lost an estimated \Sexpr{round(yrs_coef * 31, 0)} kg
+ha\textsuperscript{-1} grain yield over the study period.
 
-Increasing the size of the data to the full data set (N =
-\Sexpr{nrow(all_vt_result)}), allows the simultaneous estimation
-and results in similar estimated effect sizes for each of these
-mechanisms. However, this necessitates the inclusion of many entries
-which have characteristics not similar to conventional cultivars.
+However, the goal of a breeding program is to not only recover the
+benchmark yield of the old standard cultivar but to surpass it. Given
+the estimated yield decreases over time, to hit an ambitious target of
+500 kg ha\textsuperscript{-1} (approximately 5\%) increase in yield
+per release (with roughly 5 years between releases), a new cultivar
+needs to yield \Sexpr{round(yrs_coef * -5, -1)} more
+in side-by-side comparison on top of the 500 kg ha\textsuperscript{-1}
+target improvement. Often side-by-side yield improvements (not
+considering the difference in years since release) are closer to 2 --
+3\% per release, in which case no absolute improvements might be made
+beyond maintenance. The evidence here is consistent with the latter
+situation occuring in CA, which suggests that the genetic yield
+potential has not increased substantially in the last 30+ years (Fig. \ref{fig:release}).
+
+There has been some yield improvements over the last 30 years in the
+CA state average yield (Fig. \ref{fig:state_yield}), but the evidence
+here suggests mechanisms other than increased genetic yield potential
+are responsible (Fig. \ref{fig:compare_yrs}). Improvement in agronomic
+management, including more efficient nutrient management
+\citep{Lundy2012, Lundy2015, linquist2009}, weed management
+\citep{Saito2008, Pittelkow2012, linquist2008}, etc. may all be
+contributing to further increases. However, the gains here are less
+than those experienced during the ``Green revolution'', and by their
+nature cannot surpass the genetic yield potential
+\citep{Fischer2015}. Judging from yields achieved at the RES
+(Figs. \ref{fig:RES_year} \& \ref{fig:compare_yrs}), a
+researcher-managed production system, state average yields
+(approximately 10 Mg ha\textsuperscript{-1} at the time of writing;
+Fig. \ref{fig:state_yield}) may be approaching the agronomic limit.
+
+To be clear, our investigation here is narrowly focused on the
+rough-rice/field grain yield. Other improvements due to plant
+improvement, including qualtity \citep{mckenzie1993} and milling yield
+\citep{mckenzie1994} have been made over the study period and are no
+less important. However, planning to meet future demand necessitates
+increases in grain yield \citep{Godfray2010, Foley2011a,
+Mueller2012}. The evidence here suggests that to accomplish true
+increases over time, the improvement target needs take decreases in
+the current cultivars' yield over time into account. While side-by-side
+comparisions will continue to be useful, comparisons across years to
+each cultivar's benchmark yield at the time of release should provide
+information on the absolute improvements being made in yield potential over time.
+
+\begin{figure}[ht]
+<<fig = TRUE, echo = FALSE>>=
+
+plot(1, type = "n",
+     xaxt = "n",
+     xlab = "Time",
+     ylab = expression("Grain yield ( Mg ha"^-1~")"),
+     ylim = c(8.5,11.5),
+     xlim = c(0,40),
+     xaxs = "i")
+
+arrows(12, 8.2, 27, 8.2, xpd = TRUE, lwd = 2)
+abline(a = 10, b = fixef(RES_full)[3], lty = 2, lwd = 2)
+text(25, 10.75, "(3) target above maintenance", cex = 0.8,
+     srt = 15)#((atan(fixef(RES_full)[3] + 0.06) * 180))/pi)
+abline(h = 10, lty = 3, lwd = 2)
+text(25, 10.1, "(1) current situation", cex = 0.8)
+abline(a = 10, b = fixef(RES_full)[3] + 0.06, lty = 4, lwd = 2)
+text(25, 9.25, "(2) no maintenance breeding", cex = 0.8,
+     srt = -20)
+
+@ 
+\caption{A conceptual diagram highlighting three scenarios:
+  (1) maintenance breeding, where losses over time are offset by
+  improvements via plant improvement, (2) no maintenance breeding,
+  where grain yield slowly decreases over time, and (3) absolute
+  improvement in grain yield, where gains in grain yield exceed the
+  losses over time.}
+\label{fig:theory}
+\end{figure}
 
@@
 
- We postulate that true
-yield increases are only possible if the improvement in yield of each
-new cultivar exceeds former's benchmark yield at the time of release,
-and therefore side-by-side comparisons are potentially misleading.
-
-There is evidence of a small decrease in yield over time once a cultivar is
-released. This complicates comparisons between new entries and checks,
-since the check will decrease over time. This decrease will manifest
-as an apparent larger yield advantage of the new entry over the check.
-There is evidence of a small increase in yield with each new released
-cultivar. Therefore, it appears that successive cultivar releases are
-not merely recovering to the old benchmark but are surpassing
-it. However, progress is slower than previously estimated
-(\Sexpr{round(fixef(RES_full)[2] * scl, 1)} lbs
-acre\textsuperscript{-1} year\textsuperscript{-1} increase verses a
-previous estimate of 54 lbs acre\textsuperscript{-1}
-year\textsuperscript{-1}).
-
-The practical impact of these results is: Breeders need to be aware of
-the potential for the yield of checks to decrease over time. Due to
-this, comparisons to older checks might give a false sense of
-improvement. In order to have greater improvements in yield, the
-target yield improvement for a new cultivar needs to be higher to
-offset the impact of yield decrease once a cultivar is
-released. Breeding is essential to maintaining yield in rice. Yield is
-not stable and without continuous effort, yield potential will
-decrease over time.
+Improvement in grain yield is essential to meeting the challenges of
+feeding a growing population in a changing world. To achieve these
+improvements, we show here that (1) continuous maintenance via plant
+improvement is required, without which yields would decrease over
+time, (2) the challenge of improving genetic yield potential is
+substantial, since absolute gains require improvement in yield in
+excess of the maintenance threshold, and (3) increasing yield through
+improved management (agronomy) cannot exceed the ceiling of genetic
+yield potential. Taken together, these conclusions underscore both the
+necessity and substantial challenges facing rice breeding. We suggest
+here that historical benchmarks be used in to evaluate promising rice
+cultivars rather than side-by-side comparisions to maintain
+perspective of the absolute gains being made.
 
@@
 <<echo=FALSE, results=tex>>=
+
 CIs = posterior_interval(RES_full)[1:3,] * scl
@@
 tars <- c("M202","M206")
+comp_yrs = 2002:2005
 
-boxplot(yield_kg/1000 ~ id, data = mod_data[mod_data$id %in% tars,],
+boxplot(yield_kg/1000 ~ id, data = mod_data[(mod_data$id %in% tars) &
+                                            (mod_data$year %in% comp_yrs) &
+                                            mod_data$site == "RES"  ,],
         ylab = expression("Grain yield ( Mg ha"^-1~")"))
@@
 @
-\caption{The performance of two major cultivars at the CA Rice
-Experiment Station (RES) over the period 1995 -- 2015. The newer
-cultivar, M-206, appears to have a slight yield advantage (an
-estimated \Sexpr{m202_v_206} kg ha\textsuperscript{-1}) over the older
+\caption{An example of a typical comparison between a older (M-202)
+  and newer (M-206) cultivar. Shown is the performance of these cultivars at the CA Rice
+  Experiment Station (RES) over the period \Sexpr{min(comp_yrs)} --
+  \Sexpr{max(comp_yrs)}, approximately the comparision made to
+  determine if a new cultivar performs sufficiently better. The newer
+  cultivar, M-206, appears to have a yield advantage over the older
   cultivar, M-202, though this comparison does not account for the 10

### 2017-06-21 output/breed_paper.Rnw
Commit before messing with figures

@@
 \usepackage{etoolbox}
+\usepackage{float}
 \doublespacing
@@
 }
+
+\newcommand{\tsups}{\textsuperscript}
 \usepackage{Sweave}
@@
 \author[ucd]{Matthew B. Espe\corref{corresauthor}}
+\cortext[corresauthor]{Corresponding author} \ead{mespe@ucdavis.edu}
 \author[ucd]{David MacKill}
@@
 \author[res]{Kent McKenzie}
+\author[ucd]{Chris van Kessel}
 \author[ucd]{Bruce A. Linquist}
-
-\cortext[corresauthor]{Corresponding author} \ead{mespe@ucdavis.edu}
 
@@
 
-  Using the variety trial data, we find consistent estimates of the
-  reduction of yield once a variety is released, estimated to be between
-  -12 to -22 kg per year. This effect persists using all data, only
-  named/released cultivars, or only released cultivars only at the Rice
-  experiment station. This effect is offset by yield improvements over
-  the years, estimated to be small but positive effect of
+Meeting the challenge of feeding an growing population with
+increasingly limited resources will require increasing the yield
+potential of staple crops, such as rice. Yet many high-yielding,
+intensive production systems have experienced slow rates of yield
+improvement in recent years despite demonstrated increased yield
+potential of new genetics over old in side-by-side comparisons. We
+analyzed experimental data from one such system, California (CA) rice
+production, in order to better understand what improvements have been
+made in the genetic yield potential via plant breeding. Specifically,
+the hypothesis that a rice variety's yield erodes over time after the
+initial selection was tested as an explanation for how side-by-side
+comparisons show increased yield while no substantive increases have
+been seen in farmers' fields. We confirmed an apparent increase in
+yield in side-by-side comparisons between old and newer varieties when
+the difference between the years since release was not
+considered. However, we demonstrate this apparent increase in yield
+can be explained by the differences in the time since selection
+between cultivars. The impact of yield erosion was estimated to be
+29.8 kg ha\tsups{-1} year\tsups{-1} after initial
+selection. Once this effect was considered, the apparent yield
+advantage of newer cultivars over old was uncertain. This underscores
+the importance of breeding to maintain yield and the challenges of
+increasing the genetic yield potential.
 
@@
 
-\begin{keyword} rice, yield potential, temperature
+\begin{keyword} rice, yield potential, breeding
 \end{keyword}
@@
 
-\linenumbers
+% \linenumbers
 
@@
 load("h1_fit.Rda")
+load("bivar_fit.Rda")
 load("../data/all_vt_weather.Rda")
-
-RES = res = mod_data$site == "RES"
 
@@
 yrs_coef = round(fixef(RES_full)[3] * scl, 1)
+only_release_coef = round(fixef(RES_rel_fit)[2]*scl, 1)
 
+RES_CI <- round(posterior_interval(RES_full, pars = "yrs_in_trial") * scl, 1)
 rel_CI = round(posterior_interval(RES_full, pars = "release_yr_c") * scl, 1)
-m202_v_206 = round(release_yr_coef * 19, 0)
+only_release_CI = round(posterior_interval(RES_rel_fit, pars = "release_yr_c") * scl, 1)
 
+m202_v_206 = round(yrs_coef * -19, -1)
+
+id_est = ranef(RES_full)$id
+id_est = cbind(id_est, posterior_interval(RES_full, regex_pars = "id:M"))
+idx = order(id_est[,1], decreasing = FALSE)
+id_est = id_est[idx,]
 @
@@
 
-Constraints on ariable land are increasing simultaneous with the need
+Constraints on arable land are increasing simultaneous with the need
 to increase total production to meet growing demand \citep{Foley2011a,
@@
 Tittonell2014}. Historically, agricultural research has been
-successful in staving off the Mathesian disaster through continued
-yield improvement of staple crops, however more and more production
-systems are experiencing plateaus in grain yield \citep{Grassini2013,
-Peng2010}. If continued improvements per unit land area are no longer
-possible, increasing the area under cultivation will be needed to meet
-demand, which carries with it undesirable ecological
-implications. Therefore, it is of critical importance to better
-understand why grain yields have not continued to increase in
-instensified production systems.
+successful in staving off the ``Malthusian catastrophe'' of demand
+surpassing supply via continued yield improvement of staple crops,
+however more and more production systems are experiencing plateaus in
+grain yield \citep{Grassini2013, Peng2010}. If continued improvements
+per unit land area are no longer possible, increasing the area under
+cultivation will be needed to meet demand, which carries with it
+undesirable ecological implications. Therefore, it is of critical
+importance to better understand why grain yields have slowed or
+plateaued in intensified production systems.
 
-California (CA) rice production represents one such rice production
+California (CA) rice production represents one such production
 system. Rice yields in CA during the period from 1920 -- 1990
-increased rapidly, in part due to improvements in rice genetics (i.e.,
-the Green Revolution) through the efforts of the California
-Cooperative Rice Research Foundation (CCRRF; a collaboration between
-the University of California, the USDA-Agricultural Research Service,
-and farmer-funded research). However, since 1990 CA state average
-yields have increased more slowly (Fig. \ref{fig:state_yield}) despite
-demonstrated improvements of grain yield as the result of plant
-breeding \citep{ccrrf}, with each new cultivar having a yield
-advantage over the previous cultivars in side-by-side comparisons in
-the annual Statewide Variety Trials. Yet, taken across years, these
-same reports show stagnant yields; for example, at the year of release
-(1985), cultivar M-202 averaged 11 Mg ha\textsuperscript{-1}
-\citep[14\% MC;][]{m202}. In 2015, the average yield of the newest
-released cultivar M-209 was 10.8 Mg ha\textsuperscript{-1}
-\citep{m209}. In other systems, it has been shown that rice cultivars
-experience a lack of adaptation to changing biotic and abiotic
-conditions, resulting in a decrease of a cultivar's performance over
-time \citep{Peng2010, Peng2000, Datta1995}. We sought to explore this
-phenomonon and better understand how grain yields have increased in
-the breeders' fields without a corresponding increase in yield in the
-farmers' fields. Two competing hypotheses were tested: (1) the yield
-advantage offered by newly developed cultivars is not greater than
-previous cultivars, or (2) yield increases gained through plant
-breeding are being offset by decreases in the yield of released
-cultivars over time.
+increased rapidly (Fig. \ref{fig:state_yield}), in part due to
+improvements in rice genetics (i.e., the ``Green Revolution'') through
+the efforts of the California Cooperative Rice Research Foundation
+(CCRRF; a collaboration between the University of California, the
+USDA-Agricultural Research Service, and farmer-funded
+research). However, since 1990 CA state average yields have increased
+more slowly (Fig. \ref{fig:state_yield}) despite continuous plant
+improvement \citep{ccrrf}. For the last 30 years, each newly developed
+cultivar has demonstrated a yield advantage over the previous
+cultivars in the Statewide Variety Trials (an annual multi-site
+variety trial program that compares experimental entries side-by-side
+with ``check'' cultivars). Yet, taken across years, these same reports
+show stagnant yields. For example, at the year of release (1985),
+cultivar M-202 averaged 11 Mg ha\tsups{-1} \citep[14\%
+MC;][]{m202}. In 2015, the reported average yield of the newest released
+cultivar M-209 was 10.8 Mg ha\tsups{-1} \citep{m209}. Yet this same
+report, M-209 demonstrated a X\% yield advantage over M-202. In other rice
+systems, a decrease in a cultivar's yield performance over time has
+been observed \citep{Peng1998,
+Peng2010, Peng2000, Datta1995}, often termed yield erosion
+\citep{Peng1998}. It has been speculated that this is due to the
+inability of a selected cultivar's genetics to adapt to changing  biotic and
+abiotic conditions. We sought to quantify this phenomenon to better
+understand how plant improvement and yield erosion are contributing to
+the slow yield improvement in farmers' fields over the last
+30 years in CA. Specifically, two competing hypotheses were
+tested: (1) the yield advantage of each newly developed cultivars is
+greater than previous cultivars, and (2) the erosion of yield over
+time is creating the illusion of improved yields (i.e., the new
+cultivar is recovering the old benchmark as the older cultivar's yield
+erodes).
 
-\begin{figure}[ht]
+\begin{figure}[H]
 <<fig=TRUE, echo=FALSE>>=
@@
      xlab = "Year",
-     ylab = expression("Average grain yield ( Mg ha"^-1~")"))
+     ylab = expression("State avg. grain yield ( Mg ha"^-1~")"))
 mv_avg = filter(c(rep(mean(CA_data$Value[1:10] * 1.12), 5),
@@
 \caption{Average rice yields over the period \Sexpr{min(CA_data$Year)}
-  -- \Sexpr{max(CA_data$Year)} for California, USA. Rapid gains in
-    grain yield were experienced from approximately 1950 -- 1990, but
-    annual gains since 1990 have been slower. The line represents the
-    10 year moving average.}
+  -- \Sexpr{max(CA_data$Year)} for California, USA, as reported by the
+  US Dept. of Agriculture National Agricultural Statistics
+  Service. Rapid gains in grain yield were experienced from
+  approximately 1950 -- 1990, but annual gains since 1990 have been
+  slower. The line represents the 10 year moving average.}
 \label{fig:state_yield}
@@
 
-The Rice Experiment Station (RES) of the CCRRF is located north-west of
+The CCRRF Rice Experiment Station (RES) is located north-west of
 Biggs, CA (39.4648, -121.7342), and has been the central location for
-CCRRF's efforts to develop improved rice varieties adapted for CA since
-1969. The climate at the RES is Mediterranian, characterized by mild
+CCRRF's efforts to develop improved rice varieties adapted to CA since
+1969. The climate at the RES is Mediterranean, characterized by mild
 winters during which most of the annual precipitation occurs
@@
 as Esquon-Neerdobe clays with roughly 2\% SOM \citep{ssurgo} in the
-top 15cm. As the central location for the breeding program, the RES is
+top 15cm. As the central location for the CCRRF breeding program, the RES is
 researcher-managed (as opposed to farmer-managed trials) and conducts
@@
 
-Plots at the RES are managed similar to perscribed best management
+Plots at the RES are managed similar to prescribed best management
 practice, including land preparation, fertility management, and pest
@@
 to direct-seeding into pre-flooded fields. Plots were either 14.0 or
-18.5 m\textsuperscript{2}, and were harvested after physiological
+18.5 m\tsups{2}, and were harvested after physiological
 maturity using a small-plot combine. All yields were converted to 14\%
@@
 Yield was modeled as a response to the number of years since the
-variety was officially released using a mixed-effects linear
-model. Differences between years and cultivars were included in the
-
-%
-% model as random effects according to:
+variety was officially released and the year of release using a
+mixed-effects linear regression. To account for similarities between years
+and between cultivars, year and cultivar were included in the model as
+random effects.
 
@@
 Data was processed and models were fit in R, an environment for
-statistical computing \citep{R}. Models were fit using the 'rstanarm'
+statistical computing \citep{R}. Models were fit using the `rstanarm'
 package \citep{rstanarm}, an interface to Stan, a language for
@@
 including $\hat{R}$, effective size, and posterior predictive checks
-were examined before reporting results.
+were examined before reporting results. To test goodness of fit, the
+full model was compared to models containing only the years since
+release or the year of release using the `loo' package \citep{loo},
+an efficient means of conducting leave-one-out cross validation
+\citep{loo_paper}, with the full model estimated to have slightly
+better predictive accuracy compared to the reduced models.
 
@@
 
+\subsection{Genetic yield improvements}
+
+There was no conclusive evidence that the yields of rice cultivars
+selected by CCRRF have, on average, increased over the last 30
+years. The estimated change in grain yield per year was estimated to
+be a negligible \Sexpr{release_yr_coef} kg ha\tsups{-1}
+year\tsups{-1}, with both small positive and negative trends within
+the credible interval (90\% CI: \Sexpr{rel_CI[1]} --
+\Sexpr{rel_CI[2]}). Hence, the evidence here is consistent with little
+to no increase in yield potential over the study period. When yield
+erosion is not corrected for in the comparison (i.e., omitted from the
+model), there is an apparent small increase
+per year (approximately \Sexpr{only_release_coef} kg ha\tsups{-1}
+year\tsups{-1}; 90\% CI \Sexpr{only_release_CI[1]} --
+\Sexpr{only_release_CI[2]}) (Fig. \ref{fig:release}), though this
+estimate ignores the confounding effect of yield erosion over
+time.
+
+There do appear to be some differences between cultivars, with some
+cultivars showing higher yield than the overall average. The marginal
+estimate for the performance of a cultivar (averaged over the
+uncertainty in the other parameters in the model) shows three
+cultivars with higher than average yields:
+\Sexpr{paste(gsub('M','M-',rev(row.names(id_est))[c(1,2,4)]),collapse=', ')}
+(Fig. \ref{fig:ids_comp}). A fourth, M-209, might also have higher
+than average yield, though the CI overlaps with zero.
+
+\begin{figure}[H]
+
+<<fig=TRUE, echo=FALSE>>=
+yrs <- 1975:2025
+pp <- posterior_linpred(RES_full, re.form = NA,
+                        newdata = data.frame(release_yr_c = yrs - 2000,
+                                             yrs_in_trial = 0),
+                        prob = 0.5)
+pp <-  (pp * scl + center) / 1000
+ints = apply(pp, 2, quantile, p = c(0.05, 0.95))
+
+res = mod_data$site == "RES"
+comp_yrs = 2010:2015
+idx = mod_data$year %in% comp_yrs
+
+plot_data = prep_plot_data(mod_data$yield_kg[idx] / 1000 ,
+                           mod_data$id[idx])
+
+plot_data$tmp = tapply(mod_data$release_yr[idx],
+                       mod_data$id[idx], unique)
+
+colnames(plot_data) = c("yield_kg", "yield_se", "release_yr")
+
+plot(yield_kg ~ release_yr, data = plot_data, pch = 16, #col = rgb(0,0,0,0.25),
+     xlab = "Release year",
+     ylab = expression("Grain yield ( Mg ha"^-1~")"),
+     ylim = c(8, 12.5),
+     xlim = c(1980,2019))
+abline(a = ((((fixef(RES_full)[1] * scl) + center) ) -
+           (2000 * release_yr_coef)) / 1000,
+       b = release_yr_coef/1000)
+abline(lm(yield_kg/1000 ~ release_yr, data = mod_data[idx,]), lty = 2)
+with(plot_data, segments(x0 = release_yr, y0=yield_kg + yield_se,
+                         y1 = yield_kg - yield_se))
+text(x = plot_data$release_yr-0.4, y = plot_data$yield_kg + 0.07,
+     labels = gsub("M", "M-", row.names(plot_data)),
+     pos = 4, cex = 0.8)
+
+# legend("bottomright", legend = c("w/ years since release adj.",
+#                                  "w/o years since release adj.",
+#                                  "90% Credible interval"),
+#        lty = c(1,2, NA),
+#        bty = "n",
+#        fill = c(NA,NA, "grey"),
+#        border = NA,
+#        merge = TRUE,
+#        cex = 0.8)
+
+legend("bottomright", legend = c("w/ years since release adj.",
+                                 "w/o years since release adj."),
+       lty = c(1,2),
+       bty = "n",
+       border = NA,
+       merge = TRUE,
+       cex = 0.8)
+# polygon(x = c(yrs, rev(yrs)),
+        # y = c(ints[1,],rev(ints[2,])),
+        # border = NA,
+        # col = rgb(0,0,0,0.1))
+
+# text(1990, 8.0, bquote(.(round(release_yr_coef))~kg~ha^-1~year^-1))
+@
+\caption{The average yield ($\bullet$) and standard error (|) of
+medium grain rice cultivars by year of release. For display purposes,
+only data for experimental plots planted at the California Cooperative
+Rice Research Foundation Rice Experiment Station (California, USA)
+over the years \Sexpr{min(comp_yrs)} -- \Sexpr{max(comp_yrs)} are
+included. Ignoring difference in years since release (dashed line)
+suggests an increase in yield potential over time. However, correcting
+the comparison to account for yield erosion over time (solid line)
+suggests the apparent increase might be due to differences in years
+since release.}
+\label{fig:release}
+\end{figure}
+
 \subsection{Yield decreases over time}
 
-\begin{figure}[ht]
+\begin{figure}[H]
 
@@
 
-plot_data = prep_plot_data(mod_data$yield_kg[res] / 1000,
-                           mod_data$yrs_in_trial[res])
+plot_data = prep_plot_data(mod_data$yield_kg / 1000,
+                           mod_data$yrs_in_trial)
 
@@
   for medium grain rice varieties planted at the Rice Experiment Station
-  (RES), California, USA. The estimated decrease over time once a
-  cultivar is release is estimated to be
-  \Sexpr{yrs_coef} kg ha\textsuperscript{-1}
-  year\textsuperscript{-1}. The shaded area is the 90\% credible
-  interval.}
+  (RES), California, USA. The estimated erosion of yield over time
+  (averaged over differences in cultivar and year) is
+  estimated to be \Sexpr{yrs_coef} kg ha\tsups{-1}
+  year\tsups{-1}. Points ($bullet$) and bars (|) represent the mean and standard
+  error of the raw data, respectively. The
+  shaded area is the 90\% credible interval.}
 \label{fig:RES_year}
@@
 
-\begin{figure}[ht]
-<<fig=TRUE, echo=FALSE>>=
-yrs <- 1975:2025
-pp <- posterior_linpred(RES_full, re.form = NA,
-                        newdata = data.frame(release_yr_c = yrs - 2000,
-                                             yrs_in_trial = 0),
-                        prob = 0.5)
-pp <-  (pp * scl + center) / 1000
-ints = apply(pp, 2, quantile, p = c(0.05, 0.95))
+There is estimated to be a small loss of yield over time once a
+cultivar is selected (i.e., yield erosion) (\Sexpr{yrs_coef} kg
+ha\tsups{-1} year\tsups{-1}, 90\% CI:
+\Sexpr{RES_CI[2]} -- \Sexpr{RES_CI[1]}). The oldest cultivar in the
+trials is estimated to have lost \Sexpr{round(yrs_coef*-36/1000,1)} Mg
+ha\tsups{-1} since release. Comparing two dominate
+medium grain rice cultivars (M-202, which was widely planted from the late 1980s until
+the early 2000s, and M-206, the most widely planted CRRFS
+cultivar at approximately 50\% of planted area at the time of
+writing), the two cultivars experienced similar trends over time to
+each other (Fig. \ref{fig:compare_yrs}) and to the estimate from all
+released medium grains (Table \ref{tbl:coef}). There is an estimated
+\Sexpr{m202_v_206} kg ha\tsups{-1} yield advantage of M-206
+over M-202 at the time of writing due solely to the erosion of M-202's
+yield over time (Figs \ref{fig:release} \&
+\ref{fig:compare_vars}). Further supporting this estimate, both cultivars yielded
+similarly at their respective year of release
+(Fig. \ref{fig:compare_yrs}), despite the fact that M-202 was released
+19 years prior to M-206 (Fig. \ref{fig:release}, Table \ref{tbl:ids}).
 
-res = mod_data$site == "RES"
-plot_data = prep_plot_data(mod_data$yield_kg[res]/1000 ,
-                           mod_data$id[res])
-
-plot_data$tmp = tapply(mod_data$release_yr[res],
-                       mod_data$id[res], unique)
-
-colnames(plot_data) = c("yield_kg", "yield_se", "release_yr")
-
-plot(yield_kg ~ release_yr, data = plot_data, pch = 16, #col = rgb(0,0,0,0.25),
-     xlab = "Release year",
-     ylab = expression("Grain yield ( Mg ha"^-1~")"),
-     ylim = c(8, 12.5),
-     xlim = c(1980,2019))
-abline(a = ((((fixef(RES_full)[1] * scl) + center) ) -
-           (2000 * release_yr_coef)) / 1000,
-       b = release_yr_coef/1000)
-abline(lm(yield_kg/1000 ~ release_yr, data = mod_data[res,]), lty = 2)
-with(plot_data, segments(x0 = release_yr, y0=yield_kg + yield_se,
-                         y1 = yield_kg - yield_se))
-text(x = plot_data$release_yr, y = plot_data$yield_kg + 0.15,
-     labels = gsub("M", "M-", row.names(plot_data)),
-     pos = 4, cex = 0.8)
-
-legend("bottomright", legend = c("w/ years since release adj.",
-                                 "w/o years since release adj.",
-                                 "90% Credible interval"),
-       lty = c(1,2, NA),
-       bty = "n",
-       fill = c(NA,NA, "grey"),
-       border = NA,
-       merge = TRUE,
-       cex = 0.8)
-polygon(x = c(yrs, rev(yrs)),
-        y = c(ints[1,],rev(ints[2,])),
-        border = NA,
-        col = rgb(0,0,0,0.1))
-
-# text(1990, 8.0, bquote(.(round(release_yr_coef))~kg~ha^-1~year^-1))
-
-@
-
-\caption{The yield increase over time of medium grain rice varieties
-  in California, USA. Ignoring the years since release (dashed line)
-  suggests an increase in yield potential over time. However, correcting
-  the comparison to account for differences in the length of time since
-  each cultivar was released (solid line) shows no increase in yield over time.}
-\label{fig:release}
-\end{figure}
-
-\begin{figure}[ht]
+\begin{figure}[H]
 <<fig=TRUE, width = 8, height = 5, out.width = '85%', echo=FALSE>>=
@@
 \caption{The performance of two major cultivars at the CA Rice
-  Experiment Station (RES) over the period 1995 -- 2015. The solid line
-  represents a simple linear regression of yield for each cultivar over
-  the period which it was planted at the RES (including only that
-  cultivar's data), while the dashed line is the average yield of M-202
-  at its year of release (1985). Both cultivars have a similar yield at
-  their respective year of release, and similar decreases over
-  time. The slope for the individual regressions are comparible to the
-slope estimated from all medium grain cultivars
-(\Sexpr{yrs_coef} kg ha\textsuperscript{-1} year\textsuperscript{-1}).}
+  Experiment Station (RES) over the period
+  \Sexpr{min(mod_data[,'year'])} -- \Sexpr{max(mod_data[,'year'])}. The
+  solid line represents a simple linear regression of yield for each
+  cultivar over the period which it was planted at the RES (fit to only
+  that cultivar's data), while the dashed line is the published 3-year
+  average yield of M-202 at its release (1982-1984). Both cultivars have
+  a similar yield at their respective year of release, and similar
+  decreases over time. The slope for the individual regressions are
+  comparable to the slope estimated from all medium grain cultivars
+  (\Sexpr{yrs_coef} kg ha\tsups{-1} year\tsups{-1}).}
 \label{fig:compare_yrs}
@@
 
-<<echo=FALSE>>=
-RES_CI <- round(posterior_interval(RES_full, pars = "yrs_in_trial") * scl, 1)
-@
-
-When yields are compared accounting for differences in the years since
-release at the time of comparision, there is a small decrease in yield
-over time once a cultivar is released
-(\Sexpr{yrs_coef}
-kg ha\textsuperscript{-1} year\textsuperscript{-1}, 90\%
-CI: \Sexpr{RES_CI[2]} -- \Sexpr{RES_CI[1]}). The oldest cultivar in
-the trials is estimated to have lost \Sexpr{round(yrs_coef*36,1)}
-kg ha\textsuperscript{-1} since release. Again
-comparing the yield of two prominant cultivars, M-202 and M-206, the
-two cultivars are experiencing a similar trend over time to eachother
-(Fig. \ref{fig:compare_yrs}) and the estimate from all released medium
-grains. The estimated yield advantage of the newer cultivar (M-206)
-(\Sexpr{m202_v_206} kg ha\textsuperscript{-1}) can be explained by the
-differences in the amount of time that has passed since each cultivar
-was released (Fig. \ref{fig:compare_yrs} \&
-\ref{fig:compare_since_release})
-
-\subsection{Yield improvements over time}
-
-<<echo=FALSE>>=
-# m202_v_206 = round((posterior_interval(RES_full, pars = "release_yr_c") * scl) * 19, 0)
-
-# m202_v_206 = ((posterior_predict(RES_full, newdata = data.frame(release_yr_c = c(-15, 4)),
-#                                  re.form = NA) * scl) + center)
-# m202_v_206 = m202_v_206[,2] - m202_v_206[,1]
-# m202_v_206 = round(quantile(m202_v_206, p = c(0.05, 0.5, 0.95)), 1)
-
-@
-
-Contrary to the estimates for decreases once a cultivar is released,
-there was no increase in yield potential over time once the
-differences between the years since release were accounted for
-(Fig. \ref{fig:release}) (\Sexpr{release_yr_coef}
-kg ha\textsuperscript{-1} year\textsuperscript{-1}). This suggests
-that the overall progress on yield over the period from 1980 -- 2016
-is approximately \Sexpr{round(release_yr_coef * 36, 0)} kg
-ha\textsuperscript{-1}.
-
-When differences between the number of years since each cultivar's
-release are taken into account, the estimated increase in grain yield
-over time is estimated to be \Sexpr{release_yr_coef}
-kg ha\textsuperscript{-1} year\textsuperscript{-1} (90\% CI:
-\Sexpr{rel_CI[1]} -- \Sexpr{rel_CI[2]}). Therefore, the evidence is
-consistent with no increase in yield potential over the study
-period. If the differences in years since release of each cultivar is
-ignored, there appears to be a small increase over time (approximately
-\Sexpr{round(coef(lm(yield_kg ~ release_yr, mod_data[res,])), 1)})
-(Fig. \ref{fig:release}), much less than previous estimates of
-\Sexpr{54 * 1.12} kg ha\textsuperscript{-1} year\textsuperscript{-1}
-(CITE).  Comparing the two dominate cultivars M-202, which was widely
-planted from the late 1980s until the early 2000s, and M-206,
-currently the most widely planted CRRFS cultivar (approximately 50\%
-of planted area), there is an estimated \Sexpr{m202_v_206} kg
-ha\textsuperscript{-1} yield advantage of M-206 over M-202 (Figs
-\ref{fig:release} \& \ref{fig:compare_vars}), though both cultivars
-yielded similarly at their respective year of release
-(Fig. \ref{fig:compare_yrs}).
-
 \section{Discussion}
 
-\subsection{Improving yield potential over time}
+Overall, we do not find conclusive evidence of increased yield
+potential increasing in medium grain rice cultivars released by
+CCRRF's breeding improvement program over the period
+\Sexpr{min(mod_data[,'year'])} to \Sexpr{max(mod_data[,'year'])}. After correcting yield comparisons
+for yield erosion over time in the older cultivar, the estimated yield
+improvement spans from \Sexpr{rel_CI[1]} -- \Sexpr{rel_CI[2]} kg
+ha\tsups{-1} year\tsups{-1}.  Even when the
+comparison is not corrected for differences in years since release
+between cultivars, the estimated improvements over time
+(\Sexpr{only_release_coef} kg ha\tsups{-1}
+year\tsups{-1}; Fig. \ref{fig:release} dashed line) are
+still much less than previous estimates of 37 -- 60 kg
+ha\tsups{-1} year\tsups{-1}
+\citep{Omar2014}. Given this evidence, we conclude that the breeding
+program has likely been maintaining but not substantially improving
+yield potential in the face of changing biotic and abiotic
+pressures. These biotic and abiotic forces are expected to erode yield
+over time \citep{Peng1998, Peng2000, Peng2010}. In fact, the trend
+seen here is extremely consistent with previous reports of erosion of
+genetic gains of cultivar IR8 at IRRI \citep{Peng1998, Peng2000}. The
+evidence here further supports that the erosion of yield over time is
+a phenomenon that affects all rice breeding programs. This further
+highlights both the need for maintenance breeding and the difficulty
+of increasing genetic yield potential as yield approaches the
+physiological limit. Without maintenance breeding continuously
+selecting cultivars well adapted to current conditions, CA rice
+production systems are predicted to have lost an estimated
+\Sexpr{round(yrs_coef * (-36/1000), 1)} Mg ha\tsups{-1}
+grain yield over the study period.
 
-Overall, we found evidence consistent with a no improvements in yield
-potential in medium grain rice cultivars released by CCRRF's breeding
-improvement program over the period \Sexpr{min(mod_data[,'year'])} --
-\Sexpr{max(mod_data[,'year'])}. Rather, it seems possible that the
-breeding program has been maintaining yield potential in the face of
-changing biotic and abiotic pressures, which are expected to decrease
-yield over time \citep{Peng2010}. This highlights both the need
-for maintanence breeding and the difficulty of increasing genetic
-yield potential as yield approaches the physiological limit. Without
-maintanence breeding continuously selecting cultivars well adapted to
-current conditions, CA rice production systems are predicted to have
-lost an estimated \Sexpr{round(yrs_coef * 31, 0)} kg
-ha\textsuperscript{-1} grain yield over the study period.
+The goal of a breeding program is to not only maintain the
+benchmark yield of the previous ``standard'' but to surpass it. Given
+the estimated yield decreases over time, to hit an ambitious target of
+0.5 Mg ha\tsups{-1} (approximately 5\%) increase in yield
+per release (with roughly 4 -- 5 years between releases), a new cultivar
+needs to yield an additional \Sexpr{round(yrs_coef*-4,-1)} --
+\Sexpr{round(yrs_coef*-5,-1)} kg ha\tsups{-1} in
+side-by-side comparison on top of the 0.5 Mg ha\tsups{-1}
+target improvement. Often realized side-by-side yield
+improvements 
+are closer to 2 -- 3\% per release, in which case no absolute
+improvements are made beyond maintenance. The evidence here is
+consistent with the latter situation occurring in CA, further suggesting
+that the genetic yield potential has not increased substantially in
+the last 30+ years (Fig. \ref{fig:release}).
 
-However, the goal of a breeding program is to not only recover the
-benchmark yield of the old standard cultivar but to surpass it. Given
-the estimated yield decreases over time, to hit an ambitious target of
-500 kg ha\textsuperscript{-1} (approximately 5\%) increase in yield
-per release (with roughly 5 years between releases), a new cultivar
-needs to yield \Sexpr{round(yrs_coef * -5, -1)} more
-in side-by-side comparison on top of the 500 kg ha\textsuperscript{-1}
-target improvement. Often side-by-side yield improvements (not
-considering the difference in years since release) are closer to 2 --
-3\% per release, in which case no absolute improvements might be made
-beyond maintenance. The evidence here is consistent with the latter
-situation occuring in CA, which suggests that the genetic yield
-potential has not increased substantially in the last 30+ years (Fig. \ref{fig:release}).
-
-There has been some yield improvements over the last 30 years in the
-CA state average yield (Fig. \ref{fig:state_yield}), but the evidence
-here suggests mechanisms other than increased genetic yield potential
-are responsible (Fig. \ref{fig:compare_yrs}). Improvement in agronomic
-management, including more efficient nutrient management
-\citep{Lundy2012, Lundy2015, linquist2009}, weed management
-\citep{Saito2008, Pittelkow2012, linquist2008}, etc. may all be
-contributing to further increases. However, the gains here are less
-than those experienced during the ``Green revolution'', and by their
-nature cannot surpass the genetic yield potential
-\citep{Fischer2015}. Judging from yields achieved at the RES
-(Figs. \ref{fig:RES_year} \& \ref{fig:compare_yrs}), a
-researcher-managed production system, state average yields
-(approximately 10 Mg ha\textsuperscript{-1} at the time of writing;
-Fig. \ref{fig:state_yield}) may be approaching the agronomic limit.
-
-To be clear, our investigation here is narrowly focused on the
-rough-rice/field grain yield. Other improvements due to plant
-improvement, including qualtity \citep{mckenzie1993} and milling yield
-\citep{mckenzie1994} have been made over the study period and are no
-less important. However, planning to meet future demand necessitates
-increases in grain yield \citep{Godfray2010, Foley2011a,
-Mueller2012}. The evidence here suggests that to accomplish true
-increases over time, the improvement target needs take decreases in
-the current cultivars' yield over time into account. While side-by-side
-comparisions will continue to be useful, comparisons across years to
-each cultivar's benchmark yield at the time of release should provide
-information on the absolute improvements being made in yield potential over time.
-
-\begin{figure}[ht]
-<<fig = TRUE, echo = FALSE>>=
+\begin{figure}[H]
+<<fig = TRUE, echo = FALSE, width=5, height=5>>=
 
 plot(1, type = "n",
-     xaxt = "n",
-     xlab = "Time",
+     # xaxt = "n",
+     xlab = "Years",
      ylab = expression("Grain yield ( Mg ha"^-1~")"),
@@
 
-arrows(12, 8.2, 27, 8.2, xpd = TRUE, lwd = 2)
+# arrows(12, 8.2, 27, 8.2, xpd = TRUE, lwd = 2)
 abline(a = 10, b = fixef(RES_full)[3], lty = 2, lwd = 2)
-text(25, 10.75, "(3) target above maintenance", cex = 0.8,
-     srt = 15)#((atan(fixef(RES_full)[3] + 0.06) * 180))/pi)
+text(25, 10.85, "(3) target above maintenance", cex = 0.8,
+     srt = 18)#((atan(fixef(RES_full)[3] + 0.06) * 180))/pi)
 abline(h = 10, lty = 3, lwd = 2)
@@
 abline(a = 10, b = fixef(RES_full)[3] + 0.06, lty = 4, lwd = 2)
-text(25, 9.25, "(2) no maintenance breeding", cex = 0.8,
-     srt = -20)
+text(25, 9.35, "(2) no maintenance breeding", cex = 0.8,
+     srt = -18)
 
@@
   improvements via plant improvement, (2) no maintenance breeding,
-  where grain yield slowly decreases over time, and (3) absolute
+  where grain yield slowly erodes over time, and (3) absolute
   improvement in grain yield, where gains in grain yield exceed the
@@
 \end{figure}
+
+That is not to say there have not been yield improvements over the
+last 30 years. California state average yields have increased slowly,
+even as genetic gains have been small
+(Fig. \ref{fig:state_yield}). The evidence here suggests mechanisms
+other than increased genetic yield potential are responsible
+(Fig. \ref{fig:compare_yrs}). Improvement in agronomic management,
+including more efficient nutrient management \citep{Lundy2012,
+Lundy2015, linquist2009}, weed management \citep{Saito2008,
+Pittelkow2012, linquist2008}, etc. may all be contributing to
+increases in the state average yield. However, these gains are less
+than those experienced during the ``Green revolution'', and by their
+nature agronomic gains cannot surpass the genetic yield potential
+\citep{Fischer2015}. Judging from yields achieved at the RES
+(Figs. \ref{fig:RES_year} \& \ref{fig:compare_yrs}), a
+researcher-managed production system implementing best management
+practices in more controlled situations compared to farmers, state
+average yields (approximately 10 Mg ha\tsups{-1} at the time
+of writing; Fig. \ref{fig:state_yield}) may be approaching the current
+genetic limit. Assuming that the level of management possible in small
+experimental plots cannot be attained in farmers' fields and hence
+experimental yields represent the attainable yield ceiling
+\citep{Fischer2015}, agronomic gains in farmer's fields are nearing
+their potential, consistent with physiological model-based estimates
+of the attainable yield gap in CA rice production systems
+\citep{espe16a}.
+
+To be clear, our investigation here is narrowly focused on the
+rough-rice/field grain yield. Other advancements due to plant
+improvement, including increased quality \citep{mckenzie1993} and milling yield
+\citep{mckenzie1994} have been made over the study period and are no
+less important. However, planning to meet future demand necessitates
+increases in grain yield \citep{Godfray2010, Foley2011a, Mueller2012}.
+Meeting the challenge to produce more grain without expansion of land
+under cultivation will rely on new genetics with increased yield
+potential. The evidence here suggests that to accomplish true
+increases over time, the improvement target needs take into account
+decreases in the current ``standard'' cultivars' yield over
+time (Fig. \ref{fig:theory}). While side-by-side comparisons will continue to be useful,
+comparisons across years to each cultivar's benchmark yield at the
+time of release should provide information on the absolute
+improvements being made in yield potential over time. New technologies
+may need to be adapted to further increase yield potential, including
+hybrid rice.
 
@@
 here that historical benchmarks be used in to evaluate promising rice
-cultivars rather than side-by-side comparisions to maintain
-perspective of the absolute gains being made.
+cultivars in addition to side-by-side comparisons to maintain
+perspective of the absolute yield gains being made.
 
@@
 RES_coef = cbind(med_eff, paste(CIs[,1],CIs[,2], sep = " -- "))
-rownames(RES_coef) = c("Intercept (kg ha\\textsuperscript{-1})",
-                       "Release year (kg ha\\textsuperscript{-1} year\\textsuperscript{-1})",
-                       "Years since release (kg ha\\textsuperscript{-1} year\\textsuperscript{-1})")
+rownames(RES_coef) = c("Intercept (kg ha\\tsups{-1})",
+                       "Release year (kg ha\\tsups{-1} year\\tsups{-1})",
+                       "Years since release (kg ha\\tsups{-1} year\\tsups{-1})")
 colnames(RES_coef) = c("Median", "90\\% Credible interval")
@@
 \newpage
+
 \section{References}\label{references}
@@
 
-\begin{figure}[ht]
+\begin{figure}[H]
 
@@
 
-\begin{figure}[ht]
+\begin{figure}[H]
 <<fig=TRUE, echo=FALSE>>=
 tars <- c("M202","M206")
-comp_yrs = 2002:2005
+comp_yrs = 2004:2005
 
 boxplot(yield_kg/1000 ~ id, data = mod_data[(mod_data$id %in% tars) &
-                                            (mod_data$year %in% comp_yrs) &
-                                            mod_data$site == "RES"  ,],
+                                            (mod_data$year %in% comp_yrs),],
         ylab = expression("Grain yield ( Mg ha"^-1~")"))
@@
 @
-\caption{An example of a typical comparison between a older (M-202)
+\caption{An example of a typical side-by-side comparison between a older (M-202)
   and newer (M-206) cultivar. Shown is the performance of these cultivars at the CA Rice
   Experiment Station (RES) over the period \Sexpr{min(comp_yrs)} --
-  \Sexpr{max(comp_yrs)}, approximately the comparision made to
-  determine if a new cultivar performs sufficiently better. The newer
+  \Sexpr{max(comp_yrs)}. The newer
   cultivar, M-206, appears to have a yield advantage over the older
@@
 \label{fig:compare_vars}
+\end{figure}
+\newpage
+
+\begin{figure}[H]
+<<echo = FALSE, fig = TRUE, width = 7, height = 6>>=
+library(ggplot2)
+library(ggthemes)
+ggplot(mod_data, aes(x = year, y = yield_kg/1000)) +
+    geom_point(size = 0.5, alpha = 0.5) +
+    geom_smooth(method = "lm", color = "black") +
+    theme_few() +
+    xlab("Year") +
+    ylab(expression("Grain yield ( Mg ha"^-1~")")) + 
+    facet_wrap(~id, ncol = 4)
+
+@
+\caption{The yield of all medium grain released cultivars at the
+  California Cooperative Rice Research Foundation Rice Experiment
+  Station over the period \Sexpr{min(mod_data[,'year'])} --
+  \Sexpr{max(mod_data[,'year'])}.}
+\label{fig:all_cult}
+\end{figure}
+
+\newpage
+\begin{figure}[ht]
+<<echo=FALSE, fig = TRUE>>=
+
+dotchart(id_est[,1],
+         labels = gsub("M", "M-", row.names(id_est)),
+         pch = 16,
+         xlim = c(min(id_est[,2]),max(id_est[,3])),
+         xlab = expression("Estimated difference from average ( Mg ha"^-1~")"))
+segments(x0 = id_est[,2], x1 = id_est[,3], y0 = 1:nrow(id_est))
+abline(v = 0, lty = 2)
+@
+\caption{The marginal estimates for the difference between cultivars,
+  averaged over the uncertainty in the other parameters.}
+\label{fig:ids_comp}
 \end{figure}
@@
 @
+
 % \begin{figure}[h]

### 2017-06-21 src/h1_model.R

@@ -1,10 +1,15 @@
 ## Run model and save output as text files
 ##.args = list(commandArgs(trailingOnly=TRUE))
 
+# Set to TRUE to compare between models
+compare = TRUE
+
 library(rstan)
 library(rstanarm)
+
 # source("prep_model_data.R")
 load("../data/model_data.rda")
+
 ## mm <- stan_model(file = "yield_model.stan", verbose = TRUE)
 
 ## n.level <- function(x){
@@ -44,16 +49,24 @@ load("../data/model_data.rda")
 #                               (1|id) + (1|yr_site_fact) + release_yr_c,
 #                           data = mod_data, iter = 1000, cores = 2L)
 
-i <- mod_data$site == "RES"
-
-# RES_yrs_fit <- stan_glmer(yield_cs ~ (1|year) + (1|id) + yrs_in_trial,
-#                           data = mod_data[i,], iter = 1000, cores = 2L)
-
-# RES_rel_fit <- stan_glmer(yield_cs ~ (1|year) + (1|id) + release_yr_c,
-#                           data = mod_data[i,], iter = 1000, cores = 2L)
 
 RES_full = stan_glmer(yield_cs ~ (1|year) + (1|id) + release_yr_c + yrs_in_trial,
-                          data = mod_data[i,], iter = 1000, cores = 2L)
+                      data = mod_data, iter = 2000, cores = 2L,
+                      seed = 789)
+
+if(compare){
+    RES_yrs_fit <- stan_glmer(yield_cs ~ (1|year) + (1|id) + yrs_in_trial,
+                              data = mod_data, iter = 2000, cores = 2L,
+                              seed = 789)
+
+    RES_rel_fit <- stan_glmer(yield_cs ~ (1|year) + (1|id) + release_yr_c,
+                              data = mod_data, iter = 2000, cores = 2L,
+                              seed = 789)
+
+    library(loo)
+    compare(loo(RES_full), loo(RES_yrs_fit), loo(RES_rel_fit))
+    save(RES_yrs_fit, RES_rel_fit, file = "../output/bivar_fit.Rda")
+}
 
 
 save(RES_full, file = "../output/h1_fit.Rda")

### 2017-06-21 src/prep_model_data.R

@@ -1,5 +1,8 @@
 ## Prepares data for write-up
-load("../data/all_vt_weather.Rda")
+load("../data/all_w_2016.Rda")
+
+# Replace "-" in variety names
+all_vt_result$id = gsub("M-", "M", all_vt_result$id)
 
 ## Replace 08Y3269 with M209
 all_vt_result$id[all_vt_result$id == "08Y3269"] <- "M209"
@@ -10,21 +13,13 @@ source("../src/munge_pre95.R")
 ## Subset to medium grains
 vars <- grepl("^M[0-9]{3}$", all_data$id)
 
-## Do not include San Joaquin sites - known to hammer with cold and
-## Not managed similar to most of CA rice production
-no_sj <- all_data$county != "SAN JOAQUIN"
-mod_data <- all_data[(vars & no_sj),]
+mod_data <- all_data[vars,]
 
-## Exclude M-401/402 and Tucker site, per conversation with Bruce
-## 10 Feb 2017:
-## M-401 and M-402 are often mis-managed in trials, however they are also old varieties
-## Question if excluding them is not creating more issues
-## not4s <- !mod_data$id %in% c("M401","M402")
+# To simplify model interpretation, only analyze RES
+# since this is the best managed site
+# with the most data
 
-## Tucker was a problematic site with high variability and low yields
-notTuck <- mod_data$site != "TUCKER"
-
-mod_data <- mod_data[(notTuck),]
+mod_data = mod_data[mod_data$site == "RES",]
 
 ## Use officially published years instead of years in trial
 load("../data/yrTbl.Rda")
@@ -47,4 +42,5 @@ mod_data$yr_fact <- factor(mod_data$year)
 mod_data$yr_site_fact <- factor(paste(mod_data$year, mod_data$site))
 
 save(mod_data, scl, center, file = "../data/model_data.rda")
+
 ################################################################################

### 2017-06-23 output/breed_paper.Rnw
Adjusts figure 2 to be panel plot

@@
 
-<<fig=TRUE, echo=FALSE>>=
+<<fig=TRUE, echo=FALSE, width = 9, height = 6>>=
 yrs <- 1975:2025
+
+pp <- posterior_linpred(RES_rel_fit, re.form = NA,
+                        newdata = data.frame(release_yr_c = yrs - 2000,
+                                             yrs_in_trial = 0),
+                        prob = 0.5)
+pp <-  (pp * scl + center) / 1000
+ints1 = apply(pp, 2, quantile, p = c(0.05, 0.95))
+
 pp <- posterior_linpred(RES_full, re.form = NA,
@@
 pp <-  (pp * scl + center) / 1000
-ints = apply(pp, 2, quantile, p = c(0.05, 0.95))
+ints2 = apply(pp, 2, quantile, p = c(0.05, 0.95))
 
@@
 
+# get corrected yield
+wo_erosion = expand.grid(id = row.names(plot_data),
+            year = comp_yrs,
+            release_yr_c = 0,
+            yrs_in_trial = 0)
+wo_erosion$corr_yield = colMeans(posterior_predict(RES_full,
+                               newdata = wo_erosion)) + center/1000
+plot_data2 = prep_plot_data(wo_erosion$corr_yield,
+                            wo_erosion$id)
+plot_data2$release_yr = plot_data$release_yr
+
+par(mfrow = c(1,2),
+    mar = c(4,0,2,0.75),
+    oma = c(4,4,0.5,0.5))
+
 plot(yield_kg ~ release_yr, data = plot_data, pch = 16, #col = rgb(0,0,0,0.25),
      xlab = "Release year",
-     ylab = expression("Grain yield ( Mg ha"^-1~")"),
+     ylab = "", 
      ylim = c(8, 12.5),
-     xlim = c(1980,2019))
-abline(a = ((((fixef(RES_full)[1] * scl) + center) ) -
-           (2000 * release_yr_coef)) / 1000,
-       b = release_yr_coef/1000)
-abline(lm(yield_kg/1000 ~ release_yr, data = mod_data[idx,]), lty = 2)
+     xaxt = "n",
+     xlim = c(1980,2020))
+axis(1, at = seq(1980, 2020, by = 10))
+mtext(text = expression("Grain yield ( Mg ha"^-1~")"), side = 2, line = 2.5)
+abline(a = ((fixef(RES_rel_fit)[1] * scl) + center) / 1000 -
+      (2000 * fixef(RES_rel_fit)[2]),
+       b = fixef(RES_rel_fit)[2],
+       lty = 2)
+polygon(x = c(yrs, rev(yrs)),
+        y = c(ints1[1,],rev(ints1[2,])),
+        border = NA,
+        col = rgb(0,0,0,0.1))
+
 with(plot_data, segments(x0 = release_yr, y0=yield_kg + yield_se,
@@
      labels = gsub("M", "M-", row.names(plot_data)),
-     pos = 4, cex = 0.8)
+     pos = 4, cex = 0.75)
+text(x = 2000, y = 8, bquote(.(round(fixef(RES_rel_fit)[2] * scl, 1))~kg~ha^-1~yr^-1))
+mtext("a.", adj = 0)
 
-# legend("bottomright", legend = c("w/ years since release adj.",
-#                                  "w/o years since release adj.",
-#                                  "90% Credible interval"),
-#        lty = c(1,2, NA),
-#        bty = "n",
-#        fill = c(NA,NA, "grey"),
-#        border = NA,
-#        merge = TRUE,
-#        cex = 0.8)
+plot(mean ~ release_yr, data = plot_data2, pch = 16, #col = rgb(0,0,0,0.25),
+     xlab = "Release year",
+     ylab = "", #expression("Grain yield -corrected( Mg ha"^-1~")"),
+     yaxt = "n",
+     ylim = c(8, 12.5),
+     xaxt = "n",
+     xlim = c(1980,2020))
+axis(1, at = seq(1980, 2020, by = 10))
+abline(a = ((((fixef(RES_full)[1] * scl) + center) ) -
+           (2000 * release_yr_coef)) / 1000,
+       b = release_yr_coef/1000,
+       lty = 2)
+with(plot_data2, segments(x0 = release_yr, y0 = mean + se,
+                         y1 = mean - se))
+text(x = plot_data2$release_yr-0.4, y = plot_data2$mean + 0.07,
+     labels = gsub("M", "M-", row.names(plot_data2)),
+     pos = 4, cex = 0.75)
+text(x = 2000, y = 8, bquote(.(release_yr_coef)~kg~ha^-1~yr^-1))
+mtext("b.", adj = 0)
+polygon(x = c(yrs, rev(yrs)),
+        y = c(ints2[1,],rev(ints2[2,])),
+        border = NA,
+        col = rgb(0,0,0,0.1))
 
-legend("bottomright", legend = c("w/ years since release adj.",
-                                 "w/o years since release adj."),
-       lty = c(1,2),
-       bty = "n",
-       border = NA,
-       merge = TRUE,
-       cex = 0.8)
-# polygon(x = c(yrs, rev(yrs)),
-        # y = c(ints[1,],rev(ints[2,])),
-        # border = NA,
-        # col = rgb(0,0,0,0.1))
-
-# text(1990, 8.0, bquote(.(round(release_yr_coef))~kg~ha^-1~year^-1))
 @
 \caption{The average yield ($\bullet$) and standard error (|) of
-medium grain rice cultivars by year of release. For display purposes,
-only data for experimental plots planted at the California Cooperative
-Rice Research Foundation Rice Experiment Station (California, USA)
-over the years \Sexpr{min(comp_yrs)} -- \Sexpr{max(comp_yrs)} are
-included. Ignoring difference in years since release (dashed line)
-suggests an increase in yield potential over time. However, correcting
-the comparison to account for yield erosion over time (solid line)
-suggests the apparent increase might be due to differences in years
-since release.}
+medium grain rice cultivars by year of release (a.) without correcting
+for yield erosion over time and (b.) accounting for yield erosion over
+time. The trend without correcting for yield erosion (a.) was estimated by
+omitting the year of each cultivar's release from the
+model. Yield erosion was corrected for by including the years since
+release (b.) in the model. predictions assuming no yield erosion
+were generated from the model which included both the year of release
+and the years since release. For display purposes, only data for
+experimental plots planted at the California Cooperative Rice Research
+Foundation Rice Experiment Station (California, USA) over the years
+\Sexpr{min(comp_yrs)} -- \Sexpr{max(comp_yrs)} are included.}
 \label{fig:release}
@@
   estimated to be \Sexpr{yrs_coef} kg ha\tsups{-1}
-  year\tsups{-1}. Points ($bullet$) and bars (|) represent the mean and standard
+  year\tsups{-1}. Points ($\bullet$) and bars (|) represent the mean and standard
   error of the raw data, respectively. The

### 2017-07-03 output/breed_paper.Rnw
Update sequence of paper

@@
 \usepackage{Sweave}
-\SweaveOpts{prefix.string=graphics/report}
+\SweaveOpts{prefix.string=graphics/report, png=TRUE}
 
@@
 yield in side-by-side comparisons between old and newer varieties when
-the difference between the years since release was not
+the yield erosion was not
 considered. However, we demonstrate this apparent increase in yield
@@
 id_est = id_est[idx,]
+
 @
@@
      ylab = expression("State avg. grain yield ( Mg ha"^-1~")"))
-mv_avg = filter(c(rep(mean(CA_data$Value[1:10] * 1.12), 5),
-                  CA_data$Value * 1.12,
-                  rep(mean(CA_data$Value[(nrow(CA_data) - 10):nrow(CA_data)] * 1.12), 5)),
-                rep(1/11,11))
-lines(x = c(1907:1911, CA_data$Year, 2017:2021), y = mv_avg)
+# mv_avg = filter(c(rep(mean(CA_data$Value[1:10] * 1.12), 5),
+#                   CA_data$Value * 1.12,
+#                   rep(mean(CA_data$Value[(nrow(CA_data) - 10):nrow(CA_data)] * 1.12), 5)),
+#                 rep(1/11,11))
+# lines(x = c(1907:1911, CA_data$Year, 2017:2021), y = mv_avg)
 
@@
   approximately 1950 -- 1990, but annual gains since 1990 have been
-  slower. The line represents the 10 year moving average.}
+  slower.}%The line represents the 10 year moving average.}
+% Remove 10 year moving average - Bruce does not like. 
 \label{fig:state_yield}
@@
 entries are tested against named and released cultivars (checks) at
-the RES in multiple trials spanning multiple planting dates per
+the RES in several trials spanning multiple planting dates per
 year. For the purposes of this study, we concentrated on released
 medium grain cultivars, as medium grain cultivars constitute roughly
-90\% of the planted area in CA. For these cultivars planted in
+90\% of the planted area in CA. These cultivars are typcically planted
+in the Statewide Variety Trials for 2 -- 4 years prior to their
+official release. For these cultivars planted in
 breeding trials at the RES from \Sexpr{min(mod_data[,'year'])} --
 \Sexpr{max(mod_data[,'year'])} plot level observations were
-collected. This lead to a final data set of
+collected leading to a final data set of
 \Sexpr{format(nrow(mod_data), big.mark = ',')} observations with
@@
 trials but prior to widespread commercial seed availability) and the
-time elapse (in years) since release was calculated.
+time elapsed (in years) since release was calculated.
 
-Plots at the RES are managed similar to prescribed best management
+Plots at the RES are managed similar to University of California
+Cooperative Extension prescribed best management
 practice, including land preparation, fertility management, and pest
-and disease control. For planting, rice seed is pre-germinated prior
+and disease control \citep{ucce}. For planting, rice seed is pre-germinated prior
 to direct-seeding into pre-flooded fields. Plots were either 14.0 or
@@
 
-% \begin{equation}
-%  yield_i = \alpha + \beta_{years\ since\ release\ i}* x_{years\ since\
-%    release\ i} + \beta_{release\ year\ i}*x_{release\ year\ i} + \beta_{year\ i} + \beta_{cultivar\ i}
-% \end{equation}
+\begin{equation}
+\begin{split}
+ yield_i = Intercept + (Rate_{yield\ erosion\ i}* {years\ since\
+   release\ i}) \\+ (Rate_{yield\ improvement\ i}* {release\ year\ i}) 
+ + \beta_{year\ i} + \beta_{cultivar\ i} \\
+\end{split}
+\end{equation}
 
-% Where $yield_i$ is the grain yield for plot $i$, $\alpha$ is the
-% average yield for the RES across varieties and years (intercept),
-% $\beta_{years\ since\ release\ i}$ is effect of $x_{years\ since\
-% release\ i}$, the number of years $i$'s cultivar has been in the
-% trials (negative for years prior to release, zero at the year of
-% release), $\beta_{release\ year\ i}$ is the effect of $x_{release\
-% year\ i}$, the year of release for $i$'s cultivar, $\beta_{year}$ is
-% the random effect of year, and $\beta_{cultivar}$ is the random effect
-% of cultivar.
+Where $yield_i$ is the grain yield for plot $i$, the Intercept is the
+average yield for the RES across varieties and years, $Rate_{yield\
+erosion\ i}$ is rate of yield erosion per ${years\ since\ release\ i}$
+(negative for years prior to release, zero at the year of release),
+$Rate_{yield\ improvement\ i}$ is the rate of yield improvement per
+year (${release\ year\ i}$), $\beta_{year}$ is the random effect of
+year, and $\beta_{cultivar}$ is the random effect of cultivar. For
+comparison to estimates where yield erosion is not taken into account,
+the model was as above except omitting the $Rate_{yield\ erosion\ i}$ term.
+
+% Add more detail to the methods section - tell how you get these
+% numbers. "Evidence for erosion" - what does this mean for yield
+% increases? Erosion 1st, 2nd implications of erosion. 
+% Drop off -5 and -6
 
@@
 full model was compared to models containing only the years since
-release or the year of release using the `loo' package \citep{loo},
-an efficient means of conducting leave-one-out cross validation
-\citep{loo_paper}, with the full model estimated to have slightly
-better predictive accuracy compared to the reduced models.
+release or the year of release using the `loo' package \citep{loo}, an
+efficient means of conducting leave-one-out cross validation
+\citep{loo_paper}. The 90\% credible interval (90\% CI), defined as
+the interval containing 90\% of the distribution of estimated credible
+parameter values given the data, was calculated as a measure of
+uncertainty in the parameter estimates using the quantile method. The
+90\% CI is more stable to perpetuation's and hence is preferable to the 95\%
+interval \citep{rstanarm}. Complete data and code used for this
+analysis is available at \url{github.com/mespe/rice_advances.git}.
 
 \section{Results}\label{results}
-
-\subsection{Genetic yield improvements}
-
-There was no conclusive evidence that the yields of rice cultivars
-selected by CCRRF have, on average, increased over the last 30
-years. The estimated change in grain yield per year was estimated to
-be a negligible \Sexpr{release_yr_coef} kg ha\tsups{-1}
-year\tsups{-1}, with both small positive and negative trends within
-the credible interval (90\% CI: \Sexpr{rel_CI[1]} --
-\Sexpr{rel_CI[2]}). Hence, the evidence here is consistent with little
-to no increase in yield potential over the study period. When yield
-erosion is not corrected for in the comparison (i.e., omitted from the
-model), there is an apparent small increase
-per year (approximately \Sexpr{only_release_coef} kg ha\tsups{-1}
-year\tsups{-1}; 90\% CI \Sexpr{only_release_CI[1]} --
-\Sexpr{only_release_CI[2]}) (Fig. \ref{fig:release}), though this
-estimate ignores the confounding effect of yield erosion over
-time.
-
-There do appear to be some differences between cultivars, with some
-cultivars showing higher yield than the overall average. The marginal
-estimate for the performance of a cultivar (averaged over the
-uncertainty in the other parameters in the model) shows three
-cultivars with higher than average yields:
-\Sexpr{paste(gsub('M','M-',rev(row.names(id_est))[c(1,2,4)]),collapse=', ')}
-(Fig. \ref{fig:ids_comp}). A fourth, M-209, might also have higher
-than average yield, though the CI overlaps with zero.
-
-\begin{figure}[H]
-
-<<fig=TRUE, echo=FALSE, width = 9, height = 6>>=
-yrs <- 1975:2025
-
-pp <- posterior_linpred(RES_rel_fit, re.form = NA,
-                        newdata = data.frame(release_yr_c = yrs - 2000,
-                                             yrs_in_trial = 0),
-                        prob = 0.5)
-pp <-  (pp * scl + center) / 1000
-ints1 = apply(pp, 2, quantile, p = c(0.05, 0.95))
-
-pp <- posterior_linpred(RES_full, re.form = NA,
-                        newdata = data.frame(release_yr_c = yrs - 2000,
-                                             yrs_in_trial = 0),
-                        prob = 0.5)
-pp <-  (pp * scl + center) / 1000
-ints2 = apply(pp, 2, quantile, p = c(0.05, 0.95))
-
-res = mod_data$site == "RES"
-comp_yrs = 2010:2015
-idx = mod_data$year %in% comp_yrs
-
-plot_data = prep_plot_data(mod_data$yield_kg[idx] / 1000 ,
-                           mod_data$id[idx])
-
-plot_data$tmp = tapply(mod_data$release_yr[idx],
-                       mod_data$id[idx], unique)
-
-colnames(plot_data) = c("yield_kg", "yield_se", "release_yr")
-
-# get corrected yield
-wo_erosion = expand.grid(id = row.names(plot_data),
-            year = comp_yrs,
-            release_yr_c = 0,
-            yrs_in_trial = 0)
-wo_erosion$corr_yield = colMeans(posterior_predict(RES_full,
-                               newdata = wo_erosion)) + center/1000
-plot_data2 = prep_plot_data(wo_erosion$corr_yield,
-                            wo_erosion$id)
-plot_data2$release_yr = plot_data$release_yr
-
-par(mfrow = c(1,2),
-    mar = c(4,0,2,0.75),
-    oma = c(4,4,0.5,0.5))
-
-plot(yield_kg ~ release_yr, data = plot_data, pch = 16, #col = rgb(0,0,0,0.25),
-     xlab = "Release year",
-     ylab = "", 
-     ylim = c(8, 12.5),
-     xaxt = "n",
-     xlim = c(1980,2020))
-axis(1, at = seq(1980, 2020, by = 10))
-mtext(text = expression("Grain yield ( Mg ha"^-1~")"), side = 2, line = 2.5)
-abline(a = ((fixef(RES_rel_fit)[1] * scl) + center) / 1000 -
-      (2000 * fixef(RES_rel_fit)[2]),
-       b = fixef(RES_rel_fit)[2],
-       lty = 2)
-polygon(x = c(yrs, rev(yrs)),
-        y = c(ints1[1,],rev(ints1[2,])),
-        border = NA,
-        col = rgb(0,0,0,0.1))
-
-with(plot_data, segments(x0 = release_yr, y0=yield_kg + yield_se,
-                         y1 = yield_kg - yield_se))
-text(x = plot_data$release_yr-0.4, y = plot_data$yield_kg + 0.07,
-     labels = gsub("M", "M-", row.names(plot_data)),
-     pos = 4, cex = 0.75)
-text(x = 2000, y = 8, bquote(.(round(fixef(RES_rel_fit)[2] * scl, 1))~kg~ha^-1~yr^-1))
-mtext("a.", adj = 0)
-
-plot(mean ~ release_yr, data = plot_data2, pch = 16, #col = rgb(0,0,0,0.25),
-     xlab = "Release year",
-     ylab = "", #expression("Grain yield -corrected( Mg ha"^-1~")"),
-     yaxt = "n",
-     ylim = c(8, 12.5),
-     xaxt = "n",
-     xlim = c(1980,2020))
-axis(1, at = seq(1980, 2020, by = 10))
-abline(a = ((((fixef(RES_full)[1] * scl) + center) ) -
-           (2000 * release_yr_coef)) / 1000,
-       b = release_yr_coef/1000,
-       lty = 2)
-with(plot_data2, segments(x0 = release_yr, y0 = mean + se,
-                         y1 = mean - se))
-text(x = plot_data2$release_yr-0.4, y = plot_data2$mean + 0.07,
-     labels = gsub("M", "M-", row.names(plot_data2)),
-     pos = 4, cex = 0.75)
-text(x = 2000, y = 8, bquote(.(release_yr_coef)~kg~ha^-1~yr^-1))
-mtext("b.", adj = 0)
-polygon(x = c(yrs, rev(yrs)),
-        y = c(ints2[1,],rev(ints2[2,])),
-        border = NA,
-        col = rgb(0,0,0,0.1))
-
-@
-\caption{The average yield ($\bullet$) and standard error (|) of
-medium grain rice cultivars by year of release (a.) without correcting
-for yield erosion over time and (b.) accounting for yield erosion over
-time. The trend without correcting for yield erosion (a.) was estimated by
-omitting the year of each cultivar's release from the
-model. Yield erosion was corrected for by including the years since
-release (b.) in the model. predictions assuming no yield erosion
-were generated from the model which included both the year of release
-and the years since release. For display purposes, only data for
-experimental plots planted at the California Cooperative Rice Research
-Foundation Rice Experiment Station (California, USA) over the years
-\Sexpr{min(comp_yrs)} -- \Sexpr{max(comp_yrs)} are included.}
-\label{fig:release}
-\end{figure}
-
-\subsection{Yield decreases over time}
+% Invert this section, then talk about yield improvement accounting
+% for yield erosion.
 
@@
   (RES), California, USA. The estimated erosion of yield over time
-  (averaged over differences in cultivar and year) is
-  estimated to be \Sexpr{yrs_coef} kg ha\tsups{-1}
-  year\tsups{-1}. Points ($\bullet$) and bars (|) represent the mean and standard
-  error of the raw data, respectively. The
-  shaded area is the 90\% credible interval.}
+  (averaged over differences in cultivar and year) is estimated to be
+  \Sexpr{yrs_coef} kg ha\tsups{-1} year\tsups{-1}. Points ($\bullet$)
+  and bars (|) represent the mean and standard error of the raw data,
+  respectively. The shaded area is the 90\% credible interval.}
+% For each year, point size by n (open circles)
 \label{fig:RES_year}
@@
 
-There is estimated to be a small loss of yield over time once a
+\subsection{Yield erosion}
+
+There is estimated to be an decrease of yield over time once a
 cultivar is selected (i.e., yield erosion) (\Sexpr{yrs_coef} kg
-ha\tsups{-1} year\tsups{-1}, 90\% CI:
-\Sexpr{RES_CI[2]} -- \Sexpr{RES_CI[1]}). The oldest cultivar in the
-trials is estimated to have lost \Sexpr{round(yrs_coef*-36/1000,1)} Mg
-ha\tsups{-1} since release. Comparing two dominate
-medium grain rice cultivars (M-202, which was widely planted from the late 1980s until
-the early 2000s, and M-206, the most widely planted CRRFS
-cultivar at approximately 50\% of planted area at the time of
-writing), the two cultivars experienced similar trends over time to
-each other (Fig. \ref{fig:compare_yrs}) and to the estimate from all
-released medium grains (Table \ref{tbl:coef}). There is an estimated
-\Sexpr{m202_v_206} kg ha\tsups{-1} yield advantage of M-206
-over M-202 at the time of writing due solely to the erosion of M-202's
-yield over time (Figs \ref{fig:release} \&
-\ref{fig:compare_vars}). Further supporting this estimate, both cultivars yielded
-similarly at their respective year of release
-(Fig. \ref{fig:compare_yrs}), despite the fact that M-202 was released
-19 years prior to M-206 (Fig. \ref{fig:release}, Table \ref{tbl:ids}).
+ha\tsups{-1} year\tsups{-1}, 90\% CI: \Sexpr{RES_CI[2]} --
+\Sexpr{RES_CI[1]}). The oldest cultivar in the trials is estimated to
+have lost an average of \Sexpr{round(yrs_coef*-36/1000,1)} Mg
+ha\tsups{-1} since release. Comparing two dominate medium grain rice
+cultivars M-202 (widely planted from the late 1980s until the early
+2000s), and M-206 (the most widely planted CCRRFS cultivar at
+approximately 50\% of planted area at the time of writing), the two
+cultivars experienced similar trends over time to each other
+(Fig. \ref{fig:compare_yrs}) and to the estimate from all released
+medium grains (Table \ref{tbl:coef}). At the time of writing, an
+estimated \Sexpr{m202_v_206} kg ha\tsups{-1} yield advantage of M-206
+over M-202 can be attributed solely to the erosion of M-202's yield
+over time (Figs \ref{fig:release} \& \ref{fig:compare_vars}). Further
+supporting this estimate, both cultivars yielded similarly at their
+respective year of release (Fig. \ref{fig:compare_yrs}), despite the
+fact that M-202 was released 19 years prior to M-206
+(Fig. \ref{fig:release}, Table \ref{tbl:ids}).
 
@@
 
+\subsection{Genetic yield improvements}
+
+The full model which corrected for both the year of release and yield
+erosion was estimated to have better out-of-sample predictive accuracy
+compared to reduced models containing only one predictor. As estimated
+by the best model, there is no conclusive evidence that the yields of
+medium grain rice cultivars selected and released by CCRRF's breeding
+program have, on average, increased over the last 30 years. The
+estimated change in grain yield per year in the CCRRF's breeding
+trials at the RES was estimated to be a negligible
+\Sexpr{release_yr_coef} kg ha\tsups{-1} year\tsups{-1}, with both
+small positive and negative trends within the credible interval (90\%
+CI: \Sexpr{rel_CI[1]} -- \Sexpr{rel_CI[2]})
+(Fig. \ref{fig:release}b.). Hence, the evidence here is consistent
+with little to no increase in yield potential over the study
+period. When yield erosion is not corrected for in the comparison
+(i.e., omitted from the model), there is an apparent small increase
+per year (approximately \Sexpr{only_release_coef} kg ha\tsups{-1}
+year\tsups{-1}; 90\% CI \Sexpr{only_release_CI[1]} --
+\Sexpr{only_release_CI[2]}) (Fig. \ref{fig:release}a.), though this
+estimate ignores the confounding effect of yield erosion over time and
+performed worse in out-of-sample predictive accuracy.
+
+After correcting for the other variables, some cultivars are estimated
+to have greater yield compared to the overall average. The marginal
+estimate for the performance of a cultivar averaged over the
+uncertainty in the other parameters in the model (i.e., the random
+effect) shows three cultivars with higher than average yields:
+\Sexpr{paste(gsub('M','M-',rev(row.names(id_est))[c(1,2,4)]),collapse=',')}
+(Figs. \ref{fig:release} \& \ref{fig:ids_comp}). A fourth, M-209,
+might also have higher than average yield, though the CI for M-209
+overlaps with zero (Fig. \ref{fig:ids_comp}).
+
+\begin{figure}[H]
+
+<<fig=TRUE, echo=FALSE, width = 9, height = 6>>=
+yrs <- 1975:2025
+
+pp <- posterior_linpred(RES_rel_fit, re.form = NA,
+                        newdata = data.frame(release_yr_c = yrs - 2000,
+                                             yrs_in_trial = 0),
+                        prob = 0.5)
+pp <-  (pp * scl + center) / 1000
+ints1 = apply(pp, 2, quantile, p = c(0.05, 0.95))
+
+pp <- posterior_linpred(RES_full, re.form = NA,
+                        newdata = data.frame(release_yr_c = yrs - 2000,
+                                             yrs_in_trial = 0),
+                        prob = 0.5)
+pp <-  (pp * scl + center) / 1000
+ints2 = apply(pp, 2, quantile, p = c(0.05, 0.95))
+
+res = mod_data$site == "RES"
+comp_yrs = 2011:2016
+idx = mod_data$year %in% comp_yrs
+
+plot_data = prep_plot_data(mod_data$yield_kg[idx] / 1000 ,
+                           mod_data$id[idx])
+
+plot_data$tmp = tapply(mod_data$release_yr[idx],
+                       mod_data$id[idx], unique)
+
+colnames(plot_data) = c("yield_kg", "yield_se", "release_yr")
+
+# get corrected yield
+wo_erosion = expand.grid(id = row.names(plot_data),
+            year = comp_yrs,
+            release_yr_c = 0,
+            yrs_in_trial = 0)
+wo_erosion$corr_yield = colMeans(posterior_predict(RES_full,
+                               newdata = wo_erosion)) + center/1000
+plot_data2 = prep_plot_data(wo_erosion$corr_yield,
+                            wo_erosion$id)
+plot_data2$release_yr = plot_data$release_yr
+
+par(mfrow = c(1,2),
+    mar = c(4,0,2,0.75),
+    oma = c(4,4,0.5,0.5))
+
+plot(yield_kg ~ release_yr, data = plot_data, pch = 16, #col = rgb(0,0,0,0.25),
+     xlab = "Release year",
+     ylab = "", 
+     ylim = c(8, 12.5),
+     xaxt = "n",
+     xlim = c(1980,2020))
+axis(1, at = seq(1980, 2020, by = 10))
+mtext(text = expression("Grain yield ( Mg ha"^-1~")"), side = 2, line = 2.5)
+abline(a = ((fixef(RES_rel_fit)[1] * scl) + center) / 1000 -
+      (2000 * fixef(RES_rel_fit)[2]),
+       b = fixef(RES_rel_fit)[2],
+       lty = 2)
+polygon(x = c(yrs, rev(yrs)),
+        y = c(ints1[1,],rev(ints1[2,])),
+        border = NA,
+        col = rgb(0,0,0,0.1))
+
+with(plot_data, segments(x0 = release_yr, y0=yield_kg + yield_se,
+                         y1 = yield_kg - yield_se))
+text(x = plot_data$release_yr-0.4, y = plot_data$yield_kg + 0.07,
+     labels = gsub("M", "M-", row.names(plot_data)),
+     pos = 4, cex = 0.75)
+text(x = 2000, y = 8, bquote(.(round(fixef(RES_rel_fit)[2] * scl, 1))~kg~ha^-1~yr^-1))
+mtext("Without adjusting for yield erosion", adj = 0)
+
+plot(mean ~ release_yr, data = plot_data2, pch = 16, #col = rgb(0,0,0,0.25),
+     xlab = "Release year",
+     ylab = "", #expression("Grain yield -corrected( Mg ha"^-1~")"),
+     yaxt = "n",
+     ylim = c(8, 12.5),
+     xaxt = "n",
+     xlim = c(1980,2020))
+axis(1, at = seq(1980, 2020, by = 10))
+abline(a = ((((fixef(RES_full)[1] * scl) + center) ) -
+           (2000 * release_yr_coef)) / 1000,
+       b = release_yr_coef/1000,
+       lty = 2)
+with(plot_data2, segments(x0 = release_yr, y0 = mean + se,
+                         y1 = mean - se))
+text(x = plot_data2$release_yr-0.4, y = plot_data2$mean + 0.07,
+     labels = gsub("M", "M-", row.names(plot_data2)),
+     pos = 4, cex = 0.75)
+text(x = 2000, y = 8, bquote(.(release_yr_coef)~kg~ha^-1~yr^-1))
+mtext("Adjusting for yield erosion", adj = 0)
+polygon(x = c(yrs, rev(yrs)),
+        y = c(ints2[1,],rev(ints2[2,])),
+        border = NA,
+        col = rgb(0,0,0,0.1))
+
+@
+% Change from a./b. to accounting for erosion, More details about how
+% yield was corrected - 
+\caption{The relationship between yield and the release year of medium
+  grain rice cultivars developed by the California Cooperative Rice
+  Research Foundation (California, USA), for the last 6 years (\Sexpr{min(comp_yrs)} -- \Sexpr{max(comp_yrs)}) as estimated by two methods;
+  without correcting for yield erosion over time (left) and 
+  correcting for yield erosion over time (right). The trend without correcting
+  for yield erosion (left) was estimated by omitting the year of each
+  cultivar's release from the model. Yield erosion was corrected for by
+  including the years since release (right) in the model.}
+\label{fig:release}
+\end{figure}
+
 \section{Discussion}
 
-Overall, we do not find conclusive evidence of increased yield
-potential increasing in medium grain rice cultivars released by
-CCRRF's breeding improvement program over the period
-\Sexpr{min(mod_data[,'year'])} to \Sexpr{max(mod_data[,'year'])}. After correcting yield comparisons
-for yield erosion over time in the older cultivar, the estimated yield
-improvement spans from \Sexpr{rel_CI[1]} -- \Sexpr{rel_CI[2]} kg
-ha\tsups{-1} year\tsups{-1}.  Even when the
-comparison is not corrected for differences in years since release
-between cultivars, the estimated improvements over time
-(\Sexpr{only_release_coef} kg ha\tsups{-1}
-year\tsups{-1}; Fig. \ref{fig:release} dashed line) are
-still much less than previous estimates of 37 -- 60 kg
-ha\tsups{-1} year\tsups{-1}
-\citep{Omar2014}. Given this evidence, we conclude that the breeding
-program has likely been maintaining but not substantially improving
-yield potential in the face of changing biotic and abiotic
-pressures. These biotic and abiotic forces are expected to erode yield
-over time \citep{Peng1998, Peng2000, Peng2010}. In fact, the trend
-seen here is extremely consistent with previous reports of erosion of
-genetic gains of cultivar IR8 at IRRI \citep{Peng1998, Peng2000}. The
-evidence here further supports that the erosion of yield over time is
-a phenomenon that affects all rice breeding programs. This further
-highlights both the need for maintenance breeding and the difficulty
-of increasing genetic yield potential as yield approaches the
-physiological limit. Without maintenance breeding continuously
-selecting cultivars well adapted to current conditions, CA rice
-production systems are predicted to have lost an estimated
-\Sexpr{round(yrs_coef * (-36/1000), 1)} Mg ha\tsups{-1}
-grain yield over the study period.
+We conclude that yield erosion has decreased the yield of cultivars
+released by the CCRRF over the last 30+ years. Biotic and abiotic
+forces have been speculated to be responsible for this
+yield erosion since the cultivar's genetics remain
+``frozen'' at the time of selection but the conditions the cultivar
+may experience over time are continuously changing \citep{Peng1998,
+Peng2000, Peng2010}. The trend observed here is consistent with
+previous reports of the erosion of genetic gains of cultivar IR8 at
+IRRI \citep{Peng1998, Peng2000}, suggesting the
+erosion of yield over time is a phenomenon that widely affects rice
+breeding programs. This highlights both the need for maintenance
+breeding and the difficulty of increasing genetic yield potential,
+especially in systems such as CA where yield is approaching the
+physiological limit \citep{espe16a}. Without maintenance breeding
+continuously selecting cultivars well adapted to current conditions,
+CA rice production systems are predicted to have lost an estimated
+\Sexpr{round(yrs_coef * (-36/1000), 1)} Mg ha\tsups{-1} grain yield
+over the study period.
 
-The goal of a breeding program is to not only maintain the
+Overall, we do not find conclusive evidence of increased genetic yield
+potential in medium grain rice cultivars released by CCRRF's breeding
+improvement program over the period \Sexpr{min(mod_data[,'year'])} to
+\Sexpr{max(mod_data[,'year'])}. After correcting yield comparisons for
+yield erosion over time, the estimated yield improvement spans from
+\Sexpr{rel_CI[1]} -- \Sexpr{rel_CI[2]} kg ha\tsups{-1} year\tsups{-1}.
+Even when the comparison is not corrected for yield erosion, the
+estimated improvements over time (\Sexpr{only_release_coef} kg
+ha\tsups{-1} year\tsups{-1}; Fig. \ref{fig:release}a.) are still much
+less than previous estimates of 37 -- 60 kg ha\tsups{-1}
+year\tsups{-1} \citep{Omar2014}. We conclude that the breeding program
+has likely been maintaining but not substantially improving yield
+potential in the face of changing biotic and abiotic pressures.
+
+However, the goal of a breeding program is to not only maintain the
 benchmark yield of the previous ``standard'' but to surpass it. Given
 the estimated yield decreases over time, to hit an ambitious target of
-0.5 Mg ha\tsups{-1} (approximately 5\%) increase in yield
-per release (with roughly 4 -- 5 years between releases), a new cultivar
-needs to yield an additional \Sexpr{round(yrs_coef*-4,-1)} --
-\Sexpr{round(yrs_coef*-5,-1)} kg ha\tsups{-1} in
-side-by-side comparison on top of the 0.5 Mg ha\tsups{-1}
-target improvement. Often realized side-by-side yield
-improvements 
-are closer to 2 -- 3\% per release, in which case no absolute
-improvements are made beyond maintenance. The evidence here is
-consistent with the latter situation occurring in CA, further suggesting
-that the genetic yield potential has not increased substantially in
-the last 30+ years (Fig. \ref{fig:release}).
+0.5 Mg ha\tsups{-1} (approximately 5\%) increase in yield per release
+(with roughly 4 -- 5 years between releases), a new cultivar needs to
+yield an additional \Sexpr{round(yrs_coef*-4,-1)} --
+\Sexpr{round(yrs_coef*-5,-1)} kg ha\tsups{-1} in side-by-side
+comparison on top of the 0.5 Mg ha\tsups{-1} target improvement. Often
+realized side-by-side yield improvements are closer to 2 -- 3\% per
+release, in which case no absolute improvements are made beyond
+maintenance. The evidence here is consistent with the latter situation
+occurring in CA, further suggesting that the genetic yield potential
+has not increased substantially in the last 30+ years
+(Fig. \ref{fig:release}).
 
@@
 
-That is not to say there have not been yield improvements over the
-last 30 years. California state average yields have increased slowly,
-even as genetic gains have been small
+California state average yields have increased slowly
+even as little to no gains have been seen in the breeding trials
 (Fig. \ref{fig:state_yield}). The evidence here suggests mechanisms
-other than increased genetic yield potential are responsible
+other than increased genetic yield potential are responsible for these gains
 (Fig. \ref{fig:compare_yrs}). Improvement in agronomic management,
@@
 researcher-managed production system implementing best management
-practices in more controlled situations compared to farmers, state
-average yields (approximately 10 Mg ha\tsups{-1} at the time
-of writing; Fig. \ref{fig:state_yield}) may be approaching the current
+practices in more controlled situations compared to farmers' fields, state
+average yields (approximately 10 Mg ha\tsups{-1} at the time of
+writing; Fig. \ref{fig:state_yield}) may be approaching the current
 genetic limit. Assuming that the level of management possible in small
@@
 
+There are several cultivars that were estimated to have better than
+average yields, even after correcting for yield erosion. However,
+these cultivars (M-201, M-204, M-205, and possibly M-209;
+Fig. \ref{fig:comp_ids})) may not be widely adapted. It is known that
+cool temperatures during the booting stage can have a large impact on
+grain yields \citep{espe2017}, and it is also known that the RES
+experiences warmer temperatures than many other rice growing areas in
+CA \citep{espe2017}. These cultivars may perform well at the RES but not widely in
+CA. Therefore, it should be noted that while our investigation has
+been concerned about increasing yield potential, the estimates here
+might not capture yield stability. Yield stability is notoriously
+difficult to quantify (CITE), hence further investigation is needed to
+explore how yield stability has changed over the study period.
+
 To be clear, our investigation here is narrowly focused on the
 rough-rice/field grain yield. Other advancements due to plant
-improvement, including increased quality \citep{mckenzie1993} and milling yield
-\citep{mckenzie1994} have been made over the study period and are no
-less important. However, planning to meet future demand necessitates
-increases in grain yield \citep{Godfray2010, Foley2011a, Mueller2012}.
-Meeting the challenge to produce more grain without expansion of land
-under cultivation will rely on new genetics with increased yield
-potential. The evidence here suggests that to accomplish true
-increases over time, the improvement target needs take into account
-decreases in the current ``standard'' cultivars' yield over
-time (Fig. \ref{fig:theory}). While side-by-side comparisons will continue to be useful,
-comparisons across years to each cultivar's benchmark yield at the
-time of release should provide information on the absolute
-improvements being made in yield potential over time. New technologies
-may need to be adapted to further increase yield potential, including
-hybrid rice.
+improvement, including increased quality \citep{mckenzie1993} and
+milling yield \citep{mckenzie1994} have been made over the study
+period and are no less important. However, planning to meet future
+demand necessitates increases in grain yield \citep{Godfray2010,
+Foley2011a, Mueller2012}. Meeting the challenge to produce more grain
+without expansion of land under cultivation will rely on new genetics
+with increased yield potential. The evidence here suggests that to
+accomplish true increases over time, the improvement target needs take
+into account decreases in the current ``standard'' cultivars' yield
+over time (Fig. \ref{fig:theory}). While side-by-side comparisons will
+continue to be useful, comparisons across years to each cultivar's
+benchmark yield at the time of release should provide information on
+the absolute improvements being made in yield potential over time. New
+technologies may need to be adapted to further increase yield
+potential, including hybrid rice.
 
@@
 feeding a growing population in a changing world. To achieve these
-improvements, we show here that (1) continuous maintenance via plant
+improvements, we discuss here that (1) continuous maintenance via plant
 improvement is required, without which yields would decrease over
@@
 CIs = posterior_interval(RES_full)[1:3,] * scl
-CIs[1,] = round(CIs[1,] + center, -2)
+CIs[1,] = round((CIs[1,] + center)/1000, 1)
 CIs = round(CIs, 1)
 med_eff = fixef(RES_full) * scl
-med_eff[1] = round(med_eff[1] + center, -2)
+med_eff[1] = round((med_eff[1] + center)/1000, 1)
 med_eff = round(med_eff, 1)
@@
 RES_coef = cbind(med_eff, paste(CIs[,1],CIs[,2], sep = " -- "))
-rownames(RES_coef) = c("Intercept (kg ha\\tsups{-1})",
+rownames(RES_coef) = c("Intercept (Mg ha\\tsups{-1})",
                        "Release year (kg ha\\tsups{-1} year\\tsups{-1})",
-                       "Years since release (kg ha\\tsups{-1} year\\tsups{-1})")
+                       "Yield erosion (kg ha\\tsups{-1} year\\tsups{-1})")
 colnames(RES_coef) = c("Median", "90\\% Credible interval")
@@
 print(xtable(RES_coef,
-             caption = "Estimated median effects and 90\\% credible intervals.",
+             caption = "The estimated median fixed effects and 90\\% credible intervals for the improvement in yield of medium grain rice cultivars developed by the California Cooperative Rice Research Foundation, California, USA over the time period 1984 -- 2016.",
              label = "tbl:coef",
@@
 @
+\newpage
+
+<<echo=FALSE, results=tex>>=
+ids = aggregate(release_yr ~ id, data = mod_data, unique)
+ids$dth = as.integer(round(aggregate(dth ~ id, data = mod_data, mean, na.rm = TRUE)$dth, 0))
+colnames(ids) = c("Cultivar", "Release Year", "Days to 50% Heading")
+ids = ids[order(ids[,2]),]
+ids[,2] = as.integer(ids[,2])
+ids[,1] = gsub("M", "M-", ids[,1])
+
+print(xtable(ids,
+             caption = "All named medium grain rice cultivars and their associated release year planted at the Rice Experiment Station (RES), Biggs, California from 1984 -- 2015.",
+             label = "tbl:ids"),
+      include.rownames = FALSE,
+      booktabs = TRUE,
+      caption.placement = "top")
+
+@
+
+% Notes: Duration has gotten shorter. Etc. Rearrange - End
+% introduction with evidence of erosion, quatify the effect on yield
+% increases.
+
 \newpage
@@
 
-<<echo=FALSE, results=tex>>=
-ids = aggregate(release_yr ~ id, data = mod_data, unique)
-colnames(ids) = c("Cultivar", "Release Year")
-ids = ids[order(ids[,2]),]
-ids[,2] = as.integer(ids[,2])
-ids[,1] = gsub("M", "M-", ids[,1])
-
-print(xtable(ids,
-             caption = "All named medium grain rice cultivars and their associated release year planted at the Rice Experiment Station (RES), Biggs, California from 1984 -- 2015.",
-             label = "tbl:ids"),
-      include.rownames = FALSE,
-      booktabs = TRUE,
-      caption.placement = "top")
-@
-
-% \begin{figure}[h]
-
-% <<label=plot1, fig=TRUE, echo=FALSE>>=
-% boxplot(yield_kg ~ year, data = mod_data,
-%         xlab = "Year", ylab = expression("Yield ( kg ha"^-1~")"))
-% @
-%   \caption{Distribution of yield over the period 1995 -- 2015.} \label{fig:plot}
-% \end{figure}
-
-% \begin{figure}[h]
-% <<label=plot2, fig=TRUE, echo=FALSE>>=
-
-%   boxplot(yield_kg ~ id, data = mod_data,
-%           xlab = "Cultivar",
-%           ylab = "Yield ( kg/ha")
-% @
-% \caption{Boxplot of yields for released medium grain cultivars for
-%   the period 1995--2015.}
-% \label{fig:raw_yield}
-% \end{figure}
-
 \end{document}

### 2017-07-03 src/prep_model_data.R

@@ -18,7 +18,6 @@ mod_data <- all_data[vars,]
 # To simplify model interpretation, only analyze RES
 # since this is the best managed site
 # with the most data
-
 mod_data = mod_data[mod_data$site == "RES",]
 
 ## Use officially published years instead of years in trial
@@ -29,6 +28,10 @@ mod_data$release_yr <- yrTbl$Year[i]
 mod_data$release_yr_c <- mod_data$release_yr - 2000
 
 mod_data$yrs_in_trial <- mod_data$year - mod_data$release_yr
+
+# Omit where the variety was present for more than 4 years before release
+tooOld = mod_data$yrs_in_trial < -4
+mod_data = mod_data[!tooOld,]
 
 # convert yield to kg/ha here once, rather than repeatedly in paper
 mod_data$yield_kg = mod_data$yield_lb * 1.12

### 2017-08-06 output/breed_paper.Rnw
Commit before rearranging the paper

@@
 \usepackage{longtable,booktabs}
+\usepackage{comment}
 \usepackage{graphicx,grffile, textcomp}
@@
 comparisons show increased yield while no substantive increases have
-been seen in farmers' fields. We confirmed an apparent increase in
-yield in side-by-side comparisons between old and newer varieties when
-the yield erosion was not
-considered. However, we demonstrate this apparent increase in yield
-can be explained by the differences in the time since selection
+been seen in farmers' fields. Side-by-side comparisons under yield
+erosions will show ``false'' increases due to the older cultivar
+decreasing, not due to increased genetic yield potential by the newer
+cultivar. We confirmed an apparent increase in yield in side-by-side
+comparisons between old and newer varieties when the yield erosion was
+not considered. However, we demonstrate this apparent increase in
+yield can be explained by the differences in the time since selection
 between cultivars. The impact of yield erosion was estimated to be
-29.8 kg ha\tsups{-1} year\tsups{-1} after initial
-selection. Once this effect was considered, the apparent yield
-advantage of newer cultivars over old was uncertain. This underscores
-the importance of breeding to maintain yield and the challenges of
-increasing the genetic yield potential.
+29.3 kg ha\tsups{-1} year\tsups{-1} (90\% credible interal -4.4 to
+-53.3) after initial selection. Once this effect was considered, the
+apparent yield advantage of newer cultivars over old was uncertain
+(-3.3 kg ha\tsups{-1} year\tsups{-1}, 90\% credible interval -36.1 to
+31.5). This underscores the importance of breeding to maintain yield
+and the challenges of increasing the genetic yield potential.
 
@@
 30 years in CA. Specifically, two competing hypotheses were
-tested: (1) the yield advantage of each newly developed cultivars is
-greater than previous cultivars, and (2) the erosion of yield over
+tested: (1) the genetic yield advantage has increased over the last
+30+ years, and (2) the genetic yield advantage has not increased over
+the last 30+ and the erosion of yield over
 time is creating the illusion of improved yields (i.e., the new
@@
 \caption{Average rice yields over the period \Sexpr{min(CA_data$Year)}
-  -- \Sexpr{max(CA_data$Year)} for California, USA, as reported by the
+  to \Sexpr{max(CA_data$Year)} for California, USA, as reported by the
   US Dept. of Agriculture National Agricultural Statistics
@@
 winters during which most of the annual precipitation occurs
-(\Sexpr{18*24.5} -- \Sexpr{25*24.5} mm annually) and warm summers
+(\Sexpr{18*24.5} to \Sexpr{25*24.5} mm annually) and warm summers
 largely free of precipitation events. Soils at the RES are classified
@@
 the bulk of the trials for the Statewide Variety Trials (approximately
-3 -- 4 times the number of plots compared to any other location). For
+3 to 4 times the number of plots compared to any other location). For
 these reasons, we focused our investigation on the RES to avoid
 potential complications due to site, climate, and management
-differences at the other 6 -- 8 farmer-managed trials per year.
+differences at the other 6 to 8 farmer-managed trials per year.
 
@@
 90\% of the planted area in CA. These cultivars are typcically planted
-in the Statewide Variety Trials for 2 -- 4 years prior to their
+in the Statewide Variety Trials for 2 to 4 years prior to their
 official release. For these cultivars planted in
@@
 \begin{figure}[H]
+<<fig=TRUE, width = 8, height = 5, out.width = '85%', echo=FALSE>>=
+tars <- c("M202","M206")
+par(mfrow = c(1, length(tars)),
+    mar = c(4,0,2,0),
+    oma = c(4,4,0.5,0.5))
+
+plot_data = lapply(tars, function(x) {
+    dd = mod_data[mod_data$id == x & mod_data$site == "RES",]
+    tmp = prep_plot_data(dd$yield_kg/1000 , dd$year)
+    tmp$year = as.numeric(row.names(tmp))
+    tmp
+    })
+
+invisible(sapply(1:2, function(i){
+    plot(mean ~ year,
+         data = plot_data[[i]],
+         xlim = c(1985, 2016), ylim = c(7.5,13.5),
+         pch = 16, main = gsub("M","M-", tars[i]),
+         ylab = "",
+         yaxt = "n",
+         xlab = "Year",
+         cex.axis = 0.8)
+    segments(x0 = plot_data[[i]]$year,
+             y0 = plot_data[[i]]$mean + plot_data[[i]]$se,
+             y1 = plot_data[[i]]$mean - plot_data[[i]]$se)
+    if(i == 1){
+        axis(2, cex.axis = 0.8,)
+        mtext(expression("Yield ( Mg ha"^-1~")"), 2, line = 2.25, cex.lab = 0.6)
+    }
+    # abline(a = ((fixef(RES_full)[1] * scl) + center) / 1000 -
+    #            (2000 * yrs_coef/1000),
+    #        b = yrs_coef/1000,
+    #        lty = 1)
+    lines(as.numeric(rownames(plot_data[[i]])),
+          11.2 + (1:nrow(plot_data[[i]])) * (yrs_coef/1000))
+    # abline(a = ((fixef(RES_full)[1] * scl) + center) / 1000,
+           # b = yrs_coef/1000)
+
+    # abline(h = mean(mod_data$yield_kg[(mod_data$id == tars[i] &
+    #                                    mod_data$site == "RES")] ),
+    #        lty = 1, col = "grey50")
+    # mod = lm(mean ~ year, data = plot_data[[i]])
+    # tmp = predict(mod)
+    # if(i == 1){
+    #     abline(mod)
+    # }else{
+    #     lines(as.numeric(names(tmp)), tmp)
+    # }
+    abline(h = 11.2, lty = 2)
+    # text(2000, 7.5, bquote(.(round(coef(mod)[2] * 1000, 1))~kg~ha^-1~year^-1))
+}))
+
+@
+\caption{The performance of two major cultivars at the CA Rice
+Experiment Station (RES) over the period
+\Sexpr{min(mod_data[,'year'])} -- \Sexpr{max(mod_data[,'year'])}. The
+solid line is the estimated yield erosion (\Sexpr{yrs_coef} kg
+ha\tsups{-1} year\tsups{-1}), while the dashed line is the published
+3-year average yield of M-202 at its release (1982-1984). Both
+cultivars have a similar yield at their respective year of release,
+and similar decreases over time.}
+\label{fig:compare_yrs}
+\end{figure}
+
+\subsection{Genetic yield improvements}
+
+The full model which corrected for both the year of release and yield
+erosion was estimated to have better out-of-sample predictive accuracy
+compared to reduced models containing only one predictor. As estimated
+by the best model, there is no conclusive evidence that the yields of
+medium grain rice cultivars selected and released by CCRRF's breeding
+program have, on average, increased over the last 30 years. The
+estimated change in grain yield per year in the CCRRF's breeding
+trials at the RES was estimated to be a negligible
+\Sexpr{release_yr_coef} kg ha\tsups{-1} year\tsups{-1}, with both
+small positive and negative trends within the credible interval (90\%
+CI: \Sexpr{rel_CI[1]} to \Sexpr{rel_CI[2]})
+(Fig. \ref{fig:w_erosion}). Hence, the evidence here is consistent
+with little to no increase in yield potential over the study
+period. When yield erosion is not corrected for in the comparison
+(i.e., omitted from the model), there is an apparent small increase
+per year (approximately \Sexpr{only_release_coef} kg ha\tsups{-1}
+year\tsups{-1}; 90\% CI \Sexpr{only_release_CI[1]} to
+\Sexpr{only_release_CI[2]}) (Fig. \ref{fig:wo_erosion}), though this
+estimate ignores the confounding effect of yield erosion over time and
+performed worse in out-of-sample predictive accuracy.
+
+After correcting for the other variables, some cultivars are estimated
+to have greater yield compared to the overall average. The marginal
+estimate for the performance of a cultivar averaged over the
+uncertainty in the other parameters in the model (i.e., the random
+effect) shows three cultivars with higher than average yields:
+\Sexpr{paste(gsub('M','M-',rev(row.names(id_est))[c(1,2,4)]),collapse=',')}
+(Figs. \ref{fig:w_erosion} \& \ref{fig:ids_comp}). A fourth, M-209,
+might also have higher than average yield, though the CI for M-209
+overlaps with zero (Fig. \ref{fig:ids_comp}).
+
+\begin{comment}
+\begin{figure}[H]
+
+<<fig=TRUE, echo=FALSE, width = 6, height = 5>>=
+yrs <- 1975:2025
+
+pp <- posterior_linpred(RES_rel_fit, re.form = NA,
+                        newdata = data.frame(release_yr_c = yrs - 2000,
+                                             yrs_in_trial = 0),
+                        prob = 0.5)
+pp <-  (pp * scl + center) / 1000
+ints1 = apply(pp, 2, quantile, p = c(0.05, 0.95))
+
+pp <- posterior_linpred(RES_full, re.form = NA,
+                        newdata = data.frame(release_yr_c = yrs - 2000,
+                                             yrs_in_trial = 0),
+                        prob = 0.5)
+pp <-  (pp * scl + center) / 1000
+ints2 = apply(pp, 2, quantile, p = c(0.05, 0.95))
+
+res = mod_data$site == "RES"
+comp_yrs = 2011:2016
+idx = mod_data$year %in% comp_yrs
+
+plot_data = prep_plot_data(mod_data$yield_kg[idx] / 1000 ,
+                           mod_data$id[idx])
+
+plot_data$tmp = tapply(mod_data$release_yr[idx],
+                       mod_data$id[idx], unique)
+
+colnames(plot_data) = c("yield_kg", "yield_se", "release_yr")
+
+# get corrected yield
+wo_erosion = expand.grid(id = row.names(plot_data),
+            year = comp_yrs,
+            release_yr_c = 0,
+            yrs_in_trial = 0)
+wo_erosion$corr_yield = colMeans(posterior_predict(RES_full,
+                               newdata = wo_erosion)) + center/1000
+plot_data2 = prep_plot_data(wo_erosion$corr_yield,
+                            wo_erosion$id)
+plot_data2$release_yr = plot_data$release_yr
+
+# par(mfrow = c(1,2),
+#     mar = c(4,0,2,0.75),
+#     oma = c(4,4,0.5,0.5))
+
+plot(yield_kg ~ release_yr, data = plot_data, pch = 16, #col = rgb(0,0,0,0.25),
+     xlab = "Release year",
+     ylab = "", 
+     ylim = c(8, 12.5),
+     xaxt = "n",
+     xlim = c(1980,2020))
+axis(1, at = seq(1980, 2020, by = 10))
+mtext(text = expression("Grain yield ( Mg ha"^-1~")"), side = 2, line = 2.5)
+abline(a = ((fixef(RES_rel_fit)[1] * scl) + center) / 1000 -
+      (2000 * fixef(RES_rel_fit)[2]),
+       b = fixef(RES_rel_fit)[2],
+       lty = 2)
+polygon(x = c(yrs, rev(yrs)),
+        y = c(ints1[1,],rev(ints1[2,])),
+        border = NA,
+        col = rgb(0,0,0,0.1))
+
+with(plot_data, segments(x0 = release_yr, y0=yield_kg + yield_se,
+                         y1 = yield_kg - yield_se))
+text(x = plot_data$release_yr-0.4, y = plot_data$yield_kg + 0.07,
+     labels = gsub("M", "M-", row.names(plot_data)),
+     pos = 4, cex = 0.75)
+text(x = 2000, y = 8, bquote(.(round(fixef(RES_rel_fit)[2] * scl, 1))~kg~ha^-1~yr^-1))
+mtext("Without adjusting for yield erosion", adj = 0)
+
+@
+\caption{The relationship between yield and the release year of medium
+grain rice cultivars developed by the California Cooperative Rice
+Research Foundation (California, USA), for the last 6 years
+(\Sexpr{min(comp_yrs)} -- \Sexpr{max(comp_yrs)}) as estimated without
+correcting for yield erosion over time.}
+\label{fig:wo_erosion}
+\end{figure}
+\end{comment}
+
+\begin{figure}[H]
+<<fig=TRUE, echo=FALSE, width = 6, height = 5>>=
+
+plot(mean ~ release_yr, data = plot_data2, pch = 16, #col = rgb(0,0,0,0.25),
+     xlab = "Release year",
+     ylab = "", #expression("Grain yield -corrected( Mg ha"^-1~")"),
+     yaxt = "n",
+     ylim = c(8, 12.5),
+     xaxt = "n",
+     xlim = c(1980,2020))
+axis(1, at = seq(1980, 2020, by = 10))
+abline(a = ((((fixef(RES_full)[1] * scl) + center) ) -
+           (2000 * release_yr_coef)) / 1000,
+       b = release_yr_coef/1000,
+       lty = 2)
+with(plot_data2, segments(x0 = release_yr, y0 = mean + se,
+                         y1 = mean - se))
+text(x = plot_data2$release_yr-0.4, y = plot_data2$mean + 0.07,
+     labels = gsub("M", "M-", row.names(plot_data2)),
+     pos = 4, cex = 0.75)
+text(x = 2000, y = 8, bquote(.(release_yr_coef)~kg~ha^-1~yr^-1))
+mtext("Adjusting for yield erosion", adj = 0)
+polygon(x = c(yrs, rev(yrs)),
+        y = c(ints2[1,],rev(ints2[2,])),
+        border = NA,
+        col = rgb(0,0,0,0.1))
+
+@
+% Change from a./b. to accounting for erosion, More details about how
+% yield was corrected - 
+\caption{The relationship between yield and the release year of medium
+grain rice cultivars developed by the California Cooperative Rice
+Research Foundation (California, USA), for the last 6 years
+(\Sexpr{min(comp_yrs)} -- \Sexpr{max(comp_yrs)}) as estimated by 
+correcting for yield erosion over time. Yield erosion was corrected
+for by adjusting each observed yield by estimated amount lost 
+  (\Sexpr{yrs_coef} kg ha\tsups{-1} year\tsups{-1} after release).}
+\label{fig:w_erosion}
+\end{figure}
+
+\subsection{Yield erosion}
+\begin{figure}[H]
 
@@
 
-\subsection{Yield erosion}
-
 There is estimated to be an decrease of yield over time once a
 cultivar is selected (i.e., yield erosion) (\Sexpr{yrs_coef} kg
-ha\tsups{-1} year\tsups{-1}, 90\% CI: \Sexpr{RES_CI[2]} --
+ha\tsups{-1} year\tsups{-1}, 90\% CI: \Sexpr{RES_CI[2]} to
 \Sexpr{RES_CI[1]}). The oldest cultivar in the trials is estimated to
@@
 over M-202 can be attributed solely to the erosion of M-202's yield
-over time (Figs \ref{fig:release} \& \ref{fig:compare_vars}). Further
+over time (Figs \ref{fig:w_erosion} \& \ref{fig:compare_yrs}). Further
 supporting this estimate, both cultivars yielded similarly at their
@@
 fact that M-202 was released 19 years prior to M-206
-(Fig. \ref{fig:release}, Table \ref{tbl:ids}).
-
-\begin{figure}[H]
-<<fig=TRUE, width = 8, height = 5, out.width = '85%', echo=FALSE>>=
-tars <- c("M202","M206")
-par(mfrow = c(1, length(tars)),
-    mar = c(4,0,2,0),
-    oma = c(4,4,0.5,0.5))
-
-plot_data = lapply(tars, function(x) {
-    dd = mod_data[mod_data$id == x & mod_data$site == "RES",]
-    tmp = prep_plot_data(dd$yield_kg/1000 , dd$year)
-    tmp$year = as.numeric(row.names(tmp))
-    tmp
-    })
-
-invisible(sapply(1:2, function(i){
-    plot(mean ~ year,
-         data = plot_data[[i]],
-         xlim = c(1985, 2016), ylim = c(7.5,13.5),
-         pch = 16, main = gsub("M","M-", tars[i]),
-         ylab = "",
-         yaxt = "n",
-         xlab = "Year",
-         cex.axis = 0.8)
-    segments(x0 = plot_data[[i]]$year,
-             y0 = plot_data[[i]]$mean + plot_data[[i]]$se,
-             y1 = plot_data[[i]]$mean - plot_data[[i]]$se)
-    if(i == 1){
-        axis(2, cex.axis = 0.8,)
-        mtext(expression("Yield ( Mg ha"^-1~")"), 2, line = 2.25, cex.lab = 0.6)
-    }
-
-    # abline(h = mean(mod_data$yield_kg[(mod_data$id == tars[i] &
-    #                                    mod_data$site == "RES")] ),
-    #        lty = 1, col = "grey50")
-    mod = lm(mean ~ year, data = plot_data[[i]])
-    tmp = predict(mod)
-    if(i == 1){
-        abline(mod)
-    }else{
-        lines(as.numeric(names(tmp)), tmp)
-    }
-    abline(h = 11.2, lty = 2)
-    text(2000, 7.5, bquote(.(round(coef(mod)[2] * 1000, 1))~kg~ha^-1~year^-1))
-}))
-
-@
-\caption{The performance of two major cultivars at the CA Rice
-  Experiment Station (RES) over the period
-  \Sexpr{min(mod_data[,'year'])} -- \Sexpr{max(mod_data[,'year'])}. The
-  solid line represents a simple linear regression of yield for each
-  cultivar over the period which it was planted at the RES (fit to only
-  that cultivar's data), while the dashed line is the published 3-year
-  average yield of M-202 at its release (1982-1984). Both cultivars have
-  a similar yield at their respective year of release, and similar
-  decreases over time. The slope for the individual regressions are
-  comparable to the slope estimated from all medium grain cultivars
-  (\Sexpr{yrs_coef} kg ha\tsups{-1} year\tsups{-1}).}
-\label{fig:compare_yrs}
-\end{figure}
-
-\subsection{Genetic yield improvements}
-
-The full model which corrected for both the year of release and yield
-erosion was estimated to have better out-of-sample predictive accuracy
-compared to reduced models containing only one predictor. As estimated
-by the best model, there is no conclusive evidence that the yields of
-medium grain rice cultivars selected and released by CCRRF's breeding
-program have, on average, increased over the last 30 years. The
-estimated change in grain yield per year in the CCRRF's breeding
-trials at the RES was estimated to be a negligible
-\Sexpr{release_yr_coef} kg ha\tsups{-1} year\tsups{-1}, with both
-small positive and negative trends within the credible interval (90\%
-CI: \Sexpr{rel_CI[1]} -- \Sexpr{rel_CI[2]})
-(Fig. \ref{fig:release}b.). Hence, the evidence here is consistent
-with little to no increase in yield potential over the study
-period. When yield erosion is not corrected for in the comparison
-(i.e., omitted from the model), there is an apparent small increase
-per year (approximately \Sexpr{only_release_coef} kg ha\tsups{-1}
-year\tsups{-1}; 90\% CI \Sexpr{only_release_CI[1]} --
-\Sexpr{only_release_CI[2]}) (Fig. \ref{fig:release}a.), though this
-estimate ignores the confounding effect of yield erosion over time and
-performed worse in out-of-sample predictive accuracy.
-
-After correcting for the other variables, some cultivars are estimated
-to have greater yield compared to the overall average. The marginal
-estimate for the performance of a cultivar averaged over the
-uncertainty in the other parameters in the model (i.e., the random
-effect) shows three cultivars with higher than average yields:
-\Sexpr{paste(gsub('M','M-',rev(row.names(id_est))[c(1,2,4)]),collapse=',')}
-(Figs. \ref{fig:release} \& \ref{fig:ids_comp}). A fourth, M-209,
-might also have higher than average yield, though the CI for M-209
-overlaps with zero (Fig. \ref{fig:ids_comp}).
-
-\begin{figure}[H]
-
-<<fig=TRUE, echo=FALSE, width = 9, height = 6>>=
-yrs <- 1975:2025
-
-pp <- posterior_linpred(RES_rel_fit, re.form = NA,
-                        newdata = data.frame(release_yr_c = yrs - 2000,
-                                             yrs_in_trial = 0),
-                        prob = 0.5)
-pp <-  (pp * scl + center) / 1000
-ints1 = apply(pp, 2, quantile, p = c(0.05, 0.95))
-
-pp <- posterior_linpred(RES_full, re.form = NA,
-                        newdata = data.frame(release_yr_c = yrs - 2000,
-                                             yrs_in_trial = 0),
-                        prob = 0.5)
-pp <-  (pp * scl + center) / 1000
-ints2 = apply(pp, 2, quantile, p = c(0.05, 0.95))
-
-res = mod_data$site == "RES"
-comp_yrs = 2011:2016
-idx = mod_data$year %in% comp_yrs
-
-plot_data = prep_plot_data(mod_data$yield_kg[idx] / 1000 ,
-                           mod_data$id[idx])
-
-plot_data$tmp = tapply(mod_data$release_yr[idx],
-                       mod_data$id[idx], unique)
-
-colnames(plot_data) = c("yield_kg", "yield_se", "release_yr")
-
-# get corrected yield
-wo_erosion = expand.grid(id = row.names(plot_data),
-            year = comp_yrs,
-            release_yr_c = 0,
-            yrs_in_trial = 0)
-wo_erosion$corr_yield = colMeans(posterior_predict(RES_full,
-                               newdata = wo_erosion)) + center/1000
-plot_data2 = prep_plot_data(wo_erosion$corr_yield,
-                            wo_erosion$id)
-plot_data2$release_yr = plot_data$release_yr
-
-par(mfrow = c(1,2),
-    mar = c(4,0,2,0.75),
-    oma = c(4,4,0.5,0.5))
-
-plot(yield_kg ~ release_yr, data = plot_data, pch = 16, #col = rgb(0,0,0,0.25),
-     xlab = "Release year",
-     ylab = "", 
-     ylim = c(8, 12.5),
-     xaxt = "n",
-     xlim = c(1980,2020))
-axis(1, at = seq(1980, 2020, by = 10))
-mtext(text = expression("Grain yield ( Mg ha"^-1~")"), side = 2, line = 2.5)
-abline(a = ((fixef(RES_rel_fit)[1] * scl) + center) / 1000 -
-      (2000 * fixef(RES_rel_fit)[2]),
-       b = fixef(RES_rel_fit)[2],
-       lty = 2)
-polygon(x = c(yrs, rev(yrs)),
-        y = c(ints1[1,],rev(ints1[2,])),
-        border = NA,
-        col = rgb(0,0,0,0.1))
-
-with(plot_data, segments(x0 = release_yr, y0=yield_kg + yield_se,
-                         y1 = yield_kg - yield_se))
-text(x = plot_data$release_yr-0.4, y = plot_data$yield_kg + 0.07,
-     labels = gsub("M", "M-", row.names(plot_data)),
-     pos = 4, cex = 0.75)
-text(x = 2000, y = 8, bquote(.(round(fixef(RES_rel_fit)[2] * scl, 1))~kg~ha^-1~yr^-1))
-mtext("Without adjusting for yield erosion", adj = 0)
-
-plot(mean ~ release_yr, data = plot_data2, pch = 16, #col = rgb(0,0,0,0.25),
-     xlab = "Release year",
-     ylab = "", #expression("Grain yield -corrected( Mg ha"^-1~")"),
-     yaxt = "n",
-     ylim = c(8, 12.5),
-     xaxt = "n",
-     xlim = c(1980,2020))
-axis(1, at = seq(1980, 2020, by = 10))
-abline(a = ((((fixef(RES_full)[1] * scl) + center) ) -
-           (2000 * release_yr_coef)) / 1000,
-       b = release_yr_coef/1000,
-       lty = 2)
-with(plot_data2, segments(x0 = release_yr, y0 = mean + se,
-                         y1 = mean - se))
-text(x = plot_data2$release_yr-0.4, y = plot_data2$mean + 0.07,
-     labels = gsub("M", "M-", row.names(plot_data2)),
-     pos = 4, cex = 0.75)
-text(x = 2000, y = 8, bquote(.(release_yr_coef)~kg~ha^-1~yr^-1))
-mtext("Adjusting for yield erosion", adj = 0)
-polygon(x = c(yrs, rev(yrs)),
-        y = c(ints2[1,],rev(ints2[2,])),
-        border = NA,
-        col = rgb(0,0,0,0.1))
-
-@
-% Change from a./b. to accounting for erosion, More details about how
-% yield was corrected - 
-\caption{The relationship between yield and the release year of medium
-  grain rice cultivars developed by the California Cooperative Rice
-  Research Foundation (California, USA), for the last 6 years (\Sexpr{min(comp_yrs)} -- \Sexpr{max(comp_yrs)}) as estimated by two methods;
-  without correcting for yield erosion over time (left) and 
-  correcting for yield erosion over time (right). The trend without correcting
-  for yield erosion (left) was estimated by omitting the year of each
-  cultivar's release from the model. Yield erosion was corrected for by
-  including the years since release (right) in the model.}
-\label{fig:release}
-\end{figure}
+(Fig. \ref{fig:compare_yrs}, Table \ref{tbl:ids}).
 
@@
 
-We conclude that yield erosion has decreased the yield of cultivars
-released by the CCRRF over the last 30+ years. Biotic and abiotic
-forces have been speculated to be responsible for this
-yield erosion since the cultivar's genetics remain
-``frozen'' at the time of selection but the conditions the cultivar
-may experience over time are continuously changing \citep{Peng1998,
-Peng2000, Peng2010}. The trend observed here is consistent with
-previous reports of the erosion of genetic gains of cultivar IR8 at
-IRRI \citep{Peng1998, Peng2000}, suggesting the
-erosion of yield over time is a phenomenon that widely affects rice
-breeding programs. This highlights both the need for maintenance
-breeding and the difficulty of increasing genetic yield potential,
-especially in systems such as CA where yield is approaching the
-physiological limit \citep{espe16a}. Without maintenance breeding
-continuously selecting cultivars well adapted to current conditions,
-CA rice production systems are predicted to have lost an estimated
-\Sexpr{round(yrs_coef * (-36/1000), 1)} Mg ha\tsups{-1} grain yield
-over the study period.
+We estimate that yield erosion has decreased the yield of medium
+grain rice cultivars released by the CCRRF over the last 30+ years by
+an average of \Sexpr{yrs_coef} Mg ha\tsups{-1},
+potentially offsetting any gains made by the breeding program. Biotic
+and abiotic forces have previously been speculated to be responsible
+for yield erosion, since the cultivar's genetics remain stable from the
+time of selection but the conditions the cultivar may experience over
+time are continuously changing \citep{Peng1998, Peng2000,
+Peng2010}. The trend observed here is consistent with previous reports
+of the erosion of genetic gains of cultivar IR8 at IRRI
+\citep{Peng1998, Peng2000}, suggesting the erosion of yield over time
+is a phenomenon that widely affects rice breeding programs. This
+highlights both the need for maintenance breeding and the difficulty
+of increasing genetic yield potential, especially in systems such as
+CA where yield is approaching the physiological limit
+\citep{espe16a}. Without maintenance breeding continuously selecting
+cultivars well adapted to current conditions, CA rice production
+systems are predicted to have lost an estimated
+\Sexpr{round(yrs_coef*(-36/1000),1)} Mg ha\tsups{-1} grain yield over
+the study period.
 
@@
 yield erosion over time, the estimated yield improvement spans from
-\Sexpr{rel_CI[1]} -- \Sexpr{rel_CI[2]} kg ha\tsups{-1} year\tsups{-1}.
-Even when the comparison is not corrected for yield erosion, the
-estimated improvements over time (\Sexpr{only_release_coef} kg
-ha\tsups{-1} year\tsups{-1}; Fig. \ref{fig:release}a.) are still much
-less than previous estimates of 37 -- 60 kg ha\tsups{-1}
-year\tsups{-1} \citep{Omar2014}. We conclude that the breeding program
-has likely been maintaining but not substantially improving yield
-potential in the face of changing biotic and abiotic pressures.
+\Sexpr{rel_CI[1]} to \Sexpr{rel_CI[2]} kg ha\tsups{-1}
+year\tsups{-1}. This suggests that both increases and decreases in the
+genetic yield potential over the study period. However, even when the
+comparison is not corrected for yield erosion, the estimated
+improvements over time (\Sexpr{only_release_coef} kg ha\tsups{-1}
+year\tsups{-1}; Fig. \ref{fig:wo_erosion}) are still much less than
+previous estimates of 37 to 60 kg ha\tsups{-1} year\tsups{-1}
+\citep{Omar2014}. We conclude that the breeding program has likely
+been maintaining but not substantially improving yield potential in
+the face of changing biotic and abiotic pressures.
 
-However, the goal of a breeding program is to not only maintain the
-benchmark yield of the previous ``standard'' but to surpass it. Given
-the estimated yield decreases over time, to hit an ambitious target of
-0.5 Mg ha\tsups{-1} (approximately 5\%) increase in yield per release
-(with roughly 4 -- 5 years between releases), a new cultivar needs to
-yield an additional \Sexpr{round(yrs_coef*-4,-1)} --
-\Sexpr{round(yrs_coef*-5,-1)} kg ha\tsups{-1} in side-by-side
-comparison on top of the 0.5 Mg ha\tsups{-1} target improvement. Often
-realized side-by-side yield improvements are closer to 2 -- 3\% per
-release, in which case no absolute improvements are made beyond
-maintenance. The evidence here is consistent with the latter situation
-occurring in CA, further suggesting that the genetic yield potential
-has not increased substantially in the last 30+ years
-(Fig. \ref{fig:release}).
+However, a goal of many breeding programs is to not only maintain the
+benchmark yield of the previous ``standard'' but to surpass it,
+especially if we are to meet the target of 50\% increased food
+production by 2050 \citep{Foley2011a}. Given the estimated yield
+erosion, to hit an ambitious target of 0.5 Mg ha\tsups{-1}
+(approximately 5\%) increase in yield per release (with roughly 4 to 5
+years between releases), a new cultivar needs to yield an additional
+\Sexpr{round(yrs_coef*-4,-1)} to \Sexpr{round(yrs_coef*-5,-1)} kg
+ha\tsups{-1} in side-by-side comparison on top of the 0.5 Mg
+ha\tsups{-1} target improvement. Often realized side-by-side yield
+improvements are closer to 2 to 3\% per release, in which case no
+absolute improvements are made beyond maintenance. The evidence here
+is consistent with the latter situation occurring in CA, further
+suggesting that the genetic yield potential has not increased
+substantially in the last 30+ years (Fig. \ref{fig:w_erosion}).
 
@@
 @ 
-\caption{A conceptual diagram highlighting three scenarios:
-  (1) maintenance breeding, where losses over time are offset by
-  improvements via plant improvement, (2) no maintenance breeding,
-  where grain yield slowly erodes over time, and (3) absolute
-  improvement in grain yield, where gains in grain yield exceed the
-  losses over time.}
+\caption{A conceptual diagram highlighting three scenarios: (1)
+maintenance breeding, where losses over time are offset by
+improvements via plant improvement, (2) no maintenance breeding, where
+grain yield slowly erodes over time (as estimated from this study),
+and (3) absolute improvement in grain yield (based on ``typical''
+yield goals of 3\% increase per release), where gains in grain yield
+exceed the losses over time.}
 \label{fig:theory}
@@
 Pittelkow2012, linquist2008}, etc. may all be contributing to
-increases in the state average yield. However, these gains are less
-than those experienced during the ``Green revolution'', and by their
+increases in the state average yield. However, these gains are far less
+than those experienced during the preceeding era, and by their
 nature agronomic gains cannot surpass the genetic yield potential
@@
 these cultivars (M-201, M-204, M-205, and possibly M-209;
-Fig. \ref{fig:comp_ids})) may not be widely adapted. It is known that
+Fig. \ref{fig:ids_comp})) may not be widely adapted. It is known that
 cool temperatures during the booting stage can have a large impact on
@@
 experiences warmer temperatures than many other rice growing areas in
-CA \citep{espe2017}. These cultivars may perform well at the RES but not widely in
-CA. Therefore, it should be noted that while our investigation has
-been concerned about increasing yield potential, the estimates here
-might not capture yield stability. Yield stability is notoriously
-difficult to quantify (CITE), hence further investigation is needed to
-explore how yield stability has changed over the study period.
+CA \citep{espe2017}. These cultivars may perform well at the RES but
+not widely in CA. Therefore, it should be noted that while our
+investigation has been concerned about increasing yield potential, the
+estimates here might not capture yield stability. Yield stability is
+notoriously difficult to quantify \citep[e.g.,][]{lobell2011}, hence
+further investigation is needed to explore how yield stability has
+changed over the study period.
 
@@
 improvement, including increased quality \citep{mckenzie1993} and
-milling yield \citep{mckenzie1994} have been made over the study
+milling yield \citep{mckenzie1994}, have been made over the study
 period and are no less important. However, planning to meet future
@@
 
-RES_coef = cbind(med_eff, paste(CIs[,1],CIs[,2], sep = " -- "))
+RES_coef = cbind(med_eff, paste(CIs[,1],CIs[,2], sep = " to "))
 rownames(RES_coef) = c("Intercept (Mg ha\\tsups{-1})",
@@
 
-\begin{figure}[H]
-<<fig=TRUE, echo=FALSE>>=
-tars <- c("M202","M206")
-comp_yrs = 2004:2005
+% \begin{figure}[H]
+% <<fig=TRUE, echo=FALSE>>=
+% tars <- c("M202","M206")
+% comp_yrs = 2004:2005
 
-boxplot(yield_kg/1000 ~ id, data = mod_data[(mod_data$id %in% tars) &
-                                            (mod_data$year %in% comp_yrs),],
-        ylab = expression("Grain yield ( Mg ha"^-1~")"))
+% boxplot(yield_kg/1000 ~ id, data = mod_data[(mod_data$id %in% tars) &
+%                                             (mod_data$year %in% comp_yrs),],
+%         ylab = expression("Grain yield ( Mg ha"^-1~")"))
 
-@
-\caption{An example of a typical side-by-side comparison between a older (M-202)
-  and newer (M-206) cultivar. Shown is the performance of these cultivars at the CA Rice
-  Experiment Station (RES) over the period \Sexpr{min(comp_yrs)} --
-  \Sexpr{max(comp_yrs)}. The newer
-  cultivar, M-206, appears to have a yield advantage over the older
-  cultivar, M-202, though this comparison does not account for the 10
-  year difference in years since release between the cultivars.}
-\label{fig:compare_vars}
-\end{figure}
-\newpage
+% @
+% \caption{An example of a typical side-by-side comparison between a older (M-202)
+%   and newer (M-206) cultivar. Shown is the performance of these cultivars at the CA Rice
+%   Experiment Station (RES) over the period \Sexpr{min(comp_yrs)} --
+%   \Sexpr{max(comp_yrs)}. The newer
+%   cultivar, M-206, appears to have a yield advantage over the older
+%   cultivar, M-202, though this comparison does not account for the 10
+%   year difference in years since release between the cultivars.}
+% \label{fig:compare_vars}
+% \end{figure}
+% \newpage
 

### 2017-09-08 output/breed_paper.Rnw
Latest edits

@@
 
-\title{Rice yield improvements by plant breeding are offset by
-  decreases over time}
+\title{Rice yield improvements through plant breeding are offset by
+  inherent yield declines over time}
 \author[ucd]{Matthew B. Espe\corref{corresauthor}}
@@
 \author[res]{Kent McKenzie}
+\author[ucce2]{Michelle Leinfelder-Miles}
+\author[ucce]{Luis A. Espino}
+\author[ucce]{Randall Mutters}
 \author[ucd]{Chris van Kessel}
@@
 \address[ucd]{Dept. of Plant Sciences, University of California --
-Davis, Davis CA 95616}
+Davis, Davis CA 95616, USA}
 
 \address[res]{Rice Experiment Station, California Cooperative Rice
-Research Foundation, Biggs, CA 95917}
+Research Foundation, Biggs, CA 95917, USA}
+
+\address[ucce]{University of California Cooperative Extension,
+Oroville, CA 95965, USA}
+
+\address[ucce2]{University of California Cooperative Extension,
+Stockton, CA 95206, USA}
 
@@
 
-Meeting the challenge of feeding an growing population with
+Meeting the challenge of feeding a growing population with
 increasingly limited resources will require increasing the yield
@@
 intensive production systems have experienced slow rates of yield
-improvement in recent years despite demonstrated increased yield
-potential of new genetics over old in side-by-side comparisons. We
-analyzed experimental data from one such system, California (CA) rice
-production, in order to better understand what improvements have been
-made in the genetic yield potential via plant breeding. Specifically,
-the hypothesis that a rice variety's yield erodes over time after the
-initial selection was tested as an explanation for how side-by-side
-comparisons show increased yield while no substantive increases have
-been seen in farmers' fields. Side-by-side comparisons under yield
-erosions will show ``false'' increases due to the older cultivar
-decreasing, not due to increased genetic yield potential by the newer
-cultivar. We confirmed an apparent increase in yield in side-by-side
-comparisons between old and newer varieties when the yield erosion was
-not considered. However, we demonstrate this apparent increase in
-yield can be explained by the differences in the time since selection
-between cultivars. The impact of yield erosion was estimated to be
-29.3 kg ha\tsups{-1} year\tsups{-1} (90\% credible interal -4.4 to
--53.3) after initial selection. Once this effect was considered, the
-apparent yield advantage of newer cultivars over old was uncertain
-(-3.3 kg ha\tsups{-1} year\tsups{-1}, 90\% credible interval -36.1 to
-31.5). This underscores the importance of breeding to maintain yield
-and the challenges of increasing the genetic yield potential.
+improvement in recent years despite a demonstrated increase in the
+yield potential of new crop cultivars. We analyzed experimental data
+from one such cropping system, i.e., California (CA) rice, in order to
+better understand what improvements have been made in the genetic
+yield potential obtained through plant breeding. Specifically, the
+hypothesis was tested that the yield of a rice variety declines over
+time after the initial selection was tested. This would explain how
+side-by-side comparisons show increased yield while no substantive
+increases have been seen in farmers' fields. Side-by-side comparisons
+under yield erosion will show ``false'' yield increases due to the
+yield of the older cultivar decreasing, not due to the increased
+genetic yield potential of the newer cultivar. An apparent increase in
+yield in side-by-side comparisons between old and new rice varieties
+was observed when the yield erosion was not considered. However, this
+apparent increase in yield could be explained by the differences in
+the time when these cultivars were selected. The impact of yield
+erosion was estimated to be 29.3 kg ha\tsups{-1} year\tsups{-1} (90\%
+credible interval -4.4 to -53.3) after initial selection. Once this
+effect was considered, the apparent yield advantage of newer cultivars
+over old was uncertain (-3.3 kg ha\tsups{-1} year\tsups{-1}, 90\%
+credible interval -36.1 to 31.5). This underscores the importance of
+breeding to maintain yield and the challenges of increasing the
+genetic yield potential.
+% Add more implications
 
@@
 
-\begin{keyword} rice, yield potential, breeding
+\begin{keyword} rice, yield potential, breeding, yield erosion
 \end{keyword}
@@
 Constraints on arable land are increasing simultaneous with the need
-to increase total production to meet growing demand \citep{Foley2011a,
-Godfray2010, Mueller2012}, which has necessitated harvesting more
-grain per unit land area \citep{Lobell2009,
+to increase total food production to meet a growing demand
+\citep{Foley2011a, Godfray2010, Mueller2012}, which has necessitated
+harvesting more grain per unit land area \citep{Lobell2009,
 Tittonell2014}. Historically, agricultural research has been
 successful in staving off the ``Malthusian catastrophe'' of demand
-surpassing supply via continued yield improvement of staple crops,
-however more and more production systems are experiencing plateaus in
-grain yield \citep{Grassini2013, Peng2010}. If continued improvements
-per unit land area are no longer possible, increasing the area under
-cultivation will be needed to meet demand, which carries with it
-undesirable ecological implications. Therefore, it is of critical
-importance to better understand why grain yields have slowed or
-plateaued in intensified production systems.
+surpassing supply via continued yield improvement of staple
+crops. However production systems are experiencing plateaus in grain
+yield \citep{Grassini2013, Peng2010}. If this trend continues,
+improvements per unit land area are no longer possible, and an
+increase in the area under cultivation will be needed to meet food
+demand, which carries with it undesirable ecological
+implications. Therefore, it is of critical importance to better
+understand why the rate of increase in grain yields has declined or
+leveled off in intensified production systems.
 
@@
 research). However, since 1990 CA state average yields have increased
-more slowly (Fig. \ref{fig:state_yield}) despite continuous plant
-improvement \citep{ccrrf}. For the last 30 years, each newly developed
+more slowly (Fig. \ref{fig:state_yield}) despite continuous crop improvement \citep{ccrrf}. For the past 30 years, each newly developed
 cultivar has demonstrated a yield advantage over the previous
@@
 variety trial program that compares experimental entries side-by-side
-with ``check'' cultivars). Yet, taken across years, these same reports
-show stagnant yields. For example, at the year of release (1985),
-cultivar M-202 averaged 11 Mg ha\tsups{-1} \citep[14\%
-MC;][]{m202}. In 2015, the reported average yield of the newest released
-cultivar M-209 was 10.8 Mg ha\tsups{-1} \citep{m209}. Yet this same
-report, M-209 demonstrated a X\% yield advantage over M-202. In other rice
-systems, a decrease in a cultivar's yield performance over time has
-been observed \citep{Peng1998,
-Peng2010, Peng2000, Datta1995}, often termed yield erosion
-\citep{Peng1998}. It has been speculated that this is due to the
-inability of a selected cultivar's genetics to adapt to changing  biotic and
-abiotic conditions. We sought to quantify this phenomenon to better
-understand how plant improvement and yield erosion are contributing to
-the slow yield improvement in farmers' fields over the last
-30 years in CA. Specifically, two competing hypotheses were
-tested: (1) the genetic yield advantage has increased over the last
-30+ years, and (2) the genetic yield advantage has not increased over
-the last 30+ and the erosion of yield over
-time is creating the illusion of improved yields (i.e., the new
-cultivar is recovering the old benchmark as the older cultivar's yield
-erodes).
+with ``check'' cultivars). However, in a seeming contradiction these same
+reports show stagnant yields across release announcements (i.e.,
+reports accompanying the office release) from the
+various years. For example, at the year of release
+(1985), cultivar M-202 was reported to yield an average of 11 Mg
+ha\tsups{-1} \citep[14\% moisture content (MC);][]{m202}. In 2015, the reported average
+yield of the newest released cultivar M-209 was 10.8 Mg ha\tsups{-1}
+\citep]14\% MC;][]{m209}. In other rice systems, a decrease in a
+cultivar's yield performance over time has been observed
+\citep{Peng1998, Peng2010, Peng2000, Datta1995}, often termed yield
+erosion \citep{Peng1998}. It has been speculated that this yield erosion is due to
+the inability of a selected cultivar's genetics to adapt to changing
+biotic and abiotic conditions. To better understand how
+plant improvement and yield erosion are contributing to the
+improvement of rice yield in farmers' fields, we teseted two competing hypotheses: (1) the
+genetic yield advantage increased over the last 30+ years, and (2)
+the genetic yield advantage has not increased over the last 30+
+because the erosion of yield over time creates an illusion of improved
+yields.
 
@@
 largely free of precipitation events. Soils at the RES are classified
-as Esquon-Neerdobe clays with roughly 2\% SOM \citep{ssurgo} in the
-top 15cm. As the central location for the CCRRF breeding program, the RES is
-researcher-managed (as opposed to farmer-managed trials) and conducts
-the bulk of the trials for the Statewide Variety Trials (approximately
-3 to 4 times the number of plots compared to any other location). For
-these reasons, we focused our investigation on the RES to avoid
-potential complications due to site, climate, and management
-differences at the other 6 to 8 farmer-managed trials per year.
+as Esquon-Neerdobe clays with roughly 2\% soil organic matter
+\citep{ssurgo} in the top 15cm. As the central location for the CCRRF
+breeding program, the RES is researcher-managed (as opposed to
+farmer-managed trials) and conducts the majority of the trials for the
+Statewide Variety Trials (approximately 3 to 4 times the number of
+experimental plots compared to any other location). To avoid potential
+complications due to site, climate, and management differences at the
+other 6 to 8 farmer-managed trials per year we focused our
+investigation on the RES.
 
 As part of the annual Statewide Variety Trials, newly developed
-entries are tested against named and released cultivars (checks) at
+entries are tested against officially named and released cultivars (checks) at
 the RES in several trials spanning multiple planting dates per
 year. For the purposes of this study, we concentrated on released
-medium grain cultivars, as medium grain cultivars constitute roughly
-90\% of the planted area in CA. These cultivars are typcically planted
+medium grain cultivars, as medium grain cultivars constitute approximately
+90\% of the planted rice area in CA. These cultivars are typcically planted
 in the Statewide Variety Trials for 2 to 4 years prior to their
@@
 collected leading to a final data set of
-\Sexpr{format(nrow(mod_data), big.mark = ',')} observations with
-\Sexpr{length(unique(mod_data[,'id']))} cultivars represented,
+\Sexpr{format(nrow(mod_data), big.mark = ',')} observations representing
+\Sexpr{length(unique(mod_data[,'id']))} cultivars,
 including every publicly released medium grain variety developed by
@@
 \Sexpr{max(mod_data[,'release_yr'])}. For each cultivar, the official
-year of release (typically 3 -- 5 years after introduction to the
-trials but prior to widespread commercial seed availability) and the
+year of release, typically 3 -- 5 years after introduction to the
+trials but prior to widespread commercial seed availability and the
 time elapsed (in years) since release was calculated.
 
-Plots at the RES are managed similar to University of California
-Cooperative Extension prescribed best management
+Experimental plots at the RES are managed similar to University of California
+Cooperative Extension following prescribed best management
 practice, including land preparation, fertility management, and pest
@@
 
-Yield was modeled as a response to the number of years since the
-variety was officially released and the year of release using a
-mixed-effects linear regression. To account for similarities between years
-and between cultivars, year and cultivar were included in the model as
-random effects.
+To test each hypothesis, the influence of breeding and yield erosion
+were quantified on yield over time. Yield was modeled as a response to
+the number of years since the variety was officially released and the
+year of release using a mixed-effects linear regression. To account
+for similarities between years and between cultivars, year and
+cultivar were included in the model as random effects.
 
@@
 comparison to estimates where yield erosion is not taken into account,
-the model was as above except omitting the $Rate_{yield\ erosion\ i}$ term.
+the model was used as described above except omitting the $Rate_{yield\ erosion\ i}$ term.
 
@@
 
-Data was processed and models were fit in R, an environment for
+Data were processed and models were fit in R, an environment for
 statistical computing \citep{R}. Models were fit using the `rstanarm'
@@
 % for yield erosion.
+
+\begin{figure}[H]
+<<fig=TRUE, width = 6, height = 5, out.width = '85%', echo=FALSE>>=
+
+idx = mod_data[,"yrs_in_trial"] %in% -2:0
+
+plot_data = prep_plot_data(mod_data[idx,"yield_kg"] / 1000 ,
+                           mod_data[idx, "id"])
+
+plot_data[,"yr"] = tapply(mod_data[idx, "release_yr"],
+                          mod_data[idx,"id"], unique)
+
+plot(mean ~ yr, data = plot_data,
+     pch = 16,
+     xlab = "Year of release",
+     ylab = expression("Yield ( Mg ha"^-1~")"),
+     ylim = c(9,13),
+     xlim = c(1984, 2018.5))
+
+segments(x0 = plot_data[,"yr"],
+         y0 = plot_data[,"mean"] + plot_data[,"se"],
+         y1 = plot_data[,"mean"] - plot_data[,"se"])
+
+#Used locator() it interactively find label points 
+aa = readRDS("f2_text_pos.rds")
+text(x = aa$x, y = aa$y,
+     labels = gsub("M", "M-", row.names(plot_data)),
+     pos = 4, cex = 0.75)
+#### Polygons
+yrs <- 1975:2025
+
+pp <- posterior_linpred(RES_full, re.form = NA,
+                        newdata = data.frame(release_yr_c = yrs - 2000,
+                                             yrs_in_trial = 0),
+                        prob = 0.5)
+pp <-  (pp * scl + center) / 1000
+ints2 = apply(pp, 2, quantile, p = c(0.05, 0.95))
+
+abline(a = ((((fixef(RES_full)[1] * scl) + center) ) -
+           (2000 * release_yr_coef)) / 1000,
+       b = release_yr_coef/1000,
+       lty = 2)
+polygon(x = c(yrs, rev(yrs)),
+        y = c(ints2[1,],rev(ints2[2,])),
+        border = NA,
+        col = rgb(0,0,0,0.1))
+
+# text(x = plot_data$yr, y = plot_data$mean,
+#      labels = 1:nrow(plot_data), #gsub("M", "M-", row.names(plot_data)),
+#      pos = 4, cex = 0.75)
+# mm = lm(yield_kg/1000 ~ release_yr, data = mod_data[idx,])
+# abline(mm, lty = 2)
+# yrs = 1980:2020
+# int = predict(mm, newdata = data.frame(release_yr = yrs), se.fit = TRUE)
+
+# polygon(x = c(yrs, rev(yrs)),
+#         y = c(int$fit + int$se.fit,rev(int$fit - int$se.fit)),
+#         border = NA,
+#         col = rgb(0,0,0,0.1))
+
+@
+
+\caption{The yield of released medium grain rice varieties for only
+the two years prior to each cultivar's official release. Points
+($\bullet$) and bars (|) represent the mean and standard error of the
+raw data, respectively. The dashed line and shaded area are the
+genetic yield improvement and credible interval as estimated by the model.}
+\label{fig:first_yrs}
+\end{figure}
+
+\subsection{Genetic yield improvements}
+
+Comparing the cultivars' yield during each's initial time in the
+experimental trials confirmed that genetic yield improvements have
+been consistant over the study period
+(Fig. \ref{fig:first_yrs}). As estimated by the full model, there is
+no conclusive evidence that the yields of medium grain rice cultivars
+selected and released by CCRRF's breeding program have, on average,
+increased over the last 30 years (Table \ref{tbl:coef}). The full
+model which corrected for both the year of release and yield erosion
+was estimated to have better out-of-sample predictive accuracy
+compared to reduced models containing only one predictor. The
+estimated change in grain yield per year in the CCRRF's breeding
+trials at the RES was estimated to be a negligible
+\Sexpr{release_yr_coef} kg ha\tsups{-1} year\tsups{-1}, with both
+small positive and negative trends within the credible interval (90\%
+CI: \Sexpr{rel_CI[1]} to \Sexpr{rel_CI[2]})
+(Fig. \ref{fig:first_yrs}).
+
+When yield erosion was not corrected for in the comparison
+(i.e., omitted from the model), there is an apparent small increase
+per year (approximately \Sexpr{only_release_coef} kg ha\tsups{-1}
+year\tsups{-1}; 90\% CI \Sexpr{only_release_CI[1]} to
+\Sexpr{only_release_CI[2]}) (Fig. \ref{fig:wo_erosion}). Although this
+estimate ignores the confounding effect of yield erosion over time and
+performed worse in out-of-sample predictive accuracy, it confirms that
+yields in side-by-side comparison would support the conclusion that newer
+cultivars out-preform older ones if yield erosion is not considered.
+
+After correcting for the other variables, some cultivars were estimated
+to have higher yield compared to the overall average (as estimated by
+random intercepts). The marginal estimate for the performance of a
+cultivar averaged over the uncertainty in the other parameters in the
+model (i.e., the random effect) shows three cultivars with higher than
+average yields:
+\Sexpr{paste(gsub('M','M-',rev(row.names(id_est))[c(1,2,4)]),collapse=',')}
+(Figs. \ref{fig:first_yrs} \& \ref{fig:ids_comp}). A fourth, M-209,
+might also have higher than average yield, though the CI for M-209
+overlaps with zero (Fig. \ref{fig:ids_comp}).
+
+\begin{figure}[H]
+
+<<fig=TRUE, echo=FALSE, width = 6, height = 5>>=
+yrs <- 1975:2025
+
+pp <- posterior_linpred(RES_rel_fit, re.form = NA,
+                        newdata = data.frame(release_yr_c = yrs - 2000,
+                                             yrs_in_trial = 0),
+                        prob = 0.5)
+pp <-  (pp * scl + center) / 1000
+ints1 = apply(pp, 2, quantile, p = c(0.05, 0.95))
+
+res = mod_data$site == "RES"
+comp_yrs = 2011:2016
+idx = mod_data$year %in% comp_yrs
+
+plot_data = prep_plot_data(mod_data$yield_kg[idx] / 1000 ,
+                           mod_data$id[idx])
+
+plot_data$tmp = tapply(mod_data$release_yr[idx],
+                       mod_data$id[idx], unique)
+
+colnames(plot_data) = c("yield_kg", "yield_se", "release_yr")
+
+# get corrected yield
+wo_erosion = expand.grid(id = row.names(plot_data),
+            year = comp_yrs,
+            release_yr_c = 0,
+            yrs_in_trial = 0)
+wo_erosion$corr_yield = colMeans(posterior_predict(RES_full,
+                               newdata = wo_erosion)) + center/1000
+plot_data2 = prep_plot_data(wo_erosion$corr_yield,
+                            wo_erosion$id)
+plot_data2$release_yr = plot_data$release_yr
+
+# par(mfrow = c(1,2),
+#     mar = c(4,0,2,0.75),
+#     oma = c(4,4,0.5,0.5))
+
+plot(yield_kg ~ release_yr, data = plot_data, pch = 16, #col = rgb(0,0,0,0.25),
+     xlab = "Release year",
+     ylab = "", 
+     ylim = c(8, 12.5),
+     xaxt = "n",
+     xlim = c(1980,2020))
+axis(1, at = seq(1980, 2020, by = 10))
+mtext(text = expression("Grain yield ( Mg ha"^-1~")"), side = 2, line = 2.5)
+abline(a = ((fixef(RES_rel_fit)[1] * scl) + center) / 1000 -
+      (2000 * fixef(RES_rel_fit)[2]),
+       b = fixef(RES_rel_fit)[2],
+       lty = 2)
+polygon(x = c(yrs, rev(yrs)),
+        y = c(ints1[1,],rev(ints1[2,])),
+        border = NA,
+        col = rgb(0,0,0,0.1))
+
+with(plot_data, segments(x0 = release_yr, y0=yield_kg + yield_se,
+                         y1 = yield_kg - yield_se))
+text(x = plot_data$release_yr-0.4, y = plot_data$yield_kg + 0.07,
+     labels = gsub("M", "M-", row.names(plot_data)),
+     pos = 4, cex = 0.75)
+text(x = 2000, y = 8, bquote(.(round(fixef(RES_rel_fit)[2] * scl, 1))~kg~ha^-1~yr^-1))
+mtext("Without adjusting for yield erosion", adj = 0)
+
+@
+\caption{The relationship between yield and the release year of medium
+grain rice cultivars for the last 6 years
+(\Sexpr{min(comp_yrs)} -- \Sexpr{max(comp_yrs)}) as estimated without
+correcting for yield erosion over time.}
+\label{fig:wo_erosion}
+\end{figure}
+
+\begin{comment}
+\begin{figure}[H]
+<<fig=TRUE, echo=FALSE, width = 6, height = 5>>=
+
+plot(mean ~ release_yr, data = plot_data2, pch = 16, #col = rgb(0,0,0,0.25),
+     xlab = "Release year",
+     ylab = "", #expression("Grain yield -corrected( Mg ha"^-1~")"),
+     yaxt = "n",
+     ylim = c(8, 12.5),
+     xaxt = "n",
+     xlim = c(1980,2020))
+axis(1, at = seq(1980, 2020, by = 10))
+abline(a = ((((fixef(RES_full)[1] * scl) + center) ) -
+           (2000 * release_yr_coef)) / 1000,
+       b = release_yr_coef/1000,
+       lty = 2)
+with(plot_data2, segments(x0 = release_yr, y0 = mean + se,
+                         y1 = mean - se))
+text(x = plot_data2$release_yr-0.4, y = plot_data2$mean + 0.07,
+     labels = gsub("M", "M-", row.names(plot_data2)),
+     pos = 4, cex = 0.75)
+text(x = 2000, y = 8, bquote(.(release_yr_coef)~kg~ha^-1~yr^-1))
+mtext("Adjusting for yield erosion", adj = 0)
+polygon(x = c(yrs, rev(yrs)),
+        y = c(ints2[1,],rev(ints2[2,])),
+        border = NA,
+        col = rgb(0,0,0,0.1))
+
+@
+% Change from a./b. to accounting for erosion, More details about how
+% yield was corrected - 
+\caption{The relationship between yield and the release year of medium
+grain rice cultivars for the last 6 years
+(\Sexpr{min(comp_yrs)} -- \Sexpr{max(comp_yrs)}) as estimated by 
+correcting for yield erosion over time. Yield erosion was corrected
+for by adjusting each observed yield by estimated amount lost 
+  (\Sexpr{yrs_coef} kg ha\tsups{-1} year\tsups{-1} after release).}
+\label{fig:w_erosion}
+\end{figure}
+\end{comment}
+
+\subsection{Yield erosion}
+\begin{figure}[H]
+
+<<fig=TRUE, echo=FALSE>>=
+yrs_in <- -10:40
+pp <- posterior_linpred(RES_full, re.form = NA,
+                        newdata = data.frame(yrs_in_trial = yrs_in,
+                                             release_yr_c = 0))
+pp <-  (pp * scl + center) / 1000 
+ints = apply(pp, 2, quantile, p = c(0.05, 0.95))
+
+plot_data = prep_plot_data(mod_data$yield_kg / 1000,
+                           mod_data$yrs_in_trial)
+
+plot_data$tmp = as.numeric(row.names(plot_data))
+colnames(plot_data) = c("yield_kg", "yield_se", "yrs_in_trial")
+
+plot(yield_kg ~ yrs_in_trial, data = plot_data, pch = 16, #col = rgb(0,0,0,0.25),
+     xlab = "Year since release",
+     ylab = expression("Grain yield ( kg ha"^-1~")"),
+     ylim = c(8.0, 12.5))
+with(plot_data, segments(x0 = yrs_in_trial, y0=yield_kg + yield_se,
+                         y1 = yield_kg - yield_se))
+abline(a = ((fixef(RES_full)[1] * scl) + center) / 1000 ,
+       b = yrs_coef/1000)
+polygon(x = c(yrs_in, rev(yrs_in)),
+        y = c(ints[1,],rev(ints[2,])),
+        border = NA,
+        col = rgb(0,0,0,0.1))
+text(5, 8.0, bquote(.(yrs_coef)~kg~ha^-1~year^-1))
+
+@
+\caption{The relationship between yield and the years since release
+  for medium grain rice varieties. The estimated erosion of yield over time
+  (averaged over differences in cultivar and year) is estimated to be
+  \Sexpr{yrs_coef} kg ha\tsups{-1} year\tsups{-1}. Points ($\bullet$)
+  and bars (|) represent the mean and standard error of the raw data,
+  respectively. The shaded area is the 90\% credible interval.}
+% For each year, point size by n (open circles)
+\label{fig:RES_year}
+\end{figure}
 
@@
 @
-\caption{The performance of two major cultivars at the CA Rice
-Experiment Station (RES) over the period
+\caption{The performance of two major cultivars over the period
 \Sexpr{min(mod_data[,'year'])} -- \Sexpr{max(mod_data[,'year'])}. The
@@
 
-\subsection{Genetic yield improvements}
-
-The full model which corrected for both the year of release and yield
-erosion was estimated to have better out-of-sample predictive accuracy
-compared to reduced models containing only one predictor. As estimated
-by the best model, there is no conclusive evidence that the yields of
-medium grain rice cultivars selected and released by CCRRF's breeding
-program have, on average, increased over the last 30 years. The
-estimated change in grain yield per year in the CCRRF's breeding
-trials at the RES was estimated to be a negligible
-\Sexpr{release_yr_coef} kg ha\tsups{-1} year\tsups{-1}, with both
-small positive and negative trends within the credible interval (90\%
-CI: \Sexpr{rel_CI[1]} to \Sexpr{rel_CI[2]})
-(Fig. \ref{fig:w_erosion}). Hence, the evidence here is consistent
-with little to no increase in yield potential over the study
-period. When yield erosion is not corrected for in the comparison
-(i.e., omitted from the model), there is an apparent small increase
-per year (approximately \Sexpr{only_release_coef} kg ha\tsups{-1}
-year\tsups{-1}; 90\% CI \Sexpr{only_release_CI[1]} to
-\Sexpr{only_release_CI[2]}) (Fig. \ref{fig:wo_erosion}), though this
-estimate ignores the confounding effect of yield erosion over time and
-performed worse in out-of-sample predictive accuracy.
-
-After correcting for the other variables, some cultivars are estimated
-to have greater yield compared to the overall average. The marginal
-estimate for the performance of a cultivar averaged over the
-uncertainty in the other parameters in the model (i.e., the random
-effect) shows three cultivars with higher than average yields:
-\Sexpr{paste(gsub('M','M-',rev(row.names(id_est))[c(1,2,4)]),collapse=',')}
-(Figs. \ref{fig:w_erosion} \& \ref{fig:ids_comp}). A fourth, M-209,
-might also have higher than average yield, though the CI for M-209
-overlaps with zero (Fig. \ref{fig:ids_comp}).
-
-\begin{comment}
-\begin{figure}[H]
-
-<<fig=TRUE, echo=FALSE, width = 6, height = 5>>=
-yrs <- 1975:2025
-
-pp <- posterior_linpred(RES_rel_fit, re.form = NA,
-                        newdata = data.frame(release_yr_c = yrs - 2000,
-                                             yrs_in_trial = 0),
-                        prob = 0.5)
-pp <-  (pp * scl + center) / 1000
-ints1 = apply(pp, 2, quantile, p = c(0.05, 0.95))
-
-pp <- posterior_linpred(RES_full, re.form = NA,
-                        newdata = data.frame(release_yr_c = yrs - 2000,
-                                             yrs_in_trial = 0),
-                        prob = 0.5)
-pp <-  (pp * scl + center) / 1000
-ints2 = apply(pp, 2, quantile, p = c(0.05, 0.95))
-
-res = mod_data$site == "RES"
-comp_yrs = 2011:2016
-idx = mod_data$year %in% comp_yrs
-
-plot_data = prep_plot_data(mod_data$yield_kg[idx] / 1000 ,
-                           mod_data$id[idx])
-
-plot_data$tmp = tapply(mod_data$release_yr[idx],
-                       mod_data$id[idx], unique)
-
-colnames(plot_data) = c("yield_kg", "yield_se", "release_yr")
-
-# get corrected yield
-wo_erosion = expand.grid(id = row.names(plot_data),
-            year = comp_yrs,
-            release_yr_c = 0,
-            yrs_in_trial = 0)
-wo_erosion$corr_yield = colMeans(posterior_predict(RES_full,
-                               newdata = wo_erosion)) + center/1000
-plot_data2 = prep_plot_data(wo_erosion$corr_yield,
-                            wo_erosion$id)
-plot_data2$release_yr = plot_data$release_yr
-
-# par(mfrow = c(1,2),
-#     mar = c(4,0,2,0.75),
-#     oma = c(4,4,0.5,0.5))
-
-plot(yield_kg ~ release_yr, data = plot_data, pch = 16, #col = rgb(0,0,0,0.25),
-     xlab = "Release year",
-     ylab = "", 
-     ylim = c(8, 12.5),
-     xaxt = "n",
-     xlim = c(1980,2020))
-axis(1, at = seq(1980, 2020, by = 10))
-mtext(text = expression("Grain yield ( Mg ha"^-1~")"), side = 2, line = 2.5)
-abline(a = ((fixef(RES_rel_fit)[1] * scl) + center) / 1000 -
-      (2000 * fixef(RES_rel_fit)[2]),
-       b = fixef(RES_rel_fit)[2],
-       lty = 2)
-polygon(x = c(yrs, rev(yrs)),
-        y = c(ints1[1,],rev(ints1[2,])),
-        border = NA,
-        col = rgb(0,0,0,0.1))
-
-with(plot_data, segments(x0 = release_yr, y0=yield_kg + yield_se,
-                         y1 = yield_kg - yield_se))
-text(x = plot_data$release_yr-0.4, y = plot_data$yield_kg + 0.07,
-     labels = gsub("M", "M-", row.names(plot_data)),
-     pos = 4, cex = 0.75)
-text(x = 2000, y = 8, bquote(.(round(fixef(RES_rel_fit)[2] * scl, 1))~kg~ha^-1~yr^-1))
-mtext("Without adjusting for yield erosion", adj = 0)
-
-@
-\caption{The relationship between yield and the release year of medium
-grain rice cultivars developed by the California Cooperative Rice
-Research Foundation (California, USA), for the last 6 years
-(\Sexpr{min(comp_yrs)} -- \Sexpr{max(comp_yrs)}) as estimated without
-correcting for yield erosion over time.}
-\label{fig:wo_erosion}
-\end{figure}
-\end{comment}
-
-\begin{figure}[H]
-<<fig=TRUE, echo=FALSE, width = 6, height = 5>>=
-
-plot(mean ~ release_yr, data = plot_data2, pch = 16, #col = rgb(0,0,0,0.25),
-     xlab = "Release year",
-     ylab = "", #expression("Grain yield -corrected( Mg ha"^-1~")"),
-     yaxt = "n",
-     ylim = c(8, 12.5),
-     xaxt = "n",
-     xlim = c(1980,2020))
-axis(1, at = seq(1980, 2020, by = 10))
-abline(a = ((((fixef(RES_full)[1] * scl) + center) ) -
-           (2000 * release_yr_coef)) / 1000,
-       b = release_yr_coef/1000,
-       lty = 2)
-with(plot_data2, segments(x0 = release_yr, y0 = mean + se,
-                         y1 = mean - se))
-text(x = plot_data2$release_yr-0.4, y = plot_data2$mean + 0.07,
-     labels = gsub("M", "M-", row.names(plot_data2)),
-     pos = 4, cex = 0.75)
-text(x = 2000, y = 8, bquote(.(release_yr_coef)~kg~ha^-1~yr^-1))
-mtext("Adjusting for yield erosion", adj = 0)
-polygon(x = c(yrs, rev(yrs)),
-        y = c(ints2[1,],rev(ints2[2,])),
-        border = NA,
-        col = rgb(0,0,0,0.1))
-
-@
-% Change from a./b. to accounting for erosion, More details about how
-% yield was corrected - 
-\caption{The relationship between yield and the release year of medium
-grain rice cultivars developed by the California Cooperative Rice
-Research Foundation (California, USA), for the last 6 years
-(\Sexpr{min(comp_yrs)} -- \Sexpr{max(comp_yrs)}) as estimated by 
-correcting for yield erosion over time. Yield erosion was corrected
-for by adjusting each observed yield by estimated amount lost 
-  (\Sexpr{yrs_coef} kg ha\tsups{-1} year\tsups{-1} after release).}
-\label{fig:w_erosion}
-\end{figure}
-
-\subsection{Yield erosion}
-\begin{figure}[H]
-
-<<fig=TRUE, echo=FALSE>>=
-yrs_in <- -10:40
-pp <- posterior_linpred(RES_full, re.form = NA,
-                        newdata = data.frame(yrs_in_trial = yrs_in,
-                                             release_yr_c = 0))
-pp <-  (pp * scl + center) / 1000 
-ints = apply(pp, 2, quantile, p = c(0.05, 0.95))
-
-plot_data = prep_plot_data(mod_data$yield_kg / 1000,
-                           mod_data$yrs_in_trial)
-
-plot_data$tmp = as.numeric(row.names(plot_data))
-colnames(plot_data) = c("yield_kg", "yield_se", "yrs_in_trial")
-
-plot(yield_kg ~ yrs_in_trial, data = plot_data, pch = 16, #col = rgb(0,0,0,0.25),
-     xlab = "Year since release",
-     ylab = expression("Grain yield ( kg ha"^-1~")"),
-     ylim = c(8.0, 12.5))
-with(plot_data, segments(x0 = yrs_in_trial, y0=yield_kg + yield_se,
-                         y1 = yield_kg - yield_se))
-abline(a = ((fixef(RES_full)[1] * scl) + center) / 1000 ,
-       b = yrs_coef/1000)
-polygon(x = c(yrs_in, rev(yrs_in)),
-        y = c(ints[1,],rev(ints[2,])),
-        border = NA,
-        col = rgb(0,0,0,0.1))
-text(5, 8.0, bquote(.(yrs_coef)~kg~ha^-1~year^-1))
-
-@
-\caption{The relationship between yield and the years since release
-  for medium grain rice varieties planted at the Rice Experiment Station
-  (RES), California, USA. The estimated erosion of yield over time
-  (averaged over differences in cultivar and year) is estimated to be
-  \Sexpr{yrs_coef} kg ha\tsups{-1} year\tsups{-1}. Points ($\bullet$)
-  and bars (|) represent the mean and standard error of the raw data,
-  respectively. The shaded area is the 90\% credible interval.}
-% For each year, point size by n (open circles)
-\label{fig:RES_year}
-\end{figure}
-
-There is estimated to be an decrease of yield over time once a
-cultivar is selected (i.e., yield erosion) (\Sexpr{yrs_coef} kg
-ha\tsups{-1} year\tsups{-1}, 90\% CI: \Sexpr{RES_CI[2]} to
-\Sexpr{RES_CI[1]}). The oldest cultivar in the trials is estimated to
-have lost an average of \Sexpr{round(yrs_coef*-36/1000,1)} Mg
-ha\tsups{-1} since release. Comparing two dominate medium grain rice
-cultivars M-202 (widely planted from the late 1980s until the early
-2000s), and M-206 (the most widely planted CCRRFS cultivar at
-approximately 50\% of planted area at the time of writing), the two
-cultivars experienced similar trends over time to each other
-(Fig. \ref{fig:compare_yrs}) and to the estimate from all released
-medium grains (Table \ref{tbl:coef}). At the time of writing, an
-estimated \Sexpr{m202_v_206} kg ha\tsups{-1} yield advantage of M-206
-over M-202 can be attributed solely to the erosion of M-202's yield
-over time (Figs \ref{fig:w_erosion} \& \ref{fig:compare_yrs}). Further
-supporting this estimate, both cultivars yielded similarly at their
+The average decrease in yield over time once a cultivar is selected
+(i.e., yield erosion) was estimated as \Sexpr{yrs_coef} kg
+ha\tsups{-1} year\tsups{-1} (90\% CI: \Sexpr{RES_CI[2]} to
+\Sexpr{RES_CI[1]}). The oldest cultivar in the trials had lost an
+average of \Sexpr{round(yrs_coef*-36/1000,1)} Mg ha\tsups{-1} since
+its release. Comparing the two dominate medium grain rice cultivars M-202
+(widely planted from the late 1980s until the early 2000s), and M-206
+(the most widely planted CCRRFS cultivar at approximately 50\% of
+planted area at the time of writing), these two cultivars experienced
+similar trends in yield decline (Fig. \ref{fig:compare_yrs}; Table
+\ref{tbl:coef}). The currently estimated
+\Sexpr{m202_v_206} kg ha\tsups{-1} yield advantage of M-206 over M-202
+can be attributed solely to the erosion of M-202's yield over time
+(Figs \ref{fig:first_yrs} \& \ref{fig:compare_yrs}). Further
+supporting this estimate of yield erosion, both cultivars yielded similarly at their
 respective year of release (Fig. \ref{fig:compare_yrs}), despite the
@@
 
-We estimate that yield erosion has decreased the yield of medium
-grain rice cultivars released by the CCRRF over the last 30+ years by
-an average of \Sexpr{yrs_coef} Mg ha\tsups{-1},
-potentially offsetting any gains made by the breeding program. Biotic
-and abiotic forces have previously been speculated to be responsible
-for yield erosion, since the cultivar's genetics remain stable from the
-time of selection but the conditions the cultivar may experience over
-time are continuously changing \citep{Peng1998, Peng2000,
-Peng2010}. The trend observed here is consistent with previous reports
-of the erosion of genetic gains of cultivar IR8 at IRRI
-\citep{Peng1998, Peng2000}, suggesting the erosion of yield over time
-is a phenomenon that widely affects rice breeding programs. This
-highlights both the need for maintenance breeding and the difficulty
-of increasing genetic yield potential, especially in systems such as
-CA where yield is approaching the physiological limit
-\citep{espe16a}. Without maintenance breeding continuously selecting
-cultivars well adapted to current conditions, CA rice production
-systems are predicted to have lost an estimated
+We do not find conclusive evidence of increased genetic yield
+potential in medium grain rice cultivars released by CCRRF's breeding
+improvement program over the period \Sexpr{min(mod_data[,'year'])} to
+\Sexpr{max(mod_data[,'year'])}. Since the estimated yield improvement
+spans from \Sexpr{rel_CI[1]} to \Sexpr{rel_CI[2]} kg ha\tsups{-1}
+year\tsups{-1}, both increases and decreases in the genetic yield
+potential over the study period are plausible. However, even the
+estimated yield improvement that does not account for yield erosion
+(\Sexpr{only_release_coef} kg ha\tsups{-1} year\tsups{-1};
+Fig. \ref{fig:wo_erosion}) is much lower than previous estimated,
+i.e., 37 to 60 kg ha\tsups{-1} year\tsups{-1} \citep{Omar2014}. Based on the
+estimates from this study, there is less than a 1\% probability that
+the true value is higher than 45 kg ha\tsups{-1}
+year\tsups{-1}. Therefore, we conclude that the CCRRF breeding program
+has likely been maintaining but not substantially improved the yield
+potential in the face of changing biotic and abiotic pressures.
+
+We estimate that yield erosion has decreased the yield of medium grain
+rice cultivars released by the CCRRF over the last 30+ years by an
+average of \Sexpr{yrs_coef} Mg ha\tsups{-1}, potentially offsetting
+any gains made by the crop improvement program. Biotic and abiotic forces have
+previously been speculated to be responsible for yield erosion, since
+the cultivar's genetics remain stable from the time of selection but
+the conditions the cultivar may experience over time are continuously
+changing \citep{Peng1998, Peng2000, Peng2010}. The trend observed here
+is consistent with previous reports of the erosion of genetic gains of
+cultivar IR8 at IRRI \citep{Peng1998, Peng2000}, suggesting the
+erosion of yield over time is a phenomenon that widely affects rice
+breeding programs. This observation highlights both the need for maintenance
+of plant breeding and the difficulty of increasing genetic yield potential,
+especially in systems such as CA where yield is approaching the
+physiological limit \citep{espe16a}. Without maintenance a breeding
+program and continuously selecting cultivars well adapted to current conditions,
+CA rice production systems are predicted to have lost an estimated
 \Sexpr{round(yrs_coef*(-36/1000),1)} Mg ha\tsups{-1} grain yield over
 the study period.
-
-Overall, we do not find conclusive evidence of increased genetic yield
-potential in medium grain rice cultivars released by CCRRF's breeding
-improvement program over the period \Sexpr{min(mod_data[,'year'])} to
-\Sexpr{max(mod_data[,'year'])}. After correcting yield comparisons for
-yield erosion over time, the estimated yield improvement spans from
-\Sexpr{rel_CI[1]} to \Sexpr{rel_CI[2]} kg ha\tsups{-1}
-year\tsups{-1}. This suggests that both increases and decreases in the
-genetic yield potential over the study period. However, even when the
-comparison is not corrected for yield erosion, the estimated
-improvements over time (\Sexpr{only_release_coef} kg ha\tsups{-1}
-year\tsups{-1}; Fig. \ref{fig:wo_erosion}) are still much less than
-previous estimates of 37 to 60 kg ha\tsups{-1} year\tsups{-1}
-\citep{Omar2014}. We conclude that the breeding program has likely
-been maintaining but not substantially improving yield potential in
-the face of changing biotic and abiotic pressures.
 
@@
 \Sexpr{round(yrs_coef*-4,-1)} to \Sexpr{round(yrs_coef*-5,-1)} kg
-ha\tsups{-1} in side-by-side comparison on top of the 0.5 Mg
-ha\tsups{-1} target improvement. Often realized side-by-side yield
-improvements are closer to 2 to 3\% per release, in which case no
-absolute improvements are made beyond maintenance. The evidence here
-is consistent with the latter situation occurring in CA, further
-suggesting that the genetic yield potential has not increased
-substantially in the last 30+ years (Fig. \ref{fig:w_erosion}).
+ha\tsups{-1} in side-by-side comparison in addition to the 0.5 Mg
+ha\tsups{-1} target improvement. Often side-by-side yield improvements
+show an improvement in grain yield of 2 to 3\% per release, in which
+case no absolute improvements in yield have been made beyond
+maintaining yield. The evidence here is consistent with the
+observation that the genetic yield potential has not increased
+substantially in the last 30+ years (Fig. \ref{fig:first_yrs}).
 
@@
 California state average yields have increased slowly
-even as little to no gains have been seen in the breeding trials
+even as little to no gains have been obtained in the breeding trials
 (Fig. \ref{fig:state_yield}). The evidence here suggests mechanisms
@@
 Lundy2015, linquist2009}, weed management \citep{Saito2008,
-Pittelkow2012, linquist2008}, etc. may all be contributing to
+Pittelkow2012, linquist2008}, among other management factors. may all be contributing to
 increases in the state average yield. However, these gains are far less
@@
 
-There are several cultivars that were estimated to have better than
+There are several cultivars that were found to have better than
 average yields, even after correcting for yield erosion. However,
@@
 cool temperatures during the booting stage can have a large impact on
-grain yields \citep{espe2017}, and it is also known that the RES
+grain yields \citep{espe2017}, and the RES
 experiences warmer temperatures than many other rice growing areas in
@@
 milling yield \citep{mckenzie1994}, have been made over the study
-period and are no less important. However, planning to meet future
+period and are important. However, planning to meet future
 demand necessitates increases in grain yield \citep{Godfray2010,

### 2017-10-01 output/breed_paper.Rnw
Revisions from Bruce, Dave, Chris, and Jim

@@
 \cortext[corresauthor]{Corresponding author} \ead{mespe@ucdavis.edu}
-\author[ucd]{David MacKill}
+\author[ucd]{David Mackill}
 \author[ucd]{Jim E. Hill}
-\author[res]{Kent McKenzie}
 \author[ucce2]{Michelle Leinfelder-Miles}
@@
 Davis, Davis CA 95616, USA}
-
-\address[res]{Rice Experiment Station, California Cooperative Rice
-Research Foundation, Biggs, CA 95917, USA}
 
@@
 yield potential obtained through plant breeding. Specifically, the
-hypothesis was tested that the yield of a rice variety declines over
-time after the initial selection was tested. This would explain how
+hypothesis was tested that the yield of a rice cultivar declines over
+time after its introduction. This would explain how
 side-by-side comparisons show increased yield while no substantive
-increases have been seen in farmers' fields. Side-by-side comparisons
-under yield erosion will show ``false'' yield increases due to the
+increases have been seen in farmers' fields. If there is yield
+erosion, side-by-side comparisons will show ``false'' yield increases due to the
 yield of the older cultivar decreasing, not due to the increased
@@
 apparent increase in yield could be explained by the differences in
-the time when these cultivars were selected. The impact of yield
-erosion was estimated to be 29.3 kg ha\tsups{-1} year\tsups{-1} (90\%
+the time when these cultivars were selected. Yield
+eroded at an estimated rate of 29.3 kg ha\tsups{-1} year\tsups{-1} (90\%
 credible interval -4.4 to -53.3) after initial selection. Once this
-effect was considered, the apparent yield advantage of newer cultivars
+effect was considered, the yield advantage of newer cultivars
 over old was uncertain (-3.3 kg ha\tsups{-1} year\tsups{-1}, 90\%
-credible interval -36.1 to 31.5). This underscores the importance of
-breeding to maintain yield and the challenges of increasing the
-genetic yield potential.
-% Add more implications
+credible interval -36.1 to 31.5). Thus, without continuous plant improvement and
+deployment of new cultivars, yield will not only remain stagnant but
+will decline over time. To increasing the genetic yield potential of
+rice in the face of this phenomenon, higher yield targets are needed
+to counter-act the decrease over time in the current cultivars.
 
@@
 USDA-Agricultural Research Service, and farmer-funded
-research). However, since 1990 CA state average yields have increased
-more slowly (Fig. \ref{fig:state_yield}) despite continuous crop improvement \citep{ccrrf}. For the past 30 years, each newly developed
-cultivar has demonstrated a yield advantage over the previous
-cultivars in the Statewide Variety Trials (an annual multi-site
-variety trial program that compares experimental entries side-by-side
-with ``check'' cultivars). However, in a seeming contradiction these same
-reports show stagnant yields across release announcements (i.e.,
-reports accompanying the office release) from the
-various years. For example, at the year of release
-(1985), cultivar M-202 was reported to yield an average of 11 Mg
-ha\tsups{-1} \citep[14\% moisture content (MC);][]{m202}. In 2015, the reported average
-yield of the newest released cultivar M-209 was 10.8 Mg ha\tsups{-1}
-\citep]14\% MC;][]{m209}. In other rice systems, a decrease in a
-cultivar's yield performance over time has been observed
-\citep{Peng1998, Peng2010, Peng2000, Datta1995}, often termed yield
-erosion \citep{Peng1998}. It has been speculated that this yield erosion is due to
-the inability of a selected cultivar's genetics to adapt to changing
-biotic and abiotic conditions. To better understand how
-plant improvement and yield erosion are contributing to the
-improvement of rice yield in farmers' fields, we teseted two competing hypotheses: (1) the
-genetic yield advantage increased over the last 30+ years, and (2)
-the genetic yield advantage has not increased over the last 30+
-because the erosion of yield over time creates an illusion of improved
-yields.
+research). However, since 1990 the rate of increase has slowed
+(Fig. \ref{fig:state_yield}) despite continuous crop improvement
+\citep{ccrrf}. For the past 30 years, each newly developed cultivar
+has demonstrated a yield advantage over the previous cultivars in the
+Statewide Variety Trials (an annual multi-site variety trial program
+that compares experimental entries side-by-side with ``check''
+cultivars). However, in a seeming contradiction these same reports
+show stagnant yields across release announcements (i.e., reports
+accompanying the official public release) from the previous years. For
+example, when cultivar M-202 was released in 1985, it was reported to
+yield an average of 11 Mg ha\tsups{-1} \citep[14\% moisture content
+(MC);][]{m202}. In 2015, the reported average yield of the newest
+released cultivar M-209 was 10.8 Mg ha\tsups{-1} \citep]14\%
+MC;][]{m209}. In other rice systems, a decrease in a cultivar's yield
+performance over time has been observed \citep{Peng1998, Peng2010,
+Peng2000, Datta1995}, often termed yield erosion \citep{Peng1998}. It
+has been speculated that this yield erosion is due to the inability of
+a cultivar to adapt to changing biotic and abiotic conditions. To
+better understand how plant improvement and yield erosion are
+contributing to the improvement of rice yield in farmers' fields, we
+teseted the hypotheses: either (1) the genetic yield advantage
+increased over the last 30+ years, or (2) the genetic yield advantage
+has not increased over the last 30+ because the erosion of yield over
+time creates an illusion of improved yields.
 
@@
 \caption{Average rice yields over the period \Sexpr{min(CA_data$Year)}
-  to \Sexpr{max(CA_data$Year)} for California, USA, as reported by the
-  US Dept. of Agriculture National Agricultural Statistics
-  Service. Rapid gains in grain yield were experienced from
-  approximately 1950 -- 1990, but annual gains since 1990 have been
-  slower.}%The line represents the 10 year moving average.}
-% Remove 10 year moving average - Bruce does not like. 
+to \Sexpr{max(CA_data$Year)} for California, USA (US Dept. of
+Agriculture National Agricultural Statistics Service, 2017). Rapid
+gains in grain yield were experienced from approximately 1950 -- 1990,
+but annual gains since 1990 have been slower.}
 \label{fig:state_yield}
@@
 breeding program, the RES is researcher-managed (as opposed to
-farmer-managed trials) and conducts the majority of the trials for the
+farmer-managed trials) and plants the majority of experimental plots for the
 Statewide Variety Trials (approximately 3 to 4 times the number of
@@
 As part of the annual Statewide Variety Trials, newly developed
-entries are tested against officially named and released cultivars (checks) at
+entries are tested against officially released cultivars (checks) at
 the RES in several trials spanning multiple planting dates per
 year. For the purposes of this study, we concentrated on released
-medium grain cultivars, as medium grain cultivars constitute approximately
-90\% of the planted rice area in CA. These cultivars are typcically planted
-in the Statewide Variety Trials for 2 to 4 years prior to their
-official release. For these cultivars planted in
-breeding trials at the RES from \Sexpr{min(mod_data[,'year'])} --
-\Sexpr{max(mod_data[,'year'])} plot level observations were
-collected leading to a final data set of
-\Sexpr{format(nrow(mod_data), big.mark = ',')} observations representing
-\Sexpr{length(unique(mod_data[,'id']))} cultivars,
+medium grain cultivars, as these cultivars constitute approximately
+90\% of the planted rice area in CA. These cultivars are typcically
+evaluated in the Statewide Variety Trials for 3 to 5 years prior to
+their official release. For these cultivars planted in breeding trials
+at the RES from \Sexpr{min(mod_data[,'year'])} --
+\Sexpr{max(mod_data[,'year'])} plot level observations (replicated
+3--4 times in a CRB design) were collected leading to a final data set
+of \Sexpr{format(nrow(mod_data), big.mark = ',')} observations
+representing \Sexpr{length(unique(mod_data[,'id']))} cultivars,
 including every publicly released medium grain variety developed by
@@
 \Sexpr{max(mod_data[,'release_yr'])}. For each cultivar, the official
-year of release, typically 3 -- 5 years after introduction to the
-trials but prior to widespread commercial seed availability and the
-time elapsed (in years) since release was calculated.
+year of public release and the time elapsed (in years) since release was
+calculated.
 
@@
 
-\caption{The yield of released medium grain rice varieties for only
-the two years prior to each cultivar's official release. Points
+\caption{The experimental yield of medium grain rice varieties for the
+  two years prior to each cultivar's public release. Points
 ($\bullet$) and bars (|) represent the mean and standard error of the
@@
 
-Comparing the cultivars' yield during each's initial time in the
+Comparing the cultivars' yield during each's initial time (average 3 years) in the
 experimental trials confirmed that genetic yield improvements have
-been consistant over the study period
-(Fig. \ref{fig:first_yrs}). As estimated by the full model, there is
+been consistant over the 30 year study period
+(Fig. \ref{fig:first_yrs}). As estimated by the model including both
+yield erosion and yield improvement over time, there is
 no conclusive evidence that the yields of medium grain rice cultivars
@@
 was estimated to have better out-of-sample predictive accuracy
-compared to reduced models containing only one predictor. The
+compared to reduced models containing only one predictor (See
+supplemental material for further details). The
 estimated change in grain yield per year in the CCRRF's breeding
@@
 
-When yield erosion was not corrected for in the comparison
-(i.e., omitted from the model), there is an apparent small increase
-per year (approximately \Sexpr{only_release_coef} kg ha\tsups{-1}
+When yield erosion was not corrected for in the analysis (i.e.,
+omitted from the model), there is an apparent small increase per year
+(approximately \Sexpr{only_release_coef} kg ha\tsups{-1}
 year\tsups{-1}; 90\% CI \Sexpr{only_release_CI[1]} to
 \Sexpr{only_release_CI[2]}) (Fig. \ref{fig:wo_erosion}). Although this
-estimate ignores the confounding effect of yield erosion over time and
-performed worse in out-of-sample predictive accuracy, it confirms that
-yields in side-by-side comparison would support the conclusion that newer
-cultivars out-preform older ones if yield erosion is not considered.
-
-After correcting for the other variables, some cultivars were estimated
-to have higher yield compared to the overall average (as estimated by
-random intercepts). The marginal estimate for the performance of a
-cultivar averaged over the uncertainty in the other parameters in the
-model (i.e., the random effect) shows three cultivars with higher than
+estimate is flawed (i.e., it ignores the confounding effect of yield
+erosion over time and performed worse in out-of-sample predictive
+accuracy), it confirms previous analyses of side-by-side comparisons
+which concluded that newer cultivars out-preform older ones.
+
+After correcting for yield erosion, some cultivars were higher
+yielding compared to the overall average (as estimated by random
+intercepts). The marginal estimate for the performance of a cultivar
+averaged over the uncertainty in the other parameters in the model
+(i.e., the random effect) shows three cultivars with higher than
 average yields:
@@
 average of \Sexpr{round(yrs_coef*-36/1000,1)} Mg ha\tsups{-1} since
-its release. Comparing the two dominate medium grain rice cultivars M-202
-(widely planted from the late 1980s until the early 2000s), and M-206
-(the most widely planted CCRRFS cultivar at approximately 50\% of
-planted area at the time of writing), these two cultivars experienced
+its release. The two most dominate medium grain rice cultivars, M-202
+(widely planted from the late 1980s until the early 2000s) and M-206
+(currently the most widely planted), experienced
 similar trends in yield decline (Fig. \ref{fig:compare_yrs}; Table
@@
 year\tsups{-1}. Therefore, we conclude that the CCRRF breeding program
-has likely been maintaining but not substantially improved the yield
+has likely maintained but not substantially improved the yield
 potential in the face of changing biotic and abiotic pressures.
@@
 especially in systems such as CA where yield is approaching the
-physiological limit \citep{espe16a}. Without maintenance a breeding
-program and continuously selecting cultivars well adapted to current conditions,
-CA rice production systems are predicted to have lost an estimated
-\Sexpr{round(yrs_coef*(-36/1000),1)} Mg ha\tsups{-1} grain yield over
-the study period.
+physiological limit \citep{espe16a}. Without maintenance breeding
+(i.e., continuously selecting cultivars well adapted to current
+conditions), CA rice production systems are predicted to have lost an
+estimated \Sexpr{round(yrs_coef*(-36/1000),1)} Mg ha\tsups{-1} grain
+yield over the study period.
 
@@
 California state average yields have increased slowly
-even as little to no gains have been obtained in the breeding trials
+despite there being little to no gains in the breeding trials
 (Fig. \ref{fig:state_yield}). The evidence here suggests mechanisms
@@
 Lundy2015, linquist2009}, weed management \citep{Saito2008,
-Pittelkow2012, linquist2008}, among other management factors. may all be contributing to
+Pittelkow2012, linquist2008}, and other factors may all be contributing to
 increases in the state average yield. However, these gains are far less
@@
 these cultivars (M-201, M-204, M-205, and possibly M-209;
-Fig. \ref{fig:ids_comp})) may not be widely adapted. It is known that
+Fig. \ref{fig:ids_comp})) are not widely adapted. It is known that
 cool temperatures during the booting stage can have a large impact on
@@
 milling yield \citep{mckenzie1994}, have been made over the study
-period and are important. However, planning to meet future
+period and are important. Also, crop duration has been reduced. For
+example, during this study period, M-202 and M-206 have been the
+predominate varieties grown in CA, and while their yield potentials
+were similar at the time of release (Fig. 5), M-206 matures about 6
+days earlier than M-202 (Table 2). Shorter duration is very important
+in CA due to an already short growing season and water limitations
+(shorter duration varieties would presumably require less irrigation
+water).
+
+Planning to meet future
 demand necessitates increases in grain yield \citep{Godfray2010,
@@
 rownames(RES_coef) = c("Intercept (Mg ha\\tsups{-1})",
-                       "Release year (kg ha\\tsups{-1} year\\tsups{-1})",
+                       "Yield improvement (kg ha\\tsups{-1} year\\tsups{-1})",
                        "Yield erosion (kg ha\\tsups{-1} year\\tsups{-1})")
@@
 \caption{The marginal estimates for the difference between cultivars,
-  averaged over the uncertainty in the other parameters.}
+  averaged over the uncertainty in the fixed effect parameters (i.e.,
+  yield erosion and annual yield gains).}
 \label{fig:ids_comp}

### 2017-10-01 output/breed_paper.Rnw
Adds secondary analysis

@@
 Oroville, CA 95965, USA}
+
+\address[ucce]{University of California Cooperative Extension,
+Colusa, CA 95932, USA}
 
@@
 California (CA) rice production represents one such production
-system. Rice yields in CA during the period from 1920 -- 1990
-increased rapidly (Fig. \ref{fig:state_yield}), in part due to
-improvements in rice genetics (i.e., the ``Green Revolution'') through
-the efforts of the California Cooperative Rice Research Foundation
+system. Production in CA is predominately focused on premium quality
+medium grain \textit{japonica} cultivars (e.g., CalRose rice), and
+rice from CA is recognized globally for its quality. Rice yields in CA
+increased rapidly during the period from 1920 -- 1990
+(Fig. \ref{fig:state_yield}), in part due to improvements in rice
+genetics (i.e., the ``Green Revolution''). Most cultivars in use in CA
+are developed by the California Cooperative Rice Research Foundation
 (CCRRF; a collaboration between the University of California, the
 USDA-Agricultural Research Service, and farmer-funded
-research). However, since 1990 the rate of increase has slowed
-(Fig. \ref{fig:state_yield}) despite continuous crop improvement
-\citep{ccrrf}. For the past 30 years, each newly developed cultivar
-has demonstrated a yield advantage over the previous cultivars in the
-Statewide Variety Trials (an annual multi-site variety trial program
-that compares experimental entries side-by-side with ``check''
-cultivars). However, in a seeming contradiction these same reports
-show stagnant yields across release announcements (i.e., reports
-accompanying the official public release) from the previous years. For
-example, when cultivar M-202 was released in 1985, it was reported to
-yield an average of 11 Mg ha\tsups{-1} \citep[14\% moisture content
-(MC);][]{m202}. In 2015, the reported average yield of the newest
-released cultivar M-209 was 10.8 Mg ha\tsups{-1} \citep]14\%
-MC;][]{m209}. In other rice systems, a decrease in a cultivar's yield
-performance over time has been observed \citep{Peng1998, Peng2010,
-Peng2000, Datta1995}, often termed yield erosion \citep{Peng1998}. It
-has been speculated that this yield erosion is due to the inability of
-a cultivar to adapt to changing biotic and abiotic conditions. To
-better understand how plant improvement and yield erosion are
-contributing to the improvement of rice yield in farmers' fields, we
-teseted the hypotheses: either (1) the genetic yield advantage
-increased over the last 30+ years, or (2) the genetic yield advantage
-has not increased over the last 30+ because the erosion of yield over
-time creates an illusion of improved yields.
+research). Although grain quality has been maintained or even improved, rate of yield
+increase has slowed since the 1990s (Fig. \ref{fig:state_yield})
+despite continuous crop improvement \citep{ccrrf}. For the past 30
+years, each newly developed cultivar has demonstrated a yield
+advantage over the previous cultivars in the Statewide Variety Trials
+(an annual multi-site variety trial program that compares experimental
+entries side-by-side with ``check'' cultivars). However, in a seeming
+contradiction these same reports show stagnant yields across release
+announcements (i.e., reports accompanying the official public release)
+from the previous years. For example, when cultivar M-202 was released
+in 1985, it was reported to yield an average of 11 Mg ha\tsups{-1}
+\citep[14\% moisture content (MC);][]{m202}. In 2015, the reported
+average yield of the newest released cultivar M-209 was 10.8 Mg
+ha\tsups{-1} \citep]14\% MC;][]{m209}. In other rice systems, a
+decrease in a cultivar's yield performance over time has been observed
+\citep{Peng1998, Peng2010, Peng2000, Datta1995}, often termed yield
+erosion \citep{Peng1998}. It has been speculated that this yield
+erosion is due to the inability of a cultivar to adapt to changing
+biotic and abiotic conditions. To better understand how plant
+improvement and yield erosion are contributing to the improvement of
+rice yield in farmers' fields, we teseted the hypotheses: either (1)
+the genetic yield advantage increased over the last 30+ years, or (2)
+the genetic yield advantage has not increased over the last 30+
+because the erosion of yield over time creates an illusion of improved
+yields.
 
@@
 Biggs, CA (39.4648, -121.7342), and has been the central location for
-CCRRF's efforts to develop improved rice varieties adapted to CA since
+CCRRF's efforts to develop improved rice cultivars adapted to CA since
 1969. The climate at the RES is Mediterranean, characterized by mild
@@
 representing \Sexpr{length(unique(mod_data[,'id']))} cultivars,
-including every publicly released medium grain variety developed by
+including every publicly released medium grain cultivar developed by
 the CCRRF from \Sexpr{min(mod_data[,'release_yr'])} --
@@
 were quantified on yield over time. Yield was modeled as a response to
-the number of years since the variety was officially released and the
+the number of years since the cultivar was officially released and the
 year of release using a mixed-effects linear regression. To account
@@
 Where $yield_i$ is the grain yield for plot $i$, the Intercept is the
-average yield for the RES across varieties and years, $Rate_{yield\
+average yield for the RES across cultivars and years, $Rate_{yield\
 erosion\ i}$ is rate of yield erosion per ${years\ since\ release\ i}$
@@
 
-\caption{The experimental yield of medium grain rice varieties for the
+\caption{The experimental yield of medium grain rice cultivars for the
   two years prior to each cultivar's public release. Points
@@
 \caption{The relationship between yield and the years since release
-  for medium grain rice varieties. The estimated erosion of yield over time
+  for medium grain rice cultivars. The estimated erosion of yield over time
   (averaged over differences in cultivar and year) is estimated to be
@@
 example, during this study period, M-202 and M-206 have been the
-predominate varieties grown in CA, and while their yield potentials
+predominate cultivars grown in CA, and while their yield potentials
 were similar at the time of release (Fig. 5), M-206 matures about 6
@@
 in CA due to an already short growing season and water limitations
-(shorter duration varieties would presumably require less irrigation
+(shorter duration cultivars would presumably require less irrigation
 water).

### 2017-10-01 output/freq.Rmd
Adds secondary analysis

---
output: pdf_document
title: "Supplemental material"
date: "Oct. 1, 2017"
geometry: margin=1in
---

```{r setup, echo = FALSE}
load("../data/model_data.rda")
library(lme4)

```

This document replicates the analysis in "Rice yield improvements
through plant breeding are offset by inherent yield declines over
time" (Espe et al, 2017) using classical/frequentist statistics to
verify the results are consistent regardless of statistical method.

The complete data, code, and manuscript is available from \url{https://github.com/mespe/rice_advances}.

# Model excluding yield erosion

```{r}
no_erosion = lmer(yield_kg ~ release_yr +
                      (1|id) + (1|year), data = mod_data)

summary(no_erosion)
confint(no_erosion, parm = c("release_yr"),
        method = "boot", level = 0.9)
```

# Model including yield erosion

```{r}
erosion = lmer(yield_kg ~ release_yr + yrs_in_trial +
                   (1|id) + (1|year), data = mod_data)

summary(erosion)
confint(erosion, parm = c("release_yr","yrs_in_trial"),
        method = "boot", level = 0.9)
```

### 2017-10-29 output/breed_paper.Rnw
Commit before inverting the sections in results.

@@
 from one such cropping system, i.e., California (CA) rice, in order to
-better understand what improvements have been made in the genetic
-yield potential obtained through plant breeding. Specifically, the
-hypothesis was tested that the yield of a rice cultivar declines over
-time after its introduction. This would explain how
-side-by-side comparisons show increased yield while no substantive
-increases have been seen in farmers' fields. If there is yield
-erosion, side-by-side comparisons will show ``false'' yield increases due to the
-yield of the older cultivar decreasing, not due to the increased
-genetic yield potential of the newer cultivar. An apparent increase in
-yield in side-by-side comparisons between old and new rice varieties
-was observed when the yield erosion was not considered. However, this
-apparent increase in yield could be explained by the differences in
-the time when these cultivars were selected. Yield
-eroded at an estimated rate of 29.3 kg ha\tsups{-1} year\tsups{-1} (90\%
-credible interval -4.4 to -53.3) after initial selection. Once this
-effect was considered, the yield advantage of newer cultivars
-over old was uncertain (-3.3 kg ha\tsups{-1} year\tsups{-1}, 90\%
-credible interval -36.1 to 31.5). Thus, without continuous plant improvement and
-deployment of new cultivars, yield will not only remain stagnant but
-will decline over time. To increasing the genetic yield potential of
-rice in the face of this phenomenon, higher yield targets are needed
-to counter-act the decrease over time in the current cultivars.
+quantify improvements made in the genetic yield potential obtained
+through plant breeding. California rice systems are among the highest
+in the world and close to maximum yield potential. Specifically, the
+hypothesis was tested that if rice cultivar yields decline over time
+(termed yield erosion) then apparent yield increases in side-by-side
+yield comparison tests will not reflect increases in yield
+potential. This hypothesis was tested using experimental yield data
+over 30+ years at the California Cooperative Rice Research Foundation
+Rice Experiment Station. Results show an apparent increase in yield in
+side-by-side comparisons between old and new rice varieties was
+observed when the yield erosion was not considered. However, the
+yields of older varieties eroded at an estimated rate of 29.3 kg
+ha\tsups{-1} year\tsups{-1} (90\% credible interval -4.4 to -53.3)
+after initial selection. Once this effect was considered, the yield
+advantage of newer cultivars over old was uncertain (-3.3 kg
+ha\tsups{-1} year\tsups{-1}, 90\% credible interval -36.1 to
+31.5). These results highlight the importance of continuous crop
+improvement and deployment of new cultivars as yield will decline over
+time without it. To increase the genetic yield potential of rice in
+the face of this phenomenon, higher yield targets are needed to
+counter-act the decrease over time in the current cultivars.
 
@@
 
-% \linenumbers
+%\linenumbers
 
@@
 surpassing supply via continued yield improvement of staple
-crops. However production systems are experiencing plateaus in grain
+crops. However, many production systems are experiencing plateaus in grain
 yield \citep{Grassini2013, Peng2010}. If this trend continues,
@@
 demand, which carries with it undesirable ecological
-implications. Therefore, it is of critical importance to better
+implications~\citep{Foley2011a}. Therefore, it is of critical importance to better
 understand why the rate of increase in grain yields has declined or
@@
 
-California (CA) rice production represents one such production
-system. Production in CA is predominately focused on premium quality
-medium grain \textit{japonica} cultivars (e.g., CalRose rice), and
-rice from CA is recognized globally for its quality. Rice yields in CA
-increased rapidly during the period from 1920 -- 1990
-(Fig. \ref{fig:state_yield}), in part due to improvements in rice
-genetics (i.e., the ``Green Revolution''). Most cultivars in use in CA
-are developed by the California Cooperative Rice Research Foundation
-(CCRRF; a collaboration between the University of California, the
-USDA-Agricultural Research Service, and farmer-funded
-research). Although grain quality has been maintained or even improved, rate of yield
-increase has slowed since the 1990s (Fig. \ref{fig:state_yield})
-despite continuous crop improvement \citep{ccrrf}. For the past 30
-years, each newly developed cultivar has demonstrated a yield
-advantage over the previous cultivars in the Statewide Variety Trials
-(an annual multi-site variety trial program that compares experimental
-entries side-by-side with ``check'' cultivars). However, in a seeming
-contradiction these same reports show stagnant yields across release
-announcements (i.e., reports accompanying the official public release)
-from the previous years. For example, when cultivar M-202 was released
-in 1985, it was reported to yield an average of 11 Mg ha\tsups{-1}
-\citep[14\% moisture content (MC);][]{m202}. In 2015, the reported
-average yield of the newest released cultivar M-209 was 10.8 Mg
-ha\tsups{-1} \citep]14\% MC;][]{m209}. In other rice systems, a
-decrease in a cultivar's yield performance over time has been observed
-\citep{Peng1998, Peng2010, Peng2000, Datta1995}, often termed yield
-erosion \citep{Peng1998}. It has been speculated that this yield
-erosion is due to the inability of a cultivar to adapt to changing
-biotic and abiotic conditions. To better understand how plant
-improvement and yield erosion are contributing to the improvement of
-rice yield in farmers' fields, we teseted the hypotheses: either (1)
-the genetic yield advantage increased over the last 30+ years, or (2)
-the genetic yield advantage has not increased over the last 30+
-because the erosion of yield over time creates an illusion of improved
-yields.
+California (CA) rice represents one such production system. Rice is
+grown primarily in the Sacramento Valley which is characterized by
+having a Mediterranean climate, with long days, dry growing season,
+and relatively free of pests and diseases, which lead to some of the
+highest yields in the world~\citep{FAO2016}. Production in CA is
+predominately focused on premium quality medium grain
+\textit{japonica} cultivars (e.g., CalRose rice), and rice from CA is
+recognized globally for its quality
+(\url{http://agfax.com/2015/11/03/rice-calrose-wins-best-rice-world-competition/}). Most
+cultivars in use in CA are developed by the California Cooperative
+Rice Research Foundation (CCRRF; a collaboration between the
+University of California, the USDA-Agricultural Research Service, and
+farmer-funded research). In part due to improvements in rice genetics,
+rice yields in CA increased rapidly during the period from 1920 --
+1990; however, since the 1990s, the rate of yield increase has slowed
+(Fig. \ref{fig:state_yield}) despite continuous crop
+improvement. Based on a yield-gap analysis, on-farm yields in the
+major rice growing region of CA is 73 to 76\% of maximum yield
+potential~\citep{espe16a}. In highly intensive systems such as this,
+\citet{grassini2011} have shown that farmers are capable of attaining
+85\% of the maximum yield potential. Thus, given that farmers are near
+the attainable yield potential, increases in yield are likely to be
+slow and \citet{espe16a} reported that indeed this was the case
+with yields on average increasing by about 50 to 62 kg ha-1 between
+1999 and 2014.
+
+Certainly one question in looking at this situation is, “do new
+varieties have increased yield potential”? Like many breeding
+programs, one of the goals of the CCRRF is to increase yields through
+breeding varieties with increased yield potential. However, when the
+popular cultivar M-202 was
+released in 1985, it was reported to yield an average of 11 Mg
+ha\tsups{-1} \citep[14\% moisture content (MC);][]{m202}. In 2015, the
+reported average yield of the newest released cultivar M-209 was
+10.8 Mg ha\tsups{-1} \citep[14\% MC;][]{m209}. However, in annual
+breeding reports and varietal release announcements, based on yield
+performance in side-by-side variety trails, data show that new
+varieties usually out yield the older varieties. In other rice
+systems, a decrease in a cultivar's yield performance over time has
+been observed \citep{Peng1998, Peng2010, Peng2000, Datta1995}, often
+termed yield erosion \citep{Peng1998}. It has been speculated that
+this yield erosion is due to the inability of a cultivar to adapt to
+changing biotic and abiotic conditions~\citep{Peng1998, Peng2010,
+Peng2000, Datta1995}. Yield erosion may explain the apparent
+contradiction above, in which side-by-side yield comparisons show
+yield improvements in varieties while overall there may be little
+change in yield potential. To address this issue we teseted the
+hypothesis that if rice cultivar yields erode over time then apparent
+yield increases in side-by-side yield comparison tests will not
+reflect true yield increases.
 
@@
 
-The CCRRF Rice Experiment Station (RES) is located north-west of
-Biggs, CA (39.4648, -121.7342), and has been the central location for
-CCRRF's efforts to develop improved rice cultivars adapted to CA since
-1969. The climate at the RES is Mediterranean, characterized by mild
-winters during which most of the annual precipitation occurs
-(\Sexpr{18*24.5} to \Sexpr{25*24.5} mm annually) and warm summers
-largely free of precipitation events. Soils at the RES are classified
-as Esquon-Neerdobe clays with roughly 2\% soil organic matter
-\citep{ssurgo} in the top 15cm. As the central location for the CCRRF
-breeding program, the RES is researcher-managed (as opposed to
-farmer-managed trials) and plants the majority of experimental plots for the
-Statewide Variety Trials (approximately 3 to 4 times the number of
-experimental plots compared to any other location). To avoid potential
-complications due to site, climate, and management differences at the
-other 6 to 8 farmer-managed trials per year we focused our
-investigation on the RES.
+The CCRRF Rice Experiment Station (RES) is located near Biggs, CA
+(39.4648, -121.7342), and has been the central location for CCRRF's
+efforts to develop improved rice cultivars adapted to CA since
+1969. The California Statewide Variety Trails evaluates current and
+promising varieties at the RES and six to eight on-farm trials around
+the state each year. To avoid potential complications due to site,
+climate, and management differences at the farmer-managed trials per
+year, we focused our investigation on the RES. The RES is
+researcher-managed and plants the majority of the experimental plots
+for the statewide program (approximately 3 to 4 times the number of
+experimental plots compared to other on-farm locations). The climate
+at the RES is Mediterranean, characterized by mild winters during
+which most of the annual precipitation occurs (\Sexpr{18*24.5} to
+\Sexpr{25*24.5} mm annually) and warm summers largely free of
+precipitation events. Soils at the RES are classified as
+Esquon-Neerdobe clays with roughly 2\% soil organic matter
+\citep{ssurgo} in the top 15cm.
 
@@
 medium grain cultivars, as these cultivars constitute approximately
-90\% of the planted rice area in CA. These cultivars are typcically
-evaluated in the Statewide Variety Trials for 3 to 5 years prior to
+90\% of the planted rice area in CA. Newly developed entries are typcically
+evaluated in these trials for 3 to 5 years prior to
 their official release. For these cultivars planted in breeding trials
@@
 of \Sexpr{format(nrow(mod_data), big.mark = ',')} observations
-representing \Sexpr{length(unique(mod_data[,'id']))} cultivars,
-including every publicly released medium grain cultivar developed by
+representing \Sexpr{length(unique(mod_data[,'id']))} cultivars and
+included every publicly released medium grain cultivar developed by
 the CCRRF from \Sexpr{min(mod_data[,'release_yr'])} --
@@
 year of public release and the time elapsed (in years) since release was
-calculated.
+determined.
 
@@
 Cooperative Extension following prescribed best management
-practice, including land preparation, fertility management, and pest
-and disease control \citep{ucce}. For planting, rice seed is pre-germinated prior
+practice for land preparation, fertility rates, and pest
+and disease control \citep{ucce}. As is typical for water-seeded
+systems, for planting, rice seed is pre-germinated prior
 to direct-seeding into pre-flooded fields. Plots were either 14.0 or
@@
 
-To test each hypothesis, the influence of breeding and yield erosion
-were quantified on yield over time. Yield was modeled as a response to
-the number of years since the cultivar was officially released and the
-year of release using a mixed-effects linear regression. To account
-for similarities between years and between cultivars, year and
-cultivar were included in the model as random effects.
+To test the hypothesis, the influence of breeding and yield erosion
+were quantified on yield over time using a Bayesian hierarchical
+model. Yield was modeled as a response to the number of years since
+the cultivar was officially released and the year of release using a
+mixed-effects linear regression. To account for similarities between
+years and between cultivars, year and cultivar were included in the
+model as random effects.
 
@@
 90\% CI is more stable to perpetuation's and hence is preferable to the 95\%
-interval \citep{rstanarm}. Complete data and code used for this
+interval \citep{rstanarm}. To verify the estimates, the analysis was
+also conducted using classical methods (frequentist mixed-effects
+model fit by maximum likelihood; See supplemental material for
+details). Complete data and code used for this
 analysis is available at \url{github.com/mespe/rice_advances.git}.
@@
 
-Comparing the cultivars' yield during each's initial time (average 3 years) in the
-experimental trials confirmed that genetic yield improvements have
-been consistant over the 30 year study period
+There was agreement between the estimates from the Bayesian and
+classical approaches. The Bayesian estimates are presented here,
+though the classical estimates can be found in the supplemental
+material. Comparing the cultivars' yield during each's initial time
+(average 3 years) in the experimental trials confirmed that genetic
+yield improvements have been consistant over the 30 year study period
 (Fig. \ref{fig:first_yrs}). As estimated by the model including both
-yield erosion and yield improvement over time, there is
-no conclusive evidence that the yields of medium grain rice cultivars
-selected and released by CCRRF's breeding program have, on average,
-increased over the last 30 years (Table \ref{tbl:coef}). The full
-model which corrected for both the year of release and yield erosion
-was estimated to have better out-of-sample predictive accuracy
-compared to reduced models containing only one predictor (See
-supplemental material for further details). The
-estimated change in grain yield per year in the CCRRF's breeding
-trials at the RES was estimated to be a negligible
+yield erosion and yield improvement over time, there is no conclusive
+evidence that the yields of medium grain rice cultivars selected and
+released by CCRRF's breeding program have, on average, increased over
+the last 30 years (Table \ref{tbl:coef}). The full model which
+corrected for both the year of release and yield erosion was estimated
+to have better out-of-sample predictive accuracy compared to reduced
+models containing only one predictor (See supplemental material for
+further details). The estimated change in grain yield per year in the
+CCRRF's breeding trials at the RES was estimated to be a negligible
 \Sexpr{release_yr_coef} kg ha\tsups{-1} year\tsups{-1}, with both
@@
 mtext("Adjusting for yield erosion", adj = 0)
+
 polygon(x = c(yrs, rev(yrs)),
@@
 text(5, 8.0, bquote(.(yrs_coef)~kg~ha^-1~year^-1))
+mtext("Adjusting for yield erosion", adj = 0)
 
@@
 were similar at the time of release (Fig. 5), M-206 matures about 6
-days earlier than M-202 (Table 2). Shorter duration is very important
+days earlier than M-202 (Table \ref{tbl:ids}). Shorter duration is very important
 in CA due to an already short growing season and water limitations

### 2017-10-29 output/breed_paper.Rnw
Updated with Bruce's comments - need to revise.

@@
 surpassing supply via continued yield improvement of staple
-crops. However, many production systems are experiencing plateaus in grain
-yield \citep{Grassini2013, Peng2010}. If this trend continues,
+crops. However, many production systems are experiencing plateaus in
+grain yield \citep{Grassini2013, Peng2010}. If this trend continues,
 improvements per unit land area are no longer possible, and an
@@
 demand, which carries with it undesirable ecological
-implications~\citep{Foley2011a}. Therefore, it is of critical importance to better
-understand why the rate of increase in grain yields has declined or
-leveled off in intensified production systems.
+implications~\citep{Foley2011a, tilman1999}. Therefore, it is of
+critical importance to better understand why the rate of increase in
+grain yields has declined or leveled off in intensified production
+systems.
 
@@
 slow and \citet{espe16a} reported that indeed this was the case
-with yields on average increasing by about 50 to 62 kg ha-1 between
+with yields on average increasing by about 50 to 62 kg ha\tsups{-1} between
 1999 and 2014.
 
-Certainly one question in looking at this situation is, “do new
-varieties have increased yield potential”? Like many breeding
+Certainly one question in looking at this situation is, ``do new
+varieties have increased yield potential''?  Like many breeding
 programs, one of the goals of the CCRRF is to increase yields through
 breeding varieties with increased yield potential. However, when the
-popular cultivar M-202 was
-released in 1985, it was reported to yield an average of 11 Mg
-ha\tsups{-1} \citep[14\% moisture content (MC);][]{m202}. In 2015, the
-reported average yield of the newest released cultivar M-209 was
-10.8 Mg ha\tsups{-1} \citep[14\% MC;][]{m209}. However, in annual
-breeding reports and varietal release announcements, based on yield
-performance in side-by-side variety trails, data show that new
-varieties usually out yield the older varieties. In other rice
-systems, a decrease in a cultivar's yield performance over time has
-been observed \citep{Peng1998, Peng2010, Peng2000, Datta1995}, often
-termed yield erosion \citep{Peng1998}. It has been speculated that
-this yield erosion is due to the inability of a cultivar to adapt to
-changing biotic and abiotic conditions~\citep{Peng1998, Peng2010,
-Peng2000, Datta1995}. Yield erosion may explain the apparent
-contradiction above, in which side-by-side yield comparisons show
-yield improvements in varieties while overall there may be little
-change in yield potential. To address this issue we teseted the
-hypothesis that if rice cultivar yields erode over time then apparent
-yield increases in side-by-side yield comparison tests will not
-reflect true yield increases.
+popular cultivar M-202 was released in 1985, it was reported to yield
+an average of 11 Mg ha\tsups{-1} \citep[14\% moisture content
+(MC);][]{m202}. In 2015, the reported average yield of the newest
+released cultivar M-209 was 10.8 Mg ha\tsups{-1} \citep[14\%
+MC;][]{m209}. However, in annual breeding reports and varietal release
+announcements, based on yield performance in side-by-side variety
+trails, data show that new varieties usually out yield the older
+varieties. In other rice systems, a decrease in a cultivar's yield
+performance over time has been observed \citep{Peng1998, Peng2010,
+Peng2000, Datta1995}, often termed yield erosion \citep{Peng1998}. It
+has been speculated that this yield erosion is due to the inability of
+a cultivar to adapt to changing biotic and abiotic
+conditions~\citep{Peng1998, Peng2010, Peng2000, Datta1995}. Yield
+erosion may explain the apparent contradiction above, in which
+side-by-side yield comparisons show yield improvements in varieties
+while overall there may be little change in yield potential. To
+address this issue we tested the hypothesis that if rice cultivar
+yields erode over time then apparent yield increases in side-by-side
+yield comparison tests will not reflect true yield increases.o
 
@@
 medium grain cultivars, as these cultivars constitute approximately
-90\% of the planted rice area in CA. Newly developed entries are typcically
+90\% of the planted rice area in CA. Newly developed entries are typically
 evaluated in these trials for 3 to 5 years prior to
@@
 
-\subsection{Genetic yield improvements}
+\subsection{Cultivar yields at time of release}
 
@@
 material.  Comparing the cultivars' yield during each's initial time
-(average 3 years) in the experimental trials confirmed that genetic
-yield improvements have been consistant over the 30 year study period
-(Fig. \ref{fig:first_yrs}). As estimated by the model including both
-yield erosion and yield improvement over time, there is no conclusive
-evidence that the yields of medium grain rice cultivars selected and
-released by CCRRF's breeding program have, on average, increased over
-the last 30 years (Table \ref{tbl:coef}). The full model which
-corrected for both the year of release and yield erosion was estimated
-to have better out-of-sample predictive accuracy compared to reduced
-models containing only one predictor (See supplemental material for
-further details). The estimated change in grain yield per year in the
-CCRRF's breeding trials at the RES was estimated to be a negligible
+(average 3 years) in the experimental trials just before being
+released showed that across cultivars, yields ranged from
+\Sexpr{round(plot_data["M207", "mean"], 2)} (M-207) to
+\Sexpr{round(plot_data["M205", "mean"], 2)} Mg ha\tsups{-1}
+(M-205). Evaluating these initial yields over time indicates there was
+no evidence of genetic yield improvements during the 30-year study
+period (Fig. 2). Individual exceptions to this were two varieties
+(M-204 and M-205) that had higher yields than the others.
+
+\subsection{Yield erosion}
+\begin{figure}[H]
+
+<<fig=TRUE, echo=FALSE>>=
+yrs_in <- -10:40
+pp <- posterior_linpred(RES_full, re.form = NA,
+                        newdata = data.frame(yrs_in_trial = yrs_in,
+                                             release_yr_c = 0))
+pp <-  (pp * scl + center) / 1000 
+ints = apply(pp, 2, quantile, p = c(0.05, 0.95))
+
+plot_data = prep_plot_data(mod_data$yield_kg / 1000,
+                           mod_data$yrs_in_trial)
+
+plot_data$tmp = as.numeric(row.names(plot_data))
+colnames(plot_data) = c("yield_kg", "yield_se", "yrs_in_trial")
+
+plot(yield_kg ~ yrs_in_trial, data = plot_data, pch = 16, #col = rgb(0,0,0,0.25),
+     xlab = "Year since release",
+     ylab = expression("Grain yield ( kg ha"^-1~")"),
+     ylim = c(8.0, 12.5))
+with(plot_data, segments(x0 = yrs_in_trial, y0=yield_kg + yield_se,
+                         y1 = yield_kg - yield_se))
+abline(a = ((fixef(RES_full)[1] * scl) + center) / 1000 ,
+       b = yrs_coef/1000)
+polygon(x = c(yrs_in, rev(yrs_in)),
+        y = c(ints[1,],rev(ints[2,])),
+        border = NA,
+        col = rgb(0,0,0,0.1))
+text(5, 8.0, bquote(.(yrs_coef)~kg~ha^-1~year^-1))
+mtext("Adjusting for yield erosion", adj = 0)
+
+@
+\caption{The relationship between yield and the years since release
+  for medium grain rice cultivars. Negative values on the X axis
+  indicate experimental yields prior to the official release. The
+  estimated erosion of yield over time (averaged over differences in
+  cultivar and year) is estimated to be \Sexpr{yrs_coef} kg ha\tsups{-1}
+  year\tsups{-1}. Points ($\bullet$) and bars (|) represent the mean and
+  standard error of the raw data, respectively. The shaded area is the
+  90\% credible interval.}
+\label{fig:RES_year}
+\end{figure}
+
+The yields of each cultivar were evaluated over time relative to the
+time when the cultivar was released using the full model which
+accounts for both yield erosion and yield improvements over time. The
+average decrease in yield over time (i.e., yield erosion) once a
+cultivar was selected was estimated \Sexpr{yrs_coef} kg ha\tsups{-1}
+year\tsups{-1} (90\% CI: \Sexpr{RES_CI[2]} to \Sexpr{RES_CI[1]}) (Fig
+\ref{fig:RES_year}; Table \ref{tbl:coef}). The oldest cultivar in the
+trials (\Sexpr{unique(mod_data[which.min(mod_data[,"year"]),"id"])}
+had lost an average of \Sexpr{round(yrs_coef*-36/1000,1)} Mg
+ha\tsups{-1} since its release. The most dominate medium grain rice
+cultivar, M-202 (widely planted from the late 1980s until the early
+2000s) experienced similar trends in yield decline
+(Fig. \ref{fig:RES_year}; Table \ref{tbl:coef}). M-206 replaced M-202
+as the most widely grown cultivar in CA and the current estimated 560
+kg ha-1 yield advantage of M-206 over M-202 can be attributed solely
+to the erosion of M-202's yield over time (Figs \ref{fig:first_yrs} \&
+\ref{fig:compare_yrs}). Further supporting this estimate of yield
+erosion, both cultivars yielded similarly at their respective year of
+release (Fig. \ref{fig:compare_yrs}), despite the fact that M-202 was
+released 19 years prior to M-206 (Fig. \ref{fig:compare_yrs}, Table
+\ref{tbl:ids}).
+
+When yield erosion was not corrected for in the analysis (i.e.,
+omitted from the model), there is an apparent small increase per year
+(approximately \Sexpr{only_release_coef} kg ha\tsups{-1}
+year\tsups{-1}; 90\% CI \Sexpr{only_release_CI[1]} to
+\Sexpr{only_release_CI[2]}) (Fig. \ref{fig:wo_erosion}).  However,
+using the full model which corrected for both the year of release and
+yield erosion, indicates that the estimated change in grain yield per
+year in the CCRRF's breeding trials at the RES was
 \Sexpr{release_yr_coef} kg ha\tsups{-1} year\tsups{-1}, with both
@@
 (Fig. \ref{fig:first_yrs}).
-
-When yield erosion was not corrected for in the analysis (i.e.,
-omitted from the model), there is an apparent small increase per year
-(approximately \Sexpr{only_release_coef} kg ha\tsups{-1}
-year\tsups{-1}; 90\% CI \Sexpr{only_release_CI[1]} to
-\Sexpr{only_release_CI[2]}) (Fig. \ref{fig:wo_erosion}). Although this
-estimate is flawed (i.e., it ignores the confounding effect of yield
-erosion over time and performed worse in out-of-sample predictive
-accuracy), it confirms previous analyses of side-by-side comparisons
-which concluded that newer cultivars out-preform older ones.
 
@@
 \Sexpr{paste(gsub('M','M-',rev(row.names(id_est))[c(1,2,4)]),collapse=',')}
-(Figs. \ref{fig:first_yrs} \& \ref{fig:ids_comp}). A fourth, M-209,
-might also have higher than average yield, though the CI for M-209
-overlaps with zero (Fig. \ref{fig:ids_comp}).
+(Figs. \ref{fig:first_yrs} \& \ref{fig:ids_comp}).
 
@@
 
-\subsection{Yield erosion}
+\section{Discussion}
+
+In California, the cultivars with broad adaptation and adoption over
+the 30+ year period of this study are M-202, M-206, and M-105. Cold
+tolerance is a major problem for much of the California rice producing
+region ~\citep{espe2017}, although it is a bigger problem in the
+southern portion of the Sacramento Valley due to the influence of the
+ocean winds through the Delta ~\citep{espe16a}. These three
+cultivars have a higher degree of cold tolerance, high yield
+potential, and relatively short duration and are thus broadly grown
+with over 70\% of the area in one of these cultivars during the study
+period ~citep{ucce}. Importantly, not all medium grain cultivars that were
+released were purported to have higher yields than previous
+cultivars. Some cultivars had special traits, but upon release had
+similar yields to more broadly adapted cultivars. Examples of this
+include M-208, which had resistance to blast disease (scientific name)
+and M-402 which is a premium quality variety. Other varieties such as
+M-205 and M-209, while high yielding, do not have the cold tolerance
+of the other cultivars and are thus only recommended for the northern
+portion of the Sacramento Valley.
+
+Looking at all the cultivars that have been released, there is no
+conclusive evidence of increased genetic yield potential in medium
+grain rice cultivars released by CCRRF's breeding improvement program
+over the period \Sexpr{min(mod_data[,'year'])} to
+\Sexpr{max(mod_data[,'year'])}, which has stayed relatively constant
+at \Sexpr{round(mean(mod_data[,"yield_kg"]), 2)} Mg
+ha\tsups{-1}. Since the estimated yield improvement spans from
+\Sexpr{rel_CI[1]} to \Sexpr{rel_CI[2]} kg ha\tsups{-1} year\tsups{-1}
+(Fig. \ref{fig:RES_year}; Table \ref{tbl:coef}),
+both increases and decreases in the genetic yield potential over the
+study period are plausible. The estimated yield
+improvement of \Sexpr{only_release_coef} kg ha\tsups{-1}
+year\tsups{-1} (Fig. \ref{fig:wo_erosion}) that does not account for
+yield erosion is lower than the 37 to 60 kg ha\tsups{-1}
+year\tsups{-1} estimated by~\citet{Omar2014} in side-by-side
+comparisons. Rather, this analysis suggests that the CCRRF breeding
+program has maintained but not substantially improved the yield
+potential in the face of changing biotic and abiotic pressures. This
+conclusion holds true even when evaluating the cultivars which have
+been breed for broad adaption and high yield potential (M-202, M-206
+and M-105). Yields of these varieties at time of release have all
+yielded between
+\Sexpr{tmp=mod_data[mod_data[,"yrs_in_trial"]<=0&mod_data[,"id"]%in%c("M202","M206","M105"),];round(min(by(tmp,tmp[,"id"],function(x)mean(x[,"yield_kg"]))/1000),2)}
+and
+\Sexpr{round(min(by(tmp,tmp[,"id"],function(x)mean(x[,"yield_kg"]))/1000),2)}
+Mg ha\tsups{-1} with no indication of yields increasing
+(Fig. \ref{fig:first_yrs}). While these yields have not increased,
+M-205 did have higher yield potential; however, this variety is not
+broadly adapted due to poor cold tolerance.
+
+Yields of a particular cultivar may decline overtime for a number of
+reasons. First, ~\citet{Nie2009} reported yield declines in rice due
+to continuously flooded conditions over a long time period which could
+be overcome with improved N management. This is likely not the case
+here as the fields used in this study grow rice only once season a
+year and tend to be non-flooded for the remainder of the
+year. Furthermore, at this location and unlike most CA rice fields,
+the fields are left fallow every other year. Second, changing climate
+may influence yields with either warmer~\citep{van2013} or
+cooler (Espe et al., 2017) temperatures reducing yields. Again, this
+is unlikely to be the case in this situation as Espe et al. (2017)
+found no significant changes in growing season temperatures (either
+high, low or average temps) the Sacramento Valley during the study
+period.  Thirdly, yield erosion of a cultivar as has been reported for
+the cultivar IR8 at IRRI~\citep{Peng1998, Peng2000, Peng2010}. Biotic
+and abiotic forces have previously been speculated to be responsible
+for yield erosion, since the cultivar's genetics remain stable from
+the time of selection but the conditions the cultivar may experience
+over time are continuously changing~\citep{Peng1998, Peng2000,
+Peng2010}. We estimate that yield erosion has decreased the yield of
+medium grain rice cultivars released by the CCRRF over the last 30+
+years by an average of \Sexpr{yrs_coef} Mg ha\tsups{-1}, potentially
+offsetting any gains made by the crop improvement program. However, it
+is not clear what is responsible for this erosion. It is unlikely
+biotic factors as pest and disease pressure are low in this
+Mediterranean climate. Furthermore, these fields are fallowed every
+other year and the residues are burned, further lowering the potential
+for pest and disease pressure. While efforts should be made to
+identify the cause of erosion in this environment, this observation
+highlights both the need for maintenance of plant breeding and the
+difficulty of increasing genetic yield potential, especially in
+systems such as CA where yield is approaching the physiological limit
+\citep{espe16a}. Without maintenance breeding (i.e., continuously
+selecting cultivars well adapted to current conditions), CA rice
+production systems are predicted to have lost an estimated
+\Sexpr{round(yrs_coef*(-36/1000),1)} Mg ha\tsups{-1} grain yield over
+the study period.
+
+A goal of many breeding programs is to not only maintain the benchmark
+yield of the previous ``standard'' but to surpass it, especially if we
+are to meet the target of 50\% increased food production by 2050
+\citep{Foley2011a}. Given the estimated yield erosion, to hit an
+ambitious target of 0.5 Mg ha\tsups{-1} (approximately 5\%) increase
+in yield per release (with roughly 4 years between releases), a new
+cultivar needs to yield an additional \Sexpr{round(yrs_coef*-4,-1)} kg
+ha\tsups{-1} in side-by-side comparison in addition to the 0.5 Mg
+ha\tsups{-1} target improvement. Often side-by-side yield improvements
+show an improvement in grain yield of 2\% per release, in which case
+no absolute improvements in yield have been made beyond maintaining
+yield. The evidence here is consistent with the observation that the
+genetic yield potential has not increased substantially in the last
+30+ years (Fig. \ref{fig:first_yrs}).
+
 \begin{figure}[H]
-
-<<fig=TRUE, echo=FALSE>>=
-yrs_in <- -10:40
-pp <- posterior_linpred(RES_full, re.form = NA,
-                        newdata = data.frame(yrs_in_trial = yrs_in,
-                                             release_yr_c = 0))
-pp <-  (pp * scl + center) / 1000 
-ints = apply(pp, 2, quantile, p = c(0.05, 0.95))
-
-plot_data = prep_plot_data(mod_data$yield_kg / 1000,
-                           mod_data$yrs_in_trial)
-
-plot_data$tmp = as.numeric(row.names(plot_data))
-colnames(plot_data) = c("yield_kg", "yield_se", "yrs_in_trial")
-
-plot(yield_kg ~ yrs_in_trial, data = plot_data, pch = 16, #col = rgb(0,0,0,0.25),
-     xlab = "Year since release",
-     ylab = expression("Grain yield ( kg ha"^-1~")"),
-     ylim = c(8.0, 12.5))
-with(plot_data, segments(x0 = yrs_in_trial, y0=yield_kg + yield_se,
-                         y1 = yield_kg - yield_se))
-abline(a = ((fixef(RES_full)[1] * scl) + center) / 1000 ,
-       b = yrs_coef/1000)
-polygon(x = c(yrs_in, rev(yrs_in)),
-        y = c(ints[1,],rev(ints[2,])),
-        border = NA,
-        col = rgb(0,0,0,0.1))
-text(5, 8.0, bquote(.(yrs_coef)~kg~ha^-1~year^-1))
-mtext("Adjusting for yield erosion", adj = 0)
-
-@
-\caption{The relationship between yield and the years since release
-  for medium grain rice cultivars. The estimated erosion of yield over time
-  (averaged over differences in cultivar and year) is estimated to be
-  \Sexpr{yrs_coef} kg ha\tsups{-1} year\tsups{-1}. Points ($\bullet$)
-  and bars (|) represent the mean and standard error of the raw data,
-  respectively. The shaded area is the 90\% credible interval.}
-% For each year, point size by n (open circles)
-\label{fig:RES_year}
+<<fig = TRUE, echo = FALSE, width=5, height=5>>=
+
+plot(1, type = "n",
+     # xaxt = "n",
+     xlab = "Years",
+     ylab = expression("Grain yield ( Mg ha"^-1~")"),
+     ylim = c(8.5,11.5),
+     xlim = c(0,40),
+     xaxs = "i")
+
+# arrows(12, 8.2, 27, 8.2, xpd = TRUE, lwd = 2)
+abline(a = 10, b = fixef(RES_full)[3], lty = 2, lwd = 2)
+text(25, 10.85, "(3) target above maintenance", cex = 0.8,
+     srt = 18)#((atan(fixef(RES_full)[3] + 0.06) * 180))/pi)
+abline(h = 10, lty = 3, lwd = 2)
+text(25, 10.1, "(1) current situation", cex = 0.8)
+abline(a = 10, b = fixef(RES_full)[3] + 0.06, lty = 4, lwd = 2)
+text(25, 9.35, "(2) no maintenance breeding", cex = 0.8,
+     srt = -18)
+
+@ 
+\caption{A conceptual diagram highlighting three scenarios: (1)
+maintenance breeding, where losses over time are offset by
+improvements via plant improvement, (2) no maintenance breeding, where
+grain yield slowly erodes over time (as estimated from this study),
+and (3) absolute improvement in grain yield (based on ``typical''
+yield goals of 3\% increase per release), where gains in grain yield
+exceed the losses over time.}
+\label{fig:theory}
 \end{figure}
+
+There are several cultivars that were found to have better than
+average yields, even after correcting for yield erosion. However,
+these cultivars (M-201, M-204, M-205, and possibly M-209;
+Fig. \ref{fig:ids_comp})) are not widely adopted. It is known that
+cool temperatures during the booting stage can have a large impact on
+grain yields \citep{espe2017}, and the RES
+experiences warmer temperatures than many other rice growing areas in
+CA \citep{espe2017}. These cultivars may perform well at the RES but
+not widely in CA. Therefore, it should be noted that while our
+investigation has been concerned about increasing yield potential, the
+estimates here might not capture yield stability. Yield stability is
+notoriously difficult to quantify \citep[e.g.,][]{lobell2011}, hence
+further investigation is needed to explore how yield stability has
+changed over the study period.
+
+To be clear, our investigation here is narrowly focused on the
+rough-rice/field grain yield. Other advancements due to plant
+improvement, including increased quality \citep{mckenzie1993} and
+milling yield \citep{mckenzie1994}, have been made over the study
+period and are important. Also, crop duration has been reduced. For
+example, during this study period, M-202 and M-206 have been the
+predominate cultivars grown in CA, and while their yield potentials
+were similar at the time of release (Fig. 5), M-206 matures about 6
+days earlier than M-202 (Table \ref{tbl:ids}). Shorter duration is very important
+in CA due to an already short growing season and water limitations
+(shorter duration cultivars would presumably require less irrigation
+water).
+
+California state average yields have increased slowly
+despite there being little to no gains in the breeding trials
+(Fig. \ref{fig:state_yield}). The evidence here suggests mechanisms
+other than increased genetic yield potential are responsible for these gains
+(Fig. \ref{fig:compare_yrs}). Improvement in agronomic management,
+including more efficient nutrient management \citep{Lundy2012,
+Lundy2015, linquist2009}, weed management \citep{Saito2008,
+Pittelkow2012, linquist2008}, and other factors may all be contributing to
+increases in the state average yield. However, these gains are far less
+than those experienced during the preceding era, and by their
+nature agronomic gains cannot surpass the genetic yield potential
+\citep{Fischer2015}. Judging from yields achieved at the RES
+(Figs. \ref{fig:RES_year} \& \ref{fig:compare_yrs}), a
+researcher-managed production system implementing best management
+practices in more controlled situations compared to farmers' fields, state
+average yields (approximately 10 Mg ha\tsups{-1} at the time of
+writing; Fig. \ref{fig:state_yield}) may be approaching the current
+genetic limit. Assuming that the level of management possible in small
+experimental plots cannot be attained in farmers' fields and hence
+experimental yields represent the attainable yield ceiling
+\citep{Fischer2015}, agronomic gains in farmer's fields are nearing
+their potential, consistent with physiological model-based estimates
+of the attainable yield gap in CA rice production systems
+\citep{espe16a}.
+
+Planning to meet future
+demand necessitates increases in grain yield \citep{Godfray2010,
+Foley2011a, Mueller2012}. Meeting the challenge to produce more grain
+without expansion of land under cultivation will rely on new genetics
+with increased yield potential. The evidence here suggests that to
+accomplish true increases over time, the improvement target needs take
+into account erosion in the current ``standard'' cultivars' yield
+over time (Fig. \ref{fig:theory}). While side-by-side comparisons will
+continue to be useful, comparisons across years to each cultivar's
+benchmark yield at the time of release should provide information on
+the absolute improvements being made in yield potential over time. New
+technologies may need to be adapted to further increase yield
+potential, including hybrid rice.
+
+\section{Conclusions}
+
+Here we show that side-by-side cultivar comparisons do not necessarily
+indicate gains in yield potential as they do not account for yield
+erosion. Quantifying changes in yield potential requires also
+examining historical benchmarks. Improvement in grain yield is
+essential to meeting the challenges of feeding a growing population in
+a changing world. To achieve these improvements, we discuss here that
+(1) continuous maintenance via plant improvement is required, without
+which yields would decrease over time, (2) the challenge of improving
+genetic yield potential is substantial, since absolute gains require
+improvement in yield in excess of the maintenance threshold, and (3)
+increasing yield through improved management (agronomy) cannot exceed
+the ceiling of genetic yield potential. Taken together, these
+conclusions underscore both the necessity and substantial challenges
+facing rice breeding. We suggest here that historical benchmarks be
+used in to evaluate promising rice cultivars in addition to
+side-by-side comparisons to maintain perspective of the absolute yield
+gains being made.
+
+\newpage
+
+\section{Tables}
+
+<<echo=FALSE, results=tex>>=
+
+CIs = posterior_interval(RES_full)[1:3,] * scl
+CIs[1,] = round((CIs[1,] + center)/1000, 1)
+CIs = round(CIs, 1)
+med_eff = fixef(RES_full) * scl
+med_eff[1] = round((med_eff[1] + center)/1000, 1)
+med_eff = round(med_eff, 1)
+
+RES_coef = cbind(med_eff, paste(CIs[,1],CIs[,2], sep = " to "))
+rownames(RES_coef) = c("Intercept (Mg ha\\tsups{-1})",
+                       "Yield improvement (kg ha\\tsups{-1} year\\tsups{-1})",
+                       "Yield erosion (kg ha\\tsups{-1} year\\tsups{-1})")
+colnames(RES_coef) = c("Median", "90\\% Credible interval")
+
+print(xtable(RES_coef,
+             caption = "The estimated median fixed effects and 90\\% credible intervals for the improvement in yield of medium grain rice cultivars developed by the California Cooperative Rice Research Foundation, California, USA over the time period 1984 -- 2016.",
+             label = "tbl:coef",
+             align = "lcc"),
+      caption.placement="top",
+      booktabs=TRUE,
+      sanitize.colnames.function = function(x){x},
+      sanitize.rownames.function = function(x){x})
+@
+\newpage
+
+<<echo=FALSE, results=tex>>=
+ids = aggregate(release_yr ~ id, data = mod_data, unique)
+ids$dth = as.integer(round(aggregate(dth ~ id, data = mod_data, mean, na.rm = TRUE)$dth, 0))
+colnames(ids) = c("Cultivar", "Release Year", "Days to 50% Heading")
+ids = ids[order(ids[,2]),]
+ids[,2] = as.integer(ids[,2])
+ids[,1] = gsub("M", "M-", ids[,1])
+
+print(xtable(ids,
+             caption = "All named medium grain rice cultivars and their associated release year planted at the Rice Experiment Station (RES), Biggs, California from 1984 -- 2015.",
+             label = "tbl:ids"),
+      include.rownames = FALSE,
+      booktabs = TRUE,
+      caption.placement = "top")
+
+@
+
+% Notes: Duration has gotten shorter. Etc. Rearrange - End
+% introduction with evidence of erosion, quatify the effect on yield
+% increases.
+
+\newpage
+
+\section{References}\label{references}
+
+\setlength{\parindent}{-0.2in} \setlength{\leftskip}{0.2in}
+\setlength{\parskip}{8pt}
+
+\bibliography{breeding}
+
+\newpage
+
+\section{Supplemental tables and
+figures}\label{supplemental-tables-and-figures}
+
+\beginsupplement
+
+\begin{figure}[H]
+
+<<fig=TRUE, echo=FALSE, width = 6, height = 5>>=
+tars <- c("M202","M206")
+
+plot_data = lapply(tars, function(x) {
+    dd = mod_data[mod_data$id == x & mod_data$site == "RES",]
+    tmp = prep_plot_data(dd$yield_kg / 1000, dd$yrs_in_trial)
+    tmp$yrs_in_trial = as.numeric(row.names(tmp))
+    tmp
+    })
+
+plot(mean ~ yrs_in_trial,
+     data = plot_data[[1]],
+     xlim = c(-6, 32), ylim = c(6.5,14.5),
+     pch = 16, #main = gsub("M","M-", tars[i]),
+     ylab = expression("Yield ( kg ha"^-1~")"),
+     xlab = "Years after release")
+segments(x0 = plot_data[[1]]$yrs_in_trial,
+         y0 = plot_data[[1]]$mean + plot_data[[1]]$se,
+         y1 = plot_data[[1]]$mean - plot_data[[1]]$se)
+points(mean ~ yrs_in_trial,
+       data = plot_data[[2]],
+       pch = 17, col = "grey40")
+segments(x0 = plot_data[[2]]$yrs_in_trial,
+         y0 = plot_data[[2]]$mean + plot_data[[2]]$se,
+         y1 = plot_data[[2]]$mean - plot_data[[2]]$se,
+         col = "grey40")
+abline(lm(yield_kg/1000 ~ yrs_in_trial,
+          data = mod_data[(mod_data$id %in% tars & mod_data$site == "RES"),]),
+       lty = 1, col = "grey50")
+legend("topright", legend = gsub("M","M-", tars),
+       pch = c(16, 17), col = c("black","grey40"))
+
+@
+\caption{The performance of two major cultivars at the CA Rice
+  Experiment Station (RES) since the year each cultivar was
+  released. Both M-202 ($\bullet$) and M-206 ($\blacktriangle$) have similar
+  intercepts, suggesting similar yield at the time of release and an
+  apparent yield advantage of the newer cultivar, M-206, might be
+  explained by a yield loss of M-202 over time. Points represent the
+  mean, and bars (|) the standard error.}
+\label{fig:compare_since_release}
+\end{figure}
+
+\newpage
 
@@
 \end{figure}
-
-The average decrease in yield over time once a cultivar is selected
-(i.e., yield erosion) was estimated as \Sexpr{yrs_coef} kg
-ha\tsups{-1} year\tsups{-1} (90\% CI: \Sexpr{RES_CI[2]} to
-\Sexpr{RES_CI[1]}). The oldest cultivar in the trials had lost an
-average of \Sexpr{round(yrs_coef*-36/1000,1)} Mg ha\tsups{-1} since
-its release. The two most dominate medium grain rice cultivars, M-202
-(widely planted from the late 1980s until the early 2000s) and M-206
-(currently the most widely planted), experienced
-similar trends in yield decline (Fig. \ref{fig:compare_yrs}; Table
-\ref{tbl:coef}). The currently estimated
-\Sexpr{m202_v_206} kg ha\tsups{-1} yield advantage of M-206 over M-202
-can be attributed solely to the erosion of M-202's yield over time
-(Figs \ref{fig:first_yrs} \& \ref{fig:compare_yrs}). Further
-supporting this estimate of yield erosion, both cultivars yielded similarly at their
-respective year of release (Fig. \ref{fig:compare_yrs}), despite the
-fact that M-202 was released 19 years prior to M-206
-(Fig. \ref{fig:compare_yrs}, Table \ref{tbl:ids}).
-
-\section{Discussion}
-
-We do not find conclusive evidence of increased genetic yield
-potential in medium grain rice cultivars released by CCRRF's breeding
-improvement program over the period \Sexpr{min(mod_data[,'year'])} to
-\Sexpr{max(mod_data[,'year'])}. Since the estimated yield improvement
-spans from \Sexpr{rel_CI[1]} to \Sexpr{rel_CI[2]} kg ha\tsups{-1}
-year\tsups{-1}, both increases and decreases in the genetic yield
-potential over the study period are plausible. However, even the
-estimated yield improvement that does not account for yield erosion
-(\Sexpr{only_release_coef} kg ha\tsups{-1} year\tsups{-1};
-Fig. \ref{fig:wo_erosion}) is much lower than previous estimated,
-i.e., 37 to 60 kg ha\tsups{-1} year\tsups{-1} \citep{Omar2014}. Based on the
-estimates from this study, there is less than a 1\% probability that
-the true value is higher than 45 kg ha\tsups{-1}
-year\tsups{-1}. Therefore, we conclude that the CCRRF breeding program
-has likely maintained but not substantially improved the yield
-potential in the face of changing biotic and abiotic pressures.
-
-We estimate that yield erosion has decreased the yield of medium grain
-rice cultivars released by the CCRRF over the last 30+ years by an
-average of \Sexpr{yrs_coef} Mg ha\tsups{-1}, potentially offsetting
-any gains made by the crop improvement program. Biotic and abiotic forces have
-previously been speculated to be responsible for yield erosion, since
-the cultivar's genetics remain stable from the time of selection but
-the conditions the cultivar may experience over time are continuously
-changing \citep{Peng1998, Peng2000, Peng2010}. The trend observed here
-is consistent with previous reports of the erosion of genetic gains of
-cultivar IR8 at IRRI \citep{Peng1998, Peng2000}, suggesting the
-erosion of yield over time is a phenomenon that widely affects rice
-breeding programs. This observation highlights both the need for maintenance
-of plant breeding and the difficulty of increasing genetic yield potential,
-especially in systems such as CA where yield is approaching the
-physiological limit \citep{espe16a}. Without maintenance breeding
-(i.e., continuously selecting cultivars well adapted to current
-conditions), CA rice production systems are predicted to have lost an
-estimated \Sexpr{round(yrs_coef*(-36/1000),1)} Mg ha\tsups{-1} grain
-yield over the study period.
-
-However, a goal of many breeding programs is to not only maintain the
-benchmark yield of the previous ``standard'' but to surpass it,
-especially if we are to meet the target of 50\% increased food
-production by 2050 \citep{Foley2011a}. Given the estimated yield
-erosion, to hit an ambitious target of 0.5 Mg ha\tsups{-1}
-(approximately 5\%) increase in yield per release (with roughly 4 to 5
-years between releases), a new cultivar needs to yield an additional
-\Sexpr{round(yrs_coef*-4,-1)} to \Sexpr{round(yrs_coef*-5,-1)} kg
-ha\tsups{-1} in side-by-side comparison in addition to the 0.5 Mg
-ha\tsups{-1} target improvement. Often side-by-side yield improvements
-show an improvement in grain yield of 2 to 3\% per release, in which
-case no absolute improvements in yield have been made beyond
-maintaining yield. The evidence here is consistent with the
-observation that the genetic yield potential has not increased
-substantially in the last 30+ years (Fig. \ref{fig:first_yrs}).
-
-\begin{figure}[H]
-<<fig = TRUE, echo = FALSE, width=5, height=5>>=
-
-plot(1, type = "n",
-     # xaxt = "n",
-     xlab = "Years",
-     ylab = expression("Grain yield ( Mg ha"^-1~")"),
-     ylim = c(8.5,11.5),
-     xlim = c(0,40),
-     xaxs = "i")
-
-# arrows(12, 8.2, 27, 8.2, xpd = TRUE, lwd = 2)
-abline(a = 10, b = fixef(RES_full)[3], lty = 2, lwd = 2)
-text(25, 10.85, "(3) target above maintenance", cex = 0.8,
-     srt = 18)#((atan(fixef(RES_full)[3] + 0.06) * 180))/pi)
-abline(h = 10, lty = 3, lwd = 2)
-text(25, 10.1, "(1) current situation", cex = 0.8)
-abline(a = 10, b = fixef(RES_full)[3] + 0.06, lty = 4, lwd = 2)
-text(25, 9.35, "(2) no maintenance breeding", cex = 0.8,
-     srt = -18)
-
-@ 
-\caption{A conceptual diagram highlighting three scenarios: (1)
-maintenance breeding, where losses over time are offset by
-improvements via plant improvement, (2) no maintenance breeding, where
-grain yield slowly erodes over time (as estimated from this study),
-and (3) absolute improvement in grain yield (based on ``typical''
-yield goals of 3\% increase per release), where gains in grain yield
-exceed the losses over time.}
-\label{fig:theory}
-\end{figure}
-
-California state average yields have increased slowly
-despite there being little to no gains in the breeding trials
-(Fig. \ref{fig:state_yield}). The evidence here suggests mechanisms
-other than increased genetic yield potential are responsible for these gains
-(Fig. \ref{fig:compare_yrs}). Improvement in agronomic management,
-including more efficient nutrient management \citep{Lundy2012,
-Lundy2015, linquist2009}, weed management \citep{Saito2008,
-Pittelkow2012, linquist2008}, and other factors may all be contributing to
-increases in the state average yield. However, these gains are far less
-than those experienced during the preceeding era, and by their
-nature agronomic gains cannot surpass the genetic yield potential
-\citep{Fischer2015}. Judging from yields achieved at the RES
-(Figs. \ref{fig:RES_year} \& \ref{fig:compare_yrs}), a
-researcher-managed production system implementing best management
-practices in more controlled situations compared to farmers' fields, state
-average yields (approximately 10 Mg ha\tsups{-1} at the time of
-writing; Fig. \ref{fig:state_yield}) may be approaching the current
-genetic limit. Assuming that the level of management possible in small
-experimental plots cannot be attained in farmers' fields and hence
-experimental yields represent the attainable yield ceiling
-\citep{Fischer2015}, agronomic gains in farmer's fields are nearing
-their potential, consistent with physiological model-based estimates
-of the attainable yield gap in CA rice production systems
-\citep{espe16a}.
-
-There are several cultivars that were found to have better than
-average yields, even after correcting for yield erosion. However,
-these cultivars (M-201, M-204, M-205, and possibly M-209;
-Fig. \ref{fig:ids_comp})) are not widely adapted. It is known that
-cool temperatures during the booting stage can have a large impact on
-grain yields \citep{espe2017}, and the RES
-experiences warmer temperatures than many other rice growing areas in
-CA \citep{espe2017}. These cultivars may perform well at the RES but
-not widely in CA. Therefore, it should be noted that while our
-investigation has been concerned about increasing yield potential, the
-estimates here might not capture yield stability. Yield stability is
-notoriously difficult to quantify \citep[e.g.,][]{lobell2011}, hence
-further investigation is needed to explore how yield stability has
-changed over the study period.
-
-To be clear, our investigation here is narrowly focused on the
-rough-rice/field grain yield. Other advancements due to plant
-improvement, including increased quality \citep{mckenzie1993} and
-milling yield \citep{mckenzie1994}, have been made over the study
-period and are important. Also, crop duration has been reduced. For
-example, during this study period, M-202 and M-206 have been the
-predominate cultivars grown in CA, and while their yield potentials
-were similar at the time of release (Fig. 5), M-206 matures about 6
-days earlier than M-202 (Table \ref{tbl:ids}). Shorter duration is very important
-in CA due to an already short growing season and water limitations
-(shorter duration cultivars would presumably require less irrigation
-water).
-
-Planning to meet future
-demand necessitates increases in grain yield \citep{Godfray2010,
-Foley2011a, Mueller2012}. Meeting the challenge to produce more grain
-without expansion of land under cultivation will rely on new genetics
-with increased yield potential. The evidence here suggests that to
-accomplish true increases over time, the improvement target needs take
-into account decreases in the current ``standard'' cultivars' yield
-over time (Fig. \ref{fig:theory}). While side-by-side comparisons will
-continue to be useful, comparisons across years to each cultivar's
-benchmark yield at the time of release should provide information on
-the absolute improvements being made in yield potential over time. New
-technologies may need to be adapted to further increase yield
-potential, including hybrid rice.
-
-\section{Conclusions}
-
-Improvement in grain yield is essential to meeting the challenges of
-feeding a growing population in a changing world. To achieve these
-improvements, we discuss here that (1) continuous maintenance via plant
-improvement is required, without which yields would decrease over
-time, (2) the challenge of improving genetic yield potential is
-substantial, since absolute gains require improvement in yield in
-excess of the maintenance threshold, and (3) increasing yield through
-improved management (agronomy) cannot exceed the ceiling of genetic
-yield potential. Taken together, these conclusions underscore both the
-necessity and substantial challenges facing rice breeding. We suggest
-here that historical benchmarks be used in to evaluate promising rice
-cultivars in addition to side-by-side comparisons to maintain
-perspective of the absolute yield gains being made.
-
 \newpage
-
-\section{Tables}
-
-<<echo=FALSE, results=tex>>=
-
-CIs = posterior_interval(RES_full)[1:3,] * scl
-CIs[1,] = round((CIs[1,] + center)/1000, 1)
-CIs = round(CIs, 1)
-med_eff = fixef(RES_full) * scl
-med_eff[1] = round((med_eff[1] + center)/1000, 1)
-med_eff = round(med_eff, 1)
-
-RES_coef = cbind(med_eff, paste(CIs[,1],CIs[,2], sep = " to "))
-rownames(RES_coef) = c("Intercept (Mg ha\\tsups{-1})",
-                       "Yield improvement (kg ha\\tsups{-1} year\\tsups{-1})",
-                       "Yield erosion (kg ha\\tsups{-1} year\\tsups{-1})")
-colnames(RES_coef) = c("Median", "90\\% Credible interval")
-
-print(xtable(RES_coef,
-             caption = "The estimated median fixed effects and 90\\% credible intervals for the improvement in yield of medium grain rice cultivars developed by the California Cooperative Rice Research Foundation, California, USA over the time period 1984 -- 2016.",
-             label = "tbl:coef",
-             align = "lcc"),
-      caption.placement="top",
-      booktabs=TRUE,
-      sanitize.colnames.function = function(x){x},
-      sanitize.rownames.function = function(x){x})
-@
-\newpage
-
-<<echo=FALSE, results=tex>>=
-ids = aggregate(release_yr ~ id, data = mod_data, unique)
-ids$dth = as.integer(round(aggregate(dth ~ id, data = mod_data, mean, na.rm = TRUE)$dth, 0))
-colnames(ids) = c("Cultivar", "Release Year", "Days to 50% Heading")
-ids = ids[order(ids[,2]),]
-ids[,2] = as.integer(ids[,2])
-ids[,1] = gsub("M", "M-", ids[,1])
-
-print(xtable(ids,
-             caption = "All named medium grain rice cultivars and their associated release year planted at the Rice Experiment Station (RES), Biggs, California from 1984 -- 2015.",
-             label = "tbl:ids"),
-      include.rownames = FALSE,
-      booktabs = TRUE,
-      caption.placement = "top")
-
-@
-
-% Notes: Duration has gotten shorter. Etc. Rearrange - End
-% introduction with evidence of erosion, quatify the effect on yield
-% increases.
-
-\newpage
-
-\section{References}\label{references}
-
-\setlength{\parindent}{-0.2in} \setlength{\leftskip}{0.2in}
-\setlength{\parskip}{8pt}
-
-\bibliography{breeding}
-
-\newpage
-
-\section{Supplemental tables and
-figures}\label{supplemental-tables-and-figures}
-
-\beginsupplement
-
-\begin{figure}[H]
-
-<<fig=TRUE, echo=FALSE, width = 6, height = 5>>=
-
-plot_data = lapply(tars, function(x) {
-    dd = mod_data[mod_data$id == x & mod_data$site == "RES",]
-    tmp = prep_plot_data(dd$yield_kg / 1000, dd$yrs_in_trial)
-    tmp$yrs_in_trial = as.numeric(row.names(tmp))
-    tmp
-    })
-
-plot(mean ~ yrs_in_trial,
-     data = plot_data[[1]],
-     xlim = c(-6, 32), ylim = c(6.5,14.5),
-     pch = 16, #main = gsub("M","M-", tars[i]),
-     ylab = expression("Yield ( kg ha"^-1~")"),
-     xlab = "Years after release")
-segments(x0 = plot_data[[1]]$yrs_in_trial,
-         y0 = plot_data[[1]]$mean + plot_data[[1]]$se,
-         y1 = plot_data[[1]]$mean - plot_data[[1]]$se)
-points(mean ~ yrs_in_trial,
-       data = plot_data[[2]],
-       pch = 17, col = "grey40")
-segments(x0 = plot_data[[2]]$yrs_in_trial,
-         y0 = plot_data[[2]]$mean + plot_data[[2]]$se,
-         y1 = plot_data[[2]]$mean - plot_data[[2]]$se,
-         col = "grey40")
-abline(lm(yield_kg/1000 ~ yrs_in_trial,
-          data = mod_data[(mod_data$id %in% tars & mod_data$site == "RES"),]),
-       lty = 1, col = "grey50")
-legend("topright", legend = gsub("M","M-", tars),
-       pch = c(16, 17), col = c("black","grey40"))
-
-@
-\caption{The performance of two major cultivars at the CA Rice
-  Experiment Station (RES) since the year each cultivar was
-  released. Both M-202 ($\bullet$) and M-206 ($\blacktriangle$) have similar
-  intercepts, suggesting similar yield at the time of release and an
-  apparent yield advantage of the newer cultivar, M-206, might be
-  explained by a yield loss of M-202 over time. Points represent the
-  mean, and bars (|) the standard error.}
-\label{fig:compare_since_release}
-\end{figure}
-
-\newpage
-
 % \begin{figure}[H]

### 2017-10-29 output/breed_paper.Rnw
Updated and revised for grammar, etc.

@@
 Rice Experiment Station. Results show an apparent increase in yield in
-side-by-side comparisons between old and new rice varieties was
-observed when the yield erosion was not considered. However, the
-yields of older varieties eroded at an estimated rate of 29.3 kg
-ha\tsups{-1} year\tsups{-1} (90\% credible interval -4.4 to -53.3)
-after initial selection. Once this effect was considered, the yield
-advantage of newer cultivars over old was uncertain (-3.3 kg
-ha\tsups{-1} year\tsups{-1}, 90\% credible interval -36.1 to
-31.5). These results highlight the importance of continuous crop
-improvement and deployment of new cultivars as yield will decline over
-time without it. To increase the genetic yield potential of rice in
-the face of this phenomenon, higher yield targets are needed to
-counter-act the decrease over time in the current cultivars.
+side-by-side comparisons between old and new rice varieties when the
+yield erosion was not considered. However, the yields of older
+varieties eroded at an estimated rate of 29.3 kg ha\tsups{-1}
+year\tsups{-1} (90\% credible interval -4.4 to -53.3) after initial
+selection. Once this effect was considered, the yield advantage of
+newer cultivars over old was uncertain (-3.3 kg ha\tsups{-1}
+year\tsups{-1}, 90\% credible interval -36.1 to 31.5). This suggests
+that side-by-side comparisons reflected the decreasing yields of older
+cultivars more than the increasing yield of newer ones. These results
+highlight the importance of continuous crop improvement and deployment
+of new cultivars as yield will decline over time without it. To
+increase the genetic yield potential of rice in the face of this
+phenomenon, higher yield targets are needed to counter-act the
+decrease over time in the current cultivars.
 
@@
 \citep{Foley2011a, Godfray2010, Mueller2012}, which has necessitated
-harvesting more grain per unit land area \citep{Lobell2009,
+harvesting more grain per unit land area~\citep{Lobell2009,
 Tittonell2014}. Historically, agricultural research has been
@@
 crops. However, many production systems are experiencing plateaus in
-grain yield \citep{Grassini2013, Peng2010}. If this trend continues,
+grain yield~\citep{Grassini2013, Peng2010}. If this trend continues,
 improvements per unit land area are no longer possible, and an
@@
 California (CA) rice represents one such production system. Rice is
-grown primarily in the Sacramento Valley which is characterized by
-having a Mediterranean climate, with long days, dry growing season,
-and relatively free of pests and diseases, which lead to some of the
-highest yields in the world~\citep{FAO2016}. Production in CA is
-predominately focused on premium quality medium grain
+grown primarily in the Sacramento Valley, which is characterized by
+having a Mediterranean climate with long days, a dry growing season,
+and relatively free of pests and diseases. These conditions lead to
+some of the highest yields in the world~\citep{FAO2016}. Production in
+CA is predominately focused on premium quality medium grain
 \textit{japonica} cultivars (e.g., CalRose rice), and rice from CA is
@@
 slow and \citet{espe16a} reported that indeed this was the case
-with yields on average increasing by about 50 to 62 kg ha\tsups{-1} between
-1999 and 2014.
+with yields on average increasing by about 50 to 62 kg ha\tsups{-1}
+year\tsups{-1} between 1999 and 2014.
 
@@
 popular cultivar M-202 was released in 1985, it was reported to yield
-an average of 11 Mg ha\tsups{-1} \citep[14\% moisture content
+an average of 11 Mg ha\tsups{-1}~\citep[14\% moisture content
 (MC);][]{m202}. In 2015, the reported average yield of the newest
-released cultivar M-209 was 10.8 Mg ha\tsups{-1} \citep[14\%
+released cultivar M-209 was 10.8 Mg ha\tsups{-1}~\citep[14\%
 MC;][]{m209}. However, in annual breeding reports and varietal release
-announcements, based on yield performance in side-by-side variety
-trails, data show that new varieties usually out yield the older
+announcements, data based on yield performance in side-by-side variety
+trails show that new varieties usually out yield the older
 varieties. In other rice systems, a decrease in a cultivar's yield
-performance over time has been observed \citep{Peng1998, Peng2010,
-Peng2000, Datta1995}, often termed yield erosion \citep{Peng1998}. It
+performance over time has been observed~\citep{Peng1998, Peng2010,
+Peng2000, Datta1995}, often termed yield erosion~\citep{Peng1998}. It
 has been speculated that this yield erosion is due to the inability of
@@
 yields erode over time then apparent yield increases in side-by-side
-yield comparison tests will not reflect true yield increases.o
+yield comparison tests will not reflect true or absolute yield
+increases over time.
 
@@
 efforts to develop improved rice cultivars adapted to CA since
-1969. The California Statewide Variety Trails evaluates current and
+1969. The California Statewide Variety Trails evaluate current and
 promising varieties at the RES and six to eight on-farm trials around
@@
 precipitation events. Soils at the RES are classified as
-Esquon-Neerdobe clays with roughly 2\% soil organic matter
-\citep{ssurgo} in the top 15cm.
+Esquon-Neerdobe clays with roughly 2\% soil organic matter in the top 15cm~\citep{ssurgo}.
 
@@
 \Sexpr{max(mod_data[,'year'])} plot level observations (replicated
-3--4 times in a CRB design) were collected leading to a final data set
+3 to 4 times in a CRB design) were collected leading to a final data set
 of \Sexpr{format(nrow(mod_data), big.mark = ',')} observations
@@
 Experimental plots at the RES are managed similar to University of California
-Cooperative Extension following prescribed best management
+Cooperative Extension prescribed best management
 practice for land preparation, fertility rates, and pest
-and disease control \citep{ucce}. As is typical for water-seeded
-systems, for planting, rice seed is pre-germinated prior
+and disease control~\citep{ucce}. As is typical for water-seeded
+systems, the plots were planted by pre-germinating rice seed prior
 to direct-seeding into pre-flooded fields. Plots were either 14.0 or
@@
 maturity using a small-plot combine. All yields were converted to 14\%
-moisture content prior to reporting.
+MC prior to reporting.
 
@@
 Data were processed and models were fit in R, an environment for
-statistical computing \citep{R}. Models were fit using the `rstanarm'
-package \citep{rstanarm}, an interface to Stan, a language for
-probabilistic programming \citep{stan}. All model diagnostics,
+statistical computing~\citep{R}. Models were fit using the `rstanarm'
+package~\citep{rstanarm}, an interface to Stan, a language for
+probabilistic programming~\citep{stan}. All model diagnostics,
 including $\hat{R}$, effective size, and posterior predictive checks
@@
 full model was compared to models containing only the years since
-release or the year of release using the `loo' package \citep{loo}, an
+release or the year of release using the `loo' package~\citep{loo}, an
 efficient means of conducting leave-one-out cross validation
@@
 90\% CI is more stable to perpetuation's and hence is preferable to the 95\%
-interval \citep{rstanarm}. To verify the estimates, the analysis was
+interval~\citep{rstanarm}. To verify the estimates, the analysis was
 also conducted using classical methods (frequentist mixed-effects
@@
 \section{Results}\label{results}
-% Invert this section, then talk about yield improvement accounting
-% for yield erosion.
 
@@
 
-There was agreement between the estimates from the Bayesian and
-classical approaches.  The Bayesian estimates are presented here,
+There was general agreement between the estimates from the Bayesian and
+classical approaches.  The Bayesian estimates are presented here due
+to the inherent flexibility of Bayesian inference, 
 though the classical estimates can be found in the supplemental
@@
 \ref{fig:RES_year}; Table \ref{tbl:coef}). The oldest cultivar in the
-trials (\Sexpr{unique(mod_data[which.min(mod_data[,"year"]),"id"])}
+trials (\Sexpr{gsub("M","M-",unique(mod_data[which.min(mod_data[,"release_yr"]),"id"]))})
 had lost an average of \Sexpr{round(yrs_coef*-36/1000,1)} Mg
@@
 as the most widely grown cultivar in CA and the current estimated 560
-kg ha-1 yield advantage of M-206 over M-202 can be attributed solely
+kg ha\tsups{-1} yield advantage of M-206 over M-202 can be attributed solely
 to the erosion of M-202's yield over time (Figs \ref{fig:first_yrs} \&
@@
 release (Fig. \ref{fig:compare_yrs}), despite the fact that M-202 was
-released 19 years prior to M-206 (Fig. \ref{fig:compare_yrs}, Table
+released 19 years prior to M-206 (Fig. \ref{fig:compare_yrs}; Table
 \ref{tbl:ids}).
 
-When yield erosion was not corrected for in the analysis (i.e.,
+When yield erosion is not corrected for in the analysis (i.e.,
 omitted from the model), there is an apparent small increase per year
@@
 with over 70\% of the area in one of these cultivars during the study
-period ~citep{ucce}. Importantly, not all medium grain cultivars that were
+period~\citep{ucce}. Importantly, not all medium grain cultivars that were
 released were purported to have higher yields than previous
@@
 similar yields to more broadly adapted cultivars. Examples of this
-include M-208, which had resistance to blast disease (scientific name)
+include M-208, which had resistance to blast disease (\textit{Magnaporthe grisea})
 and M-402 which is a premium quality variety. Other varieties such as
@@
 \Sexpr{max(mod_data[,'year'])}, which has stayed relatively constant
-at \Sexpr{round(mean(mod_data[,"yield_kg"]), 2)} Mg
+at \Sexpr{round(mean(mod_data[,"yield_kg"])/1000, 1)} Mg
 ha\tsups{-1}. Since the estimated yield improvement spans from
@@
 medium grain rice cultivars released by the CCRRF over the last 30+
-years by an average of \Sexpr{yrs_coef} Mg ha\tsups{-1}, potentially
+years by an average of \Sexpr{yrs_coef} Mg ha\tsups{-1} year\tsups{-1}, potentially
 offsetting any gains made by the crop improvement program. However, it
@@
 
+There are several cultivars that were found to have better than
+average yields, even after correcting for yield erosion. However,
+these cultivars (M-201, M-204, M-205, and possibly M-209;
+Fig. \ref{fig:ids_comp})) are not widely adopted. It is known that
+cool temperatures during the booting stage can have a large impact on
+grain yields~\citep{espe2017}, and the RES experiences warmer
+temperatures than many other rice growing areas in CA
+\citep{espe2017}. These cultivars may perform well at the RES but not
+widely in CA. Therefore, it should be noted that while our
+investigation has been concerned about increasing yield potential, the
+estimates here might not capture yield stability. Yield stability is
+notoriously difficult to quantify~\citep[e.g.,][]{lobell2011}, hence
+further investigation is needed to explore how yield stability has
+changed over the study period.
+
 A goal of many breeding programs is to not only maintain the benchmark
@@
 \citep{Foley2011a}. Given the estimated yield erosion, to hit an
-ambitious target of 0.5 Mg ha\tsups{-1} (approximately 5\%) increase
+ambitious target of 500 kg ha\tsups{-1} (approximately 5\%) increase
 in yield per release (with roughly 4 years between releases), a new
 cultivar needs to yield an additional \Sexpr{round(yrs_coef*-4,-1)} kg
-ha\tsups{-1} in side-by-side comparison in addition to the 0.5 Mg
+ha\tsups{-1} in side-by-side comparison in addition to the 500 kg
 ha\tsups{-1} target improvement. Often side-by-side yield improvements
-show an improvement in grain yield of 2\% per release, in which case
+show an improvement in grain yield of 2 to 3\% per release, in which case
 no absolute improvements in yield have been made beyond maintaining
@@
 
-There are several cultivars that were found to have better than
-average yields, even after correcting for yield erosion. However,
-these cultivars (M-201, M-204, M-205, and possibly M-209;
-Fig. \ref{fig:ids_comp})) are not widely adopted. It is known that
-cool temperatures during the booting stage can have a large impact on
-grain yields \citep{espe2017}, and the RES
-experiences warmer temperatures than many other rice growing areas in
-CA \citep{espe2017}. These cultivars may perform well at the RES but
-not widely in CA. Therefore, it should be noted that while our
-investigation has been concerned about increasing yield potential, the
-estimates here might not capture yield stability. Yield stability is
-notoriously difficult to quantify \citep[e.g.,][]{lobell2011}, hence
-further investigation is needed to explore how yield stability has
-changed over the study period.
-
 To be clear, our investigation here is narrowly focused on the
 rough-rice/field grain yield. Other advancements due to plant
-improvement, including increased quality \citep{mckenzie1993} and
-milling yield \citep{mckenzie1994}, have been made over the study
+improvement, including increased quality~\citep{mckenzie1993} and
+milling yield~\citep{mckenzie1994}, have been made over the study
 period and are important. Also, crop duration has been reduced. For
@@
 (Fig. \ref{fig:compare_yrs}). Improvement in agronomic management,
-including more efficient nutrient management \citep{Lundy2012,
-Lundy2015, linquist2009}, weed management \citep{Saito2008,
+including more efficient nutrient management~\citep{Lundy2012,
+Lundy2015, linquist2009}, weed management~\citep{Saito2008,
 Pittelkow2012, linquist2008}, and other factors may all be contributing to
@@
 
-Planning to meet future
-demand necessitates increases in grain yield \citep{Godfray2010,
-Foley2011a, Mueller2012}. Meeting the challenge to produce more grain
-without expansion of land under cultivation will rely on new genetics
-with increased yield potential. The evidence here suggests that to
-accomplish true increases over time, the improvement target needs take
-into account erosion in the current ``standard'' cultivars' yield
-over time (Fig. \ref{fig:theory}). While side-by-side comparisons will
-continue to be useful, comparisons across years to each cultivar's
-benchmark yield at the time of release should provide information on
-the absolute improvements being made in yield potential over time. New
+Planning to meet future demand necessitates increases in grain
+yield~\citep{Godfray2010, Foley2011a, Mueller2012}. Meeting the
+challenge to produce more grain without expansion of land under
+cultivation will rely on new genetics with increased yield
+potential. The evidence here suggests that to accomplish true
+increases over time, the improvement target needs take into account
+erosion in the current ``standard'' cultivars' yield over time
+(Fig. \ref{fig:theory}). While side-by-side comparisons will continue
+to be useful, comparisons across years to each cultivar's benchmark
+yield at the time of release should provide information on the
+absolute improvements being made in yield potential over time. New
 technologies may need to be adapted to further increase yield

### 2017-10-30 output/breed_paper.Rnw
Final edits.

@@
 
+MC: Moisture Content
+
 \newpage
@@
 released cultivar M-209 was 10.8 Mg ha\tsups{-1}~\citep[14\%
-MC;][]{m209}. However, in annual breeding reports and varietal release
-announcements, data based on yield performance in side-by-side variety
-trails show that new varieties usually out yield the older
-varieties. In other rice systems, a decrease in a cultivar's yield
+MC;][]{m209}. However, in annual breeding reports and cultivar release
+announcements, data based on yield performance in side-by-side cultivar
+trials show that new cultivars usually out yield the older
+cultivars. In other rice systems, a decrease in a cultivar's yield
 performance over time has been observed~\citep{Peng1998, Peng2010,
@@
 erosion may explain the apparent contradiction above, in which
-side-by-side yield comparisons show yield improvements in varieties
+side-by-side yield comparisons show yield improvements in cultivars
 while overall there may be little change in yield potential. To
@@
 1969. The California Statewide Variety Trails evaluate current and
-promising varieties at the RES and six to eight on-farm trials around
+promising cultivars at the RES and six to eight on-farm trials around
 the state each year. To avoid potential complications due to site,
@@
 
-There was general agreement between the estimates from the Bayesian and
-classical approaches.  The Bayesian estimates are presented here due
-to the inherent flexibility of Bayesian inference, 
-though the classical estimates can be found in the supplemental
+There was general agreement between the estimates from the Bayesian
+and classical approaches.  The Bayesian estimates are presented here
+due to the inherent flexibility of Bayesian inference, though the
+classical estimates can be found in the supplemental
 material. Comparing the cultivars' yield during each's initial time
@@
 \Sexpr{round(plot_data["M207", "mean"], 2)} (M-207) to
-\Sexpr{round(plot_data["M205", "mean"], 2)} Mg ha\tsups{-1}
-(M-205). Evaluating these initial yields over time indicates there was
-no evidence of genetic yield improvements during the 30-year study
-period (Fig. 2). Individual exceptions to this were two varieties
-(M-204 and M-205) that had higher yields than the others.
+\Sexpr{round(plot_data["M205", "mean"], 2)} Mg ha\tsups{-1} (M-205)
+(Fig \ref{fig:RES_year}. Evaluating these initial yields over time
+indicates there was no evidence of genetic yield improvements during
+the 30-year study period (Fig. \ref{fig:RES_year}). Individual exceptions to this were
+two cultivars (M-204 and M-205) that had higher yields than the
+others.
 
@@
 
-<<fig=TRUE, echo=FALSE>>=
+<<fig=TRUE, echo=FALSE, width = 6, height = 5>>=
 yrs_in <- -10:40
@@
 
-The yields of each cultivar were evaluated over time relative to the
-time when the cultivar was released using the full model which
-accounts for both yield erosion and yield improvements over time. The
-average decrease in yield over time (i.e., yield erosion) once a
+The yields of each cultivar over time were evaluated relative to when
+the cultivar was released using the full model which accounts for both
+yield erosion and yield improvements over time. The average decrease
+in yield over time (i.e., yield erosion) once a
 cultivar was selected was estimated \Sexpr{yrs_coef} kg ha\tsups{-1}
@@
 trials (\Sexpr{gsub("M","M-",unique(mod_data[which.min(mod_data[,"release_yr"]),"id"]))})
-had lost an average of \Sexpr{round(yrs_coef*-36/1000,1)} Mg
+has lost an average of \Sexpr{round(yrs_coef*-36/1000,1)} Mg
 ha\tsups{-1} since its release. The most dominate medium grain rice
@@
 
+\begin{figure}[h]
+
+<<fig=TRUE, echo=FALSE, width = 6, height = 5>>=
+tars <- c("M202","M206")
+
+plot_data = lapply(tars, function(x) {
+    dd = mod_data[mod_data$id == x & mod_data$site == "RES",]
+    tmp = prep_plot_data(dd$yield_kg / 1000, dd$yrs_in_trial)
+    tmp$yrs_in_trial = as.numeric(row.names(tmp))
+    tmp
+    })
+
+plot(mean ~ yrs_in_trial,
+     data = plot_data[[1]],
+     xlim = c(-6, 32), ylim = c(6.5,14.5),
+     pch = 16, #main = gsub("M","M-", tars[i]),
+     ylab = expression("Yield ( kg ha"^-1~")"),
+     xlab = "Years after release")
+segments(x0 = plot_data[[1]]$yrs_in_trial,
+         y0 = plot_data[[1]]$mean + plot_data[[1]]$se,
+         y1 = plot_data[[1]]$mean - plot_data[[1]]$se)
+points(mean ~ yrs_in_trial,
+       data = plot_data[[2]],
+       pch = 17, col = "grey40")
+segments(x0 = plot_data[[2]]$yrs_in_trial,
+         y0 = plot_data[[2]]$mean + plot_data[[2]]$se,
+         y1 = plot_data[[2]]$mean - plot_data[[2]]$se,
+         col = "grey40")
+abline(lm(yield_kg/1000 ~ yrs_in_trial,
+          data = mod_data[(mod_data$id %in% tars & mod_data$site == "RES"),]),
+       lty = 2, col = "grey50")
+legend("topright", legend = gsub("M","M-", tars),
+       pch = c(16, 17), col = c("black","grey40"))
+text(5, 6.7, bquote(.(yrs_coef)~kg~ha^-1~year^-1))
+
+@
+\caption{The performance of two popular cultivars over time. The
+  dashed line is the estimated yield erosion as estimated by the full
+  model. Points ($\bullet$) represent the mean, and bars (|) the standard
+  error of the observed data.}
+\label{fig:compare_yrs}
+\end{figure}
+
 When yield erosion is not corrected for in the analysis (i.e.,
-omitted from the model), there is an apparent small increase per year
+omitted from the model), there is estimated to be a small increase per year
 (approximately \Sexpr{only_release_coef} kg ha\tsups{-1}
 year\tsups{-1}; 90\% CI \Sexpr{only_release_CI[1]} to
-\Sexpr{only_release_CI[2]}) (Fig. \ref{fig:wo_erosion}).  However,
-using the full model which corrected for both the year of release and
-yield erosion, indicates that the estimated change in grain yield per
-year in the CCRRF's breeding trials at the RES was
-\Sexpr{release_yr_coef} kg ha\tsups{-1} year\tsups{-1}, with both
-small positive and negative trends within the credible interval (90\%
-CI: \Sexpr{rel_CI[1]} to \Sexpr{rel_CI[2]})
-(Fig. \ref{fig:first_yrs}).
+\Sexpr{only_release_CI[2]}) (Fig. \ref{fig:wo_erosion}). However, the
+full model which corrected for both the year of release and yield
+erosion indicates that the estimated change in grain yield per year
+in the CCRRF's breeding trials at the RES was \Sexpr{release_yr_coef}
+kg ha\tsups{-1} year\tsups{-1}, with both small positive and negative
+trends within the credible interval (90\% CI: \Sexpr{rel_CI[1]} to
+\Sexpr{rel_CI[2]}) (Fig. \ref{fig:first_yrs}).
 
@@
 
-\begin{figure}[H]
+\begin{figure}[h]
 
@@
 \begin{figure}[H]
-<<fig=TRUE, echo=FALSE, width = 6, height = 5>>=
+<<fig=TRUE, echo=FALSE>>=
 
@@
 @
-% Change from a./b. to accounting for erosion, More details about how
-% yield was corrected - 
 \caption{The relationship between yield and the release year of medium
@@
 include M-208, which had resistance to blast disease (\textit{Magnaporthe grisea})
-and M-402 which is a premium quality variety. Other varieties such as
+and M-402 which is a premium quality cultivar. Other cultivars such as
 M-205 and M-209, while high yielding, do not have the cold tolerance
@@
 been breed for broad adaption and high yield potential (M-202, M-206
-and M-105). Yields of these varieties at time of release have all
+and M-105). Yields of these cultivars at time of release have all
 yielded between
@@
 (Fig. \ref{fig:first_yrs}). While these yields have not increased,
-M-205 did have higher yield potential; however, this variety is not
+M-205 did have higher yield potential; however, this cultivar is not
 broadly adapted due to poor cold tolerance.
@@
 
-\begin{figure}[H]
+\begin{figure}[h]
 <<fig = TRUE, echo = FALSE, width=5, height=5>>=
@@
 \begin{figure}[H]
-
-<<fig=TRUE, echo=FALSE, width = 6, height = 5>>=
-tars <- c("M202","M206")
-
-plot_data = lapply(tars, function(x) {
-    dd = mod_data[mod_data$id == x & mod_data$site == "RES",]
-    tmp = prep_plot_data(dd$yield_kg / 1000, dd$yrs_in_trial)
-    tmp$yrs_in_trial = as.numeric(row.names(tmp))
-    tmp
-    })
-
-plot(mean ~ yrs_in_trial,
-     data = plot_data[[1]],
-     xlim = c(-6, 32), ylim = c(6.5,14.5),
-     pch = 16, #main = gsub("M","M-", tars[i]),
-     ylab = expression("Yield ( kg ha"^-1~")"),
-     xlab = "Years after release")
-segments(x0 = plot_data[[1]]$yrs_in_trial,
-         y0 = plot_data[[1]]$mean + plot_data[[1]]$se,
-         y1 = plot_data[[1]]$mean - plot_data[[1]]$se)
-points(mean ~ yrs_in_trial,
-       data = plot_data[[2]],
-       pch = 17, col = "grey40")
-segments(x0 = plot_data[[2]]$yrs_in_trial,
-         y0 = plot_data[[2]]$mean + plot_data[[2]]$se,
-         y1 = plot_data[[2]]$mean - plot_data[[2]]$se,
-         col = "grey40")
-abline(lm(yield_kg/1000 ~ yrs_in_trial,
-          data = mod_data[(mod_data$id %in% tars & mod_data$site == "RES"),]),
-       lty = 1, col = "grey50")
-legend("topright", legend = gsub("M","M-", tars),
-       pch = c(16, 17), col = c("black","grey40"))
-
-@
-\caption{The performance of two major cultivars at the CA Rice
-  Experiment Station (RES) since the year each cultivar was
-  released. Both M-202 ($\bullet$) and M-206 ($\blacktriangle$) have similar
-  intercepts, suggesting similar yield at the time of release and an
-  apparent yield advantage of the newer cultivar, M-206, might be
-  explained by a yield loss of M-202 over time. Points represent the
-  mean, and bars (|) the standard error.}
-\label{fig:compare_since_release}
-\end{figure}
-
-\newpage
-
-\begin{figure}[H]
 <<fig=TRUE, width = 8, height = 5, out.width = '85%', echo=FALSE>>=
@@
 and similar decreases over time.}
-\label{fig:compare_yrs}
+\label{fig:compare_yrs2}
 \end{figure}
 \newpage
+
 % \begin{figure}[H]

### 2017-11-01 src/h1_model.R
Updates with full random slopes on yrs_in_trial.

@@ -66,6 +66,15 @@ if(compare){
     library(loo)
     compare(loo(RES_full), loo(RES_yrs_fit), loo(RES_rel_fit))
     save(RES_yrs_fit, RES_rel_fit, file = "../output/bivar_fit.Rda")
+    # Add random slopes for ID
+    RES_fuller = stan_glmer(yield_cs ~ (1|year) + (1 + yrs_in_trial|id) +
+                                release_yr_c + yrs_in_trial,
+                      data = mod_data, iter = 2000, cores = 2L,
+                      seed = 789)
+
+    compare(loo(RES_full), loo(RES_yrs_fit), loo(RES_rel_fit), loo(RES_fuller))
+    
+
 }
 
 

### 2017-11-12 output/breed_paper.Rnw
Updates with full random slopes on yrs_in_trial.
Updated supplemental analysis.

@@
 \cortext[corresauthor]{Corresponding author} \ead{mespe@ucdavis.edu}
-\author[ucd]{David Mackill}
 \author[ucd]{Jim E. Hill}
@@
 \author[ucce]{Randall Mutters}
+\author[ucd]{David Mackill}
 \author[ucd]{Chris van Kessel}
@@
 \Sexpr{round(plot_data["M205", "mean"], 2)} Mg ha\tsups{-1} (M-205)
-(Fig \ref{fig:RES_year}. Evaluating these initial yields over time
+(Fig \ref{fig:first_yrs}. Evaluating these initial yields over time
 indicates there was no evidence of genetic yield improvements during
-the 30-year study period (Fig. \ref{fig:RES_year}). Individual exceptions to this were
+the 30+ year study period (Fig. \ref{fig:RES_year}). Individual exceptions to this were
 two cultivars (M-204 and M-205) that had higher yields than the
@@
 and
-\Sexpr{round(min(by(tmp,tmp[,"id"],function(x)mean(x[,"yield_kg"]))/1000),2)}
+\Sexpr{round(max(by(tmp,tmp[,"id"],function(x)mean(x[,"yield_kg"]))/1000),2)}
 Mg ha\tsups{-1} with no indication of yields increasing

### 2017-11-12 output/freq.Rmd
Commit before inverting the sections in results.
Updated supplemental analysis.

@@
 title: "Supplemental material"
-date: "Oct. 1, 2017"
+date: \today{}
 geometry: margin=1in
+header-includes:
+   - \usepackage{dcolumn}
 ---
 
-```{r setup, echo = FALSE}
+```{r setup, echo = FALSE, message = FALSE, warning = FALSE}
 load("../data/model_data.rda")
 library(lme4)
-
+library(stargazer)
+library(xtable)
 ```
@@
 through plant breeding are offset by inherent yield declines over
-time" (Espe et al, 2017) using classical/frequentist statistics to
-verify the results are consistent regardless of statistical method.
+time" (Espe et al, 2017) using classical/frequentist mixed-effects
+models. The purpose of this document to verify the results are
+consistent regardless of statistical method used (i.e., a "sanity
+check"). As the table below show, the estimates are indeed similar. As
+such, please refer to the main manuscript for discussion of the model
+estimates.
 
-The complete data, code, and manuscript is available from \url{https://github.com/mespe/rice_advances}.
+For each model, the structure is the same as used in the Bayesian
+analysis (see main manuscript for details). The models were fit in R
+using the 'lme4' package (Bates et al., 2015). The complete data,
+code, and manuscript is available from
+[https://github.com/mespe/rice_advances](https://github.com/mespe/rice_advances).
 
-# Model excluding yield erosion
-
-```{r}
+```{r, echo = FALSE, message = FALSE, warning = FALSE, results = "asis"}
 no_erosion = lmer(yield_kg ~ release_yr +
@@
 
-summary(no_erosion)
-confint(no_erosion, parm = c("release_yr"),
-        method = "boot", level = 0.9)
+stargazer(no_erosion, title = "Model excluding yield erosion.",
+          ci = TRUE, ci.level = 0.90, float = TRUE,
+          covariate.labels = c("Year of release", "Intercept"),
+          dep.var.caption = "Response",
+          dep.var.labels = "Yield ( kg/ha )",
+          digits = 1, header = FALSE
+          )
+
 ```
 
-# Model including yield erosion
-
-```{r}
+```{r, echo = FALSE, results = "asis"}
 erosion = lmer(yield_kg ~ release_yr + yrs_in_trial +
@@
 
-summary(erosion)
-confint(erosion, parm = c("release_yr","yrs_in_trial"),
-        method = "boot", level = 0.9)
+stargazer(erosion, title = "Model including yield erosion.",
+          ci = TRUE, ci.level = 0.90, float = TRUE,
+          covariate.labels = c("Year of release", "Yield erosion","Intercept"),
+          dep.var.caption = "Response",
+          dep.var.labels = "Yield ( kg/ha )",
+          digits = 1, header = FALSE
+          )
 ```
+
+```{r, message = FALSE, warning = FALSE, results = "asis", echo = FALSE}
+tt = anova(erosion, no_erosion)
+row.names(tt) = c("Excluding yield erosion", "Including yield erosion")
+print(xtable(tt,
+             caption = "Model comparison"), comment = FALSE,
+      caption.placement = "top")
+
+```

### 2017-11-12 output/breed_paper.Rnw
Updates with some of Bruce's comments.

@@
 yield comparison tests will not reflect increases in yield
-potential. This hypothesis was tested using experimental yield data
-over 30+ years at the California Cooperative Rice Research Foundation
-Rice Experiment Station. Results show an apparent increase in yield in
-side-by-side comparisons between old and new rice varieties when the
-yield erosion was not considered. However, the yields of older
-varieties eroded at an estimated rate of 29.3 kg ha\tsups{-1}
-year\tsups{-1} (90\% credible interval -4.4 to -53.3) after initial
-selection. Once this effect was considered, the yield advantage of
-newer cultivars over old was uncertain (-3.3 kg ha\tsups{-1}
-year\tsups{-1}, 90\% credible interval -36.1 to 31.5). This suggests
-that side-by-side comparisons reflected the decreasing yields of older
-cultivars more than the increasing yield of newer ones. These results
-highlight the importance of continuous crop improvement and deployment
-of new cultivars as yield will decline over time without it. To
-increase the genetic yield potential of rice in the face of this
-phenomenon, higher yield targets are needed to counter-act the
-decrease over time in the current cultivars.
+potential. This hypothesis was tested using 33 years of experimental
+yield data from the California Cooperative Rice Research Foundation
+Rice Experiment Station. Based on side-by-side comparisons of old and
+new rice varieties which do not consider yield erosion, there was an
+apparent increase in yield. However, the yields of older varieties
+eroded at an estimated rate of 29.3 kg ha\tsups{-1} year\tsups{-1}
+(90\% credible interval -4.4 to -53.3) after initial selection. Once
+this effect was considered, the yield advantage of newer cultivars
+over old was uncertain (-3.3 kg ha\tsups{-1} year\tsups{-1}, 90\%
+credible interval -36.1 to 31.5). These results highlight first, the
+importance of continuous crop improvement and deployment of new
+cultivars simply to maintain existing yields, and second, to increase
+the genetic yield potential, higher yield targets are
+needed. Importantly, when breeding near the yield potential, despite
+the limited yield gains, significant advances in improving quality and
+reducing crop duration have been made.
 
@@
 <<echo=FALSE>>=
-
 library(rstan)
@@
 yields erode over time then apparent yield increases in side-by-side
-yield comparison tests will not reflect true or absolute yield
-increases over time.
+yield comparison tests will not reflect the true magnitude of yield
+changes over time.
 
@@
 to \Sexpr{max(CA_data$Year)} for California, USA (US Dept. of
-Agriculture National Agricultural Statistics Service, 2017). Rapid
-gains in grain yield were experienced from approximately 1950 -- 1990,
-but annual gains since 1990 have been slower.}
+Agriculture National Agricultural Statistics Service, 2017).}
 \label{fig:state_yield}
@@
 comparison to estimates where yield erosion is not taken into account,
-the model was used as described above except omitting the $Rate_{yield\ erosion\ i}$ term.
-
-% Add more detail to the methods section - tell how you get these
-% numbers. "Evidence for erosion" - what does this mean for yield
-% increases? Erosion 1st, 2nd implications of erosion. 
-% Drop off -5 and -6
-
-Data were processed and models were fit in R, an environment for
+the model was used as described above except omitting the
+$Rate_{yield\ erosion\ i}$ term (Model 2). Lastly, a model was fit to estimate
+if there were differences between cultivars in the rate of yield
+erosion via a random slope by cultivar (Model 3).
+
+Data were processed and models fit in R, an environment for
 statistical computing~\citep{R}. Models were fit using the `rstanarm'
@@
 interval~\citep{rstanarm}. To verify the estimates, the analysis was
-also conducted using classical methods (frequentist mixed-effects
-model fit by maximum likelihood; See supplemental material for
+also conducted using classical methods, specifically a frequentist mixed-effects
+model fit by maximum likelihood (see supplemental material for
 details). Complete data and code used for this
@@
 There was general agreement between the estimates from the Bayesian
-and classical approaches.  The Bayesian estimates are presented here
-due to the inherent flexibility of Bayesian inference, though the
-classical estimates can be found in the supplemental
-material. Comparing the cultivars' yield during each's initial time
-(average 3 years) in the experimental trials just before being
-released showed that across cultivars, yields ranged from
+and classical approaches for all models, with the estimates being
+similar between approaches for similar models. The Bayesian estimates
+are presented here due to the inherent flexibility of Bayesian
+inference, though the classical estimates can be found in the
+supplemental material. Comparing the cultivars' yield during each's
+initial time (average 3 years) in the experimental trials just before
+being released showed that across cultivars, yields ranged from
 \Sexpr{round(plot_data["M207", "mean"], 2)} (M-207) to
@@
 (Fig \ref{fig:first_yrs}. Evaluating these initial yields over time
-indicates there was no evidence of genetic yield improvements during
-the 30+ year study period (Fig. \ref{fig:RES_year}). Individual exceptions to this were
-two cultivars (M-204 and M-205) that had higher yields than the
-others.
+provides evidence of genetic yield improvements during the 33-year
+study period. Individual exceptions to this were two cultivars (M-204
+and M-205) that had higher yields than the others.
 
@@
 
-The yields of each cultivar over time were evaluated relative to when
-the cultivar was released using the full model which accounts for both
-yield erosion and yield improvements over time. The average decrease
-in yield over time (i.e., yield erosion) once a
-cultivar was selected was estimated \Sexpr{yrs_coef} kg ha\tsups{-1}
-year\tsups{-1} (90\% CI: \Sexpr{RES_CI[2]} to \Sexpr{RES_CI[1]}) (Fig
-\ref{fig:RES_year}; Table \ref{tbl:coef}). The oldest cultivar in the
-trials (\Sexpr{gsub("M","M-",unique(mod_data[which.min(mod_data[,"release_yr"]),"id"]))})
-has lost an average of \Sexpr{round(yrs_coef*-36/1000,1)} Mg
+There was no evidence of consistent differences between cultivars in
+the rate of yield erosion (Model 3), hence the model excluding these
+effects (Model 2) is presented here. The yields of each cultivar over
+time were evaluated relative to when the cultivar was released using
+the full model which accounts for both yield erosion and yield
+improvements over time. The average decrease in yield over time (i.e.,
+yield erosion) once a cultivar was selected was estimated at
+\Sexpr{yrs_coef} kg ha\tsups{-1} year\tsups{-1} (90\% CI:
+\Sexpr{RES_CI[2]} to \Sexpr{RES_CI[1]}) (Fig \ref{fig:RES_year}; Table
+\ref{tbl:coef}). The oldest cultivar in the trials
+(\Sexpr{gsub("M","M-",unique(mod_data[which.min(mod_data[,"release_yr"]),"id"]))})
+is estimated to have lost an average of \Sexpr{round(yrs_coef*-36/1000,1)} Mg
 ha\tsups{-1} since its release. The most dominate medium grain rice
 cultivar, M-202 (widely planted from the late 1980s until the early
-2000s) experienced similar trends in yield decline
-(Fig. \ref{fig:RES_year}; Table \ref{tbl:coef}). M-206 replaced M-202
+2000s) experienced similar trends in yield decline. M-206 replaced M-202
 as the most widely grown cultivar in CA and the current estimated 560
-kg ha\tsups{-1} yield advantage of M-206 over M-202 can be attributed solely
-to the erosion of M-202's yield over time (Figs \ref{fig:first_yrs} \&
-\ref{fig:compare_yrs}). Further supporting this estimate of yield
-erosion, both cultivars yielded similarly at their respective year of
-release (Fig. \ref{fig:compare_yrs}), despite the fact that M-202 was
-released 19 years prior to M-206 (Fig. \ref{fig:compare_yrs}; Table
-\ref{tbl:ids}).
+kg ha\tsups{-1} yield advantage of M-206 over M-202 can be attributed
+solely to the erosion of M-202's yield over time (Figs
+\ref{fig:first_yrs} \& \ref{fig:compare_yrs}). Further supporting this
+estimate of yield erosion, both cultivars yielded similarly at their
+respective year of release (Fig. \ref{fig:compare_yrs}), despite the
+fact that M-202 was released 19 years prior to M-206
+(Fig. \ref{fig:compare_yrs}; Table \ref{tbl:ids}).
 
@@
 
-When yield erosion is not corrected for in the analysis (i.e.,
-omitted from the model), there is estimated to be a small increase per year
-(approximately \Sexpr{only_release_coef} kg ha\tsups{-1}
-year\tsups{-1}; 90\% CI \Sexpr{only_release_CI[1]} to
+When yield erosion is not accounted for in the analysis (i.e.,
+omitted from the model), there is estimated to be an increase per year
+of approximately \Sexpr{only_release_coef} kg ha\tsups{-1}
+year\tsups{-1} (90\% CI \Sexpr{only_release_CI[1]} to
 \Sexpr{only_release_CI[2]}) (Fig. \ref{fig:wo_erosion}). However, the
-full model which corrected for both the year of release and yield
+full model which accounts for both the year of release and yield
 erosion indicates that the estimated change in grain yield per year
@@
 After correcting for yield erosion, some cultivars were higher
-yielding compared to the overall average (as estimated by random
-intercepts). The marginal estimate for the performance of a cultivar
-averaged over the uncertainty in the other parameters in the model
-(i.e., the random effect) shows three cultivars with higher than
-average yields:
-\Sexpr{paste(gsub('M','M-',rev(row.names(id_est))[c(1,2,4)]),collapse=',')}
+yielding compared to the overall average (as estimated via the random
+intercept term in the model). The marginal estimate for the
+performance of a cultivar averaged over the uncertainty in the other
+parameters in the model (i.e., the random effect) shows three
+cultivars with higher than average yields:
+\Sexpr{paste(gsub('M','M-',rev(row.names(id_est))[c(1,2,4)]),colla=', ')}
 (Figs. \ref{fig:first_yrs} \& \ref{fig:ids_comp}).
@@
 In California, the cultivars with broad adaptation and adoption over
-the 30+ year period of this study are M-202, M-206, and M-105. Cold
+the 33 year period of this study are M-202, M-206, and M-105. Cold
 tolerance is a major problem for much of the California rice producing
@@
 potential, and relatively short duration and are thus broadly grown
-with over 70\% of the area in one of these cultivars during the study
+with over 70\% of the area in these cultivars at any given time during the study
 period~\citep{ucce}. Importantly, not all medium grain cultivars that were
@@
 
-Looking at all the cultivars that have been released, there is no
-conclusive evidence of increased genetic yield potential in medium
+Not accounting for yield erosion, the estimated yield improvement of
+\Sexpr{only_release_coef} kg ha\tsups{-1} year\tsups{-1}
+(Fig. \ref{fig:wo_erosion}) that does not account for yield erosion is
+lower than the 37 to 60 kg ha\tsups{-1} year\tsups{-1} estimated
+by~\citet{Omar2014} in side-by-side comparisons. However, when
+accounting for yield erosion and looking acrossall cultivars, there is
+no conclusive evidence of increased genetic yield potential in medium
 grain rice cultivars released by CCRRF's breeding improvement program
@@
 \Sexpr{rel_CI[1]} to \Sexpr{rel_CI[2]} kg ha\tsups{-1} year\tsups{-1}
-(Fig. \ref{fig:RES_year}; Table \ref{tbl:coef}),
-both increases and decreases in the genetic yield potential over the
-study period are plausible. The estimated yield
-improvement of \Sexpr{only_release_coef} kg ha\tsups{-1}
-year\tsups{-1} (Fig. \ref{fig:wo_erosion}) that does not account for
-yield erosion is lower than the 37 to 60 kg ha\tsups{-1}
-year\tsups{-1} estimated by~\citet{Omar2014} in side-by-side
-comparisons. Rather, this analysis suggests that the CCRRF breeding
+(Fig. \ref{fig:RES_year}; Table \ref{tbl:coef}), both increases and
+decreases in the genetic yield potential over the study period are
+plausible. This analysis suggests that the CCRRF breeding
 program has maintained but not substantially improved the yield
@@
 the fields are left fallow every other year. Second, changing climate
-may influence yields with either warmer~\citep{van2013} or
-cooler (Espe et al., 2017) temperatures reducing yields. Again, this
-is unlikely to be the case in this situation as Espe et al. (2017)
-found no significant changes in growing season temperatures (either
-high, low or average temps) the Sacramento Valley during the study
-period.  Thirdly, yield erosion of a cultivar as has been reported for
-the cultivar IR8 at IRRI~\citep{Peng1998, Peng2000, Peng2010}. Biotic
-and abiotic forces have previously been speculated to be responsible
-for yield erosion, since the cultivar's genetics remain stable from
-the time of selection but the conditions the cultivar may experience
-over time are continuously changing~\citep{Peng1998, Peng2000,
-Peng2010}. We estimate that yield erosion has decreased the yield of
-medium grain rice cultivars released by the CCRRF over the last 30+
-years by an average of \Sexpr{yrs_coef} Mg ha\tsups{-1} year\tsups{-1}, potentially
-offsetting any gains made by the crop improvement program. However, it
-is not clear what is responsible for this erosion. It is unlikely
-biotic factors as pest and disease pressure are low in this
-Mediterranean climate. Furthermore, these fields are fallowed every
-other year and the residues are burned, further lowering the potential
-for pest and disease pressure. While efforts should be made to
-identify the cause of erosion in this environment, this observation
-highlights both the need for maintenance of plant breeding and the
-difficulty of increasing genetic yield potential, especially in
-systems such as CA where yield is approaching the physiological limit
-\citep{espe16a}. Without maintenance breeding (i.e., continuously
-selecting cultivars well adapted to current conditions), CA rice
-production systems are predicted to have lost an estimated
-\Sexpr{round(yrs_coef*(-36/1000),1)} Mg ha\tsups{-1} grain yield over
-the study period.
+may influence yields with either warmer~\citep{van2013} or cooler
+ temperatures reducing yields. Again, this is
+unlikely to be the case in this situation as ~\citet{espe2017} found
+no significant changes in growing season temperatures (either high,
+low or average temps) the Sacramento Valley during the study period.
+Thirdly, biotic and abiotic forces have previously been speculated as
+the cause for yield erosion of the cultivar IR8 at
+IRRI~\citep{Peng1998, Peng2000, Peng2010}. They argued that since the
+cultivar's genetics remain stable from the time of selection but the
+conditions the cultivar experiences over time are continuously
+changing leading ot yield erosion~\citep{Peng1998, Peng2000,
+Peng2010}. However, it is not clear what is responsible for the yield
+erosion of \Sexpr{yrs_coef} Mg ha\tsups{-1} year\tsups{-1} found in
+this analysis. It is unlikely biotic factors as pest and disease
+pressure are low in this Mediterranean climate. Furthermore, these
+fields are fallowed every other year and the residues are burned,
+further lowering the potential for pest and disease pressure. While
+efforts should be made to identify the cause of erosion in this
+environment, this observation highlights both the need for maintenance
+of plant breeding and the difficulty of increasing genetic yield
+~\citep{espe2017} potential, especially in systems such as CA where yield is approaching
+the physiological limit \citep{espe16a}. Without maintenance breeding
+(i.e., continuously selecting cultivars well adapted to current
+conditions), CA rice production systems are predicted to have lost an
+estimated \Sexpr{round(yrs_coef*(-36/1000),1)} Mg ha\tsups{-1} grain
+yield over the study period.
 
@@
 temperatures than many other rice growing areas in CA
-\citep{espe2017}. These cultivars may perform well at the RES but not
-widely in CA. Therefore, it should be noted that while our
+\citep{espe2017}. Hence, cultivars which perform well at the RES might
+not perform well across of CA. Since these stressful temperature
+events are inconsistent across years, they influence yield stability (how consistent
+yield is year to year). Therefore, it should be noted that while our
 investigation has been concerned about increasing yield potential, the
@@
 no absolute improvements in yield have been made beyond maintaining
-yield. The evidence here is consistent with the observation that the
-genetic yield potential has not increased substantially in the last
-30+ years (Fig. \ref{fig:first_yrs}).
+yield.
 
@@
 milling yield~\citep{mckenzie1994}, have been made over the study
-period and are important. Also, crop duration has been reduced. For
-example, during this study period, M-202 and M-206 have been the
-predominate cultivars grown in CA, and while their yield potentials
-were similar at the time of release (Fig. 5), M-206 matures about 6
-days earlier than M-202 (Table \ref{tbl:ids}). Shorter duration is very important
-in CA due to an already short growing season and water limitations
-(shorter duration cultivars would presumably require less irrigation
-water).
-
-California state average yields have increased slowly
-despite there being little to no gains in the breeding trials
-(Fig. \ref{fig:state_yield}). The evidence here suggests mechanisms
-other than increased genetic yield potential are responsible for these gains
-(Fig. \ref{fig:compare_yrs}). Improvement in agronomic management,
-including more efficient nutrient management~\citep{Lundy2012,
-Lundy2015, linquist2009}, weed management~\citep{Saito2008,
-Pittelkow2012, linquist2008}, and other factors may all be contributing to
-increases in the state average yield. However, these gains are far less
-than those experienced during the preceding era, and by their
-nature agronomic gains cannot surpass the genetic yield potential
-\citep{Fischer2015}. Judging from yields achieved at the RES
-(Figs. \ref{fig:RES_year} \& \ref{fig:compare_yrs}), a
-researcher-managed production system implementing best management
-practices in more controlled situations compared to farmers' fields, state
-average yields (approximately 10 Mg ha\tsups{-1} at the time of
-writing; Fig. \ref{fig:state_yield}) may be approaching the current
-genetic limit. Assuming that the level of management possible in small
-experimental plots cannot be attained in farmers' fields and hence
-experimental yields represent the attainable yield ceiling
-\citep{Fischer2015}, agronomic gains in farmer's fields are nearing
-their potential, consistent with physiological model-based estimates
-of the attainable yield gap in CA rice production systems
+period and are important. Maintaining high quality is imperative for
+the breeding program and it is not uncommon in the breeding program
+for higher yielding cultivars to be eliminated due to concerns about
+quality. However, the goal of increasing yield may require the yield
+performance of cultivars to be assessed earlier in the selection
+process to avoid discarding these higher yielding entries without full
+consideration. Another change in CA rice cultivars is that crop
+duration has been reduced. For example, during this study period,
+M-202 and M-206 have been the predominate cultivars grown in CA, and
+while their yield potentials were similar at the time of release
+(Fig. 5), M-206 matures about 6 days earlier than M-202 (Table
+\ref{tbl:ids}). Shorter duration is very important in CA due to an
+already short growing season and water limitations (shorter duration
+cultivars would presumably require less irrigation water).
+
+California state average yields have increased slowly during the study
+period (Fig. \ref{fig:state_yield}) \citep{espe16a}. What are the
+potential causes?  It is not due to the adoption of the broadly
+adapted cultivars (M-202, M-206, and M-105) as there is no evidence of
+increased yield potential among these (Fig. \ref{fig:first_yrs} \&
+\ref{fig:compare_yrs}). While M-205 has increase yield potential, it
+is not broadly adapted. Its adoption (although limited) in the
+northern and warmer part of CA may contribute to some yield gains
+reported at the state level. Other factors include improvement in
+agronomic management, including more efficient nutrient
+management~\citep{Lundy2012, Lundy2015, linquist2009}, and weed
+management~\citep{Saito2008, Pittelkow2012, linquist2008}. However,
+these state wide yield gains are far less than those experienced
+during the preceding era, and by their nature agronomic gains cannot
+surpass the genetic yield potential \citep{Fischer2015}. Judging from
+yields achieved at the RES (Figs. \ref{fig:RES_year} \&
+\ref{fig:compare_yrs}), a researcher-managed production system
+implementing best management practices in more controlled situations
+compared to farmers' fields, state average yields (approximately 10 Mg
+ha\tsups{-1} at the time of writing; Fig. \ref{fig:state_yield}) may
+be approaching the current genetic limit. Assuming that the level of
+management possible in small experimental plots cannot be attained in
+farmers' fields and hence experimental yields represent the attainable
+yield ceiling \citep{Fischer2015}, agronomic gains in farmer's fields
+are nearing their potential, consistent with physiological model-based
+estimates of the attainable yield gap in CA rice production systems
 \citep{espe16a}.

### 2017-11-13 output/breed_paper.Rnw
Final edits.

@@
 
-Meeting the challenge of feeding a growing population with
-increasingly limited resources will require increasing the yield
-potential of staple crops, such as rice. Yet many high-yielding,
-intensive production systems have experienced slow rates of yield
-improvement in recent years despite a demonstrated increase in the
-yield potential of new crop cultivars. We analyzed experimental data
-from one such cropping system, i.e., California (CA) rice, in order to
-quantify improvements made in the genetic yield potential obtained
-through plant breeding. California rice systems are among the highest
-in the world and close to maximum yield potential. Specifically, the
-hypothesis was tested that if rice cultivar yields decline over time
-(termed yield erosion) then apparent yield increases in side-by-side
-yield comparison tests will not reflect increases in yield
-potential. This hypothesis was tested using 33 years of experimental
-yield data from the California Cooperative Rice Research Foundation
-Rice Experiment Station. Based on side-by-side comparisons of old and
-new rice varieties which do not consider yield erosion, there was an
-apparent increase in yield. However, the yields of older varieties
-eroded at an estimated rate of 29.3 kg ha\tsups{-1} year\tsups{-1}
-(90\% credible interval -4.4 to -53.3) after initial selection. Once
-this effect was considered, the yield advantage of newer cultivars
-over old was uncertain (-3.3 kg ha\tsups{-1} year\tsups{-1}, 90\%
-credible interval -36.1 to 31.5). These results highlight first, the
-importance of continuous crop improvement and deployment of new
-cultivars simply to maintain existing yields, and second, to increase
-the genetic yield potential, higher yield targets are
-needed. Importantly, when breeding near the yield potential, despite
-the limited yield gains, significant advances in improving quality and
-reducing crop duration have been made.
+Meeting the challenge of feeding a growing population with limited
+resources will require increasing the yield potential of staple crops,
+such as rice. Yet many high-yielding, intensive production systems
+have experienced slow rates of yield improvement in recent years
+despite a demonstrated increase in the yield potential of new crop
+cultivars. We analyzed experimental data from one such cropping
+system, i.e., California (CA) rice, in order to quantify improvements
+made in the genetic yield potential obtained through plant
+breeding. California rice systems are among the highest in the world
+and close to maximum yield potential. Specifically, the hypothesis was
+tested that if rice cultivar yields decline over time (termed yield
+erosion) then apparent yield increases in side-by-side yield
+comparison tests will not reflect increases in yield potential. This
+hypothesis was tested using 33 years of experimental yield data from
+the California Cooperative Rice Research Foundation Rice Experiment
+Station. Based on side-by-side comparisons of old and new rice
+cultivars which do not consider yield erosion, there was an apparent
+increase in yield. However, the yields of older cultivars eroded at an
+estimated rate of 29.3 kg ha\tsups{-1} year\tsups{-1} (90\% credible
+interval -4.4 to -53.3) after initial selection. Once this effect was
+considered, the yield advantage of newer cultivars over old was
+uncertain (-3.3 kg ha\tsups{-1} year\tsups{-1}, 90\% credible interval
+-36.1 to 31.5). These results highlight 1) the importance of
+continuous crop improvement and deployment of new cultivars simply to
+maintain existing yields, and 2) to increase the genetic yield
+potential, higher yield targets are needed. Importantly, when breeding
+near the yield potential, despite the limited yield gains, significant
+advances in improving quality and reducing crop duration have been
+made.
 
@@
 crops. However, many production systems are experiencing plateaus in
-grain yield~\citep{Grassini2013, Peng2010}. If this trend continues,
+grain yield~\citep{Grassini2013}. If this trend continues,
 improvements per unit land area are no longer possible, and an
@@
 grown primarily in the Sacramento Valley, which is characterized by
-having a Mediterranean climate with long days, a dry growing season,
-and relatively free of pests and diseases. These conditions lead to
-some of the highest yields in the world~\citep{FAO2016}. Production in
-CA is predominately focused on premium quality medium grain
+having a Mediterranean climate with long days, a dry growing season
+relatively free of pests and diseases. These conditions lead to some
+of the highest yields in the world~\citep{FAO2016}. Production in CA
+is predominately focused on premium quality medium grain
 \textit{japonica} cultivars (e.g., CalRose rice), and rice from CA is
@@
 Certainly one question in looking at this situation is, ``do new
-varieties have increased yield potential''?  Like many breeding
+cultivars have increased yield potential''?  Like many breeding
 programs, one of the goals of the CCRRF is to increase yields through
-breeding varieties with increased yield potential. However, when the
+breeding cultivars with increased yield potential. However, when the
 popular cultivar M-202 was released in 1985, it was reported to yield
@@
 the state each year. To avoid potential complications due to site,
-climate, and management differences at the farmer-managed trials per
-year, we focused our investigation on the RES. The RES is
+climate, and management differences at the farmer-managed trials
+between years, we focused our investigation on the RES. The RES is
 researcher-managed and plants the majority of the experimental plots
@@
 precipitation events. Soils at the RES are classified as
-Esquon-Neerdobe clays with roughly 2\% soil organic matter in the top 15cm~\citep{ssurgo}.
+Esquon-Neerdobe clays with roughly 2\% soil organic matter in the top
+15cm~\citep{ssurgo}.
 
@@
 medium grain cultivars, as these cultivars constitute approximately
-90\% of the planted rice area in CA. Newly developed entries are typically
-evaluated in these trials for 3 to 5 years prior to
-their official release. For these cultivars planted in breeding trials
-at the RES from \Sexpr{min(mod_data[,'year'])} --
-\Sexpr{max(mod_data[,'year'])} plot level observations (replicated
-3 to 4 times in a CRB design) were collected leading to a final data set
+90\% of the planted rice area in CA. Newly developed entries are
+typically evaluated in these trials for 3 to 5 years prior to their
+official release. For these cultivars planted in breeding trials at
+the RES from \Sexpr{min(mod_data[,'year'])} --
+\Sexpr{max(mod_data[,'year'])} plot level observations (replicated 3
+to 4 times in a CRB design) were collected leading to a final data set
 of \Sexpr{format(nrow(mod_data), big.mark = ',')} observations
@@
 \Sexpr{max(mod_data[,'release_yr'])}. For each cultivar, the official
-year of public release and the time elapsed (in years) since release was
-determined.
-
-Experimental plots at the RES are managed similar to University of California
-Cooperative Extension prescribed best management
-practice for land preparation, fertility rates, and pest
-and disease control~\citep{ucce}. As is typical for water-seeded
-systems, the plots were planted by pre-germinating rice seed prior
-to direct-seeding into pre-flooded fields. Plots were either 14.0 or
-18.5 m\tsups{2}, and were harvested after physiological
-maturity using a small-plot combine. All yields were converted to 14\%
-MC prior to reporting.
+year of public release and the time elapsed (in years) since release
+was determined.
+
+Experimental plots at the RES are managed similar to University of
+California Cooperative Extension prescribed best management practice
+for land preparation, fertility rates, and pest and disease
+control~\citep{ucce}. As is typical for water-seeded systems, the
+plots were planted by pre-germinating rice seed prior to
+direct-seeding into pre-flooded fields. Plots were either 14.0 or 18.5
+m\tsups{2}, and were harvested after physiological maturity using a
+small-plot combine. All yields were converted to 14\% MC prior to
+reporting.
 
@@
 To test the hypothesis, the influence of breeding and yield erosion
-were quantified on yield over time using a Bayesian hierarchical
-model. Yield was modeled as a response to the number of years since
+were quantified on yield over time using three nested Bayesian hierarchical
+models. First, yield was modeled as a response to the number of years since
 the cultivar was officially released and the year of release using a
@@
 year (${release\ year\ i}$), $\beta_{year}$ is the random effect of
-year, and $\beta_{cultivar}$ is the random effect of cultivar. For
-comparison to estimates where yield erosion is not taken into account,
-the model was used as described above except omitting the
-$Rate_{yield\ erosion\ i}$ term (Model 2). Lastly, a model was fit to estimate
-if there were differences between cultivars in the rate of yield
-erosion via a random slope by cultivar (Model 3).
+year, and $\beta_{cultivar}$ is the random effect of cultivar. This
+full model was modified to further explore yield dynamics over
+time. For comparison to estimates where yield erosion is not taken
+into account, the model was used as described above except omitting
+the $Rate_{yield\ erosion\ i}$ term (Model 2). Lastly, a model was fit
+to estimate if there were differences between cultivars in the rate of
+yield erosion by adding a random slope by cultivar (Model 3).
 
@@
 uncertainty in the parameter estimates using the quantile method. The
-90\% CI is more stable to perpetuation's and hence is preferable to the 95\%
-interval~\citep{rstanarm}. To verify the estimates, the analysis was
-also conducted using classical methods, specifically a frequentist mixed-effects
-model fit by maximum likelihood (see supplemental material for
-details). Complete data and code used for this
-analysis is available at \url{github.com/mespe/rice_advances.git}.
+90\% CI is more stable to sample to sample variance
+and hence is preferable to the 95\% interval~\citep{rstanarm}. To
+verify the estimates, the analysis was also conducted using classical
+methods, specifically frequentist mixed-effects models fit by maximum
+likelihood (see supplemental material for details and model
+results). Complete data and code used for this analysis is available
+at \url{github.com/mespe/rice_advances.git}.
 
@@
 
-# text(x = plot_data$yr, y = plot_data$mean,
-#      labels = 1:nrow(plot_data), #gsub("M", "M-", row.names(plot_data)),
-#      pos = 4, cex = 0.75)
-# mm = lm(yield_kg/1000 ~ release_yr, data = mod_data[idx,])
-# abline(mm, lty = 2)
-# yrs = 1980:2020
-# int = predict(mm, newdata = data.frame(release_yr = yrs), se.fit = TRUE)
-
-# polygon(x = c(yrs, rev(yrs)),
-#         y = c(int$fit + int$se.fit,rev(int$fit - int$se.fit)),
-#         border = NA,
-#         col = rgb(0,0,0,0.1))
-
 @
@@
 There was general agreement between the estimates from the Bayesian
-and classical approaches for all models, with the estimates being
-similar between approaches for similar models. The Bayesian estimates
+and classical statistical approaches for all models, with the
+estimates being similar between similar models. The Bayesian estimates
 are presented here due to the inherent flexibility of Bayesian
@@
 \Sexpr{round(plot_data["M205", "mean"], 2)} Mg ha\tsups{-1} (M-205)
-(Fig \ref{fig:first_yrs}. Evaluating these initial yields over time
-provides evidence of genetic yield improvements during the 33-year
-study period. Individual exceptions to this were two cultivars (M-204
-and M-205) that had higher yields than the others.
+(Fig \ref{fig:first_yrs}. Comparing the yields from only the first few
+years for each cultivar over the study period does not provide
+evidence of genetic yield improvements during the 33-year study
+period. Individual exceptions to this were two cultivars (M-204 and
+M-205) that had higher yields than the others.
 
@@
 There was no evidence of consistent differences between cultivars in
-the rate of yield erosion (Model 3), hence the model excluding these
-effects (Model 2) is presented here. The yields of each cultivar over
-time were evaluated relative to when the cultivar was released using
-the full model which accounts for both yield erosion and yield
-improvements over time. The average decrease in yield over time (i.e.,
-yield erosion) once a cultivar was selected was estimated at
-\Sexpr{yrs_coef} kg ha\tsups{-1} year\tsups{-1} (90\% CI:
-\Sexpr{RES_CI[2]} to \Sexpr{RES_CI[1]}) (Fig \ref{fig:RES_year}; Table
-\ref{tbl:coef}). The oldest cultivar in the trials
+the rate of yield erosion (Model 3), hence the model excluding varying
+rates of erosion by cultivar (Model 2) is presented here. The yields
+of each cultivar over time were evaluated relative to when the
+cultivar was released using the full model which accounts for both
+yield erosion and yield improvements over time. The average decrease
+in yield over time (i.e., yield erosion) once a cultivar was selected
+was estimated at \Sexpr{yrs_coef} kg ha\tsups{-1} year\tsups{-1} (90\%
+CI: \Sexpr{RES_CI[2]} to \Sexpr{RES_CI[1]}) (Fig \ref{fig:RES_year};
+Table \ref{tbl:coef}). The oldest cultivar in the trials
 (\Sexpr{gsub("M","M-",unique(mod_data[which.min(mod_data[,"release_yr"]),"id"]))})
-is estimated to have lost an average of \Sexpr{round(yrs_coef*-36/1000,1)} Mg
-ha\tsups{-1} since its release. The most dominate medium grain rice
-cultivar, M-202 (widely planted from the late 1980s until the early
-2000s) experienced similar trends in yield decline. M-206 replaced M-202
-as the most widely grown cultivar in CA and the current estimated 560
-kg ha\tsups{-1} yield advantage of M-206 over M-202 can be attributed
-solely to the erosion of M-202's yield over time (Figs
-\ref{fig:first_yrs} \& \ref{fig:compare_yrs}). Further supporting this
-estimate of yield erosion, both cultivars yielded similarly at their
-respective year of release (Fig. \ref{fig:compare_yrs}), despite the
-fact that M-202 was released 19 years prior to M-206
-(Fig. \ref{fig:compare_yrs}; Table \ref{tbl:ids}).
+is estimated to have lost an average of
+\Sexpr{round(yrs_coef*-36/1000,1)} Mg ha\tsups{-1} since its
+release. Likewise, an estimated 560 kg ha\tsups{-1} yield
+advantage of M-206 over M-202 (the current and former predominant
+cultivars, respectively), can be attributed solely to the erosion
+of M-202's yield over time (Figs \ref{fig:first_yrs} \&
+\ref{fig:compare_yrs}). Further supporting this estimate of yield
+erosion, both cultivars yielded similarly at their respective year of
+release (Fig. \ref{fig:compare_yrs}), despite the fact that M-202 was
+released 19 years prior to M-206 (Fig. \ref{fig:compare_yrs}; Table
+\ref{tbl:ids}).
 
@@
 
-When yield erosion is not accounted for in the analysis (i.e.,
-omitted from the model), there is estimated to be an increase per year
-of approximately \Sexpr{only_release_coef} kg ha\tsups{-1}
-year\tsups{-1} (90\% CI \Sexpr{only_release_CI[1]} to
-\Sexpr{only_release_CI[2]}) (Fig. \ref{fig:wo_erosion}). However, the
-full model which accounts for both the year of release and yield
-erosion indicates that the estimated change in grain yield per year
-in the CCRRF's breeding trials at the RES was \Sexpr{release_yr_coef}
-kg ha\tsups{-1} year\tsups{-1}, with both small positive and negative
-trends within the credible interval (90\% CI: \Sexpr{rel_CI[1]} to
-\Sexpr{rel_CI[2]}) (Fig. \ref{fig:first_yrs}).
+\subsection{Yield gains}
+
+When yield erosion is not accounted for in the analysis (Model 2),
+there is estimated to be an increase per year of approximately
+\Sexpr{only_release_coef} kg ha\tsups{-1} year\tsups{-1} (90\% CI
+\Sexpr{only_release_CI[1]} to \Sexpr{only_release_CI[2]})
+(Fig. \ref{fig:wo_erosion}). However, the full model (Model 1) which
+accounts for both the year of release and yield erosion indicates that
+the estimated change in grain yield per year in the CCRRF's breeding
+trials at the RES was \Sexpr{release_yr_coef} kg ha\tsups{-1}
+year\tsups{-1}, with both small positive and negative trends within
+the credible interval (90\% CI: \Sexpr{rel_CI[1]} to
+\Sexpr{rel_CI[2]}) (Fig. \ref{fig:first_yrs}). According to model
+comparison metrics (PS-LOO, WAIC), the full model (Model 1) is
+estimated to have superior predictive accuracy compared to the model
+which does not account for yield erosion (Model 2)
 
@@
 plot_data2$release_yr = plot_data$release_yr
-
-# par(mfrow = c(1,2),
-#     mar = c(4,0,2,0.75),
-#     oma = c(4,4,0.5,0.5))
 
@@
 
-In California, the cultivars with broad adaptation and adoption over
-the 33 year period of this study are M-202, M-206, and M-105. Cold
-tolerance is a major problem for much of the California rice producing
-region ~\citep{espe2017}, although it is a bigger problem in the
-southern portion of the Sacramento Valley due to the influence of the
-ocean winds through the Delta ~\citep{espe16a}. These three
-cultivars have a higher degree of cold tolerance, high yield
-potential, and relatively short duration and are thus broadly grown
-with over 70\% of the area in these cultivars at any given time during the study
-period~\citep{ucce}. Importantly, not all medium grain cultivars that were
-released were purported to have higher yields than previous
-cultivars. Some cultivars had special traits, but upon release had
-similar yields to more broadly adapted cultivars. Examples of this
-include M-208, which had resistance to blast disease (\textit{Magnaporthe grisea})
-and M-402 which is a premium quality cultivar. Other cultivars such as
-M-205 and M-209, while high yielding, do not have the cold tolerance
-of the other cultivars and are thus only recommended for the northern
-portion of the Sacramento Valley.
+In CA, the cultivars with broad adaptation and adoption over the 33
+year period of this study are M-202, M-206, and M-105. Cold tolerance
+is a major problem for much of the California rice producing region
+~\citep{espe2017}, although it is a bigger problem in the southern
+portion of the Sacramento Valley due to the influence of the ocean
+winds through the Delta ~\citep{espe16a}. These three cultivars have a
+higher degree of cold tolerance, high yield potential, and relatively
+short duration and are thus broadly grown with over 70\% of the area
+in these cultivars at any given time during the study
+period~\citep{ucce}. There are several cultivars that were found to
+have better than average yields, even after correcting for yield
+erosion. However, these cultivars (M-201, M-204, M-205, and possibly
+M-209; Fig. \ref{fig:ids_comp})) are not widely adopted. Importantly,
+not all medium grain cultivars that were released were purported to
+have higher yields than previous cultivars. Some cultivars had special
+traits, but upon release had similar yields to more broadly adapted
+cultivars. Examples of this include M-208, which had resistance to
+blast disease (\textit{Magnaporthe grisea}) and M-402 which is a
+premium quality cultivar. Other cultivars such as M-205 and M-209,
+while high yielding, do not have the cold tolerance of the other
+cultivars and are thus only recommended for the northern portion of
+the Sacramento Valley.
 
@@
 \Sexpr{only_release_coef} kg ha\tsups{-1} year\tsups{-1}
-(Fig. \ref{fig:wo_erosion}) that does not account for yield erosion is
-lower than the 37 to 60 kg ha\tsups{-1} year\tsups{-1} estimated
-by~\citet{Omar2014} in side-by-side comparisons. However, when
-accounting for yield erosion and looking acrossall cultivars, there is
-no conclusive evidence of increased genetic yield potential in medium
-grain rice cultivars released by CCRRF's breeding improvement program
-over the period \Sexpr{min(mod_data[,'year'])} to
-\Sexpr{max(mod_data[,'year'])}, which has stayed relatively constant
-at \Sexpr{round(mean(mod_data[,"yield_kg"])/1000, 1)} Mg
+(Fig. \ref{fig:wo_erosion}) is lower than the 37 to 60 kg ha\tsups{-1}
+year\tsups{-1} estimated by~\citet{Omar2014} in side-by-side
+comparisons. However, when accounting for yield erosion and looking
+across all cultivars, there is no conclusive evidence of increased
+genetic yield potential in medium grain rice cultivars released by
+CCRRF's breeding improvement program over the period
+\Sexpr{min(mod_data[,'year'])} to \Sexpr{max(mod_data[,'year'])}. 
+Grain yield over this period has stayed relatively constant at an average
+\Sexpr{round(mean(mod_data[,"yield_kg"])/1000, 1)} Mg
 ha\tsups{-1}. Since the estimated yield improvement spans from
@@
 decreases in the genetic yield potential over the study period are
-plausible. This analysis suggests that the CCRRF breeding
-program has maintained but not substantially improved the yield
+plausible. This analysis suggests that the CCRRF breeding program has
+maintained but not substantially improved the yield
 potential in the face of changing biotic and abiotic pressures. This
@@
 Mg ha\tsups{-1} with no indication of yields increasing
-(Fig. \ref{fig:first_yrs}). While these yields have not increased,
-M-205 did have higher yield potential; however, this cultivar is not
-broadly adapted due to poor cold tolerance.
+(Fig. \ref{fig:first_yrs}).
 
@@
 be overcome with improved N management. This is likely not the case
-here as the fields used in this study grow rice only once season a
-year and tend to be non-flooded for the remainder of the
-year. Furthermore, at this location and unlike most CA rice fields,
-the fields are left fallow every other year. Second, changing climate
-may influence yields with either warmer~\citep{van2013} or cooler
- temperatures reducing yields. Again, this is
-unlikely to be the case in this situation as ~\citet{espe2017} found
-no significant changes in growing season temperatures (either high,
-low or average temps) the Sacramento Valley during the study period.
-Thirdly, biotic and abiotic forces have previously been speculated as
-the cause for yield erosion of the cultivar IR8 at
-IRRI~\citep{Peng1998, Peng2000, Peng2010}. They argued that since the
-cultivar's genetics remain stable from the time of selection but the
-conditions the cultivar experiences over time are continuously
-changing leading ot yield erosion~\citep{Peng1998, Peng2000,
-Peng2010}. However, it is not clear what is responsible for the yield
-erosion of \Sexpr{yrs_coef} Mg ha\tsups{-1} year\tsups{-1} found in
-this analysis. It is unlikely biotic factors as pest and disease
-pressure are low in this Mediterranean climate. Furthermore, these
-fields are fallowed every other year and the residues are burned,
-further lowering the potential for pest and disease pressure. While
-efforts should be made to identify the cause of erosion in this
-environment, this observation highlights both the need for maintenance
-of plant breeding and the difficulty of increasing genetic yield
-~\citep{espe2017} potential, especially in systems such as CA where yield is approaching
-the physiological limit \citep{espe16a}. Without maintenance breeding
-(i.e., continuously selecting cultivars well adapted to current
-conditions), CA rice production systems are predicted to have lost an
-estimated \Sexpr{round(yrs_coef*(-36/1000),1)} Mg ha\tsups{-1} grain
-yield over the study period.
-
-There are several cultivars that were found to have better than
-average yields, even after correcting for yield erosion. However,
-these cultivars (M-201, M-204, M-205, and possibly M-209;
-Fig. \ref{fig:ids_comp})) are not widely adopted. It is known that
-cool temperatures during the booting stage can have a large impact on
-grain yields~\citep{espe2017}, and the RES experiences warmer
-temperatures than many other rice growing areas in CA
-\citep{espe2017}. Hence, cultivars which perform well at the RES might
-not perform well across of CA. Since these stressful temperature
-events are inconsistent across years, they influence yield stability (how consistent
-yield is year to year). Therefore, it should be noted that while our
-investigation has been concerned about increasing yield potential, the
-estimates here might not capture yield stability. Yield stability is
-notoriously difficult to quantify~\citep[e.g.,][]{lobell2011}, hence
-further investigation is needed to explore how yield stability has
-changed over the study period.
+here as the fields used in this study grow rice only one season a year
+and tend to be non-flooded for the remainder of the year. Furthermore,
+at this location and unlike most CA rice fields, the fields are left
+fallow every other year. Second, changing climate may influence yields
+with either warmer~\citep{van2013} or cooler~\citep{espe2017}
+temperatures reducing yields. Again, this is unlikely to be the case
+in this situation as ~\citet{espe2017} found no significant changes in
+growing season temperatures (either high, low, or average temps) the
+Sacramento Valley during the study period. Thirdly, small changes in
+biotic and abiotic forces have previously been speculated as the cause
+for yield erosion of the cultivar IR8 at IRRI~\citep{Peng1998,
+Peng2000, Peng2010}. Since the cultivar's genetics remain stable from
+the time of selection but the conditions the cultivar experiences over
+time are continuously changing, accumulated changes too small to
+reliably measure could plausibly lead to yield
+erosion~\citep{Peng1998, Peng2000, Peng2010}. However, it is not clear
+if this mechanism is responsible for the yield erosion of
+\Sexpr{yrs_coef} Mg ha\tsups{-1} year\tsups{-1} found in this
+analysis. Biotic stressors such as pests and diseases are low in this
+Mediterranean climate. Furthermore, these fields are fallowed every
+other year and the residues are burned, further lowering the potential
+for pest and disease pressure.
+
+It is striking that the estimated yield erosion is similar to that
+reported by \citet{Peng2010} over a similar time span (15\% decline
+over roughly 30 years). This suggests the mechanisms responsible could
+potentially be endemic to all rice production systems. While efforts
+should be made to identify the cause of erosion, this study highlights
+both the need for maintenance of plant breeding and the difficulty of
+increasing genetic yield ~\citep{espe2017} potential, especially in
+systems such as CA where yield is approaching the physiological limit
+\citep{espe16a}.
 
@@
 
-To be clear, our investigation here is narrowly focused on the
-rough-rice/field grain yield. Other advancements due to plant
-improvement, including increased quality~\citep{mckenzie1993} and
-milling yield~\citep{mckenzie1994}, have been made over the study
-period and are important. Maintaining high quality is imperative for
-the breeding program and it is not uncommon in the breeding program
-for higher yielding cultivars to be eliminated due to concerns about
-quality. However, the goal of increasing yield may require the yield
-performance of cultivars to be assessed earlier in the selection
-process to avoid discarding these higher yielding entries without full
-consideration. Another change in CA rice cultivars is that crop
-duration has been reduced. For example, during this study period,
-M-202 and M-206 have been the predominate cultivars grown in CA, and
-while their yield potentials were similar at the time of release
-(Fig. 5), M-206 matures about 6 days earlier than M-202 (Table
+Our investigation here is narrowly focused on the rough-rice/field
+grain yield. Other advancements due to plant improvement, including
+increased quality~\citep{mckenzie1993} and milling
+yield~\citep{mckenzie1994}, have been made over the study period and
+are important. Additionally, crop duration has been reduced in CA rice
+cultivars. For example, while the predominate cultivars M-202 and
+M-206 had similar yield potentials at the time of release (Fig. 5),
+M-206 matures about 6 days earlier than M-202 (Table
 \ref{tbl:ids}). Shorter duration is very important in CA due to an
@@
 increased yield potential among these (Fig. \ref{fig:first_yrs} \&
-\ref{fig:compare_yrs}). While M-205 has increase yield potential, it
-is not broadly adapted. Its adoption (although limited) in the
+\ref{fig:compare_yrs}). Although M-205 is not broadly adapted, it has
+increase yield potential and its adoption (although limited) in the
 northern and warmer part of CA may contribute to some yield gains
@@
 management~\citep{Saito2008, Pittelkow2012, linquist2008}. However,
-these state wide yield gains are far less than those experienced
-during the preceding era, and by their nature agronomic gains cannot
-surpass the genetic yield potential \citep{Fischer2015}. Judging from
-yields achieved at the RES (Figs. \ref{fig:RES_year} \&
-\ref{fig:compare_yrs}), a researcher-managed production system
-implementing best management practices in more controlled situations
-compared to farmers' fields, state average yields (approximately 10 Mg
-ha\tsups{-1} at the time of writing; Fig. \ref{fig:state_yield}) may
-be approaching the current genetic limit. Assuming that the level of
-management possible in small experimental plots cannot be attained in
-farmers' fields and hence experimental yields represent the attainable
-yield ceiling \citep{Fischer2015}, agronomic gains in farmer's fields
-are nearing their potential, consistent with physiological model-based
-estimates of the attainable yield gap in CA rice production systems
-\citep{espe16a}.
+recent rates of state-wide yield gains are far less than those of the
+preceding era, consistent with a system near the yield
+ceiling~\citep{Grassini2013}.
+
+It should be noted that while our investigation has been concerned
+about increasing yield potential, the estimates here might not capture
+yield stability (how consistent grain yields are from year to year). The
+occurrence of stress events (e.g., extreme temperatures) are
+unpredictable across years and are not well captured by the analysis
+here. Yield stability is notoriously difficult to
+quantify~\citep[e.g.,][]{lobell2011}, hence further investigation is
+needed to explore how yield stability has changed over the study
+period.
 
@@
 conclusions underscore both the necessity and substantial challenges
-facing rice breeding. We suggest here that historical benchmarks be
+facing rice breeding. Maintaining high quality is imperative for
+a breeding program and it is not uncommon in a breeding program
+for higher yielding cultivars to be eliminated due to concerns about
+quality. However, the goal of increasing yield may require the yield
+performance of cultivars to be assessed earlier in the selection
+process to avoid discarding these higher yielding entries without full
+consideration. We suggest here that historical benchmarks be
 used in to evaluate promising rice cultivars in addition to
@@
 print(xtable(ids,
-             caption = "All named medium grain rice cultivars and their associated release year planted at the Rice Experiment Station (RES), Biggs, California from 1984 -- 2015.",
+             caption = "All named medium grain rice cultivars and their associated release year (from 1984 -- 2015) planted at the Rice Experiment Station (RES), Biggs, California.",
              label = "tbl:ids"),
@@
 \beginsupplement
-
+\begin{comment}
 \begin{figure}[H]
@@
 \newpage
+\end{comment}
 

### 2017-12-10 output/breed_paper.Rnw
Final edits 'cause I am done

@@
 cultivars. We analyzed experimental data from one such cropping
-system, i.e., California (CA) rice, in order to quantify improvements
+system, California (CA) rice, in order to quantify improvements
 made in the genetic yield potential obtained through plant
@@
 grown primarily in the Sacramento Valley, which is characterized by
-having a Mediterranean climate with long days, a dry growing season
+having a Mediterranean climate with long days and a dry growing season
 relatively free of pests and diseases. These conditions lead to some
@@
 years and between cultivars, year and cultivar were included in the
-model as random effects.
+model as random effects (Model 1):
 
@@
 \Sexpr{round(plot_data["M205", "mean"], 2)} Mg ha\tsups{-1} (M-205)
-(Fig \ref{fig:first_yrs}. Comparing the yields from only the first few
-years for each cultivar over the study period does not provide
+(Fig \ref{fig:first_yrs}. Comparing these yields from the time of
+release for each cultivar does not provide
 evidence of genetic yield improvements during the 33-year study
@@
   estimated erosion of yield over time (averaged over differences in
-  cultivar and year) is estimated to be \Sexpr{yrs_coef} kg ha\tsups{-1}
-  year\tsups{-1}. Points ($\bullet$) and bars (|) represent the mean and
-  standard error of the raw data, respectively. The shaded area is the
-  90\% credible interval.}
+  cultivar and year) is estimated by a mixed-effects model to be
+  \Sexpr{yrs_coef} kg ha\tsups{-1} year\tsups{-1}. Points ($\bullet$)
+  and bars (|) represent the mean and standard error of the raw data,
+  respectively. The shaded area is the 90\% credible interval.}
 \label{fig:RES_year}
@@
 There was no evidence of consistent differences between cultivars in
-the rate of yield erosion (Model 3), hence the model excluding varying
-rates of erosion by cultivar (Model 2) is presented here. The yields
-of each cultivar over time were evaluated relative to when the
-cultivar was released using the full model which accounts for both
-yield erosion and yield improvements over time. The average decrease
-in yield over time (i.e., yield erosion) once a cultivar was selected
-was estimated at \Sexpr{yrs_coef} kg ha\tsups{-1} year\tsups{-1} (90\%
-CI: \Sexpr{RES_CI[2]} to \Sexpr{RES_CI[1]}) (Fig \ref{fig:RES_year};
-Table \ref{tbl:coef}). The oldest cultivar in the trials
-(\Sexpr{gsub("M","M-",unique(mod_data[which.min(mod_data[,"release_yr"]),"id"]))})
-is estimated to have lost an average of
-\Sexpr{round(yrs_coef*-36/1000,1)} Mg ha\tsups{-1} since its
-release. Likewise, an estimated 560 kg ha\tsups{-1} yield
-advantage of M-206 over M-202 (the current and former predominant
-cultivars, respectively), can be attributed solely to the erosion
-of M-202's yield over time (Figs \ref{fig:first_yrs} \&
+the rate of yield erosion (Model 3); although visual evidence suggest
+otherwise (Fig. \ref{fig:all_cult}), more data is needed. Based on
+this, the models excluding varying rates of erosion by cultivar (i.e.,
+Models 1 \& 2) are presented here. The yields of each cultivar over time were
+evaluated relative to when the cultivar was released using the full
+model (Model 1) which accounts for both yield erosion and yield improvements
+over time. The average decrease in yield over time (i.e., yield
+erosion) once a cultivar was selected was estimated at
+\Sexpr{yrs_coef} kg ha\tsups{-1} year\tsups{-1} (90\% CI:
+\Sexpr{RES_CI[2]} to \Sexpr{RES_CI[1]}) (Fig \ref{fig:RES_year}; Table
+\ref{tbl:coef}). The estimated 560 kg ha\tsups{-1} yield advantage of
+M-206 over M-202 (the current and former predominant cultivars,
+respectively), can be attributed solely to the erosion of M-202's
+yield over time (Fig. \ref{fig:first_yrs} \&
 \ref{fig:compare_yrs}). Further supporting this estimate of yield
@@
 release (Fig. \ref{fig:compare_yrs}), despite the fact that M-202 was
-released 19 years prior to M-206 (Fig. \ref{fig:compare_yrs}; Table
+released 19 years prior to M-206 (Table
 \ref{tbl:ids}).
@@
 When yield erosion is not accounted for in the analysis (Model 2),
-there is estimated to be an increase per year of approximately
+there is estimated to be a yield increase of approximately
 \Sexpr{only_release_coef} kg ha\tsups{-1} year\tsups{-1} (90\% CI
 \Sexpr{only_release_CI[1]} to \Sexpr{only_release_CI[2]})
-(Fig. \ref{fig:wo_erosion}). However, the full model (Model 1) which
-accounts for both the year of release and yield erosion indicates that
-the estimated change in grain yield per year in the CCRRF's breeding
-trials at the RES was \Sexpr{release_yr_coef} kg ha\tsups{-1}
-year\tsups{-1}, with both small positive and negative trends within
-the credible interval (90\% CI: \Sexpr{rel_CI[1]} to
-\Sexpr{rel_CI[2]}) (Fig. \ref{fig:first_yrs}). According to model
-comparison metrics (PS-LOO, WAIC), the full model (Model 1) is
-estimated to have superior predictive accuracy compared to the model
-which does not account for yield erosion (Model 2)
+(Fig. \ref{fig:wo_erosion}). This analysis is functionally similar to evaluating
+cultivars in side-by-side yield tests, as the differences in years
+since release is between cultivars is not accounted for. However, the
+full model (Model 1) which accounts for both the year of release and
+yield erosion indicates that the estimated change in grain yield per
+year in the CCRRF's breeding trials at the RES was
+\Sexpr{release_yr_coef} kg ha\tsups{-1} year\tsups{-1}, with both
+small positive and negative trends within the credible interval (90\%
+CI: \Sexpr{rel_CI[1]} to \Sexpr{rel_CI[2]})
+(Fig. \ref{fig:first_yrs}). According to model comparison metrics
+(PS-LOO, WAIC), the full model (Model 1) is estimated to have superior
+predictive accuracy compared to the model which does not account for
+yield erosion (Model 2)
 
@@
 period~\citep{ucce}. There are several cultivars that were found to
-have better than average yields, even after correcting for yield
+have higher than average yields, even after correcting for yield
 erosion. However, these cultivars (M-201, M-204, M-205, and possibly
-M-209; Fig. \ref{fig:ids_comp})) are not widely adopted. Importantly,
+M-209; Fig. \ref{fig:ids_comp}) are not widely adopted. Importantly,
 not all medium grain cultivars that were released were purported to
@@
 decreases in the genetic yield potential over the study period are
-plausible. This analysis suggests that the CCRRF breeding program has
+plausible. This analysis suggests that the breeding program has
 maintained but not substantially improved the yield
@@
 
-Yields of a particular cultivar may decline overtime for a number of
+Yields of a particular cultivar may decline over time for a number of
 reasons. First, ~\citet{Nie2009} reported yield declines in rice due
@@
 both the need for maintenance of plant breeding and the difficulty of
-increasing genetic yield ~\citep{espe2017} potential, especially in
+increasing genetic yield potential~\citep{espe2017}, especially in
 systems such as CA where yield is approaching the physiological limit
@@
 agronomic management, including more efficient nutrient
-management~\citep{Lundy2012, Lundy2015, linquist2009}, and weed
-management~\citep{Saito2008, Pittelkow2012, linquist2008}. However,
+management~\citep{Lundy2012, linquist2009}, and weed
+management~\citep{Pittelkow2012, linquist2008}. However,
 recent rates of state-wide yield gains are far less than those of the
@@
 It should be noted that while our investigation has been concerned
-about increasing yield potential, the estimates here might not capture
+with increasing yield potential, the estimates here might not capture
 yield stability (how consistent grain yields are from year to year). The
@@
 @
-\caption{The marginal estimates for the difference between cultivars,
-  averaged over the uncertainty in the fixed effect parameters (i.e.,
-  yield erosion and annual yield gains).}
+\caption{The marginal estimates (i.e., random effects) for the
+  difference between cultivars, averaged over the uncertainty in the
+  fixed effect parameters (i.e., yield erosion and annual yield gains).}
 \label{fig:ids_comp}

### 2018-02-21 output/breed_paper_revised.Rnw
Start of revisions.
Remove 'yield erosion'.

--- 2017-11-13 output/breed_paper.Rnw
@@
 \usepackage{Sweave}
-\SweaveOpts{prefix.string=graphics/report, png=TRUE}
+\SweaveOpts{prefix.string=graphics/report, png=FALSE}
 
@@
 Station. Based on side-by-side comparisons of old and new rice
-cultivars which do not consider yield erosion, there was an apparent
+cultivars which do not consider yield decline over time, there was an apparent
 increase in yield. However, the yields of older cultivars eroded at an
@@
 performance over time has been observed~\citep{Peng1998, Peng2010,
-Peng2000, Datta1995}, often termed yield erosion~\citep{Peng1998}. It
-has been speculated that this yield erosion is due to the inability of
+Peng2000, Datta1995}, sometimes termed ``yield erosion''~\citep{Peng1998}. It
+has been speculated that this yield decline over time is due to the inability of
 a cultivar to adapt to changing biotic and abiotic
@@
 \Sexpr{max(mod_data[,'year'])} plot level observations (replicated 3
-to 4 times in a CRB design) were collected leading to a final data set
+to 4 times in a RCB design) were collected leading to a final data set
 of \Sexpr{format(nrow(mod_data), big.mark = ',')} observations
@@
 
-To test the hypothesis, the influence of breeding and yield erosion
+To test the hypothesis, the influence of breeding and yield decline over time
 were quantified on yield over time using three nested Bayesian hierarchical
@@
 average yield for the RES across cultivars and years, $Rate_{yield\
-erosion\ i}$ is rate of yield erosion per ${years\ since\ release\ i}$
+erosion\ i}$ is rate of yield decline per ${years\ since\ release\ i}$
 (negative for years prior to release, zero at the year of release),
@@
 full model was modified to further explore yield dynamics over
-time. For comparison to estimates where yield erosion is not taken
+time. For comparison to estimates where yield decline over time is not taken
 into account, the model was used as described above except omitting
@@
 to estimate if there were differences between cultivars in the rate of
-yield erosion by adding a random slope by cultivar (Model 3).
+yield decline over time by adding a random slope by cultivar (Model 3).
 
@@
 
-\subsection{Yield erosion}
+\subsection{yield decline over time}
 \begin{figure}[H]
@@
 text(5, 8.0, bquote(.(yrs_coef)~kg~ha^-1~year^-1))
-mtext("Adjusting for yield erosion", adj = 0)
+mtext("Adjusting for yield decline over time", adj = 0)
 
@@
 There was no evidence of consistent differences between cultivars in
-the rate of yield erosion (Model 3), hence the model excluding varying
+the rate of yield decline over time (Model 3), hence the model excluding varying
 rates of erosion by cultivar (Model 2) is presented here. The yields
@@
 cultivar was released using the full model which accounts for both
-yield erosion and yield improvements over time. The average decrease
-in yield over time (i.e., yield erosion) once a cultivar was selected
+yield decline over time and yield improvements over time. The average decrease
+in yield over time once a cultivar was selected
 was estimated at \Sexpr{yrs_coef} kg ha\tsups{-1} year\tsups{-1} (90\%
@@
 \caption{The performance of two popular cultivars over time. The
-  dashed line is the estimated yield erosion as estimated by the full
+  dashed line is the estimated yield decline over time as estimated by the full
   model. Points ($\bullet$) represent the mean, and bars (|) the standard
@@
 
-When yield erosion is not accounted for in the analysis (Model 2),
+When yield decline over time is not accounted for in the analysis (Model 2),
 there is estimated to be an increase per year of approximately
@@
 (Fig. \ref{fig:wo_erosion}). However, the full model (Model 1) which
-accounts for both the year of release and yield erosion indicates that
+accounts for both the year of release and yield decline over time indicates that
 the estimated change in grain yield per year in the CCRRF's breeding
@@
 estimated to have superior predictive accuracy compared to the model
-which does not account for yield erosion (Model 2)
-
-After correcting for yield erosion, some cultivars were higher
+which does not account for yield decline over time (Model 2)
+
+After correcting for yield decline over time, some cultivars were higher
 yielding compared to the overall average (as estimated via the random
@@
 text(x = 2000, y = 8, bquote(.(round(fixef(RES_rel_fit)[2] * scl, 1))~kg~ha^-1~yr^-1))
-mtext("Without adjusting for yield erosion", adj = 0)
+mtext("Without adjusting for yield decline over time", adj = 0)
 
@@
 (\Sexpr{min(comp_yrs)} -- \Sexpr{max(comp_yrs)}) as estimated without
-correcting for yield erosion over time.}
+correcting for yield decline over time.}
 \label{fig:wo_erosion}
@@
 text(x = 2000, y = 8, bquote(.(release_yr_coef)~kg~ha^-1~yr^-1))
-mtext("Adjusting for yield erosion", adj = 0)
+mtext("Adjusting for yield decline over time", adj = 0)
 
@@
   (\Sexpr{min(comp_yrs)} -- \Sexpr{max(comp_yrs)}) as estimated by 
-  correcting for yield erosion over time. Yield erosion was corrected
+  correcting for yield decline over time. yield decline over time was corrected
   for by adjusting each observed yield by estimated amount lost 
@@
 
-Not accounting for yield erosion, the estimated yield improvement of
+Not accounting for yield decline over time, the estimated yield improvement of
 \Sexpr{only_release_coef} kg ha\tsups{-1} year\tsups{-1}
@@
 year\tsups{-1} estimated by~\citet{Omar2014} in side-by-side
-comparisons. However, when accounting for yield erosion and looking
+comparisons. However, when accounting for yield decline over time and looking
 across all cultivars, there is no conclusive evidence of increased
@@
 biotic and abiotic forces have previously been speculated as the cause
-for yield erosion of the cultivar IR8 at IRRI~\citep{Peng1998,
+for yield decline over time of the cultivar IR8 at IRRI~\citep{Peng1998,
 Peng2000, Peng2010}. Since the cultivar's genetics remain stable from
@@
 erosion~\citep{Peng1998, Peng2000, Peng2010}. However, it is not clear
-if this mechanism is responsible for the yield erosion of
+if this mechanism is responsible for the yield decline over time of
 \Sexpr{yrs_coef} Mg ha\tsups{-1} year\tsups{-1} found in this
@@
 
-It is striking that the estimated yield erosion is similar to that
+It is striking that the estimated yield decline over time is similar to that
 reported by \citet{Peng2010} over a similar time span (15\% decline
@@
 are to meet the target of 50\% increased food production by 2050
-\citep{Foley2011a}. Given the estimated yield erosion, to hit an
+\citep{Foley2011a}. Given the estimated yield decline over time, to hit an
 ambitious target of 500 kg ha\tsups{-1} (approximately 5\%) increase
@@
                        "Yield improvement (kg ha\\tsups{-1} year\\tsups{-1})",
-                       "Yield erosion (kg ha\\tsups{-1} year\\tsups{-1})")
+                       "Yield decline (kg ha\\tsups{-1} year\\tsups{-1})")
 colnames(RES_coef) = c("Median", "90\\% Credible interval")
@@
 \Sexpr{min(mod_data[,'year'])} -- \Sexpr{max(mod_data[,'year'])}. The
-solid line is the estimated yield erosion (\Sexpr{yrs_coef} kg
+solid line is the estimated yield decline over time (\Sexpr{yrs_coef} kg
 ha\tsups{-1} year\tsups{-1}), while the dashed line is the published
@@
   averaged over the uncertainty in the fixed effect parameters (i.e.,
-  yield erosion and annual yield gains).}
+  yield decline over time and annual yield gains).}
 \label{fig:ids_comp}

### 2018-02-21 output/breed_paper_revised.Rnw
Update with year:id interaction.

@@
 
-release_yr_coef = round(fixef(RES_full)[2] * scl, 1)
-yrs_coef = round(fixef(RES_full)[3] * scl, 1)
+release_yr_coef = round(fixef(RES_fullest)[2] * scl, 1)
+yrs_coef = round(fixef(RES_fullest)[3] * scl, 1)
 only_release_coef = round(fixef(RES_rel_fit)[2]*scl, 1)
 
-RES_CI <- round(posterior_interval(RES_full, pars = "yrs_in_trial") * scl, 1)
-rel_CI = round(posterior_interval(RES_full, pars = "release_yr_c") * scl, 1)
+RES_CI <- round(posterior_interval(RES_fullest, pars = "yrs_in_trial") * scl, 1)
+rel_CI = round(posterior_interval(RES_fullest, pars = "release_yr_c") * scl, 1)
 only_release_CI = round(posterior_interval(RES_rel_fit, pars = "release_yr_c") * scl, 1)
@@
 
-id_est = ranef(RES_full)$id
-id_est = cbind(id_est, posterior_interval(RES_full, regex_pars = "id:M"))
+id_est = ranef(RES_fullest)$id
+id_est = cbind(id_est, posterior_interval(RES_fullest, regex_pars = "id:M"))
 idx = order(id_est[,1], decreasing = FALSE)
@@
 
-\begin{figure}[H]
-<<fig=TRUE, echo=FALSE>>=
-CA_data = read.csv("../data/CA_rice_yield.csv", stringsAsFactors = FALSE)
-CA_data$Value = as.numeric(gsub(",", "", CA_data$Value))/1000
-CA_data = CA_data[order(CA_data$Year),]
-plot(Value * 1.12 ~ Year, CA_data,
-     # type = "l",
-     pch = 16,
-     xlab = "Year",
-     ylab = expression("State avg. grain yield ( Mg ha"^-1~")"))
-# mv_avg = filter(c(rep(mean(CA_data$Value[1:10] * 1.12), 5),
-#                   CA_data$Value * 1.12,
-#                   rep(mean(CA_data$Value[(nrow(CA_data) - 10):nrow(CA_data)] * 1.12), 5)),
-#                 rep(1/11,11))
-# lines(x = c(1907:1911, CA_data$Year, 2017:2021), y = mv_avg)
-
-@
-\caption{Average rice yields over the period \Sexpr{min(CA_data$Year)}
-to \Sexpr{max(CA_data$Year)} for California, USA (US Dept. of
-Agriculture National Agricultural Statistics Service, 2017).}
-\label{fig:state_yield}
-\end{figure}
-
 \section{Methods}\label{methods}
@@
 \Sexpr{max(mod_data[,'year'])} plot level observations (replicated 3
-to 4 times in a RCB design) were collected leading to a final data set
+to 4 times in a completely randomized design) were collected leading to a final data set
 of \Sexpr{format(nrow(mod_data), big.mark = ',')} observations
@@
 \begin{split}
- yield_i = Intercept + (Rate_{yield\ erosion\ i}* {years\ since\
-   release\ i}) \\+ (Rate_{yield\ improvement\ i}* {release\ year\ i}) 
- + \beta_{year\ i} + \beta_{cultivar\ i} \\
+ yield_{ijk} = Intercept + (Rate_{yield\ erosion}* {years\ since\
+   release\ ik}) \\+ (Rate_{yield\ improvement}* {release\ year\ ik}) 
+ + \beta_{year\ ij} + \beta_{cultivar\ ik} + \epsilon \\
 \end{split}
@@
 
-Where $yield_i$ is the grain yield for plot $i$, the Intercept is the
-average yield for the RES across cultivars and years, $Rate_{yield\
-erosion\ i}$ is rate of yield decline per ${years\ since\ release\ i}$
-(negative for years prior to release, zero at the year of release),
-$Rate_{yield\ improvement\ i}$ is the rate of yield improvement per
+Where $yield_i$ is the grain yield for cultivar $k$ in plot $i$ for
+year $j$,
+the Intercept is the average yield for the RES across cultivars and years, $Rate_{yield\
+erosion}$ is rate of yield decline per ${years\ since\ release\ i}$
+(mean-centered, i.e., negative for years prior to release, zero at the year of release),
+$Rate_{yield\ improvement}$ is the rate of yield improvement per
 year (${release\ year\ i}$), $\beta_{year}$ is the random effect of
@@
 
-pp <- posterior_linpred(RES_full, re.form = NA,
+pp <- posterior_linpred(RES_fullest, re.form = NA,
                         newdata = data.frame(release_yr_c = yrs - 2000,
@@
 
-abline(a = ((((fixef(RES_full)[1] * scl) + center) ) -
+abline(a = ((((fixef(RES_fullest)[1] * scl) + center) ) -
            (2000 * release_yr_coef)) / 1000,
@@
 
-\subsection{yield decline over time}
+\subsection{Yield decline over time}
+
+\begin{figure}[H]
+<<echo = FALSE, fig = TRUE, width = 7, height = 6>>=
+library(ggplot2)
+library(ggthemes)
+ggplot(mod_data, aes(x = year, y = yield_kg/1000)) +
+    geom_point(size = 0.5, alpha = 0.5) +
+    geom_smooth(method = "lm", color = "black") +
+    theme_few() +
+    xlab("Year") +
+    ylab(expression("Grain yield ( Mg ha"^-1~")")) + 
+    facet_wrap(~id, ncol = 4)
+
+@
+\caption{The yield of all medium grain released cultivars at the
+  California Cooperative Rice Research Foundation Rice Experiment
+  Station over the period \Sexpr{min(mod_data[,'year'])} --
+  \Sexpr{max(mod_data[,'year'])}.}
+\label{fig:all_cult}
+\end{figure}
+
 \begin{figure}[H]
@@
 yrs_in <- -10:40
-pp <- posterior_linpred(RES_full, re.form = NA,
+pp <- posterior_linpred(RES_fullest, re.form = NA,
                         newdata = data.frame(yrs_in_trial = yrs_in,
@@
                          y1 = yield_kg - yield_se))
-abline(a = ((fixef(RES_full)[1] * scl) + center) / 1000 ,
+abline(a = ((fixef(RES_fullest)[1] * scl) + center) / 1000 ,
        b = yrs_coef/1000)
@@
             yrs_in_trial = 0)
-wo_erosion$corr_yield = colMeans(posterior_predict(RES_full,
+wo_erosion$corr_yield = colMeans(posterior_predict(RES_fullest,
                                newdata = wo_erosion)) + center/1000
@@
 axis(1, at = seq(1980, 2020, by = 10))
-abline(a = ((((fixef(RES_full)[1] * scl) + center) ) -
+abline(a = ((((fixef(RES_fullest)[1] * scl) + center) ) -
            (2000 * release_yr_coef)) / 1000,
@@
 # arrows(12, 8.2, 27, 8.2, xpd = TRUE, lwd = 2)
-abline(a = 10, b = fixef(RES_full)[3], lty = 2, lwd = 2)
+abline(a = 10, b = fixef(RES_fullest)[3], lty = 2, lwd = 2)
 text(25, 10.85, "(3) target above maintenance", cex = 0.8,
-     srt = 18)#((atan(fixef(RES_full)[3] + 0.06) * 180))/pi)
+     srt = 18)#((atan(fixef(RES_fullest)[3] + 0.06) * 180))/pi)
 abline(h = 10, lty = 3, lwd = 2)
 text(25, 10.1, "(1) current situation", cex = 0.8)
-abline(a = 10, b = fixef(RES_full)[3] + 0.06, lty = 4, lwd = 2)
+abline(a = 10, b = fixef(RES_fullest)[3] + 0.06, lty = 4, lwd = 2)
 text(25, 9.35, "(2) no maintenance breeding", cex = 0.8,
@@
 
-CIs = posterior_interval(RES_full)[1:3,] * scl
+CIs = posterior_interval(RES_fullest)[1:3,] * scl
 CIs[1,] = round((CIs[1,] + center)/1000, 1)
 CIs = round(CIs, 1)
-med_eff = fixef(RES_full) * scl
+med_eff = fixef(RES_fullest) * scl
 med_eff[1] = round((med_eff[1] + center)/1000, 1)
@@
 \beginsupplement
+
+\begin{figure}[H]
+<<fig=TRUE, echo=FALSE>>=
+CA_data = read.csv("../data/CA_rice_yield.csv", stringsAsFactors = FALSE)
+CA_data$Value = as.numeric(gsub(",", "", CA_data$Value))/1000
+CA_data = CA_data[order(CA_data$Year),]
+plot(Value * 1.12 ~ Year, CA_data,
+     # type = "l",
+     pch = 16,
+     xlab = "Year",
+     ylab = expression("State avg. grain yield ( Mg ha"^-1~")"))
+# mv_avg = filter(c(rep(mean(CA_data$Value[1:10] * 1.12), 5),
+#                   CA_data$Value * 1.12,
+#                   rep(mean(CA_data$Value[(nrow(CA_data) - 10):nrow(CA_data)] * 1.12), 5)),
+#                 rep(1/11,11))
+# lines(x = c(1907:1911, CA_data$Year, 2017:2021), y = mv_avg)
+
+@
+\caption{Average rice yields over the period \Sexpr{min(CA_data$Year)}
+to \Sexpr{max(CA_data$Year)} for California, USA (US Dept. of
+Agriculture National Agricultural Statistics Service, 2017).}
+\label{fig:state_yield}
+\end{figure}
+
+\newpage
+
 \begin{comment}
@@
     }
-    # abline(a = ((fixef(RES_full)[1] * scl) + center) / 1000 -
+    # abline(a = ((fixef(RES_fullest)[1] * scl) + center) / 1000 -
     #            (2000 * yrs_coef/1000),
@@
           11.2 + (1:nrow(plot_data[[i]])) * (yrs_coef/1000))
-    # abline(a = ((fixef(RES_full)[1] * scl) + center) / 1000,
+    # abline(a = ((fixef(RES_fullest)[1] * scl) + center) / 1000,
            # b = yrs_coef/1000)
@@
 \newpage
+
+\begin{figure}[H]
+<<fig=TRUE, echo=FALSE>>=
+tars <- c("M202","M206")
+comp_yrs = 2004:2005
+
+boxplot(yield_kg/1000 ~ id, data = mod_data[(mod_data$id %in% tars) &
+                                            (mod_data$year %in% comp_yrs),],
+        ylab = expression("Grain yield ( Mg ha"^-1~")"))
+
+@
+\caption{An example of a typical side-by-side comparison between a older (M-202)
+  and newer (M-206) cultivar. Shown is the performance of these cultivars at the CA Rice
+  Experiment Station (RES) over the period \Sexpr{min(comp_yrs)} --
+  \Sexpr{max(comp_yrs)}. The newer
+  cultivar, M-206, appears to have a yield advantage over the older
+  cultivar, M-202, though this comparison does not account for the 10
+  year difference in years since release between the cultivars.}
+\label{fig:compare_vars}
+\end{figure}
+%% \newpage
+
 \end{comment}
 
-% \begin{figure}[H]
-% <<fig=TRUE, echo=FALSE>>=
-% tars <- c("M202","M206")
-% comp_yrs = 2004:2005
-
-% boxplot(yield_kg/1000 ~ id, data = mod_data[(mod_data$id %in% tars) &
-%                                             (mod_data$year %in% comp_yrs),],
-%         ylab = expression("Grain yield ( Mg ha"^-1~")"))
-
-% @
-% \caption{An example of a typical side-by-side comparison between a older (M-202)
-%   and newer (M-206) cultivar. Shown is the performance of these cultivars at the CA Rice
-%   Experiment Station (RES) over the period \Sexpr{min(comp_yrs)} --
-%   \Sexpr{max(comp_yrs)}. The newer
-%   cultivar, M-206, appears to have a yield advantage over the older
-%   cultivar, M-202, though this comparison does not account for the 10
-%   year difference in years since release between the cultivars.}
-% \label{fig:compare_vars}
-% \end{figure}
-% \newpage
-
-\begin{figure}[H]
-<<echo = FALSE, fig = TRUE, width = 7, height = 6>>=
-library(ggplot2)
-library(ggthemes)
-ggplot(mod_data, aes(x = year, y = yield_kg/1000)) +
-    geom_point(size = 0.5, alpha = 0.5) +
-    geom_smooth(method = "lm", color = "black") +
-    theme_few() +
-    xlab("Year") +
-    ylab(expression("Grain yield ( Mg ha"^-1~")")) + 
-    facet_wrap(~id, ncol = 4)
-
-@
-\caption{The yield of all medium grain released cultivars at the
-  California Cooperative Rice Research Foundation Rice Experiment
-  Station over the period \Sexpr{min(mod_data[,'year'])} --
-  \Sexpr{max(mod_data[,'year'])}.}
-\label{fig:all_cult}
-\end{figure}
-
-\newpage
 \begin{figure}[ht]

### 2018-02-21 src/h1_model.R

@@ -50,11 +50,17 @@ load("../data/model_data.rda")
 #                           data = mod_data, iter = 1000, cores = 2L)
 
 
-RES_full = stan_glmer(yield_cs ~ (1|year) + (1|id) + release_yr_c + yrs_in_trial,
-                      data = mod_data, iter = 2000, cores = 2L,
-                      seed = 789)
+RES_fullest = stan_glmer(yield_cs ~ (1|year) + (1|id) +
+                             (1|id:year)+ release_yr_c + yrs_in_trial,
+                         data = mod_data, iter = 2000, cores = 2L,
+                         seed = 789)
 
 if(compare){
+
+    RES_full = stan_glmer(yield_cs ~ (1|year) + (1|id) + release_yr_c + yrs_in_trial,
+                          data = mod_data, iter = 2000, cores = 2L,
+                          seed = 789)
+    
     RES_yrs_fit <- stan_glmer(yield_cs ~ (1|year) + (1|id) + yrs_in_trial,
                               data = mod_data, iter = 2000, cores = 2L,
                               seed = 789)
@@ -73,9 +79,14 @@ if(compare){
                       seed = 789)
 
     compare(loo(RES_full), loo(RES_yrs_fit), loo(RES_rel_fit), loo(RES_fuller))
-    
 
+    RES_fullest = stan_glmer(yield_cs ~ (1|year) + (1|id) +
+                                 (1|id:year)+ release_yr_c + yrs_in_trial,
+                      data = mod_data, iter = 2000, cores = 2L,
+                      seed = 789)
+
+    compare(loo(RES_full), loo(RES_fullest))
 }
 
 
-save(RES_full, file = "../output/h1_fit.Rda")
+save(RES_fullest, file = "../output/h1_fit.Rda")

### 2018-02-23 output/breed_paper_revised.Rnw
Commenting on Kents 2014 pub.

@@
 
+\citet{mckenzie2014} explored the yield of rice in this same system
+during a similar time period using similar data as these here, but
+concluded that genetic rice yield potential has increased over this
+time period at a rate of 25 kg\tsups{-1} ha\tsups{-1}
+year\tsups{-1}. There are several explanations for how this could
+be. First, \citet{mckenzie2014} did not account for yield decreases
+over time. As shown here, when these decreases are not taken into about, the
+estimated increase in yield over time is estimated to be approximately
+the same as that found by \citet{mckenzie2014}. Second,
+\citet{mckenzie2014} concluded that there was no trend in the yield
+of the major cultivars over time, yet this interpretation of the
+results of a statistical test is flawed. Failing to reject the
+hypothesis that of a trend in yield does not equate to there being no
+trend. Third, it is not surprising that no trend in yield over time
+was detected. \citet{mckenzie2014} used data from several locations, yet analyzed
+those data using a simple t-test. By using this simple analysis,
+\citet{mckenzie2014} ignore systematic errors related to differences between
+observations, including different management systems, growing
+environments, and crop timing between locations. By contrast, this
+investigation used a single, well managed system and analyzed the data
+using a mixed effects model which fully accounts for non-independence
+between similar observations. In doing this, the analysis presented
+here makes fewer assumptions and more fully captures the structure of
+the data. Hence, the estimates presented here are likely more
+reliable.
+
 It should be noted that while our investigation has been concerned

### 2018-02-23 output/breed_paper_revised.Rnw
More revisions.

--- 2018-02-21 output/breed_paper_revised.Rnw
@@
 
+raw_samples = as.data.frame(RES_fullest)
 @
@@
 increase in the area under cultivation will be needed to meet food
-demand, which carries with it undesirable ecological
+demand, which carries undesirable ecological
 implications~\citep{Foley2011a, tilman1999}. Therefore, it is of
@@
 the attainable yield potential, increases in yield are likely to be
-slow and \citet{espe16a} reported that indeed this was the case
+relatively slow and \citet{espe16a} reported that indeed this was the case
 with yields on average increasing by about 50 to 62 kg ha\tsups{-1}
@@
 efforts to develop improved rice cultivars adapted to CA since
-1969. The California Statewide Variety Trails evaluate current and
+1969. The California Statewide Variety Trials evaluate current and
 promising cultivars at the RES and six to eight on-farm trials around
@@
 
+\citet{mckenzie2014} explored the yield of rice in this same system
+during a similar time period using similar data as these here, but
+concluded that genetic rice yield potential has increased over this
+time period at a rate of 25 kg\tsups{-1} ha\tsups{-1}
+year\tsups{-1}. Yet the CI estimated from these data using the best
+model suggests that there is only a
+\Sexpr{round(sum((raw_samples$release_yr_c*scl)>=25)/nrow(raw_samples)*100,1)}\%
+probability of the rate of yield increase is equal to or greater than
+25 kg\tsups{-1} ha\tsups{-1} year\tsups{-1}. There are several
+explanations for how this could be. 1) \citet{mckenzie2014} did not
+account for yield decreases over time in their analysis. As shown
+here, when these decreases are not taken into about, the estimated
+increase in yield over time is estimated to be approximately the same
+as that found by \citet{mckenzie2014}. 2) These yield decreases over time
+were not taken into account possibly because \citet{mckenzie2014}
+concluded that there was no trend in the yield of the major cultivars
+over time. However this conclusion is based on a flawed interpretation
+of the results of their statistical test. Failing to reject the
+hypothesis that the trend in yield equals zero does not equate to the
+confirmation of no trend. Instead, failing to reject the hypothesis
+could suggest the analysis lacks the statistical power to detect the trend. 3) It
+is not surprising that no trend in yield over time was detected given
+that \citet{mckenzie2014} used data from several locations, yet
+analyzed those data using a simple t-test. By using this simple
+analysis, \citet{mckenzie2014} ignore systematic errors related to
+differences between observations from different locations, including
+different management systems, growing environments, and crop
+timing. By contrast, this investigation presented here used a single,
+well managed system and analyzed the data using a mixed effects model
+which fully accounts for non-independence between similar
+observations. In doing this, the analysis makes fewer assumptions and
+more fully captures the structural relationships in the data. Hence,
+the estimates presented here are likely more reliable.
+
 It should be noted that while our investigation has been concerned

### 2018-02-25 output/breed_paper_revised.Rnw
Added comments, formatting.

@@
     xlab("Year") +
-    ylab(expression("Grain yield ( Mg ha"^-1~")")) + 
+i    ylab(expression("Grain yield ( Mg ha"^-1~")")) + 
     facet_wrap(~id, ncol = 4)

### 2018-02-26 output/review_responses.tex
Additional changes and comments.

\documentclass{article} \usepackage[margin=1in]{geometry}
\usepackage{pdfpages}
\begin{document}

\includepdf{response_letter.pdf}

\newpage

\section{Point-by-point responses}

\subsection{Editor}

\textbf{Comment: This is a very valuable and interesting manuscript.  It is very well
  suited to Field Crops Research.  All reviewers recommend 'Minor
  revision'.  There was some uncertainty over the statistical analysis
  and so a third reviewer was asked to review this aspect of it. This
  was not able to be entered into EVISE and so I have appended it below.}\\

\textbf{Response:} Thank you for including these comments. \\

\textbf{Comment: This paper analyses 33 years worth of single-site yield data on rice
  varieties grown in California. Regression models are fitted to assess
  yield trend due to cultivars and trend due to "yield erosion".I have a
  few comments on the statistical analysis.}\\

\textbf{(1) The term yield erosion suggests that this declining trend
  component relates to properties of the cultivars, or cultivars losing
  their potential over time. I am not sure this interpretation is the
  only one possible. There are many other factors that may cause a
  decline over time, i.e. climate change, deterioration of the soil,
  agronomic practices not keeping up with changes of the environment
  etc. See my comments further below.}\\

\textbf{Response:} We were using the term ``yield erosion'' as this is
the term used to describe this phenomonom in previous work in
rice. Given that ``erosion'' is a slow, sometimes
imperceptible, process of reducing something
driven by abiotic, climatic, and biotic forces, ``erosion'' seems to
be the appropriate term here. While there is a component of ``erosion'' inherent
to the material being eroded, this term is not often interpreted as it
has been here, and there are likely other concerns which are loading
this term with additional meaning.

However, given the sensitivities involved, we have elected to 
remove the term ``yield erosion'' throughout the manuscript.\\

\textbf{Comment: (2) "CRB design" is not a proper name for a design. I suspect the
  authors are refering to a randomized complete block design (RCBD).}\\

\textbf{Response:} Thank you for pointing out this typo - the correct
design is a CRD. The text has been update accordingly.\\

\textbf{Comment: (3) I think the model statement in equation (1) needs some polishing
  up. Indexing observations just by plots is not sufficient. It appears
  from the model description that plot data were modelled. If that's the
  case, then the model misses a block effect.}\\

\textbf{Response:} As stated above, there are no blocks, hence the
model is correct in this aspect.\\

\textbf{Comment: Moreover, a plot error term is lacking.}\\

\textbf{Response:} Added, though it should be noted that when detailing
the structural aspect of the model in an article, the error term is often
not included. \\

\textbf{Comment: For clarity, the authors could add subscripts for
  years, cultivars and blocks nested within years, amounting to three
  different subscripts.}\\

\textbf{Response:} Subscript for years and cultivar added. There are
no blocks, so this subscript is not needed.\\

\textbf{Comment: If plot data are analysed, a random year-cultivar interaction
  needs to be added as well.}\\

\textbf{Response:} Added to the model. The estimates are largely
unchanged from before.\\

\textbf{Comment: I do not think
  it's appropriate to equate the intercept with an overall mean, unless
  both time covariates were mean-centered, which does not seem to be the
  case.}\\

\textbf{Response:} The time variates were mean-centered as stated in the
manuscript. We have clarified this further (lines 91).\\

\textbf{Comment: (4) If subscripts are used for years and cultivars, the covariates
  should be indexed accordingly. If i and k index cultivars and years,
  respectively, then the time elapsed since release needs to be indexed
  by both i and k (a\_ik, say), whereas the year of release would be
  indexed by i only (r\_i, say). Using this notation, the authors are
  regressing yield on a\_ik and r\_i.}\\

\textbf{Response:} Fixed\\

\textbf{Comment: There's a simple third time
  covariate that could be considered for regression: calendar year
  (t\_k). This regression would directly assess non-genetic components
  of trend, due, e.g., to climate change, advances in agronomic
  practices etc. It turns out, however, that a regression on all three
  covariates is not possible, because a\_ik = t\_k - r\_i, meaning that
  there would be confounding. Thus, only two of these three regressor
  vaiables can be fitted, so the regression done by the authors is fine
  in pinciple.  Nevertheless, this little thought exercise shows that
  there is some ambiguity in the interpretation of regression
  coefficients. This issue is discussed in some detail in
  Piepho, H.P., Laidig, F., Drobek, T., Meyer, U. (2014): Dissecting
  genetic and non-genetic sources of long-term yield trend in German
  official variety trials. Theoretical and Applied Genetics 127,
  1009-1018. The authors of that paper did a regression on t\_k and r\_i.}\\

\textbf{Response:} Yes, this is indeed the case. However, this is
largely a methodological concern, and after some consideration, we have
elected to not discuss this matter here, due to space constraints.\\

\subsection{Reviewer 1}

\textbf{Comment: This paper certainly should be published as the results are
  important even if they lack a satisfactory explanation. But I would
  like to see some changes and more attention to some issues. Much is
  annotated on the attached manuscript, and some repetitive and
  confusing, those remarks came as I read the paper. This summary is
  upon more considered reflection afterwards.}

\textbf{There seems little doubt that for many cultivars their yield
  steadily declined with time, or years after release which was used
  here, but it is the same as simply years. Figure S1 says it all
  (that’s your whole data set I presume): 8 out of the 14 cvs showed
  yield decreases, the other 6 little change with years after release;
  this Figure should be in the paper proper up front I think, it is
  more convincing that the stats modelling derived Figs 3 and 4.}\\

\textbf{Response:} On the suggestion of several reviewers, we have
moved figure S1 into the main paper.\\

\textbf{Comment: It was important to test whether the interaction cv x years after
  release was significant, because this could give a clue as to what is
  the cause of the yield decline, but it wasn’t significant according to
  the stats which I didn’t really understand. Discussion could explore
  this in more depth; I mean why some cvs didn’t lose yield and some
  did. Maybe only widely adopted cvs lost yeild, due to greater presume
  for disease to evolve virulence?}\\

\textbf{Response:} The reviewer should mindful that
stating that the estimated CI for the cultivar-year interaction
includes zero is not the same thing as ``no
effect''. Rather, the correct interpretation is that the effect, if it
exists, is smaller than our power to detect. Therefore, the correct
conclusion is that the differences between cultivars in the decline
over time, if existent, is likely small, though the sign and magnitude of the exact
effect is uncertain.

While this is indeed interesting, we
have attempted to limit our speculation on this point given that we do not have
sufficient data to actually test many of these ideas. We have expanded
the discussion to include some of the points raised (lines 210). \\

\textbf{Comment: The paper seems to be written to put the blame on the cultivars and
  the challenge on the breeders, the notion of their cvs’ yield
  advantages being “eroded”. This might be the case if a biotic stress
  is evolving virulence against the host plant resistance of any cv.
  But it is not really the case when there is a deterioration in the
  aspects of the environment (climate, agronomy, and/or soil), against
  which younger cultivars are more resistant or tolerant.  For me the 29
  kg/ha/yr could be breeding progress in the face of the deteriorating
  E; at about 0.3\% p.a. it is not unlike the rate of progress in
  advanced wheat breeding programs (anything near the 1.2\% p.a. which
  the paper mentions in discussion is very rare these days).}\\

\textbf{Response:} The term ``yield erosion'' has been removed due to
these concerns. However, we have taken great pains to specifically
point out that the evidence suggests this inherent to the system
(as the title of this manuscript clearly states), that it is similar
to observations from other rice systems (lines 211-217), and that it is plausibly
due to the environmental aspects the reviewer lists (lines 190-210).

However, the evidence from this study explicitly states that given
these data and this analysis, a rate of improvement of 29 kg/ha/yr
is unlikely. The reviewer appears to have missed one of the main
conclusions of the manuscript. We have edited for greater clarity. \\

\textbf{Comment: Really it all depends on the definition of potential yield. See van
  Ittersum and colleagues in FCR Vol 143 March 2013. Its determined by
  latest best yielding Cv and the current E in the absences of any
  manageable stresses, both biotic and abiotic, and measured in good
  experiments and models based on such experiments. A breakdown in cv
  resistance is probably best described as “erosion”, while a
  deterioration in the E due, for example, to a negative climate trend,
  is a decrease in the potential yield, but a new cv tolerating this
  deterioration would lift the potential yield (see also below).}\\

\textbf{Response:} Added discussion of these issues (lines 174-181). \\

\textbf{Comment: The authors should read McKay et al (2010, TAG 122, 225-238) where a
  similar situation in winter wheat in the UK is described. But by
  having plots with and without fungicide over the last 40 years or so,
  these authors could measure the erosion and narrow down the cause
  (foliar disease), which did vary with cultivar.}\\

\textbf{Response:} This is an interesting paper (assuming the reviewer
intended \textbf{Mackay et al. 2011}), but unfortunately
these data do not lend themselves to similar explorations. Unlike
Mackay et al., we do not have paired treated/untreated plots. In the
course of writing this paper, we discussed various means to to narrow down the
cause of yield losses, but ultimately lack the data to do so. We have
added a citation to McKay et al. to the manuscript.\\

\textbf{Comment: Also the authors need to refer to McKenzie et al (2014) in the CSSA
  Special Publicn 33 with a whole section on rice yield progress in
  Ca.}\\

\textbf{Response:} Discussion added (lines 247-269). \\

\textbf{Comment: These authors show in Fig 10-8 trial results from all over the
  rice areas of Ca, whereas the manuscript deals with trials solely
  from the RES at Briggs.This needs to be discussed and may help
  explain what is happening, as there is little or no evidence in
  McKenzie et al that, for example, the yield of M-202 is declining
  with time.}\\

\textbf{Response:} Explaination added (lines 247-269)\\

\textbf{Comment: What also worries me is that Ca farm yields (and Butte
  county yields) have been increasing since 1995 or so, at a low rate
  (about 70 kg/ha/yr) but increasing, which doesn’t agree with Fig S1
  no matter what cv you choose.}\\

\textbf{Response:} The reviewer appears to misunderstand basic
differences in the data being presented. We agree and plainly state in
the manuscript that CA farm yields have been increasing (line
). However, we do not find evidence that gains as large as 70 kg/ha/yr
are due to increasing genetic yield potential. It is entirely possible
for farm yields to increase for other reasons which, again, we outline
specifically in the manuscript (lines 241-246).\\

\textbf{Comment: If yield is declining because of increasing disease susceptibility
  with time that is erosion, well known to wheat breeders, and driving
  what is called maintenance breeding. By definition yield potential is
  measured without biotic stresses, which are manageable, so this sort
  of erosion is not a loss of yield potential. If the yield
  deterioration is due to something like increased ozone pollution or
  cooler nights at meiosis, that is less clear cut: those stresses are
  part of the environment, largely unmanageable, and amount to a
  decrease in the yield potential of the environment, but breeding can
  still overcome them, at 29 kg/ha/yr if this was the cause in this rice
  study. Therefore the possible causes at BES need more discussion and
  reference to other trial data sources. Sorry this repeats 4 a bit!}\\

\textbf{Response:} As this repeats the previous concerns, please see
above for our response.\\

\textbf{Comment:  I am not sure, however, that you need to show Ca yields (Fig 1)
  because this broadens the discussion a lot, and almost demands you
  look at what is happening across the rice region in terms of weather
  trends etc, probably using a rice simulation model (although
  modelling at just RES itself would be relevant to the paper also). It
  is useful to show how close farm yields are to RES trial yields but
  that could be covered in a text sentence to two. And if you want to
  discuss this bigger picture, please include a simple map of rice
  areas in Ca.}\\

\textbf{Response:} Fig. 1 has been moved to supplemental material.\\

\textbf{Comment: Table 2 lists the key cv data and should be up front in the paper. You
  could add to the Table brief comments on the important quality attr
  ibutes of the various releases (if there any) and finally you could
  include a column showing the important data of Fig 2, which has
  nothing to do with the modelling.}\\

\textbf{Response:} Table 1 and 2 have been swapped. Quality attributes
are not of direct relevance to this paper, so we have provided a
citation for where this information can be found for those
interested (line 235). We disagree that the data from Figure 2 should be
presented in table form. Data of this nature is best presented in
figures, and presenting these data twice is redundent.\\

\textbf{Comment: So in summary, I am not sure you need Fig 1, you could put the data of
  Fig 2 in Table 2 and bring it to the front along with Fig S1, and I
  don’t think you need Fig 6. But consider how you use the word
  “erosion” and discuss more thoroughly what could be causing the yield
  declines in Fig S1, and whether there is evidence it is happening also
  outside of the RES at Biggs.}\\

\textbf{Response:} Figure 1 has been moved to supplemental material,
and figure S1 has been moved to the main body. See above regarding
putting the data from figure 2 in tabular format.\\

\subsection{Reviewer 2}

\textbf{Comment: Using data from variety trials, the authors proved the importance of
  maintenance breeding for world food security. However, I am not very
  familiar with the statistical analysis that the authors used to derive
  the conclusions. If a reviewer who is familiar in the related
  statistical procedures has no problem with the methodology, I believe
  it is suitable for publication in FCR. Please consider the following
  suggestion during your revision.}\\

\textbf{Comment: L10: Something is wrong with “…with it undesirable…”}\\

\textbf{Response:} Edited\\

\textbf{Comment: L30: Yield increase of 50-62 kg per ha per year was not slow.}\\

\textbf{Response:} Added clarification.\\

\textbf{Comment: L51-52: What are the two numbers (39.4648, - 121.7342) ? You may use
normal way to indicate the location.}\\

\textbf{Response:} These are standard latitude/longitude coordinates.\\

\textbf{Comment: L53: “Trials” instead of “Trails”}\\

\textbf{Response:} Corrected.\\

\textbf{Comment: L120: (Fig. 2)}\\

\textbf{Response:} Corrected.

\textbf{Comment: L141: Something is wrong with “…since release is between…”}

\textbf{Response:} Corrected.\\

\textbf{Comment: Fig. S1: It appears that there are two kinds of dots in the figure
(black and grey). Please indicate their difference.}\\

\textbf{Response:} All points are black.\\

\newpage

\includepdf{diffs.pdf}

\newpage

\includepdf{Espe_rice_yield_improvement_revised.pdf}

\end{document}

### 2018-02-26 output/breed_paper_revised.Rnw
Additional changes and comments.

--- 2018-02-23 output/breed_paper_revised.Rnw
@@
 load("../data/model_data.rda")
-load("h1_fit.Rda")
+load("h1_fit_revised.Rda")
 load("bivar_fit.Rda")
@@
 raw_samples = as.data.frame(RES_fullest)
+
+# Fix ID names once
+mod_data$id_fact = factor(gsub("^M", "M-", mod_data$id))
 @
@@
  yield_{ijk} = Intercept + (Rate_{yield\ erosion}* {years\ since\
-   release\ ik}) \\+ (Rate_{yield\ improvement}* {release\ year\ ik}) 
- + \beta_{year\ ij} + \beta_{cultivar\ ik} + \epsilon \\
+   release\ ik}) \\+ (Rate_{yield\ improvement}* {release\ year\ ik})\\ 
+ + \beta_{year\ ij} + \beta_{cultivar\ ik} +
+ \beta_{cultivar\ ik:year\ ij} + \epsilon \\
 \end{split}
@@
     ylab(expression("Grain yield ( Mg ha"^-1~")")) + 
-    facet_wrap(~id, ncol = 4)
+    facet_wrap(~id_fact, ncol = 4)
 
@@
 
-In CA, the cultivars with broad adaptation and adoption over the 33
-year period of this study are M-202, M-206, and M-105. Cold tolerance
-is a major problem for much of the California rice producing region
-~\citep{espe2017}, although it is a bigger problem in the southern
-portion of the Sacramento Valley due to the influence of the ocean
-winds through the Delta ~\citep{espe16a}. These three cultivars have a
-higher degree of cold tolerance, high yield potential, and relatively
-short duration and are thus broadly grown with over 70\% of the area
-in these cultivars at any given time during the study
+In CA, the cultivars with broad adaptation and adoption over the
+33-year period of this study are M-202, M-206, and M-105. Cold
+tolerance is a major problem for much of the California rice producing
+region ~\citep{espe2017}, although it is a bigger problem in the
+southern portion of the Sacramento Valley due to the influence of the
+ocean winds through the Delta ~\citep{espe16a}. These three cultivars
+have a higher degree of cold tolerance, high yield potential, and
+relatively short duration and are thus broadly grown with over 70\% of
+the area in these cultivars at any given time during the study
 period~\citep{ucce}. There are several cultivars that were found to
@@
 
-Not accounting for yield decline over time, the estimated yield improvement of
-\Sexpr{only_release_coef} kg ha\tsups{-1} year\tsups{-1}
-(Fig. \ref{fig:wo_erosion}) is lower than the 37 to 60 kg ha\tsups{-1}
-year\tsups{-1} estimated by~\citet{Omar2014} in side-by-side
-comparisons. However, when accounting for yield decline over time and looking
-across all cultivars, there is no conclusive evidence of increased
-genetic yield potential in medium grain rice cultivars released by
-CCRRF's breeding improvement program over the period
+Not accounting for yield decline over time, the estimated yield
+improvement of \Sexpr{only_release_coef} kg ha\tsups{-1}
+year\tsups{-1} (Fig. \ref{fig:wo_erosion}) is lower than the 37 to 60
+kg ha\tsups{-1} year\tsups{-1} estimated by~\citet{Omar2014} in
+side-by-side comparisons. ~\citet{vanIttersum2013} defined potential
+yield as the yield of optimum management and cultivars given
+environmental limitations. Under this definition, potential yield
+could be reasonably estimated from the yields of 'M-202' and 'M-206'
+in experimental plots at the RES in the years 1984 and 2015,
+respectively. Given these criteria, it is clear that potential yield
+has changed little over the study period
+(Fig. \ref{fig:first_yrs}). When accounting for yield decline over
+time across all cultivars, there is no conclusive evidence of
+increased genetic yield potential in medium grain rice cultivars
+released by CCRRF's breeding improvement program over the period
 \Sexpr{min(mod_data[,'year'])} to \Sexpr{max(mod_data[,'year'])}.
-Grain yield over this period has stayed relatively constant at an average
-\Sexpr{round(mean(mod_data[,"yield_kg"])/1000, 1)} Mg
+Grain yield over this period has stayed relatively constant at an
+average \Sexpr{round(mean(mod_data[,"yield_kg"])/1000, 1)} Mg
 ha\tsups{-1}. Since the estimated yield improvement spans from
@@
 plausible. This analysis suggests that the CCRRF breeding program has
-maintained but not substantially improved the yield
-potential in the face of changing biotic and abiotic pressures. This
-conclusion holds true even when evaluating the cultivars which have
-been breed for broad adaption and high yield potential (M-202, M-206
-and M-105). Yields of these cultivars at time of release have all
-yielded between
+maintained but not substantially improved the yield potential in the
+face of changing biotic and abiotic pressures. This conclusion holds
+true even when evaluating the cultivars which have been breed for
+broad adaption and high yield potential (M-202, M-206 and
+M-105). Yields of these cultivars at time of release have all yielded
+between
 \Sexpr{tmp=mod_data[mod_data[,"yrs_in_trial"]<=0&mod_data[,"id"]%in%c("M202","M206","M105"),];round(min(by(tmp,tmp[,"id"],function(x)mean(x[,"yield_kg"]))/1000),2)}
@@
 temperatures reducing yields. Again, this is unlikely to be the case
-in this situation as ~\citet{espe2017} found no significant changes in
+in this situation as ~\citet{espe2017} found no substantial changes in
 growing season temperatures (either high, low, or average temps) the
@@
 reliably measure could plausibly lead to yield
-erosion~\citep{Peng1998, Peng2000, Peng2010}. However, it is not clear
+erosion~\citep{Peng1998, Peng2000, Peng2010, Mackay2011}. However, it is not clear
 if this mechanism is responsible for the yield decline over time of
@@
 other year and the residues are burned, further lowering the potential
-for pest and disease pressure.
+for pest and disease pressure. More data is needed to further explore
+the exact cause of the yield declines observed here, as well to
+determine if different cultivars have differential declines over time.
 
@@
 already short growing season and water limitations (shorter duration
-cultivars would presumably require less irrigation water).
+cultivars would presumably require less irrigation water). For a
+detailed discussion of cultivars' agronomic and grain quality
+characteristics, refer to \url{http://rice.ucanr.edu/Reports-Publications/Rice_Production_Workshop_Manual/}.
 
@@
 <<echo=FALSE, results=tex>>=
+ids = aggregate(release_yr ~ id, data = mod_data, unique)
+ids$dth = as.integer(round(aggregate(dth ~ id, data = mod_data, mean, na.rm = TRUE)$dth, 0))
+colnames(ids) = c("Cultivar", "Release Year", "Days to 50% Heading")
+ids = ids[order(ids[,2]),]
+ids[,2] = as.integer(ids[,2])
+ids[,1] = gsub("M", "M-", ids[,1])
+
+print(xtable(ids,
+             caption = "All named medium grain rice cultivars and their associated release year (from 1984 -- 2015) planted at the Rice Experiment Station (RES), Biggs, California.",
+             label = "tbl:ids"),
+      include.rownames = FALSE,
+      booktabs = TRUE,
+      caption.placement = "top")
+
+@
+
+\newpage
+
+<<echo=FALSE, results=tex>>=
 
@@
 @
-\newpage
-
-<<echo=FALSE, results=tex>>=
-ids = aggregate(release_yr ~ id, data = mod_data, unique)
-ids$dth = as.integer(round(aggregate(dth ~ id, data = mod_data, mean, na.rm = TRUE)$dth, 0))
-colnames(ids) = c("Cultivar", "Release Year", "Days to 50% Heading")
-ids = ids[order(ids[,2]),]
-ids[,2] = as.integer(ids[,2])
-ids[,1] = gsub("M", "M-", ids[,1])
-
-print(xtable(ids,
-             caption = "All named medium grain rice cultivars and their associated release year (from 1984 -- 2015) planted at the Rice Experiment Station (RES), Biggs, California.",
-             label = "tbl:ids"),
-      include.rownames = FALSE,
-      booktabs = TRUE,
-      caption.placement = "top")
-
-@
-
-% Notes: Duration has gotten shorter. Etc. Rearrange - End
-% introduction with evidence of erosion, quatify the effect on yield
-% increases.
 

### 2018-02-26 src/h1_model.R

@@ -88,5 +88,5 @@ if(compare){
     compare(loo(RES_full), loo(RES_fullest))
 }
 
-
-save(RES_fullest, file = "../output/h1_fit.Rda")
+save(RES_full, file = "../output/h1_fit.Rda")
+save(RES_fullest, file = "../output/h1_fit_revised.Rda")

### 2018-03-04 output/review_responses.tex
Adjust the response letter to not include the other pdfs.
Add sticky notes.

@@
 differences in the data being presented. We agree and plainly state in
-the manuscript that CA farm yields have been increasing (line
-). However, we do not find evidence that gains as large as 70 kg/ha/yr
+the manuscript that CA farm yields have been increasing (lines 236-246). However, we do not find evidence that gains as large as 70 kg/ha/yr
 are due to increasing genetic yield potential. It is entirely possible
@@
 
+\subsubsection{Sticky note comments}
+
+\textbf{Comment: would "unexplained" be a better word}\\
+
+\textbf{Response:} No - the use of inherent is intentional and conveys
+a different meaning.\\
+
+\textbf{Comment: The decline could be because the potential of the
+  environment is declining for unmanageable reasons. My point is that
+  your emphasis is all on the cultivars, when it has to be the
+  environment deteriorating, not the cultivar. }\\
+
+\textbf{Response: } See response above.\\
+
+\textbf{Comment: Thats if the stats are right. Only Fig S2 really
+  convinces me there is yield decline, but the cause could be
+  unrelated to potential yield, as would be a manageable stress to
+  which newer cvs are more resistant..}\\
+
+\textbf{Response: } It is inappropriate for the reviewer, who above plainly
+states that they do not understand the statistics, to make this
+evaluation.\\
+
+\textbf{Comment: Give sentence on location (latitudes) and area of rice in
+  CA. Even a map of the rice areas , also showing location of RES}\\
+
+\textbf{Well it went backwards for several yeasr, and then resumed
+  going up but at a slower rate. Somethink happened in mid 1990s?}\\
+
+\textbf{Response: } None needed - speculation.\\
+
+\textbf{Comment: again it may be semantics, but the newer cvs must be
+  better adapted to the deteriotating environment, thats breeding
+  progress, even potential yield increase if the deterioration is part
+  of the unmanageable environment (relative to the older cvs)}\\
+
+\textbf{Response: } See response above.\\
+
+\textbf{Comment: I cannot understand why there is no reference here in
+  Introduction or in Discussion to McKenzie et al 2014 , a Chapter in
+  the CSSA special publication No 33.}\\
+
+\textbf{Response: } Added to discussion.\\
+
+\textbf{Comment: More info on plots shape sown, and harvested please}\\
+
+\textbf{Response: } Added to methods.\\
+
+\textbf{Comment : Is this the best wording, as I understand year and
+  cultivar in the model account for what is left after the first two
+  terms. Like a bad year or a low yielding cv (maybe high quality)
+  will cause a deviation of plot yield from that explained by the
+  first 3 terms}\\
+
+\textbf{Response: } This is the correct wording.\\
+
+\textbf{Comment: is this the same as the McKay model? No, he has
+  cultivar and year effetrs (relative to grand mean). And why here the
+  same integer for year as for cultivar? Doesnt see quite correct the
+  equation?}\\
+
+\textbf{Response: } See response above.\\
+
+\textbf{Comment: Progress M202 to M204 and M 205 agree with McKenzies'
+  data from trials across farms and RES, with no correction for years
+  after release.  But farmer data Butte County show M-206 0.4 t/ha
+  over M-202. Are all their yield increases apparent because M-202 was
+  eroding for longer. I guess thats what you are saying. But McKenzie
+  doesnt show "erosion" of M-202 yields. Also where is M-201?}\\
+
+\textbf{Response: } See response above. We do not have data from the
+first two years of M-201, since that is before the study period.\\
+
+\textbf{Comment: Being suspicious of any mixed liner modelling (for
+  example why only 1 point for each year since release?) I find Fig S1
+  as much more convincing, esp cvs like M-202, M-103, M-104, M-204,
+  M-205. Why not include in the paper proper.  BUT the graphs of
+  McKenzie (Fig 10-8) show now decline in M-202 yield after about
+  1994. So what's going on at RES??}\\
+
+\textbf{Response: } Again, it is wholely inappropriate for the
+reviewer to be ``suspicious'' of the technique based solely on the
+grounds that they do not understand it.
+
+See response above regarding discrepencies between McKenzie et al.,
+2014 and our work.\\
+
+\textbf{Comment: agreed, but is there something which groups cvs with
+  declines and ones without??}\\
+
+\textbf{Response: } See response above.\\
+
+\textbf{Comment: Actually its over the time between release of M 202
+  and M 206, 20 years or about 0.6 t/ha.}\\
+
+\textbf{Response: } Or exactly 560 kg/ha. \\
+
+\textbf{Comment: Not in Fig 2 }\\
+
+\textbf{Response: } The figure reference is correct.\\
+
+\textbf{Comment: where are the 2 years before release for M-202 that
+  you used in Fig 2. Looking at the  10 years after release when the
+  two cvs can be compared (?), M-206 is no better that M-202 on this
+  basis?}\\
+
+\textbf{Response: } The data are in the figure. Yes, that is our conclusion as well.\\
+
+\textbf{Comment: M 201 not in Fig 2. And why is there not perfect
+  correspondence between Figs 2 and S2.}\\
+
+\textbf{Response: } Figure reference updated. These figure show
+different data, as indicated by the captions.\\
+
+\textbf{Comment: Again a map would help}\\
+
+\textbf{Response: } See response above.\\
+
+\textbf{Comment: This is where reference ot McKay et al (2010) TAG
+  122, 225-238 would be useful. They were able to pin down the
+  "erosion" to increasing disease susceptibility because they had plus
+  and minus fungicide. You cannot do that and I am not sure you
+  separate the effect of years after release from other year related
+  changes which may have been happening throughout (presumably
+  negative ones) and linearly correlated with year} \\
+
+\textbf{Response: } See response above.\\
+
+\textbf{Comment: Could you list these supposed stresses, they may be a
+  clue to the cause of erosion}\\
+
+\textbf{Response: } See response above.\\
+
+\textbf{Comment: But you have to argue also that breeding is
+  countering these negativities such that the newest cv is yielding
+  more that the older one in any given year. This is actually
+  potential yield progress if the "potential"" of the E has
+  deteriorated because of growing non manageable stresses like less
+  sunshine, chilling, or more ozone.}\\
+
+\textbf{Response: } See response above.\\
+
+\textbf{Comment: see McKay et al (2010) on "erosion" due to evolving
+  biotic stress organisms}\\
+
+\textbf{Response: } See response above.\\
+
+\textbf{Comment: This is the likely cause: a more thorough exploration
+  of biotic stresses is needed : blast, nematodes, etc}\\
+
+\textbf{Response: } As we do not have the data to explore this, this
+discussion would be solely speculation. For more, see our response
+above.\\
+
+\textbf{Comment: Yes, this would be alarming....a steady decrease in
+  the potential of all paddy Es, but why now after millennia of
+  paddy?}\\
+
+\textbf{Response: } As we state, more study is needed to verify and
+determine the extent and cause of this (lines ).\\
+
+\textbf{Comment: thats 1.25\% p.a. increase, almost noone is achieving
+  that in any crop! Except when you jump up to hybrids, a one off
+  jump. But your point is valid, yield progress is slowed by the
+  "erosion" . I dont think the Figure 6,  however, is needed.}\\
+
+\textbf{Response: } This is the rate of improvement which has been
+claimed by the rice breeders in this system (lines ). The figure is
+helpful to distill the major mechanisms at work here.\\
+
+\textbf{Comment: Not in Fig 2}
+
+\textbf{Response: } See response above.\\
+
+\textbf{Comment: Maybe would help to mention the problems which hit in
+  the mid 1990s, so that farm yield went backwards for several years,
+  and then started the climb you talk about.}\\
+
+\textbf{Response: } See response above.\\
+
+\textbf{Comment: As shown in Butte Co in McKenxzie et al 2014, another
+  reason for a map!}\\
+
+\textbf{Response:}\\
+
+\textbf{Comment: Again, a special period of instabiliy from early
+  1990s for a decade or so. Why} \\
+
+\textbf{Response: } See response above.\\
+
+\textbf{Comment: Not sure you need Fig 6 or the last paragraph. The
+  issue is we need to know why erosion is occuring , and only it seems
+  in some cultivars (see Fig S2?)}\\
+
+\textbf{Response: } See response above.\\
+
+\textbf{Comment: This is only possible if you account also for all
+  environmental changes (climate plus agronomy)....you need to be
+  running a rice simulation model continuously. Note climate is more
+  than just T, also radiation, ratio diffuse to total, ozone, CO2 ),
+  and with paddy in Ca there is also water quality and
+  temperature. And agronomy is also gradual soil changes, fallow may
+  not be enough.}\\
+
+\textbf{Response: } See response above.\\
+
+\textbf{Comment: Explain better what you mean}\\
+
+\textbf{Response: } Done.\\
+
+\textbf{Comment: Improved management can contribute to increased
+  potential yield, if its "new" agronomic technology.}\\
+
+\textbf{Response: } See response above, as well as the reviewer's own
+definition of yield potential.\\
+
+\textbf{Response: } See response above.\\
+
+\textbf{Comment: are some dot black and some grey or is it my eyes?}\\
+
+\textbf{Response: } See response above.\\
+
 \subsection{Reviewer 2}
@@
 
-\textbf{Response:} Corrected.
+\textbf{Response:} Corrected.\\
 
-\textbf{Comment: L141: Something is wrong with “…since release is between…”}
+\textbf{Comment: L141: Something is wrong with “…since release is between…”}\\
 
@@
 
-\newpage
-
-\includepdf{diffs.pdf}
-
-\newpage
-
-\includepdf{Espe_rice_yield_improvement_revised.pdf}
-
 \end{document}

### 2018-03-04 output/breed_paper_revised.Rnw

@@
 plots were planted by pre-germinating rice seed prior to
-direct-seeding into pre-flooded fields. Plots were either 14.0 or 18.5
+direct-seeding into pre-flooded fields. Plots were rectangular and either 14.0 or 18.5
 m\tsups{2}, and were harvested after physiological maturity using a
@@
 \Sexpr{paste(gsub('M','M-',rev(row.names(id_est))[c(1,2,4)]),colla=', ')}
-(Figs. \ref{fig:first_yrs} \& \ref{fig:ids_comp}).
+(Fig. \ref{fig:ids_comp}).
 
@@
 25 kg\tsups{-1} ha\tsups{-1} year\tsups{-1}. There are several
-explanations for how this could be. 1) \citet{mckenzie2014} did not
+explanations for this discrepancy in conclusions. First, \citet{mckenzie2014} did not
 account for yield decreases over time in their analysis. As shown
@@
 increase in yield over time is estimated to be approximately the same
-as that found by \citet{mckenzie2014}. 2) These yield decreases over time
+as that found by \citet{mckenzie2014}. Second, these yield decreases over time
 were not taken into account possibly because \citet{mckenzie2014}
@@
 confirmation of no trend. Instead, failing to reject the hypothesis
-could suggest the analysis lacks the statistical power to detect the trend. 3) It
+could suggest the analysis lacks the statistical power to detect the
+trend. Third, it
 is not surprising that no trend in yield over time was detected given
@@
 different management systems, growing environments, and crop
-timing. By contrast, this investigation presented here used a single,
+timing. By contrast, the investigation presented here used a single,
 well managed system and analyzed the data using a mixed effects model

### 2018-03-05 output/review_responses.tex
Adds map.

@@
 
+\textit{Note: as the reviewer points out, these comments are largely
+  repeated from above. Hence, for many comments, please refer to our
+  responses above.}\\
+
 \textbf{Comment: would "unexplained" be a better word}\\
@@
 
-\textbf{Response: } None needed - speculation.\\
+\textbf{Response: } This is speculation, and not directly relevant to
+this study.\\
 
@@
 
-\textbf{Response: } Or exactly 560 kg/ha. \\
+\textbf{Response: } Or, more exactly 0.56 t/ha (560 kg/ha), as we
+state in the manuscript.\\
 

### 2018-03-05 output/breed_paper_revised.Rnw
Adds map.

@@
 \usepackage{lscape} %% Harvard
+\usepackage[shortcuts]{extdash}
 \bibliographystyle{model2-names}\biboptions{authoryear}
@@
 The CCRRF Rice Experiment Station (RES) is located near Biggs, CA
-(39.4648, -121.7342), and has been the central location for CCRRF's
+(39.4648, \==121.7342; Fig. \ref{fig:map}), and has been the central location for CCRRF's
 efforts to develop improved rice cultivars adapted to CA since
@@
 \begin{figure}[H]
+  \includegraphics[]{map.pdf}
+  \caption{The locations of the Rice Experiment Station relative to
+    rice harvested area in California, USA.}
+  \label{fig:map}
+\end{figure}
+
+\newpage
+
+\begin{figure}[H]
 <<fig=TRUE, echo=FALSE>>=

### 2018-03-05 src/maps.R
data and code to make map.

@@ -1,0 +1,43 @@
+library(raster)
+library(RColorBrewer)
+library(rgdal)
+library(maps)
+map_data_loc = "../data/map_data.rda"
+
+if(!file.exists(map_data_loc)){
+    usda_rice <- raster("~/Dropbox/D--CDL-NASS_DATA_CACHE-extract_3_CDL_2015_stat_clip_20161101201010_1475574269.tif")
+    usda_rice <- aggregate(usda_rice, fact = 100)
+    usda_rice <- projectRaster(usda_rice, crs = CRS("+init=epsg:4326"))
+
+    usda_rice <- crop(usda_rice, extent(-125, -120, 36, 41))
+    usda_rice[usda_rice < 0.01] <- NA
+    usda_rice <- trim(usda_rice)
+    save(usda_rice, file = map_data_loc)
+}else{
+    load(map_data_loc)
+} 
+cols = brewer.pal(9, "YlGn")
+
+pdf(file = "../output/map.pdf")
+map("state", "California", xlim = c(-123.5,-120.5), ylim = c(37,40),
+    fill = TRUE, col = "grey80")
+map.axes()
+map.scale(x = -123.4, y = 37.3, ratio = FALSE, relwidth = 0.15,
+          cex = 0.8)
+
+plot(usda_rice, add = TRUE,
+     col = cols[6],
+     # col = "chartreuse3",     
+     # col = colorRampPalette(cols)(255),
+     legend = FALSE, interpolate = FALSE)
+points(y = 39.4648, x = -121.7342, pch = 21, col = "black", bg = "white", cex = 2)
+legend("bottomright", legend = c("Rice Experiment Station",
+                              "Rice harvested area"),
+       pch = c(21, NA),
+       bg = "white",
+       fill = c(NA,cols[6]),
+       border = "white",
+       cex = 1.1,
+       bty = "o")
+text(-123, 37.7, "San Francisco\n Bay", cex = 0.8, font = 3)                           
+dev.off()

### 2018-03-05 output/review_responses.tex
Final comments.

@@
 
+\textbf{Response: } Map added to supplemental figures.
+
 \textbf{Well it went backwards for several yeasr, and then resumed
@@
 \textbf{Response: } As we state, more study is needed to verify and
-determine the extent and cause of this (lines ).\\
+determine the extent and cause of this (lines 209-210).\\
 
@@
 \textbf{Response: } This is the rate of improvement which has been
-claimed by the rice breeders in this system (lines ). The figure is
-helpful to distill the major mechanisms at work here.\\
+claimed by the rice breeders in this system. This information has been
+added to the manuscript (line 39). The figure is
+helpful to distill the major mechanisms at work here to a simple
+conceptual description.\\
 

### 2018-03-05 output/breed_paper_revised.Rnw
Final comments.

@@
 \usepackage{lscape} %% Harvard
-\usepackage[shortcuts]{extdash}
+%% \usepackage[shortcuts]{extdash}
 \bibliographystyle{model2-names}\biboptions{authoryear}
@@
 trials show that new cultivars usually out yield the older
-cultivars. In other rice systems, a decrease in a cultivar's yield
+cultivars, typically by 3--5\% (e.g., M-209 release description at \url{http://www.crrf.org/linked/2015annualreport.pdf}). In other rice systems, a decrease in a cultivar's yield
 performance over time has been observed~\citep{Peng1998, Peng2010,
@@
 The CCRRF Rice Experiment Station (RES) is located near Biggs, CA
-(39.4648, \==121.7342; Fig. \ref{fig:map}), and has been the central location for CCRRF's
+(39.4648, -121.7342; Fig. \ref{fig:map}), and has been the central location for CCRRF's
 efforts to develop improved rice cultivars adapted to CA since

### 2018-03-11 output/breed_paper_revised.Rnw
Commit before large yank.

@@
 and close to maximum yield potential. Specifically, the hypothesis was
-tested that if rice cultivar yields decline over time (termed yield
-erosion) then apparent yield increases in side-by-side yield
-comparison tests will not reflect increases in yield potential. This
-hypothesis was tested using 33 years of experimental yield data from
-the California Cooperative Rice Research Foundation Rice Experiment
-Station. Based on side-by-side comparisons of old and new rice
-cultivars which do not consider yield decline over time, there was an apparent
-increase in yield. However, the yields of older cultivars eroded at an
-estimated rate of 29.3 kg ha\tsups{-1} year\tsups{-1} (90\% credible
-interval -4.4 to -53.3) after initial selection. Once this effect was
+tested that if rice cultivar yields decline over time then apparent
+yield increases in side-by-side yield comparison tests will not
+reflect increases in yield potential. This hypothesis was tested using
+33 years of experimental yield data from the California Cooperative
+Rice Research Foundation Rice Experiment Station. Based on
+side-by-side comparisons of old and new rice cultivars which do not
+consider yield decline over time, there was an apparent increase in
+yield. However, the yields of older cultivars declined at an estimated
+rate of 29.3 kg ha\tsups{-1} year\tsups{-1} (90\% credible interval
+-4.4 to -53.3) after initial selection. Once this effect was
 considered, the yield advantage of newer cultivars over old was
@@
 
-\begin{keyword} rice, yield potential, breeding, yield erosion
+\begin{keyword} rice, yield potential, breeding, yield declines
 \end{keyword}
@@
 
-\linenumbers
+%% \linenumbers
 
@@
 Certainly one question in looking at this situation is, ``do new
-cultivars have increased yield potential''?  Like many breeding
+cultivars have greater yield potential''?  Like many breeding
 programs, one of the goals of the CCRRF is to increase yields through
@@
 conditions~\citep{Peng1998, Peng2010, Peng2000, Datta1995}. Yield
-erosion may explain the apparent contradiction above, in which
+declines over time may explain the apparent contradiction above, in which
 side-by-side yield comparisons show yield improvements in cultivars
@@
 \begin{split}
- yield_{ijk} = Intercept + (Rate_{yield\ erosion}* {years\ since\
+ yield_{ijk} = Intercept + (Rate_{yield\ decline}* {years\ since\
    release\ ik}) \\+ (Rate_{yield\ improvement}* {release\ year\ ik})\\ 
@@
 the Intercept is the average yield for the RES across cultivars and years, $Rate_{yield\
-erosion}$ is rate of yield decline per ${years\ since\ release\ i}$
+decline}$ is rate of yield decline per ${years\ since\ release\ i}$
 (mean-centered, i.e., negative for years prior to release, zero at the year of release),
@@
 into account, the model was used as described above except omitting
-the $Rate_{yield\ erosion\ i}$ term (Model 2). Lastly, a model was fit
+the $Rate_{yield\ decline\ i}$ term (Model 2). Lastly, a model was fit
 to estimate if there were differences between cultivars in the rate of
@@
 supplemental material. Comparing the cultivars' yield during each's
-initial time (average 3 years) in the experimental trials just before
-being released showed that across cultivars, yields ranged from
+initial time (average 3 years) in the experimental trials showed that across cultivars, yields ranged from
 \Sexpr{round(plot_data["M207", "mean"], 2)} (M-207) to
 \Sexpr{round(plot_data["M205", "mean"], 2)} Mg ha\tsups{-1} (M-205)
-(Fig \ref{fig:first_yrs}. Comparing the yields from only the first few
-years for each cultivar over the study period does not provide
-evidence of genetic yield improvements during the 33-year study
+(Fig \ref{fig:first_yrs}). The relationship between the year
+a cultivar was released and yield over the 33-year study period was
+estimated to be slightly negative (\Sexpr{release_yr_coef}
+kg\tsups{-1} ha\tsups{-1} yr) and does not provide
+evidence of genetic yield improvements during this
 period. Individual exceptions to this were two cultivars (M-204 and
@@
   indicate experimental yields prior to the official release. The
-  estimated erosion of yield over time (averaged over differences in
+  estimated decline of yield over time (averaged over differences in
   cultivar and year) is estimated to be \Sexpr{yrs_coef} kg ha\tsups{-1}
@@
 
-There was no evidence of consistent differences between cultivars in
-the rate of yield decline over time (Model 3), hence the model excluding varying
-rates of erosion by cultivar (Model 2) is presented here. The yields
+Based on Model 3, there was no evidence of consistent differences between cultivars in
+the rate of yield decline over time (i.e., all cultivar specific rates
+of decline overlapped). Differences between cultivars in the decline
+of yield over time, seen in the raw data (Fig. \ref{fig:all_cult}),
+are therefore likely small and uncertain. Given this, the model excluding varying
+rates of decline by cultivar (Model 2) is presented here. The yields
 of each cultivar over time were evaluated relative to when the
 cultivar was released using the full model which accounts for both
-yield decline over time and yield improvements over time. The average decrease
-in yield over time once a cultivar was selected
+yield decline over time and yield improvements over time (Model 2). The average decrease
+in yield over time following a cultivar's release
 was estimated at \Sexpr{yrs_coef} kg ha\tsups{-1} year\tsups{-1} (90\%
@@
 advantage of M-206 over M-202 (the current and former predominant
-cultivars, respectively), can be attributed solely to the erosion
+cultivars, respectively), can be attributed solely to the decline
 of M-202's yield over time (Figs \ref{fig:first_yrs} \&
 \ref{fig:compare_yrs}). Further supporting this estimate of yield
-erosion, both cultivars yielded similarly at their respective year of
+decline, both M-202 and M-205 yielded similarly at their respective year of
 release (Fig. \ref{fig:compare_yrs}), despite the fact that M-202 was
@@
        data = plot_data[[2]],
-       pch = 17, col = "grey40")
+       pch = 1)
 segments(x0 = plot_data[[2]]$yrs_in_trial,
@@
 legend("topright", legend = gsub("M","M-", tars),
-       pch = c(16, 17), col = c("black","grey40"))
+       pch = c(16, 1))
 text(5, 6.7, bquote(.(yrs_coef)~kg~ha^-1~year^-1))
@@
 When yield decline over time is not accounted for in the analysis (Model 2),
-there is estimated to be an increase per year of approximately
+there is estimated to be a yield increase of approximately
 \Sexpr{only_release_coef} kg ha\tsups{-1} year\tsups{-1} (90\% CI
 \Sexpr{only_release_CI[1]} to \Sexpr{only_release_CI[2]})
-(Fig. \ref{fig:wo_erosion}). However, the full model (Model 1) which
+(Fig. \ref{fig:wo_erosion}). In contrast, the full model (Model 1) which
 accounts for both the year of release and yield decline over time indicates that
@@
 \Sexpr{rel_CI[2]}) (Fig. \ref{fig:first_yrs}). According to model
-comparison metrics (PS-LOO, WAIC), the full model (Model 1) is
+comparison metrics (PS-LOO, WAIC), Model 1 is
 estimated to have superior predictive accuracy compared to the model
-which does not account for yield decline over time (Model 2)
+which does not account for yield decline over time.
 
@@
 cultivars with higher than average yields:
-\Sexpr{paste(gsub('M','M-',rev(row.names(id_est))[c(1,2,4)]),colla=', ')}
-(Fig. \ref{fig:ids_comp}).
+M-201, M-205, M-204 (Fig. \ref{fig:ids_comp}).
 
@@
 
+\subsection{Cultivars within the context of the production system}
+
 In CA, the cultivars with broad adaptation and adoption over the
@@
 have better than average yields, even after correcting for yield
-erosion. However, these cultivars (M-201, M-204, M-205, and possibly
+declines over time. However, these cultivars (M-201, M-204, M-205, and possibly
 M-209; Fig. \ref{fig:ids_comp})) are not widely adopted. Importantly,
@@
 
+\subsection{Yield changes over time}
+
+The high yielding variety M-202 was released in 1985 and became the
+most widely planted cultivar for two decades~\citep{mkenzie2014}. At
+the time of release, M-202 yields were reported at 11.0 Mg ha\tsups{-1}~\citep{m202} which is almost identical to what we . This value of
+11.0 Mg ha\tsups{-1} is what we also found in this analysis based on the
+average experimental yields prior to public release (Fig. \ref{fig:first_yrs}). Since
+the release of M-202, the rice yields of newly released varieties
+cultivars have remained between 10.8 and 11.5 Mg ha\tsups{-1} (Fig. \ref{first_yrs}),
+with the exception of M-204 and M-205 (M-204 is no longer grown and
+M-205 has limited adoaptation range so is only grown in a limited
+area). The variety cultivar M-206 was released in 2004 and is
+currently still the most widely grown rice varietycultivar in
+California. At the time of its release, M-206 was reported to yield
+\Sexpr{round(9350*1.12,1)}
+Mg ha\tsups{-1} -- also similar to what we found based on our analysis (Fig. \ref{fig:first_yrs}). Based on this premise, it is clear that improvements
+in yield potential have not been made over the 33-yr period of this
+study. In contrast, both~\citet{Omar2014} and ~\citet{mckenzie2014}, report yield potential gains in California rice systems
+based on analyses that showed
+that newly released cultivars had higher yields than previously
+released cultivars. Based on their
+analyses,~\citet{Omar2014} reported yield gains from the rice
+breeding program of 37 to 60 kg ha\tsups{-1} yr\tsups{-1} and ~\citet{mckenzie2014} a
+yield increase of 25 kg ha\tsups{-1} yr\tsups{-1}. Using a similar
+approach to~\citet{Omar2014} and ~\citet{mckenzie2014}, when the compared average yields
+for the period 2011--2016 of varieties cultivars released since 1982 (and still in
+the variety trials from 2011-- 2016) was compared without accounting
+for yield declines over time, yields were
+estimated to increase by \Sexpr{only_release_coef} kg ha\tsups{-1}
+yr\tsups{-1} (Fig. \ref{fig:wo_erosion}), almost identical to what
+~\citet{mckenzie2014}.
+
+The only way to reconcile the two apparently
+contradictory findings, is that the yield of potential of varieties
+cultivars decline over time such that when newly released varieties
+cultivars are compared with the older variety cultivar (as shown by
+~\citet{Omar2014}, ~\citet{mckenzie2014}, and Fig. \ref{fig:wo_erosion}), yields of
+the new cultivar variety are higher. Here we provide evidence that
+cultivar varietal yield does decline - on average at a rate of \Sexpr{yrs_coef} kg
+ha\tsups{-1} yr\tsups{-1} (Fig. \ref{w_erosion}). ~\citet{Peng2010} also found that a cultivar’s
+varieties yield potential declines over time and reported a 15\%
+decline over a 30 yr period.  In this study, over a 30 year period
+and a \Sexpr{yrs_coef} kg ha\tsups{-1} yr\tsups{-1} yield decline rate, the yield
+decline would be approximately
+\Sexpr{round((yrs_coef*-30)/mean(mod_data$yield_kg), 3)*100}\%.
+
 Not accounting for yield decline over time, the estimated yield
@@
 kg ha\tsups{-1} year\tsups{-1} estimated by~\citet{Omar2014} in
-side-by-side comparisons. ~\citet{vanIttersum2013} defined potential
-yield as the yield of optimum management and cultivars given
-environmental limitations. Under this definition, potential yield
+side-by-side comparisons. To evaluate if 37 to 60 kg ha\tsups{-1}
+year\tsups{-1} is reasonable, the potential yields at
+the beginning and end of the study period can be compared. ~\citet{vanIttersum2013} defined potential
+yield as the yield of optimum management and well-adapted cultivars given
+current environmental limitations. Under this definition, potential yield
 could be reasonably estimated from the yields of 'M-202' and 'M-206'
@@
 reliably measure could plausibly lead to yield
-erosion~\citep{Peng1998, Peng2000, Peng2010, Mackay2011}. However, it is not clear
+declines over time~\citep{Peng1998, Peng2000, Peng2010, Mackay2011}. However, it is not clear
 if this mechanism is responsible for the yield decline over time of
@@
 potentially be endemic to all rice production systems. While efforts
-should be made to identify the cause of erosion, this study highlights
+should be made to identify the cause of yield decline over time, this study highlights
 both the need for maintenance of plant breeding and the difficulty of
@@
 improvements via plant improvement, (2) no maintenance breeding, where
-grain yield slowly erodes over time (as estimated from this study),
+grain yield slowly declines over time (as estimated from this study),
 and (3) absolute improvement in grain yield (based on ``typical''
@@
 increases over time, the improvement target needs take into account
-erosion in the current ``standard'' cultivars' yield over time
+declines in the current ``standard'' cultivars' yield over time
 (Fig. \ref{fig:theory}). While side-by-side comparisons will continue
@@
 indicate gains in yield potential as they do not account for yield
-erosion. Quantifying changes in yield potential requires also
+declines over time. Quantifying changes in yield potential requires also
 examining historical benchmarks. Improvement in grain yield is

### 2018-03-11 output/breed_paper_revised.Rnw
After large yank.

@@
 
-Not accounting for yield decline over time, the estimated yield
-improvement of \Sexpr{only_release_coef} kg ha\tsups{-1}
-year\tsups{-1} (Fig. \ref{fig:wo_erosion}) is lower than the 37 to 60
-kg ha\tsups{-1} year\tsups{-1} estimated by~\citet{Omar2014} in
-side-by-side comparisons. To evaluate if 37 to 60 kg ha\tsups{-1}
-year\tsups{-1} is reasonable, the potential yields at
-the beginning and end of the study period can be compared. ~\citet{vanIttersum2013} defined potential
-yield as the yield of optimum management and well-adapted cultivars given
-current environmental limitations. Under this definition, potential yield
-could be reasonably estimated from the yields of 'M-202' and 'M-206'
-in experimental plots at the RES in the years 1984 and 2015,
-respectively. Given these criteria, it is clear that potential yield
-has changed little over the study period
-(Fig. \ref{fig:first_yrs}). When accounting for yield decline over
-time across all cultivars, there is no conclusive evidence of
-increased genetic yield potential in medium grain rice cultivars
-released by CCRRF's breeding improvement program over the period
-\Sexpr{min(mod_data[,'year'])} to \Sexpr{max(mod_data[,'year'])}.
-Grain yield over this period has stayed relatively constant at an
-average \Sexpr{round(mean(mod_data[,"yield_kg"])/1000, 1)} Mg
-ha\tsups{-1}. Since the estimated yield improvement spans from
-\Sexpr{rel_CI[1]} to \Sexpr{rel_CI[2]} kg ha\tsups{-1} year\tsups{-1}
-(Fig. \ref{fig:RES_year}; Table \ref{tbl:coef}), both increases and
-decreases in the genetic yield potential over the study period are
-plausible. This analysis suggests that the CCRRF breeding program has
+This analysis suggests that the CCRRF breeding program has
 maintained but not substantially improved the yield potential in the
-face of changing biotic and abiotic pressures. This conclusion holds
+face of changing biotic and abiotic pressures over the 33 year time
+frame of this study. This conclusion holds
 true even when evaluating the cultivars which have been breed for

### 2018-03-11 output/breed_paper_revised.Rnw
Another big yank to conclusions.

@@
 most widely planted cultivar for two decades~\citep{mkenzie2014}. At
-the time of release, M-202 yields were reported at 11.0 Mg ha\tsups{-1}~\citep{m202} which is almost identical to what we . This value of
-11.0 Mg ha\tsups{-1} is what we also found in this analysis based on the
-average experimental yields prior to public release (Fig. \ref{fig:first_yrs}). Since
-the release of M-202, the rice yields of newly released varieties
-cultivars have remained between 10.8 and 11.5 Mg ha\tsups{-1} (Fig. \ref{first_yrs}),
-with the exception of M-204 and M-205 (M-204 is no longer grown and
-M-205 has limited adoaptation range so is only grown in a limited
-area). The variety cultivar M-206 was released in 2004 and is
-currently still the most widely grown rice varietycultivar in
-California. At the time of its release, M-206 was reported to yield
-\Sexpr{round(9350*1.12,1)}
-Mg ha\tsups{-1} -- also similar to what we found based on our analysis (Fig. \ref{fig:first_yrs}). Based on this premise, it is clear that improvements
-in yield potential have not been made over the 33-yr period of this
-study. In contrast, both~\citet{Omar2014} and ~\citet{mckenzie2014}, report yield potential gains in California rice systems
-based on analyses that showed
-that newly released cultivars had higher yields than previously
-released cultivars. Based on their
-analyses,~\citet{Omar2014} reported yield gains from the rice
-breeding program of 37 to 60 kg ha\tsups{-1} yr\tsups{-1} and ~\citet{mckenzie2014} a
-yield increase of 25 kg ha\tsups{-1} yr\tsups{-1}. Using a similar
-approach to~\citet{Omar2014} and ~\citet{mckenzie2014}, when the compared average yields
-for the period 2011--2016 of varieties cultivars released since 1982 (and still in
+the time of release, M-202 yields were reported at 11.0 Mg
+ha\tsups{-1}~\citep{m202} which is almost identical to what we . This
+value of 11.0 Mg ha\tsups{-1} is what we also found in this analysis
+based on the average experimental yields prior to public release
+(Fig. \ref{fig:first_yrs}). Since the release of M-202, the rice
+yields of newly released varieties cultivars have remained between
+10.8 and 11.5 Mg ha\tsups{-1} (Fig. \ref{first_yrs}), with the
+exception of M-204 and M-205 (M-204 is no longer grown and M-205 has
+limited adoaptation range so is only grown in a limited area). The
+variety cultivar M-206 was released in 2004 and is currently still the
+most widely grown rice varietycultivar in California. At the time of
+its release, M-206 was reported to yield \Sexpr{round(9350*1.12,1)} Mg
+ha\tsups{-1} -- also similar to what we found based on our analysis
+(Fig. \ref{fig:first_yrs}). Based on this premise, it is clear that
+improvements in yield potential have not been made over the 33-yr
+period of this study. In contrast, both~\citet{Omar2014} and
+~\citet{mckenzie2014}, report yield potential gains in California rice
+systems based on analyses that showed that newly released cultivars
+had higher yields than previously released cultivars. Based on their
+analyses,~\citet{Omar2014} reported yield gains from the rice breeding
+program of 37 to 60 kg ha\tsups{-1} yr\tsups{-1} and
+~\citet{mckenzie2014} a yield increase of 25 kg ha\tsups{-1}
+yr\tsups{-1}. Using a similar approach to~\citet{Omar2014} and
+~\citet{mckenzie2014}, when the compared average yields for the period
+2011--2016 of varieties cultivars released since 1982 (and still in
 the variety trials from 2011-- 2016) was compared without accounting
-for yield declines over time, yields were
-estimated to increase by \Sexpr{only_release_coef} kg ha\tsups{-1}
-yr\tsups{-1} (Fig. \ref{fig:wo_erosion}), almost identical to what
+for yield declines over time, yields were estimated to increase by
+\Sexpr{only_release_coef} kg ha\tsups{-1} yr\tsups{-1}
+(Fig. \ref{fig:wo_erosion}), almost identical to what
 ~\citet{mckenzie2014}.
 
-The only way to reconcile the two apparently
-contradictory findings, is that the yield of potential of varieties
-cultivars decline over time such that when newly released varieties
-cultivars are compared with the older variety cultivar (as shown by
-~\citet{Omar2014}, ~\citet{mckenzie2014}, and Fig. \ref{fig:wo_erosion}), yields of
-the new cultivar variety are higher. Here we provide evidence that
-cultivar varietal yield does decline - on average at a rate of \Sexpr{yrs_coef} kg
-ha\tsups{-1} yr\tsups{-1} (Fig. \ref{w_erosion}). ~\citet{Peng2010} also found that a cultivar’s
+The only way to reconcile the two apparently contradictory findings,
+is that the yield of potential of varieties cultivars decline over
+time such that when newly released varieties cultivars are compared
+with the older variety cultivar (as shown by ~\citet{Omar2014},
+~\citet{mckenzie2014}, and Fig. \ref{fig:wo_erosion}), yields of the
+new cultivar variety are higher. Here we provide evidence that
+cultivar varietal yield does decline - on average at a rate of
+\Sexpr{yrs_coef} kg ha\tsups{-1} yr\tsups{-1}
+(Fig. \ref{w_erosion}). ~\citet{Peng2010} also found that a cultivar’s
 varieties yield potential declines over time and reported a 15\%
-decline over a 30 yr period.  In this study, over a 30 year period
-and a \Sexpr{yrs_coef} kg ha\tsups{-1} yr\tsups{-1} yield decline rate, the yield
-decline would be approximately
+decline over a 30 yr period.  In this study, over a 30 year period and
+a \Sexpr{yrs_coef} kg ha\tsups{-1} yr\tsups{-1} yield decline rate,
+the yield decline would be approximately
 \Sexpr{round((yrs_coef*-30)/mean(mod_data$yield_kg), 3)*100}\%.
@@
 broad adaption and high yield potential (M-202, M-206 and
-M-105). Yields of these cultivars at time of release have all yielded
-between
-\Sexpr{tmp=mod_data[mod_data[,"yrs_in_trial"]<=0&mod_data[,"id"]%in%c("M202","M206","M105"),];round(min(by(tmp,tmp[,"id"],function(x)mean(x[,"yield_kg"]))/1000),2)}
-    and
-    \Sexpr{round(max(by(tmp,tmp[,"id"],function(x)mean(x[,"yield_kg"]))/1000),2)}
-    Mg ha\tsups{-1} with no indication of yields increasing
-    (Fig. \ref{fig:first_yrs}).
+M-105).
+
+\subsection{Causes of declining yields}
 
@@
 at this location and unlike most CA rice fields, the fields are left
-fallow every other year. Second, changing climate may influence yields
-with either warmer~\citep{van2013} or cooler~\citep{espe2017}
-temperatures reducing yields. Again, this is unlikely to be the case
-in this situation as ~\citet{espe2017} found no substantial changes in
-growing season temperatures (either high, low, or average temps) the
-Sacramento Valley during the study period. Thirdly, small changes in
-biotic and abiotic forces have previously been speculated as the cause
-for yield decline over time of the cultivar IR8 at IRRI~\citep{Peng1998,
-Peng2000, Peng2010}. Since the cultivar's genetics remain stable from
-the time of selection but the conditions the cultivar experiences over
-time are continuously changing, accumulated changes too small to
-reliably measure could plausibly lead to yield
-declines over time~\citep{Peng1998, Peng2000, Peng2010, Mackay2011}. However, it is not clear
-if this mechanism is responsible for the yield decline over time of
-\Sexpr{yrs_coef} Mg ha\tsups{-1} year\tsups{-1} found in this
+fallow every other year. Secondly, small changes in biotic and abiotic
+forces have previously been speculated as the cause for yield decline
+over time of the cultivar IR8 at IRRI~\citep{Peng1998, Peng2000,
+  Peng2010}. Since the cultivar's genetics remain stable from the time
+of selection but the conditions the cultivar experiences over time are
+continuously changing, accumulated changes too small to reliably
+measure could plausibly lead to yield declines over
+time~\citep{Peng1998, Peng2000, Peng2010, Mackay2011}. However, it is
+not clear if this mechanism is responsible for the yield decline over
+time of \Sexpr{yrs_coef} Mg ha\tsups{-1} year\tsups{-1} found in this
 analysis. Biotic stressors such as pests and diseases are low in this
@@
 other year and the residues are burned, further lowering the potential
-for pest and disease pressure. More data is needed to further explore
-the exact cause of the yield declines observed here, as well to
-determine if different cultivars have differential declines over time.
-
-It is striking that the estimated yield decline over time is similar to that
-reported by \citet{Peng2010} over a similar time span (15\% decline
-over roughly 30 years). This suggests the mechanisms responsible could
-potentially be endemic to all rice production systems. While efforts
-should be made to identify the cause of yield decline over time, this study highlights
-both the need for maintenance of plant breeding and the difficulty of
-increasing genetic yield ~\citep{espe2017} potential, especially in
-systems such as CA where yield is approaching the physiological limit
-\citep{espe16a}.
+for pest and disease pressure. Thirdly, changing climate may influence
+yield potential with either warmer~\citep{van2013} or
+cooler~\citep{espe2017} temperatures reducing
+yields. Yield potential can be defined as the yield of optimum
+cultivar in the absence of biotic and abiotic stresses in a given
+environment \citep{vanIttersum2013}. However, ~\citet{espe2017} found
+no substantial trends in growing season temperatures (either high,
+low, or average temps) in the Sacramento Valley during the study
+period, suggesting that older cultivars do not have lower yield
+potential due to changing climatic factors. While our analysis
+indicates that there is no interaction between cultivar and year (CI
+includes zero), visual examination suggests that cultivars may be
+different in terms of rates of yield decline
+(Fig. \ref{fig:all_cult}); however, we do not have sufficient data nor
+the proper controls in the existing data set to fully test this
+hypothesis. More data is needed to further explore the exact cause of
+the yield declines observed here, as well to determine if different
+cultivars have differential declines over time.
+
+\subsection{Causes for statewide increases over time}
+
+\begin{comment}
+It is striking that the estimated yield decline over time is similar
+to that reported by \citet{Peng2010} over a similar time span (15\%
+decline over roughly 30 years). This suggests the mechanisms
+responsible could potentially be endemic to all rice production
+systems. While efforts should be made to identify the cause of yield
+decline over time, this study highlights both the need for maintenance
+of plant breeding and the difficulty of increasing genetic yield
+~\citep{espe2017} potential, especially in systems such as CA where
+yield is approaching the physiological limit \citep{espe16a}.
 
@@
 
-Our investigation here is narrowly focused on the rough-rice/field
-grain yield. Other advancements due to plant improvement, including
-increased quality~\citep{mckenzie1993} and milling
-yield~\citep{mckenzie1994}, have been made over the study period and
-are important. Additionally, crop duration has been reduced in CA rice
-cultivars. For example, while the predominate cultivars M-202 and
-M-206 had similar yield potentials at the time of release (Fig. 5),
-M-206 matures about 6 days earlier than M-202 (Table
-\ref{tbl:ids}). Shorter duration is very important in CA due to an
-already short growing season and water limitations (shorter duration
-cultivars would presumably require less irrigation water). For a
-detailed discussion of cultivars' agronomic and grain quality
-characteristics, refer to \url{http://rice.ucanr.edu/Reports-Publications/Rice_Production_Workshop_Manual/}.
+\end{comment}
 
@@
 potential causes?  It is not due to the adoption of the broadly
-adapted cultivars (M-202, M-206, and M-105) as there is no evidence of
+adapted cultivars (M-206 and M-105) as there is no evidence of
 increased yield potential among these (Fig. \ref{fig:first_yrs} \&
-\ref{fig:compare_yrs}). Although M-205 is not broadly adapted, it has
-increase yield potential and its adoption (although limited) in the
+\ref{fig:compare_yrs}) compared to M-202 released at the start of this
+period. Although M-205 is not broadly adapted, it has
+increase yield potential (Fig. \ref{fig:first_yrs} and its adoption (although limited) in the
 northern and warmer part of CA may contribute to some yield gains
@@
 management~\citep{Lundy2012, Lundy2015, linquist2009}, and weed
-management~\citep{Saito2008, Pittelkow2012, linquist2008}. However,
+management~\citep{deforest2017, Pittelkow2012, caton2002}. However,
 recent rates of state-wide yield gains are far less than those of the
@@
 
+\begin{comment}
 \citet{mckenzie2014} explored the yield of rice in this same system
@@
 the estimates presented here are likely more reliable.
-
-It should be noted that while our investigation has been concerned
-about increasing yield potential, the estimates here might not capture
-yield stability (how consistent grain yields are from year to year). The
-occurrence of stress events (e.g., extreme temperatures) are
-unpredictable across years and are not well captured by the analysis
-here. Yield stability is notoriously difficult to
-quantify~\citep[e.g.,][]{lobell2011}, hence further investigation is
-needed to explore how yield stability has changed over the study
-period.
+\end{comment}
 
@@
 
+\subsection{Other advances from breeding}
+
+Our investigation here is narrowly focused on the rough-rice/field
+grain yield. Other advancements due to plant improvement, including
+increased quality~\citep{mckenzie1993} and milling
+yield~\citep{mckenzie1994}, have been made over the study period and
+are important. Additionally, crop duration has been reduced in CA rice
+cultivars. For example, while the predominate cultivars M-202 and
+M-206 had similar yield potentials at the time of release (Fig. 5),
+M-206 matures about 6 days earlier than M-202 (Table
+\ref{tbl:ids}). Shorter duration is very important in CA due to an
+already short growing season and water limitations (shorter duration
+cultivars would presumably require less irrigation water). For a
+detailed discussion of cultivars' agronomic and grain quality
+characteristics, refer to \url{http://rice.ucanr.edu/Reports-Publications/Rice_Production_Workshop_Manual/}.
+
+It should be noted that while our investigation has been concerned
+about increasing yield potential, the estimates here might not capture
+yield stability (how consistent grain yields are from year to year). The
+occurrence of stress events (e.g., extreme temperatures) are
+unpredictable across years and are not well captured by the analysis
+here. Yield stability is notoriously difficult to
+quantify~\citep[e.g.,][]{lobell2011}, hence further investigation is
+needed to explore how yield stability has changed over the study
+period.
+
 \section{Conclusions}
+
+Increasing grain yield is essential to meeting the challenges of
+feeding a growing population in a changing world. Here we show that
+side-by-side cultivar comparisons do not necessarily indicate gains in
+yield potential as they do not account for yield declines over time of
+the older cultivars being tested. Here we show, as have others, that
+the yield of a cultivar declines over time. Therefore, quantifying
+changes in yield potential requires comparing yields of new cultivar
+yields to yields of the older cultivars when they were first
+released. Unfortunately, our analysis did not allow us to identify the
+cause of yield declines in this case. While efforts should be made to
+identify the cause of yield decline over time, this study highlights
+both the need for plant breeding to simply maintain current yields, as
+well as the challenge of increasing genetic yield potential,
+especially in systems such as CA where yield is approaching the
+physiological limit. The goal of increasing yield may require the
+yield performance of cultivars to be assessed earlier in the selection
+process to avoid discarding higher yielding entries without full
+consideration. We suggest here that historical benchmarks be used in
+to evaluate promising rice cultivars in addition to side-by-side
+comparisons to maintain perspective of the absolute yield gains being
+made.
 

### 2018-03-11 output/breed_paper_revised.Rnw
After yanking the conclusions.

@@
 
-Here we show that side-by-side cultivar comparisons do not necessarily
-indicate gains in yield potential as they do not account for yield
-declines over time. Quantifying changes in yield potential requires also
-examining historical benchmarks. Improvement in grain yield is
-essential to meeting the challenges of feeding a growing population in
-a changing world. To achieve these improvements, we discuss here that
-(1) continuous maintenance via plant improvement is required, without
-which yields would decrease over time, (2) the challenge of improving
-genetic yield potential is substantial, since absolute gains require
-improvement in yield in excess of the maintenance threshold, and (3)
-increasing yield through improved management (agronomy) cannot exceed
-the ceiling of genetic yield potential. Taken together, these
-conclusions underscore both the necessity and substantial challenges
-facing rice breeding. Maintaining high quality is imperative for
-a breeding program and it is not uncommon in a breeding program
-for higher yielding cultivars to be eliminated due to concerns about
-quality. However, the goal of increasing yield may require the yield
-performance of cultivars to be assessed earlier in the selection
-process to avoid discarding these higher yielding entries without full
-consideration. We suggest here that historical benchmarks be
-used in to evaluate promising rice cultivars in addition to
-side-by-side comparisons to maintain perspective of the absolute yield
-gains being made.
-
 \newpage

### 2018-03-11 output/review_responses.tex
Update reviewer comments.

@@
 
-\textbf{Comment: This is a very valuable and interesting manuscript.  It is very well
-  suited to Field Crops Research.  All reviewers recommend 'Minor
-  revision'.  There was some uncertainty over the statistical analysis
-  and so a third reviewer was asked to review this aspect of it. This
-  was not able to be entered into EVISE and so I have appended it below.}\\
+\textbf{Comment: This is a very valuable and interesting manuscript.
+  It is very well suited to Field Crops Research.  All reviewers
+  recommend 'Minor revision'.  There was some uncertainty over the
+  statistical analysis and so a third reviewer was asked to review
+  this aspect of it. This was not able to be entered into EVISE and so
+  I have appended it below.}\\
 
@@
 
-\textbf{Comment: This paper analyses 33 years worth of single-site yield data on rice
-  varieties grown in California. Regression models are fitted to assess
-  yield trend due to cultivars and trend due to "yield erosion".I have a
-  few comments on the statistical analysis.}\\
+\textbf{Comment: This paper analyses 33 years worth of single-site
+  yield data on rice varieties grown in California. Regression models
+  are fitted to assess yield trend due to cultivars and trend due to
+  "yield erosion".I have a few comments on the statistical
+  analysis.}\\
 
 \textbf{(1) The term yield erosion suggests that this declining trend
-  component relates to properties of the cultivars, or cultivars losing
-  their potential over time. I am not sure this interpretation is the
-  only one possible. There are many other factors that may cause a
-  decline over time, i.e. climate change, deterioration of the soil,
-  agronomic practices not keeping up with changes of the environment
-  etc. See my comments further below.}\\
+  component relates to properties of the cultivars, or cultivars
+  losing their potential over time. I am not sure this interpretation
+  is the only one possible. There are many other factors that may
+  cause a decline over time, i.e. climate change, deterioration of the
+  soil, agronomic practices not keeping up with changes of the
+  environment etc. See my comments further below.}\\
 
@@
 the term used to describe this phenomonom in previous work in
-rice. Given that ``erosion'' is a slow, sometimes
-imperceptible, process of reducing something
-driven by abiotic, climatic, and biotic forces, ``erosion'' seems to
-be the appropriate term here. While there is a component of ``erosion'' inherent
-to the material being eroded, this term is not often interpreted as it
-has been here, and there are likely other concerns which are loading
-this term with additional meaning.
+rice. Given that ``erosion'' is a slow, sometimes imperceptible,
+process of reducing something driven by abiotic, climatic, and biotic
+forces, ``erosion'' seems to be the appropriate term here. While there
+is a component of ``erosion'' inherent to the material being eroded,
+this term is not often interpreted as it has been here, and there are
+likely other concerns which are loading this term with additional
+meaning.
 
-However, given the sensitivities involved, we have elected to 
-remove the term ``yield erosion'' throughout the manuscript.\\
+However, given the sensitivities involved, we have elected to remove
+the term ``yield erosion'' throughout the manuscript.\\
 
-\textbf{Comment: (2) "CRB design" is not a proper name for a design. I suspect the
-  authors are refering to a randomized complete block design (RCBD).}\\
+\textbf{Comment: (2) "CRB design" is not a proper name for a design. I
+  suspect the authors are refering to a randomized complete block
+  design (RCBD).}\\
 
@@
 
-\textbf{Comment: (3) I think the model statement in equation (1) needs some polishing
-  up. Indexing observations just by plots is not sufficient. It appears
-  from the model description that plot data were modelled. If that's the
-  case, then the model misses a block effect.}\\
+\textbf{Comment: (3) I think the model statement in equation (1) needs
+  some polishing up. Indexing observations just by plots is not
+  sufficient. It appears from the model description that plot data
+  were modelled. If that's the case, then the model misses a block
+  effect.}\\
 
@@
 
-\textbf{Response:} Added, though it should be noted that when detailing
-the structural aspect of the model in an article, the error term is often
-not included. \\
+\textbf{Response:} Added, though it should be noted that when
+detailing the structural aspect of the model in an article, the error
+term is often not included. \\
 
@@
 
-\textbf{Comment: If plot data are analysed, a random year-cultivar interaction
-  needs to be added as well.}\\
+\textbf{Comment: If plot data are analysed, a random year-cultivar
+  interaction needs to be added as well.}\\
 
@@
 
-\textbf{Comment: I do not think
-  it's appropriate to equate the intercept with an overall mean, unless
-  both time covariates were mean-centered, which does not seem to be the
-  case.}\\
+\textbf{Comment: I do not think it's appropriate to equate the
+  intercept with an overall mean, unless both time covariates were
+  mean-centered, which does not seem to be the case.}\\
 
-\textbf{Response:} The time variates were mean-centered as stated in the
-manuscript. We have clarified this further (lines 91).\\
+\textbf{Response:} The time variates were mean-centered as stated in
+the manuscript. We have clarified this further (lines 91).\\
 
-\textbf{Comment: (4) If subscripts are used for years and cultivars, the covariates
-  should be indexed accordingly. If i and k index cultivars and years,
-  respectively, then the time elapsed since release needs to be indexed
-  by both i and k (a\_ik, say), whereas the year of release would be
-  indexed by i only (r\_i, say). Using this notation, the authors are
-  regressing yield on a\_ik and r\_i.}\\
+\textbf{Comment: (4) If subscripts are used for years and cultivars,
+  the covariates should be indexed accordingly. If i and k index
+  cultivars and years, respectively, then the time elapsed since
+  release needs to be indexed by both i and k (a\_ik, say), whereas
+  the year of release would be indexed by i only (r\_i, say). Using
+  this notation, the authors are regressing yield on a\_ik and
+  r\_i.}\\
 
@@
 
-\textbf{Comment: There's a simple third time
-  covariate that could be considered for regression: calendar year
-  (t\_k). This regression would directly assess non-genetic components
-  of trend, due, e.g., to climate change, advances in agronomic
-  practices etc. It turns out, however, that a regression on all three
-  covariates is not possible, because a\_ik = t\_k - r\_i, meaning that
-  there would be confounding. Thus, only two of these three regressor
-  vaiables can be fitted, so the regression done by the authors is fine
-  in pinciple.  Nevertheless, this little thought exercise shows that
-  there is some ambiguity in the interpretation of regression
-  coefficients. This issue is discussed in some detail in
-  Piepho, H.P., Laidig, F., Drobek, T., Meyer, U. (2014): Dissecting
-  genetic and non-genetic sources of long-term yield trend in German
-  official variety trials. Theoretical and Applied Genetics 127,
-  1009-1018. The authors of that paper did a regression on t\_k and r\_i.}\\
+\textbf{Comment: There's a simple third time covariate that could be
+  considered for regression: calendar year (t\_k). This regression
+  would directly assess non-genetic components of trend, due, e.g., to
+  climate change, advances in agronomic practices etc. It turns out,
+  however, that a regression on all three covariates is not possible,
+  because a\_ik = t\_k - r\_i, meaning that there would be
+  confounding. Thus, only two of these three regressor vaiables can be
+  fitted, so the regression done by the authors is fine in pinciple.
+  Nevertheless, this little thought exercise shows that there is some
+  ambiguity in the interpretation of regression coefficients. This
+  issue is discussed in some detail in Piepho, H.P., Laidig, F.,
+  Drobek, T., Meyer, U. (2014): Dissecting genetic and non-genetic
+  sources of long-term yield trend in German official variety
+  trials. Theoretical and Applied Genetics 127, 1009-1018. The authors
+  of that paper did a regression on t\_k and r\_i.}\\
 
 \textbf{Response:} Yes, this is indeed the case. However, this is
-largely a methodological concern, and after some consideration, we have
-elected to not discuss this matter here, due to space constraints.\\
+largely a methodological concern, and after some consideration, we
+have elected to not discuss this matter here, due to space
+constraints.\\
 
@@
 
-\textbf{Comment: This paper certainly should be published as the results are
-  important even if they lack a satisfactory explanation. But I would
-  like to see some changes and more attention to some issues. Much is
-  annotated on the attached manuscript, and some repetitive and
-  confusing, those remarks came as I read the paper. This summary is
-  upon more considered reflection afterwards.}
+\textbf{Comment: This paper certainly should be published as the
+  results are important even if they lack a satisfactory
+  explanation. But I would like to see some changes and more attention
+  to some issues. Much is annotated on the attached manuscript, and
+  some repetitive and confusing, those remarks came as I read the
+  paper. This summary is upon more considered reflection afterwards.}
 
@@
 
-\textbf{Comment: It was important to test whether the interaction cv x years after
-  release was significant, because this could give a clue as to what is
-  the cause of the yield decline, but it wasn’t significant according to
-  the stats which I didn’t really understand. Discussion could explore
-  this in more depth; I mean why some cvs didn’t lose yield and some
-  did. Maybe only widely adopted cvs lost yeild, due to greater presume
-  for disease to evolve virulence?}\\
+\textbf{Comment: It was important to test whether the interaction cv x
+  years after release was significant, because this could give a clue
+  as to what is the cause of the yield decline, but it wasn’t
+  significant according to the stats which I didn’t really
+  understand. Discussion could explore this in more depth; I mean why
+  some cvs didn’t lose yield and some did. Maybe only widely adopted
+  cvs lost yeild, due to greater presume for disease to evolve
+  virulence?}\\
 
-\textbf{Response:} The reviewer should mindful that
-stating that the estimated CI for the cultivar-year interaction
-includes zero is not the same thing as ``no
-effect''. Rather, the correct interpretation is that the effect, if it
-exists, is smaller than our power to detect. Therefore, the correct
-conclusion is that the differences between cultivars in the decline
-over time, if existent, is likely small, though the sign and magnitude of the exact
-effect is uncertain.
+\textbf{Response:} The reviewer should mindful that stating that the
+estimated CI for the cultivar-year interaction includes zero is not
+the same thing as ``no effect''. Rather, the correct interpretation is
+that the effect, if it exists, is smaller than our power to
+detect. Therefore, the correct conclusion is that the differences
+between cultivars in the decline over time, if existent, is likely
+small, though the sign and magnitude of the exact effect is uncertain.
 
-While this is indeed interesting, we
-have attempted to limit our speculation on this point given that we do not have
-sufficient data to actually test many of these ideas. We have expanded
-the discussion to include some of the points raised (lines 210). \\
+While this is indeed interesting, we have attempted to limit our
+speculation on this point given that we do not have sufficient data to
+actually test many of these ideas. We have expanded the discussion to
+include some of the points raised (lines 210). \\
 
-\textbf{Comment: The paper seems to be written to put the blame on the cultivars and
-  the challenge on the breeders, the notion of their cvs’ yield
-  advantages being “eroded”. This might be the case if a biotic stress
-  is evolving virulence against the host plant resistance of any cv.
-  But it is not really the case when there is a deterioration in the
-  aspects of the environment (climate, agronomy, and/or soil), against
-  which younger cultivars are more resistant or tolerant.  For me the 29
-  kg/ha/yr could be breeding progress in the face of the deteriorating
-  E; at about 0.3\% p.a. it is not unlike the rate of progress in
-  advanced wheat breeding programs (anything near the 1.2\% p.a. which
-  the paper mentions in discussion is very rare these days).}\\
+\textbf{Comment: The paper seems to be written to put the blame on the
+  cultivars and the challenge on the breeders, the notion of their
+  cvs’ yield advantages being “eroded”. This might be the case if a
+  biotic stress is evolving virulence against the host plant
+  resistance of any cv.  But it is not really the case when there is a
+  deterioration in the aspects of the environment (climate, agronomy,
+  and/or soil), against which younger cultivars are more resistant or
+  tolerant.  For me the 29 kg/ha/yr could be breeding progress in the
+  face of the deteriorating E; at about 0.3\% p.a. it is not unlike
+  the rate of progress in advanced wheat breeding programs (anything
+  near the 1.2\% p.a. which the paper mentions in discussion is very
+  rare these days).}\\
 
@@
 these concerns. However, we have taken great pains to specifically
-point out that the evidence suggests this inherent to the system
-(as the title of this manuscript clearly states), that it is similar
-to observations from other rice systems (lines 211-217), and that it is plausibly
-due to the environmental aspects the reviewer lists (lines 190-210).
+point out that the evidence suggests this inherent to the system (as
+the title of this manuscript clearly states), that it is similar to
+observations from other rice systems (lines 211-217), and that it is
+plausibly due to the environmental aspects the reviewer lists (lines
+190-210).
 
 However, the evidence from this study explicitly states that given
-these data and this analysis, a rate of improvement of 29 kg/ha/yr
-is unlikely. The reviewer appears to have missed one of the main
+these data and this analysis, a rate of improvement of 29 kg/ha/yr is
+unlikely. The reviewer appears to have missed one of the main
 conclusions of the manuscript. We have edited for greater clarity. \\
 
-\textbf{Comment: Really it all depends on the definition of potential yield. See van
-  Ittersum and colleagues in FCR Vol 143 March 2013. Its determined by
-  latest best yielding Cv and the current E in the absences of any
-  manageable stresses, both biotic and abiotic, and measured in good
-  experiments and models based on such experiments. A breakdown in cv
-  resistance is probably best described as “erosion”, while a
-  deterioration in the E due, for example, to a negative climate trend,
-  is a decrease in the potential yield, but a new cv tolerating this
-  deterioration would lift the potential yield (see also below).}\\
+\textbf{Comment: Really it all depends on the definition of potential
+  yield. See van Ittersum and colleagues in FCR Vol 143 March
+  2013. Its determined by latest best yielding Cv and the current E in
+  the absences of any manageable stresses, both biotic and abiotic,
+  and measured in good experiments and models based on such
+  experiments. A breakdown in cv resistance is probably best described
+  as “erosion”, while a deterioration in the E due, for example, to a
+  negative climate trend, is a decrease in the potential yield, but a
+  new cv tolerating this deterioration would lift the potential yield
+  (see also below).}\\
 
-\textbf{Response:} Added discussion of these issues (lines 174-181). \\
+\textbf{Response:} This is a good point and we have in part addressed
+this issue by removing the term “yield erosion”. We have also,
+addressed the issue raised by van Ittersum that a changing climate may
+change yield potential (lines ).\\
 
-\textbf{Comment: The authors should read McKay et al (2010, TAG 122, 225-238) where a
-  similar situation in winter wheat in the UK is described. But by
-  having plots with and without fungicide over the last 40 years or so,
-  these authors could measure the erosion and narrow down the cause
-  (foliar disease), which did vary with cultivar.}\\
+\textbf{Comment: The authors should read McKay et al (2010, TAG 122,
+  225-238) where a similar situation in winter wheat in the UK is
+  described. But by having plots with and without fungicide over the
+  last 40 years or so, these authors could measure the erosion and
+  narrow down the cause (foliar disease), which did vary with
+  cultivar.}\\
 
 \textbf{Response:} This is an interesting paper (assuming the reviewer
-intended \textbf{Mackay et al. 2011}), but unfortunately
-these data do not lend themselves to similar explorations. Unlike
-Mackay et al., we do not have paired treated/untreated plots. In the
-course of writing this paper, we discussed various means to to narrow down the
-cause of yield losses, but ultimately lack the data to do so. We have
-added a citation to McKay et al. to the manuscript.\\
+intended \textbf{Mackay et al. 2011}), but unfortunately these data do
+not lend themselves to similar explorations. Unlike Mackay et al., we
+do not have paired treated/untreated plots. In the course of writing
+this paper, we discussed various means to to narrow down the cause of
+yield losses, but ultimately lack the data to do so. We have added a
+citation to McKay et al. to the manuscript.\\
 
-\textbf{Comment: Also the authors need to refer to McKenzie et al (2014) in the CSSA
-  Special Publicn 33 with a whole section on rice yield progress in
-  Ca.}\\
+\textbf{Comment: Also the authors need to refer to McKenzie et al
+  (2014) in the CSSA Special Publicn 33 with a whole section on rice
+  yield progress in Ca.}\\
 
-\textbf{Response:} Discussion added (lines 247-269). \\
+\textbf{Response:} We have addressed this paper more
+thoroughly (lines ). Both McKenzie et al. (2014) and Samonte et
+al (2014) fail to take yield declines into account in their analyses
+and conclusions.  When we exclude these effects, we estimate similar
+effects. We show that accounting yield declines results in better
+explanation of the data, and resolves appearant contradiction between
+short-term and long-term data. \\
 
-\textbf{Comment: These authors show in Fig 10-8 trial results from all over the
-  rice areas of Ca, whereas the manuscript deals with trials solely
-  from the RES at Briggs.This needs to be discussed and may help
-  explain what is happening, as there is little or no evidence in
+\textbf{Comment: These authors show in Fig 10-8 trial results from all
+  over the rice areas of Ca, whereas the manuscript deals with trials
+  solely from the RES at Briggs.This needs to be discussed and may
+  help explain what is happening, as there is little or no evidence in
   McKenzie et al that, for example, the yield of M-202 is declining
@@
 
-\textbf{Response:} Explaination added (lines 247-269)\\
+\textbf{Response:} Explanation added (lines 247-269)\\
 
-\textbf{Comment: What also worries me is that Ca farm yields (and Butte
-  county yields) have been increasing since 1995 or so, at a low rate
-  (about 70 kg/ha/yr) but increasing, which doesn’t agree with Fig S1
-  no matter what cv you choose.}\\
+\textbf{Comment: What also worries me is that Ca farm yields (and
+  Butte county yields) have been increasing since 1995 or so, at a low
+  rate (about 70 kg/ha/yr) but increasing, which doesn’t agree with
+  Fig S1 no matter what cv you choose.}\\
 
@@
 differences in the data being presented. We agree and plainly state in
-the manuscript that CA farm yields have been increasing (lines 236-246). However, we do not find evidence that gains as large as 70 kg/ha/yr
-are due to increasing genetic yield potential. It is entirely possible
-for farm yields to increase for other reasons which, again, we outline
-specifically in the manuscript (lines 241-246).\\
+the manuscript that CA farm yields have been increasing (lines
+236-246). However, we do not find evidence that gains as large as 70
+kg/ha/yr are due to increasing genetic yield potential. It is entirely
+possible for farm yields to increase for other reasons which, again,
+we outline specifically in the manuscript (lines 241-246).\\
 
-\textbf{Comment: If yield is declining because of increasing disease susceptibility
-  with time that is erosion, well known to wheat breeders, and driving
-  what is called maintenance breeding. By definition yield potential is
-  measured without biotic stresses, which are manageable, so this sort
-  of erosion is not a loss of yield potential. If the yield
-  deterioration is due to something like increased ozone pollution or
-  cooler nights at meiosis, that is less clear cut: those stresses are
-  part of the environment, largely unmanageable, and amount to a
-  decrease in the yield potential of the environment, but breeding can
-  still overcome them, at 29 kg/ha/yr if this was the cause in this rice
-  study. Therefore the possible causes at BES need more discussion and
-  reference to other trial data sources. Sorry this repeats 4 a bit!}\\
+\textbf{Comment: If yield is declining because of increasing disease
+  susceptibility with time that is erosion, well known to wheat
+  breeders, and driving what is called maintenance breeding. By
+  definition yield potential is measured without biotic stresses,
+  which are manageable, so this sort of erosion is not a loss of yield
+  potential. If the yield deterioration is due to something like
+  increased ozone pollution or cooler nights at meiosis, that is less
+  clear cut: those stresses are part of the environment, largely
+  unmanageable, and amount to a decrease in the yield potential of the
+  environment, but breeding can still overcome them, at 29 kg/ha/yr if
+  this was the cause in this rice study. Therefore the possible causes
+  at BES need more discussion and reference to other trial data
+  sources. Sorry this repeats 4 a bit!}\\
 
@@
 
-\textbf{Comment:  I am not sure, however, that you need to show Ca yields (Fig 1)
-  because this broadens the discussion a lot, and almost demands you
-  look at what is happening across the rice region in terms of weather
-  trends etc, probably using a rice simulation model (although
-  modelling at just RES itself would be relevant to the paper also). It
-  is useful to show how close farm yields are to RES trial yields but
-  that could be covered in a text sentence to two. And if you want to
-  discuss this bigger picture, please include a simple map of rice
-  areas in Ca.}\\
+\textbf{Comment: I am not sure, however, that you need to show Ca
+  yields (Fig 1) because this broadens the discussion a lot, and
+  almost demands you look at what is happening across the rice region
+  in terms of weather trends etc, probably using a rice simulation
+  model (although modelling at just RES itself would be relevant to
+  the paper also). It is useful to show how close farm yields are to
+  RES trial yields but that could be covered in a text sentence to
+  two. And if you want to discuss this bigger picture, please include
+  a simple map of rice areas in Ca.}\\
 
@@
 
-\textbf{Comment: Table 2 lists the key cv data and should be up front in the paper. You
-  could add to the Table brief comments on the important quality attr
-  ibutes of the various releases (if there any) and finally you could
-  include a column showing the important data of Fig 2, which has
-  nothing to do with the modelling.}\\
+\textbf{Comment: Table 2 lists the key cv data and should be up front
+  in the paper. You could add to the Table brief comments on the
+  important quality attr ibutes of the various releases (if there any)
+  and finally you could include a column showing the important data of
+  Fig 2, which has nothing to do with the modelling.}\\
 
@@
 are not of direct relevance to this paper, so we have provided a
-citation for where this information can be found for those
-interested (line 235). We disagree that the data from Figure 2 should be
+citation for where this information can be found for those interested
+(line 235). We disagree that the data from Figure 2 should be
 presented in table form. Data of this nature is best presented in
@@
 
-\textbf{Comment: So in summary, I am not sure you need Fig 1, you could put the data of
-  Fig 2 in Table 2 and bring it to the front along with Fig S1, and I
-  don’t think you need Fig 6. But consider how you use the word
-  “erosion” and discuss more thoroughly what could be causing the yield
-  declines in Fig S1, and whether there is evidence it is happening also
-  outside of the RES at Biggs.}\\
+\textbf{Comment: So in summary, I am not sure you need Fig 1, you
+  could put the data of Fig 2 in Table 2 and bring it to the front
+  along with Fig S1, and I don’t think you need Fig 6. But consider
+  how you use the word “erosion” and discuss more thoroughly what
+  could be causing the yield declines in Fig S1, and whether there is
+  evidence it is happening also outside of the RES at Biggs.}\\
 
@@
 
-\textbf{Response: } It is inappropriate for the reviewer, who above plainly
-states that they do not understand the statistics, to make this
-evaluation.\\
+\textbf{Response: } It is inappropriate for the reviewer, who above
+plainly states that they do not understand the statistics, to make
+this evaluation.\\
 
-\textbf{Comment: Give sentence on location (latitudes) and area of rice in
-  CA. Even a map of the rice areas , also showing location of RES}\\
+\textbf{Comment: Give sentence on location (latitudes) and area of
+  rice in CA. Even a map of the rice areas , also showing location of
+  RES}\\
 
@@
 
-\textbf{Comment: More info on plots shape sown, and harvested please}\\
+\textbf{Comment: More info on plots shape sown, and harvested
+  please}\\
 
@@
 
-\textbf{Response: } The data are in the figure. Yes, that is our conclusion as well.\\
+\textbf{Response: } The data are in the figure. Yes, that is our
+conclusion as well.\\
 
@@
 claimed by the rice breeders in this system. This information has been
-added to the manuscript (line 39). The figure is
-helpful to distill the major mechanisms at work here to a simple
-conceptual description.\\
+added to the manuscript (line 39). The figure is helpful to distill
+the major mechanisms at work here to a simple conceptual
+description.\\
 
@@
 
-\textbf{Comment: Using data from variety trials, the authors proved the importance of
-  maintenance breeding for world food security. However, I am not very
-  familiar with the statistical analysis that the authors used to derive
-  the conclusions. If a reviewer who is familiar in the related
-  statistical procedures has no problem with the methodology, I believe
-  it is suitable for publication in FCR. Please consider the following
-  suggestion during your revision.}\\
+\textbf{Comment: Using data from variety trials, the authors proved
+  the importance of maintenance breeding for world food
+  security. However, I am not very familiar with the statistical
+  analysis that the authors used to derive the conclusions. If a
+  reviewer who is familiar in the related statistical procedures has
+  no problem with the methodology, I believe it is suitable for
+  publication in FCR. Please consider the following suggestion during
+  your revision.}\\
 
-\textbf{Comment: L10: Something is wrong with “…with it undesirable…”}\\
+\textbf{Comment: L10: Something is wrong with “…with it
+  undesirable…”}\\
 
@@
 
-\textbf{Comment: L30: Yield increase of 50-62 kg per ha per year was not slow.}\\
+\textbf{Comment: L30: Yield increase of 50-62 kg per ha per year was
+  not slow.}\\
 
@@
 
-\textbf{Comment: L51-52: What are the two numbers (39.4648, - 121.7342) ? You may use
-normal way to indicate the location.}\\
+\textbf{Comment: L51-52: What are the two numbers (39.4648, -
+  121.7342) ? You may use normal way to indicate the location.}\\
 
-\textbf{Response:} These are standard latitude/longitude coordinates.\\
+\textbf{Response:} These are standard latitude/longitude
+coordinates.\\
 
@@
 
-\textbf{Comment: L141: Something is wrong with “…since release is between…”}\\
+\textbf{Comment: L141: Something is wrong with “…since release is
+  between…”}\\
 
@@
 
-\textbf{Comment: Fig. S1: It appears that there are two kinds of dots in the figure
-(black and grey). Please indicate their difference.}\\
+\textbf{Comment: Fig. S1: It appears that there are two kinds of dots
+  in the figure (black and grey). Please indicate their difference.}\\
 

### 2018-03-11 output/breed_paper_revised.Rnw
Modifies end of paper.
Updates with minor edits.

@@
 consider yield decline over time, there was an apparent increase in
-yield. However, the yields of older cultivars declined at an estimated
-rate of 29.3 kg ha\tsups{-1} year\tsups{-1} (90\% credible interval
--4.4 to -53.3) after initial selection. Once this effect was
-considered, the yield advantage of newer cultivars over old was
-uncertain (-3.3 kg ha\tsups{-1} year\tsups{-1}, 90\% credible interval
--36.1 to 31.5). These results highlight 1) the importance of
-continuous crop improvement and deployment of new cultivars simply to
-maintain existing yields, and 2) to increase the genetic yield
-potential, higher yield targets are needed. Importantly, when breeding
-near the yield potential, despite the limited yield gains, significant
-advances in improving quality and reducing crop duration have been
-made.
+yield. However, the yields of
+older cultivars were found to decline at an estimated rate of 29.3 kg ha\tsups{-1}
+year\tsups{-1} (90\% credible interval -4.4 to -53.3) after initial
+selection. Once this effect was considered, the yield advantage of
+newer cultivars over old was uncertain (-3.3 kg ha\tsups{-1}
+year\tsups{-1}, 90\% credible interval -36.1 to 31.5). These results
+highlight 1) the importance of continuous crop improvement and
+deployment of new cultivars simply to maintain existing yields, and 2)
+to increase the genetic yield potential, higher yield targets are
+needed. Importantly, when breeding near the yield potential, despite
+the limited yield gains, significant advances in improving quality and
+reducing crop duration have been made.
 
@@
 performance over time has been observed~\citep{Peng1998, Peng2010,
-Peng2000, Datta1995}, sometimes termed ``yield erosion''~\citep{Peng1998}. It
+Peng2000, Datta1995}. It
 has been speculated that this yield decline over time is due to the inability of
@@
 \begin{split}
- yield_{ijk} = Intercept + (Rate_{yield\ decline}* {years\ since\
-   release\ ik}) \\+ (Rate_{yield\ improvement}* {release\ year\ ik})\\ 
- + \beta_{year\ ij} + \beta_{cultivar\ ik} +
- \beta_{cultivar\ ik:year\ ij} + \epsilon \\
+  yield_{ijk} = Intercept + (Rate_{yield\ decline}* {years\ since\ release_{ik}}) \\
+  + (Rate_{yield\ improvement}* {release\ year_{k}}) \\
+  + \beta_{year_j} + \beta_{cultivar_k} + \beta_{cultivar_k:year_j} + \epsilon \\
 \end{split}
@@
 
-Where $yield_i$ is the grain yield for cultivar $k$ in plot $i$ for
+Where $yield_i$ is the grain yield for cultivar $k$ in plot $i$ during
 year $j$,
 the Intercept is the average yield for the RES across cultivars and years, $Rate_{yield\
-decline}$ is rate of yield decline per ${years\ since\ release\ i}$
+decline}$ is the rate of yield decline per
+${years\ since\ release_{ki}}$ for cultivar $k$ in plot $i$
 (mean-centered, i.e., negative for years prior to release, zero at the year of release),
 $Rate_{yield\ improvement}$ is the rate of yield improvement per
-year (${release\ year\ i}$), $\beta_{year}$ is the random effect of
-year, and $\beta_{cultivar}$ is the random effect of cultivar. This
+year (${release\ year_{ik}}$), $\beta_{year_j}$ is the random effect of
+year $j$, and $\beta_{cultivar_k}$ is the random effect of cultivar
+$k$, and lastly $\beta_{cultivar_k:year_j}$ is the interaction term
+for cultivar $k$ in year $j$. This
 full model was modified to further explore yield dynamics over
@@
 into account, the model was used as described above except omitting
-the $Rate_{yield\ decline\ i}$ term (Model 2). Lastly, a model was fit
+the $Rate_{yield\ decline}$ term (Model 2). Lastly, a model was fit
 to estimate if there were differences between cultivars in the rate of
-yield decline over time by adding a random slope by cultivar (Model 3).
+yield decline over time by adding allowing the rate of decline to vary
+by cultivar (i.e., adding
+a $Rate_{yield\ decline_k} * years\ since\ release_{ki}$ term to the
+model for the rate of yield decline for cultivar $k$) (Model 3).
 
@@
 ggplot(mod_data, aes(x = year, y = yield_kg/1000)) +
-    geom_point(size = 0.5, alpha = 0.5) +
+    geom_point(size = 0.5) +
     geom_smooth(method = "lm", color = "black") +
@@
 declines over time. However, these cultivars (M-201, M-204, M-205, and possibly
-M-209; Fig. \ref{fig:ids_comp})) are not widely adopted. Importantly,
+M-209; Fig. \ref{fig:ids_comp}) are not widely adopted. Importantly,
 not all medium grain cultivars that were released were purported to
@@
 The high yielding variety M-202 was released in 1985 and became the
-most widely planted cultivar for two decades~\citep{mkenzie2014}. At
+most widely planted cultivar for two decades~\citep{mckenzie2014}. At
 the time of release, M-202 yields were reported at 11.0 Mg
@@
 yields of newly released varieties cultivars have remained between
-10.8 and 11.5 Mg ha\tsups{-1} (Fig. \ref{first_yrs}), with the
+10.8 and 11.5 Mg ha\tsups{-1} (Fig. \ref{fig:first_yrs}), with the
 exception of M-204 and M-205 (M-204 is no longer grown and M-205 has
@@
 most widely grown rice varietycultivar in California. At the time of
-its release, M-206 was reported to yield \Sexpr{round(9350*1.12,1)} Mg
-ha\tsups{-1} -- also similar to what we found based on our analysis
+its release, M-206 was reported to yield \Sexpr{round(9350*1.12/1000,1)} Mg
+ha\tsups{-1} -- similar to what we found based on our analysis
 (Fig. \ref{fig:first_yrs}). Based on this premise, it is clear that
@@
 ~\citet{mckenzie2014} a yield increase of 25 kg ha\tsups{-1}
-yr\tsups{-1}. Using a similar approach to~\citet{Omar2014} and
-~\citet{mckenzie2014}, when the compared average yields for the period
-2011--2016 of varieties cultivars released since 1982 (and still in
-the variety trials from 2011-- 2016) was compared without accounting
-for yield declines over time, yields were estimated to increase by
+yr\tsups{-1}. The simpliest model (Model 2) in this study is similar
+to the analyses of~\citet{Omar2014} and ~\citet{mckenzie2014}, and
+results in similar findings; yields were estimated to increase by
 \Sexpr{only_release_coef} kg ha\tsups{-1} yr\tsups{-1}
-(Fig. \ref{fig:wo_erosion}), almost identical to what
+(Fig. \ref{fig:wo_erosion}), almost identical to estimates by
 ~\citet{mckenzie2014}.
 
-The only way to reconcile the two apparently contradictory findings,
-is that the yield of potential of varieties cultivars decline over
-time such that when newly released varieties cultivars are compared
-with the older variety cultivar (as shown by ~\citet{Omar2014},
+The only way to reconcile the two apparently contradictory findings above
+is that the yield of potential of cultivars decline over
+time such that when newly released cultivars are compared
+with the older cultivar (as shown by ~\citet{Omar2014},
 ~\citet{mckenzie2014}, and Fig. \ref{fig:wo_erosion}), yields of the
-new cultivar variety are higher. Here we provide evidence that
-cultivar varietal yield does decline - on average at a rate of
+new cultivar are higher. Here we provide evidence that
+cultivar yield does decline - on average at a rate of
 \Sexpr{yrs_coef} kg ha\tsups{-1} yr\tsups{-1}
-(Fig. \ref{w_erosion}). ~\citet{Peng2010} also found that a cultivar’s
-varieties yield potential declines over time and reported a 15\%
+(Fig. \ref{RES_year}). ~\citet{Peng2010} also found that a cultivar's
+yield potential declines over time and reported a 15\%
 decline over a 30 yr period.  In this study, over a 30 year period and
@@
 the yield decline would be approximately
-\Sexpr{round((yrs_coef*-30)/mean(mod_data$yield_kg), 3)*100}\%.
+\Sexpr{round((yrs_coef*-30)/mean(mod_data$yield_kg), 3)*100}\%. Given
+the estimates here using the model which accounts for these declines
+suggests that there is only a
+\Sexpr{round(sum((raw_samples$release_yr_c*scl)>=25)/nrow(raw_samples)*100,1)}\%
+probability of the rate of yield increase is equal to or greater than
+25 kg\tsups{-1} ha\tsups{-1} year\tsups{-1}.
 
@@
 true even when evaluating the cultivars which have been breed for
-broad adaption and high yield potential (M-202, M-206 and
+broad adaption and high yield potential (M-202, M-206, and
 M-105).
@@
 
-Yields of a particular cultivar may decline overtime for a number of
+Yields of a particular cultivar may decline over time for a number of
 reasons. First, ~\citet{Nie2009} reported yield declines in rice due
 to continuously flooded conditions over a long time period which could
-be overcome with improved N management. This is likely not the case
+be overcome with improved N management. This is not likely the case
 here as the fields used in this study grow rice only one season a year
@@
 period, suggesting that older cultivars do not have lower yield
-potential due to changing climatic factors. While our analysis
-indicates that there is no interaction between cultivar and year (CI
-includes zero), visual examination suggests that cultivars may be
+potential due to changing climatic factors. Our analysis
+offers inconclusive evidence of an interaction between cultivar and
+year (i.e., the CI for the interactions overlapped with each other and zero), visual examination suggests that cultivars may be
 different in terms of rates of yield decline
@@
 yield.
-
-\begin{figure}[h]
-<<fig = TRUE, echo = FALSE, width=5, height=5>>=
-
-plot(1, type = "n",
-     # xaxt = "n",
-     xlab = "Years",
-     ylab = expression("Grain yield ( Mg ha"^-1~")"),
-     ylim = c(8.5,11.5),
-     xlim = c(0,40),
-     xaxs = "i")
-
-# arrows(12, 8.2, 27, 8.2, xpd = TRUE, lwd = 2)
-abline(a = 10, b = fixef(RES_fullest)[3], lty = 2, lwd = 2)
-text(25, 10.85, "(3) target above maintenance", cex = 0.8,
-     srt = 18)#((atan(fixef(RES_fullest)[3] + 0.06) * 180))/pi)
-abline(h = 10, lty = 3, lwd = 2)
-text(25, 10.1, "(1) current situation", cex = 0.8)
-abline(a = 10, b = fixef(RES_fullest)[3] + 0.06, lty = 4, lwd = 2)
-text(25, 9.35, "(2) no maintenance breeding", cex = 0.8,
-     srt = -18)
-
-@ 
-\caption{A conceptual diagram highlighting three scenarios: (1)
-maintenance breeding, where losses over time are offset by
-improvements via plant improvement, (2) no maintenance breeding, where
-grain yield slowly declines over time (as estimated from this study),
-and (3) absolute improvement in grain yield (based on ``typical''
-yield goals of 3\% increase per release), where gains in grain yield
-exceed the losses over time.}
-\label{fig:theory}
-\end{figure}
 
@@
 
+\begin{figure}[h]
+<<fig = TRUE, echo = FALSE, width=5, height=5>>=
+
+plot(1, type = "n",
+     # xaxt = "n",
+     xlab = "Years",
+     ylab = expression("Grain yield ( Mg ha"^-1~")"),
+     ylim = c(8.5,11.5),
+     xlim = c(0,40),
+     xaxs = "i")
+
+# arrows(12, 8.2, 27, 8.2, xpd = TRUE, lwd = 2)
+abline(a = 10, b = fixef(RES_fullest)[3], lty = 2, lwd = 2)
+text(25, 10.85, "(3) target above maintenance", cex = 0.8,
+     srt = 18)#((atan(fixef(RES_fullest)[3] + 0.06) * 180))/pi)
+abline(h = 10, lty = 3, lwd = 2)
+text(25, 10.1, "(1) current situation", cex = 0.8)
+abline(a = 10, b = fixef(RES_fullest)[3] + 0.06, lty = 4, lwd = 2)
+text(25, 9.35, "(2) no maintenance breeding", cex = 0.8,
+     srt = -18)
+
+@ 
+\caption{A conceptual diagram highlighting three scenarios: (1)
+maintenance breeding, where losses over time are offset by
+improvements via plant improvement, (2) no maintenance breeding, where
+grain yield slowly declines over time (as estimated from this study),
+and (3) absolute improvement in grain yield (based on ``typical''
+yield goals of 3\% increase per release), where gains in grain yield
+exceed the losses over time.}
+\label{fig:theory}
+\end{figure}
+
 \subsection{Other advances from breeding}

### 2018-03-11 output/breed_paper_revised.Rnw
Yank section in discussion.

@@
 
-\begin{comment}
-It is striking that the estimated yield decline over time is similar
-to that reported by \citet{Peng2010} over a similar time span (15\%
-decline over roughly 30 years). This suggests the mechanisms
-responsible could potentially be endemic to all rice production
-systems. While efforts should be made to identify the cause of yield
-decline over time, this study highlights both the need for maintenance
-of plant breeding and the difficulty of increasing genetic yield
-~\citep{espe2017} potential, especially in systems such as CA where
-yield is approaching the physiological limit \citep{espe16a}.
-
-A goal of many breeding programs is to not only maintain the benchmark
-yield of the previous ``standard'' but to surpass it, especially if we
-are to meet the target of 50\% increased food production by 2050
-\citep{Foley2011a}. Given the estimated yield decline over time, to hit an
-ambitious target of 500 kg ha\tsups{-1} (approximately 5\%) increase
-in yield per release (with roughly 4 years between releases), a new
-cultivar needs to yield an additional \Sexpr{round(yrs_coef*-4,-1)} kg
-ha\tsups{-1} in side-by-side comparison in addition to the 500 kg
-ha\tsups{-1} target improvement. Often side-by-side yield improvements
-show an improvement in grain yield of 2 to 3\% per release, in which case
-no absolute improvements in yield have been made beyond maintaining
-yield.
-
-\end{comment}
-
 California state average yields have increased slowly during the study

### 2018-03-11 output/review_responses.tex
Adds non-breaking hypens.

@@
 \textbf{Response:} The time variates were mean-centered as stated in
-the manuscript. We have clarified this further (lines 91).\\
+the manuscript. We have clarified this further (lines 92).\\
 
@@
 actually test many of these ideas. We have expanded the discussion to
-include some of the points raised (lines 210). \\
+include some of the points raised (lines 237-241). \\
 
@@
 the title of this manuscript clearly states), that it is similar to
-observations from other rice systems (lines 211-217), and that it is
+observations from other rice systems (lines 206,223-228), and that it is
 plausibly due to the environmental aspects the reviewer lists (lines
-190-210).
+227-236).
 
@@
 addressed the issue raised by van Ittersum that a changing climate may
-change yield potential (lines ).\\
+change yield potential (lines 233).\\
 
@@
 \textbf{Response:} We have addressed this paper more
-thoroughly (lines ). Both McKenzie et al. (2014) and Samonte et
+thoroughly (lines 194-209). Both McKenzie et al. (2014) and Samonte et
 al (2014) fail to take yield declines into account in their analyses
@@
 
-\textbf{Response:} Explanation added (lines 247-269)\\
+\textbf{Response:} Explanation added (lines 194-209)\\
 
@@
 the manuscript that CA farm yields have been increasing (lines
-236-246). However, we do not find evidence that gains as large as 70
+245-255). However, we do not find evidence that gains as large as 70
 kg/ha/yr are due to increasing genetic yield potential. It is entirely
 possible for farm yields to increase for other reasons which, again,
-we outline specifically in the manuscript (lines 241-246).\\
+we outline specifically in the manuscript.\\
 
@@
 citation for where this information can be found for those interested
-(line 235). We disagree that the data from Figure 2 should be
+(line 261). We disagree that the data from Figure 2 should be
 presented in table form. Data of this nature is best presented in
@@
 \textbf{Response: } As we state, more study is needed to verify and
-determine the extent and cause of this (lines 209-210).\\
+determine the extent and cause of this (lines 241-243).\\
 

### 2018-03-11 output/breed_paper.Rnw
Final comments.
Adds non-breaking hypens.

@@
 \usepackage{lscape} %% Harvard
+\usepackage[shortcuts]{extdash} %% Avoid breaking on M--20X
 \bibliographystyle{model2-names}\biboptions{authoryear}
@@
 breeding cultivars with increased yield potential. However, when the
-popular cultivar M-202 was released in 1985, it was reported to yield
+popular cultivar M\=/202 was released in 1985, it was reported to yield
 an average of 11 Mg ha\tsups{-1}~\citep[14\% moisture content
 (MC);][]{m202}. In 2015, the reported average yield of the newest
-released cultivar M-209 was 10.8 Mg ha\tsups{-1}~\citep[14\%
+released cultivar M\=/209 was 10.8 Mg ha\tsups{-1}~\citep[14\%
 MC;][]{m209}. However, in annual breeding reports and cultivar release
@@
 being released showed that across cultivars, yields ranged from
-\Sexpr{round(plot_data["M207", "mean"], 2)} (M-207) to
-\Sexpr{round(plot_data["M205", "mean"], 2)} Mg ha\tsups{-1} (M-205)
+\Sexpr{round(plot_data["M207", "mean"], 2)} (M\=/207) to
+\Sexpr{round(plot_data["M205", "mean"], 2)} Mg ha\tsups{-1} (M\=/205)
 (Fig \ref{fig:first_yrs}. Comparing these yields from the time of
@@
 evidence of genetic yield improvements during the 33-year study
-period. Individual exceptions to this were two cultivars (M-204 and
-M-205) that had higher yields than the others.
+period. Individual exceptions to this were two cultivars (M\=/204 and
+M\=/205) that had higher yields than the others.
 
@@
 \ref{tbl:coef}). The estimated 560 kg ha\tsups{-1} yield advantage of
-M-206 over M-202 (the current and former predominant cultivars,
-respectively), can be attributed solely to the erosion of M-202's
+M\=/206 over M\=/202 (the current and former predominant cultivars,
+respectively), can be attributed solely to the erosion of M\=/202's
 yield over time (Fig. \ref{fig:first_yrs} \&
@@
 erosion, both cultivars yielded similarly at their respective year of
-release (Fig. \ref{fig:compare_yrs}), despite the fact that M-202 was
-released 19 years prior to M-206 (Table
+release (Fig. \ref{fig:compare_yrs}), despite the fact that M\=/202 was
+released 19 years prior to M\=/206 (Table
 \ref{tbl:ids}).
@@
 In CA, the cultivars with broad adaptation and adoption over the 33
-year period of this study are M-202, M-206, and M-105. Cold tolerance
+year period of this study are M\=/202, M\=/206, and M\=/105. Cold tolerance
 is a major problem for much of the California rice producing region
@@
 have higher than average yields, even after correcting for yield
-erosion. However, these cultivars (M-201, M-204, M-205, and possibly
-M-209; Fig. \ref{fig:ids_comp}) are not widely adopted. Importantly,
+erosion. However, these cultivars (M\=/201, M\=/204, M\=/205, and possibly
+M\=/209; Fig. \ref{fig:ids_comp}) are not widely adopted. Importantly,
 not all medium grain cultivars that were released were purported to
@@
 traits, but upon release had similar yields to more broadly adapted
-cultivars. Examples of this include M-208, which had resistance to
-blast disease (\textit{Magnaporthe grisea}) and M-402 which is a
-premium quality cultivar. Other cultivars such as M-205 and M-209,
+cultivars. Examples of this include M\=/208, which had resistance to
+blast disease (\textit{Magnaporthe grisea}) and M\=/402 which is a
+premium quality cultivar. Other cultivars such as M\=/205 and M\=/209,
 while high yielding, do not have the cold tolerance of the other
@@
 conclusion holds true even when evaluating the cultivars which have
-been breed for broad adaption and high yield potential (M-202, M-206
-and M-105). Yields of these cultivars at time of release have all
+been breed for broad adaption and high yield potential (M\=/202, M\=/206
+and M\=/105). Yields of these cultivars at time of release have all
 yielded between
@@
 are important. Additionally, crop duration has been reduced in CA rice
-cultivars. For example, while the predominate cultivars M-202 and
-M-206 had similar yield potentials at the time of release (Fig. 5),
-M-206 matures about 6 days earlier than M-202 (Table
+cultivars. For example, while the predominate cultivars M\=/202 and
+M\=/206 had similar yield potentials at the time of release (Fig. 5),
+M\=/206 matures about 6 days earlier than M\=/202 (Table
 \ref{tbl:ids}). Shorter duration is very important in CA due to an
@@
 potential causes?  It is not due to the adoption of the broadly
-adapted cultivars (M-202, M-206, and M-105) as there is no evidence of
+adapted cultivars (M\=/202, M\=/206, and M\=/105) as there is no evidence of
 increased yield potential among these (Fig. \ref{fig:first_yrs} \&
-\ref{fig:compare_yrs}). Although M-205 is not broadly adapted, it has
+\ref{fig:compare_yrs}). Although M\=/205 is not broadly adapted, it has
 increase yield potential and its adoption (although limited) in the
@@
 ha\tsups{-1} year\tsups{-1}), while the dashed line is the published
-3-year average yield of M-202 at its release (1982-1984). Both
+3-year average yield of M\=/202 at its release (1982-1984). Both
 cultivars have a similar yield at their respective year of release,

### 2018-03-11 output/breed_paper_revised.Rnw
Adds non-breaking hypens.

@@
 \usepackage{lineno}
+\usepackage[shortcuts]{extdash} %% Avoid breaking on M--20X
 \usepackage{lscape} %% Harvard
@@
 
-%% \linenumbers
+\linenumbers
 
@@
 trials show that new cultivars usually out yield the older
-cultivars, typically by 3--5\% (e.g., M-209 release description at \url{http://www.crrf.org/linked/2015annualreport.pdf}). In other rice systems, a decrease in a cultivar's yield
+cultivars, typically by 3--5\% (e.g., M\=/209 release description at \url{http://www.crrf.org/linked/2015annualreport.pdf}). In other rice systems, a decrease in a cultivar's yield
 performance over time has been observed~\citep{Peng1998, Peng2010,
@@
 results). Complete data and code used for this analysis is available
-at \url{github.com/mespe/rice_advances.git}.
+through the Open Science Framework at \url{https://osf.io/6ed5k/}.
 
@@
 initial time (average 3 years) in the experimental trials showed that across cultivars, yields ranged from
-\Sexpr{round(plot_data["M207", "mean"], 2)} (M-207) to
-\Sexpr{round(plot_data["M205", "mean"], 2)} Mg ha\tsups{-1} (M-205)
+\Sexpr{round(plot_data["M207", "mean"], 2)} (M\=/207) to
+\Sexpr{round(plot_data["M205", "mean"], 2)} Mg ha\tsups{-1} (M\=/205)
 (Fig \ref{fig:first_yrs}). The relationship between the year
@@
 estimated to be slightly negative (\Sexpr{release_yr_coef}
-kg\tsups{-1} ha\tsups{-1} yr) and does not provide
+kg ha\tsups{-1} yr\tsups{-1}) and does not provide
 evidence of genetic yield improvements during this
-period. Individual exceptions to this were two cultivars (M-204 and
-M-205) that had higher yields than the others.
+period. Individual exceptions to this were two cultivars (M\=/204 and
+M\=/205) that had higher yields than the others.
 
@@
 release. Likewise, an estimated 560 kg ha\tsups{-1} yield
-advantage of M-206 over M-202 (the current and former predominant
+advantage of M\=/206 over M\=/202 (the current and former predominant
 cultivars, respectively), can be attributed solely to the decline
-of M-202's yield over time (Figs \ref{fig:first_yrs} \&
+of M\=/202's yield over time (Figs \ref{fig:first_yrs} \&
 \ref{fig:compare_yrs}). Further supporting this estimate of yield
-decline, both M-202 and M-205 yielded similarly at their respective year of
-release (Fig. \ref{fig:compare_yrs}), despite the fact that M-202 was
-released 19 years prior to M-206 (Fig. \ref{fig:compare_yrs}; Table
+decline, both M\=/202 and M\=/205 yielded similarly at their respective year of
+release (Fig. \ref{fig:compare_yrs}), despite the fact that M\=/202 was
+released 19 years prior to M\=/206 (Fig. \ref{fig:compare_yrs}; Table
 \ref{tbl:ids}).
@@
 there is estimated to be a yield increase of approximately
-\Sexpr{only_release_coef} kg ha\tsups{-1} year\tsups{-1} (90\% CI
+\Sexpr{only_release_coef} kg ha\tsups{-1} yr\tsups{-1} (90\% CI
 \Sexpr{only_release_CI[1]} to \Sexpr{only_release_CI[2]})
@@
 trials at the RES was \Sexpr{release_yr_coef} kg ha\tsups{-1}
-year\tsups{-1}, with both small positive and negative trends within
+yr\tsups{-1}, with both small positive and negative trends within
 the credible interval (90\% CI: \Sexpr{rel_CI[1]} to
@@
 cultivars with higher than average yields:
-M-201, M-205, M-204 (Fig. \ref{fig:ids_comp}).
+M\=/201, M\=/205, M\=/204 (Fig. \ref{fig:ids_comp}).
 
@@
   for by adjusting each observed yield by estimated amount lost 
-  (\Sexpr{yrs_coef} kg ha\tsups{-1} year\tsups{-1} after release).}
+  (\Sexpr{yrs_coef} kg ha\tsups{-1} yr\tsups{-1} after release).}
 \label{fig:w_erosion}
@@
 In CA, the cultivars with broad adaptation and adoption over the
-33-year period of this study are M-202, M-206, and M-105. Cold
+33-year period of this study are M\=/202, M\=/206, and M\=/105. Cold
 tolerance is a major problem for much of the California rice producing
@@
 have better than average yields, even after correcting for yield
-declines over time. However, these cultivars (M-201, M-204, M-205, and possibly
-M-209; Fig. \ref{fig:ids_comp}) are not widely adopted. Importantly,
+declines over time. However, these cultivars (M\=/201, M\=/204, M\=/205, and possibly
+M\=/209; Fig. \ref{fig:ids_comp}) are not widely adopted. Importantly,
 not all medium grain cultivars that were released were purported to
@@
 
-The high yielding variety M-202 was released in 1985 and became the
+The high yielding variety M\=/202 was released in 1985 and became the
 most widely planted cultivar for two decades~\citep{mckenzie2014}. At
-the time of release, M-202 yields were reported at 11.0 Mg
+the time of release, M\=/202 yields were reported at 11.0 Mg
 ha\tsups{-1}~\citep{m202} which is almost identical to what we . This
@@
 based on the average experimental yields prior to public release
-(Fig. \ref{fig:first_yrs}). Since the release of M-202, the rice
+(Fig. \ref{fig:first_yrs}). Since the release of M\=/202, the rice
 yields of newly released varieties cultivars have remained between
 10.8 and 11.5 Mg ha\tsups{-1} (Fig. \ref{fig:first_yrs}), with the
-exception of M-204 and M-205 (M-204 is no longer grown and M-205 has
+exception of M\=/204 and M\=/205 (M\=/204 is no longer grown and M\=/205 has
 limited adoaptation range so is only grown in a limited area). The
-variety cultivar M-206 was released in 2004 and is currently still the
+variety cultivar M\=/206 was released in 2004 and is currently still the
 most widely grown rice varietycultivar in California. At the time of
-its release, M-206 was reported to yield \Sexpr{round(9350*1.12/1000,1)} Mg
+its release, M\=/206 was reported to yield \Sexpr{round(9350*1.12/1000,1)} Mg
 ha\tsups{-1} -- similar to what we found based on our analysis
 (Fig. \ref{fig:first_yrs}). Based on this premise, it is clear that
-improvements in yield potential have not been made over the 33-yr
+improvements in yield potential have not been made over the 33-year
 period of this study. In contrast, both~\citet{Omar2014} and
@@
 
-The only way to reconcile the two apparently contradictory findings above
-is that the yield of potential of cultivars decline over
-time such that when newly released cultivars are compared
-with the older cultivar (as shown by ~\citet{Omar2014},
-~\citet{mckenzie2014}, and Fig. \ref{fig:wo_erosion}), yields of the
-new cultivar are higher. Here we provide evidence that
-cultivar yield does decline - on average at a rate of
-\Sexpr{yrs_coef} kg ha\tsups{-1} yr\tsups{-1}
-(Fig. \ref{RES_year}). ~\citet{Peng2010} also found that a cultivar's
-yield potential declines over time and reported a 15\%
-decline over a 30 yr period.  In this study, over a 30 year period and
-a \Sexpr{yrs_coef} kg ha\tsups{-1} yr\tsups{-1} yield decline rate,
-the yield decline would be approximately
-\Sexpr{round((yrs_coef*-30)/mean(mod_data$yield_kg), 3)*100}\%. Given
-the estimates here using the model which accounts for these declines
-suggests that there is only a
+The only way to reconcile the two apparently contradictory findings
+above is that the yield of potential of cultivars decline over time
+such that when newly released cultivars are compared with the older
+cultivar (as shown by ~\citet{Omar2014}, ~\citet{mckenzie2014}, and
+Fig. \ref{fig:wo_erosion}), yields of the new cultivar are
+higher. Here we provide evidence that cultivar yield does decline - on
+average at a rate of \Sexpr{yrs_coef} kg ha\tsups{-1} yr\tsups{-1}
+(Fig. \ref{fig:RES_year}). When these declines are taken into account,
+there is only a
 \Sexpr{round(sum((raw_samples$release_yr_c*scl)>=25)/nrow(raw_samples)*100,1)}\%
-probability of the rate of yield increase is equal to or greater than
-25 kg\tsups{-1} ha\tsups{-1} year\tsups{-1}.
+probability of a rate of yield increase equal to or greater than 25
+kg ha\tsups{-1} yr\tsups{-1} during the study
+period. ~\citet{Peng2010} also found that a cultivar's yield potential
+declines over time and reported a 15\% decline over a 30 yr period.
+In this study, over a 30 year period and a \Sexpr{yrs_coef*-1} kg
+ha\tsups{-1} yr\tsups{-1} rate, the yield decline would be
+approximately \Sexpr{round((yrs_coef*-30)/mean(mod_data$yield_kg),2)*100}\%.
 
@@
 true even when evaluating the cultivars which have been breed for
-broad adaption and high yield potential (M-202, M-206, and
-M-105).
+broad adaption and high yield potential (M\=/202, M\=/206, and
+M\=/105).
 
@@
 not clear if this mechanism is responsible for the yield decline over
-time of \Sexpr{yrs_coef} Mg ha\tsups{-1} year\tsups{-1} found in this
+time of \Sexpr{yrs_coef} kg ha\tsups{-1} yr\tsups{-1} found in this
 analysis. Biotic stressors such as pests and diseases are low in this
@@
 period, suggesting that older cultivars do not have lower yield
-potential due to changing climatic factors. Our analysis
-offers inconclusive evidence of an interaction between cultivar and
-year (i.e., the CI for the interactions overlapped with each other and zero), visual examination suggests that cultivars may be
-different in terms of rates of yield decline
+potential due to changing climatic factors. While our analysis offers
+inconclusive evidence of differences between cultivars in rate of
+yield decline (i.e., the CIs for the interactions overlapped with
+each other and zero), visual examination suggests that cultivars may
+be different in terms of rates of yield decline
 (Fig. \ref{fig:all_cult}); however, we do not have sufficient data nor
@@
 potential causes?  It is not due to the adoption of the broadly
-adapted cultivars (M-206 and M-105) as there is no evidence of
+adapted cultivars (M\=/206 and M\=/105) as there is no evidence of
 increased yield potential among these (Fig. \ref{fig:first_yrs} \&
-\ref{fig:compare_yrs}) compared to M-202 released at the start of this
-period. Although M-205 is not broadly adapted, it has
-increase yield potential (Fig. \ref{fig:first_yrs} and its adoption (although limited) in the
+\ref{fig:compare_yrs}) compared to M\=/202 released at the start of this
+period. Although M\=/205 is not broadly adapted, it has
+increase yield potential (Fig. \ref{fig:first_yrs}) and its adoption (although limited) in the
 northern and warmer part of CA may contribute to some yield gains
@@
 concluded that genetic rice yield potential has increased over this
-time period at a rate of 25 kg\tsups{-1} ha\tsups{-1}
-year\tsups{-1}. Yet the CI estimated from these data using the best
+time period at a rate of 25 kg ha\tsups{-1}
+yr\tsups{-1}. Yet the CI estimated from these data using the best
 model suggests that there is only a
@@
 probability of the rate of yield increase is equal to or greater than
-25 kg\tsups{-1} ha\tsups{-1} year\tsups{-1}. There are several
+25 kg ha\tsups{-1} yr\tsups{-1}. There are several
 explanations for this discrepancy in conclusions. First, \citet{mckenzie2014} did not
@@
 
+It should be noted that while our investigation has been concerned
+about increasing yield potential, the estimates here might not capture
+yield stability (how consistent grain yields are from year to year). The
+occurrence of stress events (e.g., extreme temperatures) are
+unpredictable across years and are not well captured by the analysis
+here. Yield stability is notoriously difficult to
+quantify~\citep[e.g.,][]{lobell2011}, hence further investigation is
+needed to explore how yield stability has changed over the study
+period.
+
 Planning to meet future demand necessitates increases in grain
@@
 characteristics, refer to \url{http://rice.ucanr.edu/Reports-Publications/Rice_Production_Workshop_Manual/}.
-
-It should be noted that while our investigation has been concerned
-about increasing yield potential, the estimates here might not capture
-yield stability (how consistent grain yields are from year to year). The
-occurrence of stress events (e.g., extreme temperatures) are
-unpredictable across years and are not well captured by the analysis
-here. Yield stability is notoriously difficult to
-quantify~\citep[e.g.,][]{lobell2011}, hence further investigation is
-needed to explore how yield stability has changed over the study
-period.
 
@@
 solid line is the estimated yield decline over time (\Sexpr{yrs_coef} kg
-ha\tsups{-1} year\tsups{-1}), while the dashed line is the published
-3-year average yield of M-202 at its release (1982-1984). Both
+ha\tsups{-1} yr\tsups{-1}), while the dashed line is the published
+3-year average yield of M\=/202 at its release (1982-1984). Both
 cultivars have a similar yield at their respective year of release,
@@
 @
-\caption{An example of a typical side-by-side comparison between a older (M-202)
-  and newer (M-206) cultivar. Shown is the performance of these cultivars at the CA Rice
+\caption{An example of a typical side-by-side comparison between a older (M\=/202)
+  and newer (M\=/206) cultivar. Shown is the performance of these cultivars at the CA Rice
   Experiment Station (RES) over the period \Sexpr{min(comp_yrs)} --
   \Sexpr{max(comp_yrs)}. The newer
-  cultivar, M-206, appears to have a yield advantage over the older
-  cultivar, M-202, though this comparison does not account for the 10
+  cultivar, M\=/206, appears to have a yield advantage over the older
+  cultivar, M\=/202, though this comparison does not account for the 10
   year difference in years since release between the cultivars.}

### 2018-03-11 output/breed_paper_revised.Rnw
Fix spelling.

@@
 uncertainty in the parameter estimates using the quantile method. The
-90\% CI is more stable to sample to sample variance
+90\% CI is more stable to sample-to-sample variance
 and hence is preferable to the 95\% interval~\citep{rstanarm}. To
@@
 exception of M\=/204 and M\=/205 (M\=/204 is no longer grown and M\=/205 has
-limited adoaptation range so is only grown in a limited area). The
+limited adaptation range so is only grown in a limited area). The
 variety cultivar M\=/206 was released in 2004 and is currently still the
-most widely grown rice varietycultivar in California. At the time of
+most widely grown rice cultivar in California. At the time of
 its release, M\=/206 was reported to yield \Sexpr{round(9350*1.12/1000,1)} Mg
@@
 ~\citet{mckenzie2014} a yield increase of 25 kg ha\tsups{-1}
-yr\tsups{-1}. The simpliest model (Model 2) in this study is similar
+yr\tsups{-1}. The simplest model (Model 2) in this study is similar
 to the analyses of~\citet{Omar2014} and ~\citet{mckenzie2014}, and
@@
 declines over time and reported a 15\% decline over a 30 yr period.
-In this study, over a 30 year period and a \Sexpr{yrs_coef*-1} kg
+In this study, over a 30-year period and a \Sexpr{yrs_coef*-1} kg
 ha\tsups{-1} yr\tsups{-1} rate, the yield decline would be
@@
 maintained but not substantially improved the yield potential in the
-face of changing biotic and abiotic pressures over the 33 year time
+face of changing biotic and abiotic pressures over the 33-year time
 frame of this study. This conclusion holds
@@
 identify the cause of yield decline over time, this study highlights
-both the need for plant breeding to simply maintain current yields, as
-well as the challenge of increasing genetic yield potential,
-especially in systems such as CA where yield is approaching the
-physiological limit. The goal of increasing yield may require the
-yield performance of cultivars to be assessed earlier in the selection
-process to avoid discarding higher yielding entries without full
-consideration. We suggest here that historical benchmarks be used in
-to evaluate promising rice cultivars in addition to side-by-side
-comparisons to maintain perspective of the absolute yield gains being
-made.
+both the need for plant breeding to simply maintain current yields,
+and the challenge of increasing genetic yield potential, especially in
+systems such as CA where yield is approaching the physiological
+limit. The goal of increasing yield may require the yield performance
+of cultivars to be assessed earlier in the selection process to avoid
+discarding higher yielding entries without full consideration. We
+suggest here that historical benchmarks be used in to evaluate
+promising rice cultivars in addition to side-by-side comparisons to
+maintain perspective of the absolute yield gains being made.
 

### 2018-03-11 output/breed_paper_revised.Rnw
Final edits.

@@
 years and between cultivars, year and cultivar were included in the
-model as random effects.
+model as random effects (Model 1).
 
@@
 year $j$, and $\beta_{cultivar_k}$ is the random effect of cultivar
-$k$, and lastly $\beta_{cultivar_k:year_j}$ is the interaction term
-for cultivar $k$ in year $j$. This
-full model was modified to further explore yield dynamics over
+$k$, and lastly $\beta_{cultivar_k:year_j}$ is the a random effect for
+the interaction term
+for cultivar $k$ in year $j$. Two variations of this
+full model were used to further explore yield dynamics over
 time. For comparison to estimates where yield decline over time is not taken
@@
 period. Individual exceptions to this were two cultivars (M\=/204 and
-M\=/205) that had higher yields than the others.
+M\=/205) which were observed to have higher yields than the others in
+the raw data (Fig. \ref{fig:RES_year}).
 
@@
 the rate of yield decline over time (i.e., all cultivar specific rates
-of decline overlapped). Differences between cultivars in the decline
+of decline overlapped with eachother and zero). Differences between cultivars in the decline
 of yield over time, seen in the raw data (Fig. \ref{fig:all_cult}),
@@
 was estimated at \Sexpr{yrs_coef} kg ha\tsups{-1} year\tsups{-1} (90\%
-CI: \Sexpr{RES_CI[2]} to \Sexpr{RES_CI[1]}) (Fig \ref{fig:RES_year};
+CI: \Sexpr{RES_CI[2]} to \Sexpr{RES_CI[1]}) (Fig. \ref{fig:RES_year};
 Table \ref{tbl:coef}). The oldest cultivar in the trials
@@
 cultivars, respectively), can be attributed solely to the decline
-of M\=/202's yield over time (Figs \ref{fig:first_yrs} \&
+of M\=/202's yield over time (Figs. \ref{fig:first_yrs} \&
 \ref{fig:compare_yrs}). Further supporting this estimate of yield
@@
 southern portion of the Sacramento Valley due to the influence of the
-ocean winds through the Delta ~\citep{espe16a}. These three cultivars
+ocean winds through the San Francisco Bay ~\citep{espe16a} (see Fig. \ref{fig:map}
+for the geographic orientation of the rice growing area relative to the
+San Francisco Bay). These three cultivars
 have a higher degree of cold tolerance, high yield potential, and
@@
 yr\tsups{-1}. The simplest model (Model 2) in this study is similar
-to the analyses of~\citet{Omar2014} and ~\citet{mckenzie2014}, and
+to the procedures used by~\citet{Omar2014} and ~\citet{mckenzie2014}, and
 results in similar findings; yields were estimated to increase by
-\Sexpr{only_release_coef} kg ha\tsups{-1} yr\tsups{-1}
+an average of \Sexpr{only_release_coef} kg ha\tsups{-1} yr\tsups{-1}
 (Fig. \ref{fig:wo_erosion}), almost identical to estimates by
@@
 cultivar (as shown by ~\citet{Omar2014}, ~\citet{mckenzie2014}, and
-Fig. \ref{fig:wo_erosion}), yields of the new cultivar are
-higher. Here we provide evidence that cultivar yield does decline - on
-average at a rate of \Sexpr{yrs_coef} kg ha\tsups{-1} yr\tsups{-1}
-(Fig. \ref{fig:RES_year}). When these declines are taken into account,
-there is only a
+Fig. \ref{fig:wo_erosion}), yields of the new cultivar are greater
+than those of the older cultivar. Here we provide evidence that
+cultivar yield does decline - on average at a rate of \Sexpr{yrs_coef}
+kg ha\tsups{-1} yr\tsups{-1} (Fig. \ref{fig:RES_year}). When these
+declines are taken into account, it becomes clear that substantial
+increases in yield over the study period are unlikely; there is only a
 \Sexpr{round(sum((raw_samples$release_yr_c*scl)>=25)/nrow(raw_samples)*100,1)}\%
-probability of a rate of yield increase equal to or greater than 25
-kg ha\tsups{-1} yr\tsups{-1} during the study
-period. ~\citet{Peng2010} also found that a cultivar's yield potential
-declines over time and reported a 15\% decline over a 30 yr period.
-In this study, over a 30-year period and a \Sexpr{yrs_coef*-1} kg
-ha\tsups{-1} yr\tsups{-1} rate, the yield decline would be
-approximately \Sexpr{round((yrs_coef*-30)/mean(mod_data$yield_kg),2)*100}\%.
+probability of a rate of yield increase equal to or greater than 25 kg
+ha\tsups{-1} yr\tsups{-1} during the study period. ~\citet{Peng2010}
+also found that a cultivar's yield potential declines over time and
+reported a 15\% decline over a 30 yr period.  In this study, over a
+30-year period and a \Sexpr{yrs_coef*-1} kg ha\tsups{-1} yr\tsups{-1}
+rate, the yield decline would be approximately
+\Sexpr{round((yrs_coef*-30)/mean(mod_data$yield_kg),2)*100}\%.
 
@@
 cooler~\citep{espe2017} temperatures reducing
-yields. Yield potential can be defined as the yield of optimum
+yields, in effect decreasing the evironmental conditions required for
+higher yield potential. Yield potential can be defined as the yield of optimum
 cultivar in the absence of biotic and abiotic stresses in a given
 environment \citep{vanIttersum2013}. However, ~\citet{espe2017} found
-no substantial trends in growing season temperatures (either high,
-low, or average temps) in the Sacramento Valley during the study
-period, suggesting that older cultivars do not have lower yield
-potential due to changing climatic factors. While our analysis offers
+no evidence of sustained trends in growing season temperatures (either high,
+low, or average temperatures) in the Sacramento Valley during the study
+period, which does not support this theory that new cultivars have higher yield
+potential in the face of declining environmental conditions. While our analysis offers
 inconclusive evidence of differences between cultivars in rate of
@@
 increased yield potential among these (Fig. \ref{fig:first_yrs} \&
-\ref{fig:compare_yrs}) compared to M\=/202 released at the start of this
-period. Although M\=/205 is not broadly adapted, it has
-increase yield potential (Fig. \ref{fig:first_yrs}) and its adoption (although limited) in the
-northern and warmer part of CA may contribute to some yield gains
-reported at the state level. Other factors include improvement in
-agronomic management, including more efficient nutrient
+\ref{fig:compare_yrs}) compared to M\=/202 released at the start of
+this period. Although M\=/205 is not broadly adapted, it has increase
+yield potential (Fig. \ref{fig:first_yrs}) and its adoption (although
+limited) in the northern and warmer part of CA may contribute to some
+yield gains reported at the state level. Other factors include
+improvement in agronomic management, including more efficient nutrient
 management~\citep{Lundy2012, Lundy2015, linquist2009}, and weed
@@
 \begin{comment}
+
 \citet{mckenzie2014} explored the yield of rice in this same system
@@
 concluded that genetic rice yield potential has increased over this
-time period at a rate of 25 kg ha\tsups{-1}
-yr\tsups{-1}. Yet the CI estimated from these data using the best
-model suggests that there is only a
+time period at a rate of 25 kg ha\tsups{-1} yr\tsups{-1}. Yet the CI
+estimated from these data using the best model suggests that there is
+only a
 \Sexpr{round(sum((raw_samples$release_yr_c*scl)>=25)/nrow(raw_samples)*100,1)}\%
 probability of the rate of yield increase is equal to or greater than
-25 kg ha\tsups{-1} yr\tsups{-1}. There are several
-explanations for this discrepancy in conclusions. First, \citet{mckenzie2014} did not
+25 kg ha\tsups{-1} yr\tsups{-1}. There are several explanations for
+this discrepancy in conclusions. First, \citet{mckenzie2014} did not
 account for yield decreases over time in their analysis. As shown
@@
 increase in yield over time is estimated to be approximately the same
-as that found by \citet{mckenzie2014}. Second, these yield decreases over time
-were not taken into account possibly because \citet{mckenzie2014}
-concluded that there was no trend in the yield of the major cultivars
-over time. However this conclusion is based on a flawed interpretation
-of the results of their statistical test. Failing to reject the
-hypothesis that the trend in yield equals zero does not equate to the
-confirmation of no trend. Instead, failing to reject the hypothesis
-could suggest the analysis lacks the statistical power to detect the
-trend. Third, it
-is not surprising that no trend in yield over time was detected given
-that \citet{mckenzie2014} used data from several locations, yet
-analyzed those data using a simple t-test. By using this simple
-analysis, \citet{mckenzie2014} ignore systematic errors related to
-differences between observations from different locations, including
-different management systems, growing environments, and crop
-timing. By contrast, the investigation presented here used a single,
-well managed system and analyzed the data using a mixed effects model
-which fully accounts for non-independence between similar
-observations. In doing this, the analysis makes fewer assumptions and
-more fully captures the structural relationships in the data. Hence,
-the estimates presented here are likely more reliable.
+as that found by \citet{mckenzie2014}. Second, these yield decreases
+over time were not taken into account possibly because
+\citet{mckenzie2014} concluded that there was no trend in the yield of
+the major cultivars over time. However this conclusion is based on a
+flawed interpretation of the results of their statistical
+test. Failing to reject the hypothesis that the trend in yield equals
+zero does not equate to the confirmation of no trend. Instead, failing
+to reject the hypothesis could suggest the analysis lacks the
+statistical power to detect the trend. Third, it is not surprising
+that no trend in yield over time was detected given that
+\citet{mckenzie2014} used data from several locations, yet analyzed
+those data using a simple t-test. By using this simple analysis,
+\citet{mckenzie2014} ignore systematic errors related to differences
+between observations from different locations, including different
+management systems, growing environments, and crop timing. By
+contrast, the investigation presented here used a single, well managed
+system and analyzed the data using a mixed effects model which fully
+accounts for non-independence between similar observations. In doing
+this, the analysis makes fewer assumptions and more fully captures the
+structural relationships in the data. Hence, the estimates presented
+here are likely more reliable.
 \end{comment}

### 2018-03-11 src/h1_model.R
Fully adds interaction term to all models and double checks estimates.

@@ -61,24 +61,33 @@ if(compare){
                           data = mod_data, iter = 2000, cores = 2L,
                           seed = 789)
     
-    RES_yrs_fit <- stan_glmer(yield_cs ~ (1|year) + (1|id) + yrs_in_trial,
+    RES_yrs_fit <- stan_glmer(yield_cs ~ (1|year) + (1|id) +  (1|id:year) + yrs_in_trial,
                               data = mod_data, iter = 2000, cores = 2L,
                               seed = 789)
 
-    RES_rel_fit <- stan_glmer(yield_cs ~ (1|year) + (1|id) + release_yr_c,
+    RES_rel_fit <- stan_glmer(yield_cs ~ (1|year) + (1|id) +  (1|id:year) + release_yr_c,
                               data = mod_data, iter = 2000, cores = 2L,
                               seed = 789)
 
     library(loo)
-    compare(loo(RES_full), loo(RES_yrs_fit), loo(RES_rel_fit))
+    compare(loo(RES_fullest), loo(RES_yrs_fit), loo(RES_rel_fit))
     save(RES_yrs_fit, RES_rel_fit, file = "../output/bivar_fit.Rda")
     # Add random slopes for ID
     RES_fuller = stan_glmer(yield_cs ~ (1|year) + (1 + yrs_in_trial|id) +
-                                release_yr_c + yrs_in_trial,
+                                 (1|id:year) + release_yr_c + yrs_in_trial,
                       data = mod_data, iter = 2000, cores = 2L,
                       seed = 789)
 
-    compare(loo(RES_full), loo(RES_yrs_fit), loo(RES_rel_fit), loo(RES_fuller))
+    compare(loo(RES_fullest), loo(RES_yrs_fit), loo(RES_rel_fit), loo(RES_fuller))
+
+# 10 Mar 2018 
+# fuller and fullest are mis-labels,
+# as fullest has fewer p than fuller but not going to change to avoid causing issues
+#                  looic   se_looic elpd_loo se_elpd_loo p_loo   se_p_loo
+# loo(RES_fullest)  4300.7    60.5  -2150.4     30.2       144.4     6.1 
+# loo(RES_fuller)   4303.2    60.5  -2151.6     30.2       146.5     6.2 
+# loo(RES_yrs_fit)  4303.2    60.6  -2151.6     30.3       145.4     6.1 
[…]

### 2018-03-18 output/breed_paper_revised.Rnw
Bruce's final edits?

@@
 
-Based on Model 3, there was no evidence of consistent differences between cultivars in
+Based on Model 3, there was no evidence of differences between cultivars in
 the rate of yield decline over time (i.e., all cultivar specific rates
-of decline overlapped with eachother and zero). Differences between cultivars in the decline
+of decline overlapped with eachother). Differences between cultivars in the decline
 of yield over time, seen in the raw data (Fig. \ref{fig:all_cult}),
@@
 yield~\citep{mckenzie1994}, have been made over the study period and
-are important. Additionally, crop duration has been reduced in CA rice
+are important. Furthermore, as previously mentioned, important disease
+resistance traits have been developed. Additionally, crop duration has been reduced in CA rice
 cultivars. For example, while the predominate cultivars M\=/202 and
